### 2024-10-09 exporatory_visualisation/data_visualisation.qmd
Initial commit

---
title: "Data analysis"
format: 
  docx:
    toc: true
    number-sections: true
execute: 
  echo: false
  warning: false
  message: false
  cache: true
---

```{r}
#| include: false
library(tidyverse)
library(lubridate)
library(fpp3)
library(ggthemes)
monthly_issues_tsibble <- read_rds("../data/monthly_issues_tsibble.rds")
monthly_issues_tsibble <- monthly_issues_tsibble |> filter_index("2017 Aug" ~ "2022 Jun")

knitr::opts_chunk$set(fig.align = 'center', out.width = "90%")

monthly_issues_tsibble |> as_tibble() |> select(month) -> predictors

write_csv(predictors, "../data/predictors.csv")
getwd()
```

```{r}
#| include: false
items <- unique(monthly_issues_tsibble$item)
issue_plot_monthly <- function(my_item) {
  ggplot(data = monthly_issues_tsibble |> filter(item == my_item), mapping = aes(x = month, y = quantity_product_issued))+
    geom_point(size=1)+
    geom_line()+
    theme_few()+
    labs(title=my_item)+
    scale_x_yearmonth(date_breaks = "2 month")+
    theme(axis.text.x = element_text(angle=90))
}

my_plot <- list()
for (i in (1:40)) {
  my_plot[[i]] <- issue_plot_monthly(items[i])
}
```

# Visualising monthly time series

## Title

```{r}
my_plot[[1]]
```

The pharmaceutical supply chain plays a critical role in ensuring the availability and timely distribution of medications and healthcare products to patients and healthcare providers worldwide. It encompasses a complex network of manufacturers, distributors, wholesalers, and pharmacies, all working together to deliver safe and effective medicines to those in need. The supply chain begins with the sourcing of raw materials and active pharmaceutical ingredients (APIs), followed by manufacturing and quality control processes. Once the products are ready, they are transported

## Title

```{r}
my_plot[[2]]
```

The pharmaceutical supply chain plays a critical role in ensuring the availability and timely distribution of medications and healthcare products to patients and healthcare providers worldwide. It encompasses a complex network of manufacturers, distributors, wholesalers, and pharmacies, all working together to deliver safe and effective medicines to those in need. The supply chain begins with the sourcing of raw materials and active pharmaceutical ingredients (APIs), followed by manufacturing and quality control processes. Once the products are ready, they are transported

```{r}
my_plot[[3]]
```

```{r}
my_plot[[4]]
```

```{r}
my_plot[[5]]
```

```{r}
my_plot[[6]]
```

```{r}
my_plot[[7]]
```

```{r}
my_plot[[8]]
```

```{r}
my_plot[[9]]
```

```{r}
my_plot[[10]]
```

```{r}
my_plot[[11]]
```

```{r}
my_plot[[12]]
```

```{r}
my_plot[[13]]
```

```{r}
my_plot[[14]]
```

```{r}
my_plot[[15]]
```

```{r}
my_plot[[16]]
```

```{r}
my_plot[[17]]
```

```{r}
my_plot[[18]]
```

```{r}
my_plot[[19]]
```

```{r}
my_plot[[20]]
```

```{r}
my_plot[[21]]
```

```{r}
my_plot[[22]]
```

```{r}
my_plot[[23]]
```

```{r}
my_plot[[24]]
```

```{r}
my_plot[[25]]
```

```{r}
my_plot[[26]]
```

```{r}
my_plot[[27]]
```

```{r}
my_plot[[28]]
```

```{r}
my_plot[[29]]
```

```{r}
my_plot[[30]]
```

```{r}
my_plot[[31]]
```

```{r}
my_plot[[32]]
```

```{r}
my_plot[[33]]
```

```{r}
my_plot[[34]]
```

```{r}
my_plot[[35]]
```

```{r}
my_plot[[36]]
```

```{r}
my_plot[[37]]
```

```{r}
my_plot[[38]]
```

```{r}
my_plot[[39]]
```

```{r}
my_plot[[40]]
```

### 2024-10-09 manuscript/01-introduction.qmd
Initial commit

---
bibliography: references.bib
---
# Introduction {#sec-intro}

Universal access  essential medicines is a cornerstone of effective healthcare systems, yet ensuring equitable availability of these vital medications remains an urgent global challenge[@quick2003essential;@world2004annual]. Across continents, especially in Africa and Asia, many nations grapple with restricted access due to factors like unaffordability and suboptimal supply chain management [@world2004medicines]. A survey conducted across eight sub-Saharan African countries indicated that there were  unacceptably low availability of essential medicines.The mean availability of 12 priority essential medicines for women ranged from 22% to 40% and from 28% to 57% for children [@droti2019poor]. However, the issue of drug shortages extends beyond developing countries, with instances reported in the United States and Europe, adversely impacting healthcare systems and compromising patient care quality [@fox2003managing, @kaakeh2011impact; @johnson2011drug; @huys2013european;@le2011prevalence].

The consequences of drug shortages are felt both financially within the healthcare sector and in terms of the quality of patient care [@alspach2012drug; @kaakeh2011impact;@kaakeh2011impact; @baumer2004national].The consequences extend across patients, providers, healthcare institutions, and research programs. Patients face delays or cancellations of treatments, substandard care due to the unavailability of optimal therapies, and increased costs in a secondary "grey market." Compromised patient safety, complications, prolonged hospital stays, and even deaths have been reported [@alspach2012drug;].Addressing these shortages is crucial to preserving the quality of patient care and ensuring the effectiveness of healthcare delivery. Thus, a smoothly operating pharmaceutical supply chain is essential to ensure timely access to essential medicines, making accurate demand forecasting an indispensable component.

An efficient pharmaceutical supply chain is vital for timely access to essential health commodities. One of the most crucial components of such a supply chain is a well-defined demand forecasting process. Forecasting accurately the future demand for essential medicines provides valuable insights for anticipating future healthcare requirements including procurement and replenishment of inventory[@subramanian2021effective]. Despite the significance of forecasting for supply chain management, pharmaceutical product forecasting remains a complex endeavor influenced by a multitude of factors. Key considerations include the availability of data, the specific position within the supply chain, and the nuanced understanding of contributing elements. The position within the supply chain introduces unique challenges and dynamics, requiring tailored forecasting approaches. Additionally, a nuanced understanding of contributing elements involves grasping intricate factors that impact forecasting accuracy, such as market trends, regulatory changes, and external influences. Schneider, Wilson, and Rosenbeck (2010) and Hyndman and Athanasopoulos (2018) highlight the intricate nature of pharmaceutical forecasting, emphasizing the need for a comprehensive approach that incorporates these multifaceted aspects to enhance supply chain management effectiveness[@schneider2010pharmaceutical; @hyndman2018forecasting].

The Ethiopian pharmaceutical sector's distinct characteristics, marked by a diverse array of medicines and unique data challenges stemming from service delivery points, staff capacity limitations, communication hurdles, prolonged procurement and distribution lead times, poor forecast accuracy, and policy intricacies, necessitates tailored forecasting strategies that cater to the nation's specific context [@boche2022procurement]. In this context, a comprehensive exploration of effective forecasting models that align with Ethiopia's pharmaceutical landscape is paramount to mitigating stock outs and enhancing healthcare delivery. The implications of pharmaceutical demand forecasting extend beyond financial considerations to societal well-being [@rostami2022forecasting]. In Ethiopia, a country undergoing demographic and healthcare evolution, precise demand forecasting assumes immense importance for resource allocation, supply chain optimization, and policy formulation. With mounting interest in refining demand forecasting in Ethiopia's pharmaceutical sector, this study seeks to contribute to existing knowledge by investigating and recommending proficient demand forecasting methods. Our investigation centers on the application of mean, naive, exponential smoothing, regression, and ARIMA models to bolster forecast accuracy and effectiveness. This paper aims to forecast the demand for selected pharmaceuticals in EPSS over the next 12 months by utilizing various forecasting models and integrating predictors. The integration of predictors with the models will be examined to determine whether they enhance the forecasting accuracy. These predictors are derived from a historical analysis of events in Ethiopia over the last 5 years that have impacted pharmaceutical sales in EPSS. These events include COVID-19, conflicts in the country, disease campaigns, budget closures, budget releases, and physical inventory, all of which, in varying degrees, influence the demand for pharmaceuticals in Ethiopia.

The current paper offers the following contributions we aim at forecasting pharmaceutical sales over a 12-month period these models aim to enhance the accuracy and reliability of demand predictions within the context of EPSS. Moreover an essential contribution lies in the identification of key factors influencing pharmaceutical sales. The study develops a model that effectively incorporates these driving factors, providing a more nuanced and comprehensive approach to demand prediction. We generate forecasts and evaluate model performance using the Root Mean Squared Error (RMSE), Root Mean Squared Scaled Error (RMSSE) and Mean Absolute Error (MAE), Mean Absolute Scaled Error (MASE), Crips and Winkler are used to assess probabilistic forecast accuracy. 
#how you know the forecasting accutacy problem and provide eveidence about the forecasting problems 
# most forecast focus on point forecast and we want to focus on probablistic forecast
# i will write the contribution of the reearch rather than these objectives

This research is guided by several objectives:

1. Evaluate the robustness of various forecasting models for pharmaceutical demand prediction within EPSS.
2. Identify pharmaceuticals that exhibit superior forecast accuracy under specific models.
3. Assess the performance of forecast accuracy metrics to gauge the efficacy of selected models.
4. Test the incorporation of diverse predictors affecting forecasts and evaluate resulting forecasting accuracy.
In pursuing these objectives, this study endeavors to elevate the precision and reliability of pharmaceutical demand forecasting, thereby contributing to optimized supply chain management and enhanced healthcare outcomes in Ethiopia.

### 2024-10-09 manuscript/02-literature.qmd
Initial commit

# Research background {#sec-lit}
# focus on forecasting on pharmaceuticals products 
# 2 pages 
# how forecasting is done 
# what other researches had done 
# what are the limitation of the other papers
# 15 relevant papaers

# 02- literature

Precise sales prediction is an essential and inexpensive way for each company to augment their profits, decrease their costs, and achieve greater flexibility to changes. In other words, exact sales forecasting is utilized for capturing the tradeoff between customer demand satisfaction and inventory costs [@gupta2000mid]. Especially, for the pharmaceutical industry, successful sales forecasting systems can be very beneficial, due to the short shelf-life of many pharmaceutical products
and the importance of the product quality which is closely related to the human health [@doganis2006time].There are many decisions that depend on the quality of forecasts in the health care system, from capacity planning to layout decisions to the daily schedules. In general,the role of forecasting in health care is to inform both clinical and non-clinical decisions. While the former concerns decisions related to patients and their treatments [@makridakis2019forecasting], the latter involves policy/management, and supply chain decisions that support the delivery of high-quality care for patients.In the coming paragarphs we will review what other researchers had done in the area of pharmaceuticals forecasting.Morover today,manufacturing companies are organized following the principles of just-in-time (JIT) delivery which focuses on efficiency and eliminating the waste of excessive inventory [@van2021using]. In JIT systems, accurate Stock Keeping Unit (SKU) short-term demand forecasts, typically on a daily, weekly or monthly basis, are crucial as they are the fundamental inputs to drive the supply chain planning system.Therefore, forecasting performance can strongly affect a company’s financial results, since both under- and overestimating short-term demand are expensive as they result in supply shortages and consequently poor customer service, and excess inventories and product obsolescence, respectively (Sanders and Graman, 2009).In this section, we provide a brief review of studies on forecasting harmaceuticals products.In conclusion, the research findings indicated that the hybrid neural network approach, which combines linear and nonlinear modeling techniques, outperformed traditional methods like ARIMA and standalone neural networks in forecasting pharmaceutical sales. The hybrid approach demonstrated the ability to capture both linear and nonlinear patterns in sales data accurately, especially in scenarios with limited historical records for individual drugs.

A study by Khalil Zadeh et al. developed an intelligent sales prediction model for Pharmaceutical Distribution Companies (PDCs) by combining network analysis tools and time series forecasting methods. This approach addresses the challenge of limited historical sales data for pharmaceutical products, making traditional forecasting methods inadequate[@zhang2004neural].The researchers employed Mean Squared Error (MSE) and Mean Absolute Error (MAE) to evaluate the performance of various forecasting models, including ARIMA, neural networks, and a hybrid approach combining both. The hybrid model, inspired by Zhang's work[@zhang2004neural], used ARIMA for linear components and neural networks for nonlinear components.Three methods were used: (1) ARIMA, (2) a hybrid neural network with each drug’s past records, and (3) a hybrid neural network with both each drug’s and its group members’ past records. Grouping products and using group members' records significantly improved forecasting accuracy by addressing the issue of limited data and leveraging both linear and nonlinear modeling capabilities.The study concluded that the hybrid approach, incorporating both individual and group past records, performed better than using either method alone [@khalil2014intelligent].

A study by Van Bell et al. used various forecasting models for point of sale data to assess the impact of including sell-through data on forecast accuracy. They compared information sharing (IS) methods with non-information sharing (NIS) methods using the following approaches. Substitution Approach (ETS-W and ARIMA-W): These models use downstream wholesaler sales time series as input instead of prior manufacturer shipments, suitable when all goods pass through wholesalers. Integration Approach (Extrapolative Time Series Methods with Exogenous Variables): This includes methods that integrate downstream information using extrapolative time series techniques and machine learning (ML) approaches. Extrapolative Time Series Methods (ETS and ARIMA): Traditional forecasting methods based on modeling past time series structures and extrapolating into the future. The study evaluated these methods using Mean Squared Error (MSE) and Mean Absolute Error (MAE). Additionally, the researchers used naive methods, IS and NIS with ETS and ARIMA models, and machine learning techniques like LASSO, Support Vector Regression (SVR), and Random Forest (RF).LASSO (Least Absolute Shrinkage and Selection Operator) was identified as the best-performing method for improving forecast accuracy, particularly for one-step ahead forecasts. It outperformed other methods for longer horizons (h > 1) by effectively selecting the most relevant variables and reducing overfitting. ETS was found to be the most effective univariate baseline method, and SVR showed the best performance for h = 1. The study concluded that IS methods generally outperformed NIS methods, indicating that incorporating downstream information, such as sell-through data, can significantly enhance short-term forecast accuracy in supply chain management. Machine learning techniques like LASSO, SVR, and RF faced challenges related to overfitting and model complexity, especially with high-dimensional data. Balancing model complexity and interpretability was crucial for robust forecasting. Overall, the research suggests that LASSO, along with ETS and SVR, are effective for improving demand prediction accuracy in multi-echelon supply chains by leveraging downstream information and collaborative approaches[@van2021using].

A study by Nikolopoulos et al. forecasted UK pharmaceutical time series before and after patent expiry, a critical point when a generic version enters the market. The study used various models, including naïve and moving averages, the Bass diffusion model, repeat purchase diffusion models, three regression models, and two exponential smoothing (ES) approaches for non-trended (simple ES) and trended data (Holt ES) [@nikolopoulos2016forecasting].Holt Exponential Smoothing: This model was the most accurate for forecasting both branded and generic drugs across short-term, mid-term, and long-term horizons.Naïve Model: Used as a benchmark, the relative absolute error (RAE) was calculated to compare forecasting accuracy.Bass Diffusion Model: Although popular in marketing research, it had limitations in accurately forecasting pharmaceutical time series due to the complex dynamics between branded and generic drugs. Comparative Analysis: The study compared various forecasting models to evaluate their performance, highlighting the importance of selecting appropriate models based on the data and context.The research found that Holt’s exponential smoothing model outperformed others, including naïve models, moving averages, and ARIMA models, especially for short-term forecasts. However, forecasting accuracy varied across different horizons. Despite their limitations, diffusion models were included to provide additional insights.Overall, the study emphasized the strengths and limitations of different forecasting approaches in the context of pharmaceutical time series, with Holt’s exponential smoothing model demonstrating superior performance[@nikolopoulos2016forecasting].

Astudy by Siddiqui et al used to develop a combined forecasting model.The combination is developed using two frequent, but separately used models, i.e. Autoregressive Integrated Moving Average (ARIMA)  model and Holts Winter model and a combined model were proposed which is known as the ARHOW model.The ARIMA model is commonly used for time series forecasting, while Holt’s Winter model is effective for handling seasonality in data [@inproceedings{markovska2016comparative].The ARHOW model aims to enhance forecasting accuracy by leveraging the strengths of both ARIMA and Holt’s Winter models. The model equation for obtaining a consolidated forecast is as follows: ARHOW = β * ArimaForecast + β1 * HoltWinterForecast
In this equation:
ARHOW represents the Combined Weighted Forecast. ArimaForecast and HoltWinterForecast are the forecasts obtained by applying the ARIMA and Holt’s Winter models over the training set.β and β1 are the weights assigned to the ARIMA and Holt’s Winter forecasts, respectively. By combining the forecasts generated by these two models, the ARHOW model seeks to mitigate the impact of volatility and external factors on forecasting accuracy, ultimately providing more reliable predictions for pharmaceutical sales data [@siddiqui2022hybrid] .  Combining multiple forecasting techniques like ARIMA and Holt’s Winter to create the ARHOW model introduces complexity. Managing and interpreting the results of a hybrid model may pose challenges in practical implementation and decision-making processes .Morover External factors such as regulatory changes, technological advancements, and market dynamics can introduce forecasting errors. While the ARHOW model aims to improve accuracy, it may still be susceptible to errors resulting from unforeseen external influences [@siddiqui2022hybrid].The key forecasting metrics employed in the study include:

Mean Absolute Percentage Error (MAPE), Mean, Mean Absolute Deviation (MAD), Mean Square Error (MSE), Root Mean Square Error (RMSE).The Mean Absolute Percentage Error (MAPE) and other forecasting metrics suggest that the ARHOW model provides more accurate predictions compared to traditional forecasting techniques like ARIMA, Holt’s Winter, ETS, Theta, and Naïve models. The model's efficiency in forecasting pharmaceutical demand can help manufacturers plan inventory and prevent stockouts or overstock situations .In conclusion, the study underscores the effectiveness of the ARHOW model in improving demand forecasting accuracy in the pharmaceutical industry and provides insights for enhancing supply chain efficiency through advanced forecasting techniques .

A study by de Oliveira et al. compared five machine learning regression approaches for lead time forecasting in the pharmaceutical supply chain. The models used were k-nearest neighbors (KNN), support vector machines (SVM), random forests (RF), linear regression (LR), and multi-layer perceptrons (MLP). These models aimed to predict supplier delivery times, enhancing planning for production and logistics services in the pharmaceutical field [@de2021lead]. The study focused on using machine learning techniques to forecast lead times for purchasing products in the pharmaceutical industry. The comparison aimed to identify the most effective approach for predicting purchasing lead times, demonstrating high accuracy potential, especially with linear support vector regression. Key limitations noted included computational complexity, sensitivity to outliers, difficulty in interpretability, and overfitting. Using mean squared error (MSE) as the forecast accuracy metric, the study found that the SVM approach achieved the best performance, with an average error of less than two days. This indicates that SVM was the most effective model among those tested for forecasting purchasing lead times in the pharmaceutical supply chain[@inproceedings{de2021lead].

A study by Rathipriya et al. focused on demand forecasting for time-series pharmaceutical data using both shallow and deep neural network models[@rathipriya2023demand]. The researchers employed several models, including the Autoregressive Integrated Moving Average (ARIMA) model, known for its simplicity and ability to handle non-stationary series.The shallow neural network models used were, Radial Basis Function Neural Network (RBF_NN),Generalized Regression Neural Network (GR_NN),Probabilistic Neural Network (P_NN),For deep neural networks, the study investigated:Long Short-Term Memory (LSTM),Stacked LSTM,Additionally, the study explored hybrid models, combining ARIMA with the Adaptive Neural Fuzzy Inference System (ANFIS) to forecast Iran's annual energy usage, demonstrating higher accuracy compared to individual models. The researchers also employed methods like Genetic Algorithms (GA), Bayesian structures, and Reinforcing Artificial Neural Networks (RANN) for forecasting short-run demand.The performance of the demand forecast models (DFMs) was evaluated using Root Mean Squared Error (RMSE). The study found that shallow neural network models, such as GR_NN, RBF_NN, and P_NN, had almost identical mean RMSE values across all eight drug categories, indicating their effectiveness in predicting pharmaceutical demand. Conversely, deep neural network models like LSTM and Stacked LSTM did not perform as well due to the limited number of samples available for training. The research concluded that shallow neural network-based DFMs performed well for five out of eight drug categories, while the ARIMA model was best for the remaining three. This suggests that no single forecasting model is optimal for all drug categories[@article{rathipriya2023demand].

A study by Candan et al. on demand forecasting in the pharmaceutical industry utilized a neuro-fuzzy approach as the forecasting method[@inproceedings{hrupelj2017demand]. Neuro-fuzzy modeling combines information from empirical models, heuristics, and data, providing effective models that explain solutions better than black-box models like neural networks[@abraham2001neuro; @babuvska2003neuro]. The study also explored the emotional learning fuzzy inference system (ELFIS) and locally linear neuro-fuzzy (LLNF) methods proposed by Abdollahzade et al. (2012) for predicting the stock prices of a pharmaceutical company[@abdollahzade2012application; @hrupelj2017demand]. While the neuro-fuzzy approach combines fuzzy logic and neural networks to improve forecasting, it may still struggle with extremely complex or non-linear relationships within the data. The study used paired t-tests and mean difference analysis to evaluate forecast accuracy. Overall, the metrics supported the conclusion that the neuro-fuzzy approach is a robust method for demand forecasting in the pharmaceutical industry, providing accurate predictions closely aligned with actual demand quantities.

A study by Burinskiene discusses various forecasting techniques in the pharmaceutical retail industry. These include: Single Exponential Smoothing: Applies weighted averages with exponentially decreasing weights for older observations. Moving Average: Used for trend-based estimation to predict demand and standard deviation.
Naive Method: Assumes all predictions are equal to the last observation, serving as a benchmark. Holt's Linear Method: An extension of single exponential smoothing for trend prediction. Holt-Winters No Trend Method: A variation without a trend component[@burinskiene2022forecasting]. The study focuses on short-term forecasting, excluding long-term forecasting and safety stock considerations, which are crucial in the pharmaceutical industry due to varying demand patterns[@nikolopoulos2016forecasting]. The naive method fails to capture trends or seasonality, while the moving average method struggles with sudden changes and irregular patterns[@restyana2021analysis]. Holt's linear method is limited in handling complex seasonal patterns, and the Holt-Winters method requires careful parameter tuning and historical data analysis[@koehler2001forecasting]. Single exponential smoothing is not effective for sudden changes or outliers[@inproceedings{zahra2019forecasting]. The study used MAPE, MSE, and Theil's U2 test to evaluate forecast accuracy, with Theil's U2 test comparing results to the naive forecast. A U2 value greater than 1 indicates better performance than the naive method [@burinskiene2022forecasting]. The Holt-Winters Multiplicative method was found to improve weekly forecasts and outperform other methods, especially in addressing product shortages in pharmacies.

A study conducted by Anusha et al. in the Indian pharmaceutical retail sector utilized forecasting models such as Simple Moving Average, Exponential Smoothing, and Winter’s Exponential Smoothing[@anusha2014demand]. These models were employed to predict demand for both seasonal pharmaceutical items like Okacet 10mg tablets and random pharmaceutical items like Stamlo Beta tablets. Quantitative indicators like Mean Absolute Deviation (MAD), Mean Squared Error (MSE), and Mean Absolute Percentage Error (MAPE) were used to assess forecast error. The Winter’s Exponential Smoothing (WES) model showed the least sales forecast error for the seasonal pharmaceutical item Okacet 10mg tablets, with a MAPE of 27.50, indicating high forecast accuracy. Conversely, for the non-seasonal pharmaceutical item Stamlo Beta tablets, the Six-month Moving Average model performed best with a MAPE of 45.62[@anusha2014demand]. However, the study did not account for unexplained seasonal variations in staff, shifts in the number of potential patients, physician turnover, preferences for pharmaceuticals, and the influence of marketing representatives on surrounding clinics. These factors were not factored into the analysis, potentially affecting the accuracy of demand forecasting in the pharmaceutical retail setting. These limitations suggest that the study may not have considered all relevant variables impacting demand forecasting accuracy.

Zhu et al. conducted a study comparing various forecasting methods in the pharmaceutical industry, including Moving Average (MA), Linear Regression (LR), Exponential Smoothing (ES), Exponential State Space Model (ETS), Vector Autoregressive with exogenous variables (VARX), Symbolic regression, Graph-based analysis, Artificial Neural Networks (ANN), Clustering, and Recurrent Neural Networks (RNN)[@zhu2021demand]. Additionally, machine learning algorithms such as Linear Regression (LR), Support Vector Regression (SVR), Random Forest (RF), and RNN with cross-series training were assessed.The study evaluated forecast accuracy using metrics like Normalized Mean Squared Error (NMSE), Normalized Mean Absolute Error (NMAE), and Bias. Results indicated that the cross-series training models outperformed baseline models (Moving Average, Exponential Smoothing, Linear Regression) in terms of NMSE, NMAE, and Bias. RNN models, particularly when implemented with cross-series training, showed superiority and generalizability across datasets compared to other machine learning algorithms (LR, SVR, RF), confirming the effectiveness of the proposed forecasting framework in the pharmaceutical industry.

Ribeiro et al. conducted a study aimed at detecting outliers and predicting sales for a pharmaceutical distribution company. They employed the dampened Pegels method, a time-series data mining technique, to forecast future quantities of medicines to be ordered by the company[@article{ribeiro2017improving]. The accuracy of the forecasting model was evaluated using the Symmetric Mean Absolute Percentage Error (SMAPE) metric, which indicated that the Pegels method provided more accurate sales forecasts compared to the company's previous method. Furthermore, these forecasts effectively met demand, aiding in efficient stock management and ensuring customer satisfaction while minimizing inventory costs.

The study by Bon and Ng at the University Health Centre aimed to enhance inventory demand forecasting for Panadol 650mg through the utilization of various forecasting techniques. Their research design encompassed both qualitative and quantitative data collection methods within a descriptive framework. Historical data spanning from January 2009 to August 2014 were analyzed employing methods like single moving average, exponential smoothing, and regression analysis. Forecast accuracy was assessed using metrics including RMSE, MSE, MAD, and MAPE to identify the most precise forecasting technique. The objective was to optimize inventory management by selecting the most accurate forecasting method based on these measurements[@bon2017optimization]. Results highlighted regression analysis as the superior technique among the ten methods tested, exhibiting the least forecasting error for optimizing Panadol 650mg inventory demand at the University Health Centre[@inproceedings{bon2017optimization].

The study conducted by Newberne and MIKE focused on utilizing Holt-Winters Forecasting in Healthcare, particularly in the context of healthcare management involving Pseudoephedrine prescriptions and Upper Respiratory Infection (URI) visits [@newberne2006holt]. Data from the Military Healthcare System's M2 database were analyzed using the Holt-Winters method to demonstrate its applicability in healthcare forecasting. Forecasting accuracy was assessed using the Mean Absolute Percentage Error (MAPE) metric. The study employed various forecasting methods including Naïve, Seasonal Naïve, ARIMA, and Deep Learning Model: Long Short-Term Memory (LSTM) Recurrent Neural Network. Additionally, techniques like Simple Exponential Smoothing, Holt with linear and damped trends, Holt-Winters, ES state space model, and Theta were explored to forecast drug quantities for distribution in the Brazilian Public Health System. Among these methods, the LSTM model often exhibited the best performance based on Mean Absolute Error (MAE) as a measure of forecast accuracy. However, deep learning models like LSTM may pose challenges due to their complexity and computational intensity, requiring specialized expertise for tuning and implementation in practical settings.

The study conducted by Zahra and Putra focused on using ARIMA and Single Exponential Smoothing methods to predict and calculate EOQ (Economic Order Quantity). To measure forecast accuracy, the researchers utilized RMSE, MAPE, MAD, and MSE metrics. The findings revealed that the ARIMA (1,0,0) method exhibited the smallest error values, with a MAPE of 13% and MSE of 283,715, indicating its superior accuracy compared to the Exponential Smoothing method. Based on these results, the researchers predicted a need for 222 boxes of medicine for the next period and recommended procuring 223 boxes to maintain a service level of 95%. The study underscored the significance of accurate forecasting in optimizing medicine procurement processes and ensuring adequate stock levels to efficiently meet demand. By employing MAPE and MSE as forecast accuracy metrics, the researchers could compare the performance of different forecasting methods and determine the most effective approach for predicting medicine needs in hospital settings[@zahra2019forecasting].

The study conducted by Restyana et al. focused on analyzing drug forecasting within the Jombang Regency from 2017 to 2019. They employed the Single Moving Average and Single Exponential Smoothing methods to predict drug demand across 35 public health centers. The objective was to identify the method with the lowest forecasting error by assessing metrics like Mean Absolute Deviation (MAD) and Mean Square Error (MSE). By analyzing annual primary data and testing errors, the researchers aimed to evaluate the accuracy of the forecasting methods. The findings revealed that the Single Exponential Smoothing (SES) approach exhibited superior forecast accuracy, as evidenced by lower MAD and MSE values.The study by Kim et al. used Vector Autoregressive with Exogenous Variables (VARX) models for demand forecasting of medicines. Two models were constructed to test the effect of exogenous variables: Model 1, a pure autoregressive model without exogenous variables, and Model 2, the suggested VARX model incorporating topic trends from social media analysis. The study used sales data as endogenous variables and moving averages of topic trends as exogenous variables in the VARX models.The results showed that incorporating topic trends into the VARX models enhanced the performance of demand forecasting, particularly for medicines sensitive to environmental and seasonal changes. By analyzing topic trends related to certain diseases and incorporating them into VARX models, the researchers found significant improvement in forecast accuracy.Prediction error rates were used as the forecast accuracy measure, calculated for four different medicines for each model (Model 1 and Model 2). The comparison of error rates between the two models demonstrated that adding topic trends as exogenous variables in the VARX model significantly improved the accuracy of demand forecasts, highlighting the importance of integrating social media data into forecasting processes[@kim2015demand].

Table 1 provides a summary of some studies in the literature on forecasting in pharmaceuticals products. We have identified several limitations in the existing literature on pharmaceutical product forecasting that have motivated us to pursue this research. These limitations can be summarized as follows:

1. Almost all of the studies have primarily concentrated on point forecasts at a single level.
2. A majority of the research has not explored longer time series data periods.
3. Only a few studies have integrated domain knowledge with time series data to enhance pharmaceutical forecasting accuracy.
4. Additionally, no studies have been conducted in African settings, where challenges related to forecasting and pharmaceutical availability are prevalent.

Table 1. Summary of Some Studies on Forecasting in pharmaceuticals products

---
title: "Table 1. Summary of Some Studies on Forecasting in pharmaceuticals products"
tbl-cap-location:top

```{r}

library(knitr)
data <- data.frame(
  Header1 = c("Row 1", "Row 2", "Row 3"),
  Header2 = c("Data 1", "Data 2", "Data 3"),
  Header3 = c("Data 1", "Data 2", "Data 3")
)
kable(data, format = "html")

```

 |                       |           |               |             |          | References       |  | Year      | Variable      |             |
 | :-----------           |-----------| ------------- |-----------  |
 | Siddiqui et al            |           |               |             | 
 |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               | |
 |            |           |               |
 |            |           |               |         
 |            |           |               |
 |            |           |               |
 |            |           |               |
 |            |           |               |
 |            |           |               |
 |            |           |               |
 |            |           |               |
 |            |           |               |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |          
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           ||            |           |
 | -----------|-----------|
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |          
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |          
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |           |
 |            |
 |            |
 |            |
 |            |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |
 |

### 2024-10-09 manuscript/03-experiment.qmd
Initial commit

# Experiment setup {#sec-experiment}

# you need to talk about the data set 
# some visualization 
# make reduce to paragaphs minimize it 
# forecast accuracy measures provid ethe formulas

## Data {#sec-data}
# 2.1. Dataset
For this study, we utilized a dataset spanning five years (2018-2022) of sales data obtained from EPSS. Rigorous checks were conducted to ensure the consistency and completeness of the collected data. From the extensive pool of pharmaceutical products within EPSS, we strategically selected a set of 37 key pharmaceuticals, encompassing various programs and representing different classes of drugs. This deliberate selection enriches the research by providing a comprehensive representation across the pharmaceutical landscape.
We are presenting the graph for some of the pharmaceuticals sales for the last five years by using visualization for the readers to undesrtand the sales of the pharmaceuticals.
```{r}
#| include: false
library(tidyverse)
library(lubridate)
library(fpp3)
library(ggthemes)
monthly_issues_tsibble <- read_rds("../data/monthly_issues_tsibble.rds")
monthly_issues_tsibble <- monthly_issues_tsibble |> filter_index("2017 Aug" ~ "2022 Jun")

knitr::opts_chunk$set(fig.align = 'center', out.width = "90%")

monthly_issues_tsibble |> as_tibble() |> select(month) -> predictors

write_csv(predictors, "../data/predictors.csv")
getwd()
```

```{r}
#| include: false
items <- unique(monthly_issues_tsibble$item)
issue_plot_monthly <- function(my_item) {
  ggplot(data = monthly_issues_tsibble |> filter(item == my_item), mapping = aes(x = month, y = quantity_product_issued))+
    geom_point(size=1)+
    geom_line()+
    theme_few()+
    labs(title=my_item)+
    scale_x_yearmonth(date_breaks = "2 month")+
    theme(axis.text.x = element_text(angle=90))
}

my_plot <- list()
for (i in (1:40)) {
  my_plot[[i]] <- issue_plot_monthly(items[i])
}
```

# Visualising monthly time series

## Title

```{r}
my_plot[[1]]
```

```{r}
my_plot[[2]]
```

```{r}
my_plot[[4]]
```

```{r}
my_plot[[6]]
```
```{r}
my_plot[[9]]
```

```{r}
my_plot[[11]]
```
```{r}
my_plot[[12]]
```
```{r}
my_plot[[15]]
```
```{r}
my_plot[[17]]
```
```{r}
my_plot[[19]]
```
```{r}
my_plot[[21]]
```
```{r}
my_plot[[25]]
```
```{r}
my_plot[[26]]
```
```{r}
my_plot[[29]]
```
```{r}
my_plot[[30]]
```
```{r}
my_plot[[32]]
```
```{r}
my_plot[[36]]
```
```{r}
my_plot[[37]]
```

# 2. 2. Data Splitting and Preparation
The primary objective of this study is to forecast pharmaceutical demand within EPSS for the year 2023. To achieve this, the historical sales data for the chosen pharmaceuticals in EPSS was partitioned into distinct training and testing subsets. Employing the widely recognized 70:30 split method, 70% of the available data was designated as the training set, while the remaining 30% was reserved for testing and validation purposes [@chan2019comparison]. As a result, the dataset was segregated into a training set consisting of 42 months and a testing set spanning 18 months

# 2.3.1. Forecasting Methods
In pursuit of accurate pharmaceutical demand forecasts, we explored a spectrum of forecasting techniques. The following methods were applied:

Naive Method: A simple approach relying on the assumption that the future value will be the same as the most recent observed value.This Methods uses when there is no
seasonality [@article{burinskiene2022forecasting; @article{nikolopoulos2016forecasting]
Mean Method: A straightforward technique forecasting future values based on the mean of past observations.
Seasonal Naive Method: Similar to the naive method, this approach predicts future values using the most recent observed value from the same season.
Exponential Smoothing: A method that assigns exponentially decreasing weights to past observations to forecast future values.
Regression: Employing statistical regression models to establish relationships between variables for prediction.
Autoregressive Integrated Moving Average Model (ARIMA): A time series model incorporating autoregressive, differencing, and moving average components for accurate forecasting.
ARIMA with Predictors: Extending the ARIMA model with additional predictor variables to enhance forecasting accuracy.
Combination of Models: Exploring ensemble methods that combine multiple models for improved predictive performance.
By applying these diverse forecasting methodologies, we aimed to comprehensively assess and compare their efficacy in predicting pharmaceutical demand within EPSS. Each method was evaluated against the testing dataset to ascertain its forecasting accuracy and suitability for the specific

## Forecasting methods {#sec-method}
Model building

# Naive 
The naive forecast method is a simple approach that assumes future values will be the same as the most recent observed value, making it useful for stable or slowly changing time series. While it doesn't consider underlying patterns, it serves as a baseline for evaluating more advanced techniques[@hyndman2018forecasting]. It's particularly suitable for products with stable demand or limited seasonality, providing a straightforward reference for forecasting and inventory management.

# Mean 
The mean forecast method predicts future values by averaging historical data, making it suitable for stable time series. While it overlooks trends and seasonality, it serves as a baseline for evaluating advanced techniques [@hyndman2018forecasting]. Its simplicity makes it valuable in certain forecasting applications [@hyndman2018forecasting].

# Seasonal naive

The seasonal naive forecast method predicts future values based on the most recent observed value from the corresponding season in the previous year, making it useful for data with strong seasonal patterns. It serves as a baseline for evaluating seasonal forecasting models, capturing repetitive patterns effectively. Its simplicity and ability to capture seasonality make it valuable in seasonal forecasting applications[@hyndman2018forecasting].

# Exponential smoothing

Exponential Smoothing is regarded as one of the most accurate and robust methods for time series forecasting. Not only does it exhibit high accuracy, but it also offers versatility and robustness. What makes Exponential Smoothing particularly appealing is its intuitive nature, making it easy to understand and interpret. Furthermore, this method boasts minimal data storage and computational requirements, making it highly efficient for real-time applications. In the business realm, Exponential Smoothing finds extensive use for inventory demand forecasting, showcasing its practicality [@gardner1985exponential]. Surprisingly, even when pitted against more sophisticated techniques, Exponential Smoothing has demonstrated impressive performance in forecasting competitions [@makridakis2000m3,@makridakis1982accuracy]. Its simplicity and reliability make it a widely trusted approach in the field of forecasting.

# Regression

Regression methods play a crucial role in forecasting and optimizing the health supply chain, ensuring the availability and efficient distribution of essential medical resources. By analyzing historical data, regression models can identify patterns and relationships between variables such as patient demand, disease prevalence, and inventory levels. The basic concept is that we forecast the time series of interest assuming that it has a linear relationship with other time series. For example, we might wish to forecast monthly sales y using total advertising spend as a predictor X. Or we might forecast daily electricity y demand using temperature x1   and the day of week x2 as predictors. The forecast variable y  is sometimes also called the regress and, dependent or explained variable. The predictor variables x  are sometimes also called the regressors, independent or explanatory variables.

Simple linear regression 
In the simplest case, the regression model allows for a linear relationship between the forecast variable y and a single predictor variable x: yt = β0 + β1xt + ε 
Example of data from such a model is shown in Figure 5.1. The coefficients β0 and β1 denote the intercept and the slope of the line respectively. The intercept β0 represents the predicted value of  y when x=0 . The slope represents the average predicted change in y resulting from a one unit increase in x.

# Figure : An example of data from a simple linear regression model

Notice that the observations do not lie on the straight line but are scattered around it. We can think of each observation yt  as consisting of the systematic or explained part of the model, = β0 + β1xt , and the random “error”, εt. The “error” term does not imply a mistake, but a deviation from the underlying straight line model. It captures anything that may affect yt other than xt

Multiple linear regressions 
When there are two or more predictor variables, the model is called a multiple regression model. The general form of a multiple regression model is

yt = β0 + β1x1,t + β2x2,t + ⋯ + βkxk,t + εt, (1)

where is the variable to be forecast and are the predictor variables. Each of the predictor variables must be numerical. The coefficients β1,….. βk measure the effect  of each predictor after taking into account the effects  of all the other predictors in the model. Thus, the coefficients measure the marginal effects of the predictor variables. 
Assumptions 
When we use a linear regression model, we are implicitly making some assumptions about the variables in Equation (1).

First, we assume that the model is a reasonable approximation to reality; that is, the relationship between the forecast variable and the predictor variables satisfies this linear equation.

Second, we make the following assumptions about the errors (ε1, … , εT ) :

they have mean zero; otherwise the forecasts will be systematically biased.

they are not autocorrelated; otherwise the forecasts will be inefficient, as

there is more information in the data that can be exploited.

they are unrelated to the predictor variables; otherwise there would be more information that should be included in the systematic part of the model.

It is also useful to have the errors being normally distributed with a constant variance σ2  in order to easily produce prediction intervals. 
Another important assumption in the linear regression model is that each predictor x is not a random variable. If we were performing a controlled experiment in a laboratory, we could control the values of each x (so they would not be random) and observe the resulting values of y. With observational data (including most data in business and economics), it is not possible to control the 
value of x, we simply observe it. Hence we make this an assumption.

# Auto regressive Integrated Moving Average Model (ARIMA)

The Autoregressive Integrated Moving Average (ARIMA) model is a widely utilized approach for forecasting and analyzing time series data. It combines three key components: autoregression (AR), differencing (I), and moving average (MA). ARIMA models have been extensively studied and applied in various domains due to their flexibility and effectiveness in capturing both short-term and long-term dependencies in time series.Box and Jenkins (1970) [1] introduced the ARIMA model as a powerful tool for time series analysis. The authors provided a comprehensive framework for identifying, estimating, and diagnosing ARIMA models. They emphasized the importance of differencing to achieve stationarity in non-stationary time series data, allowing the model to capture trends and patterns effectively.

[@chase2013demand] describes the application of the ARIMA model. It is the forecasting technique that integrates vital components of time-series and methods of regression. In 1970, George Box and Gwilym Jenkins introduced the ARIMA forecasting method, and they developed a comprehensive approach for forecasting [@chase2013demand].ARIMA (p,d,q) models gather relevant factors from the historical data by using auto-correlation between them to distinguish those slacked request verifiable qualities that best anticipate future demand; whereas p is the order of Autoregressive (AR) term, q is the order of MA term

ARIMA can demonstrate cycle and regularity, and is mainly used for mid-term aggregate forecasts and pre-sent far superior estimations than time series models or even casual models; explained by [@stadtler2014supply] explicitly considers dependent demands. Additionally, different components, for example, informative factors that impact request. It is also considered a good fit for the Mean Absolute Percentage Errors (MAPE) calculation.

ARIMA model helps analyse the time series’s probabilistic properties and develop sin-gle or multi-equation models [@moritz2015comparison]. Analysing a time series data is that there is a straightforward yet extensive arrangement of useful models that can speak to numerous conceivable examples of information found in time series. For a stationary time-series, we can envision the information creating process as a weighted blend of earlier perceptions in addition to an irregular request term. Besides, identifying the ARIMA model parameters (i.e. p, q and d) is often challenging. Thereby, we use R software’s statistical package; it is widely used in the relevant literature [@bokde2016psf; @dhamo2010using]. Moreover, ‘R’ generated the best combination of p, d and q values because it estimates the value of the parameter based on AIC (Akaike Information Criterion) and BIC (Bayesian Information Criterion).

#Performance evaluation {#sec-performance}

# 2.4. Forecast accuracy measures
To evaluate the forecast accuracy and bias, several measures including Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), Mean Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE),and Winkler are used. We only present the results for RMSSE due to the space limit of the journal and also because it is the recommended error metric in M5 competition [@makridakis2022m5].

Point forecast accuracy is measured via the Mean Squared Scaled Error (MSSE) and the Mean Absolute Scaled Error (MASE). The Mean Absolute Scaled Error (MASE) [@book{hyndman2018forecasting,@article{hyndman2006another,] is calculated as
```{r}

```

### 2024-10-09 manuscript/04-results.qmd
Initial commit

# Results and discussion {#sec-results}
```{r}
#| include: false
#| label: restults
library(readr)
fc_accuracy_overall__without_predictor <- read_csv("../result/fc_accuracy_overall__without_predictor.csv")
fc_accuracy_predictor_overall <- read_csv("../result/fc_accuracy_predictor_overall.csv")
fc_accuracy_overall <- read_csv("../result/fc_accuracy_all_item.csv")
data_visualisation <- read.csv("../rscript/data_visualisation.R")
```
from the total of 37 key pharmaceuticals selected and tested with the  different forecasting models arima showed that a better forecast accuracy for most of the pharmaceuticals followed by mean and  regresstion models with RMSSE value of 0.850, 0.942 and 0.945 respectively (Table 1).

```{r}
library(kableExtra)
library(dplyr)  # Load the dplyr package
fc_accuracy_cleaned <- na.omit(fc_accuracy_overall__without_predictor)
fc_accuracy_table <- fc_accuracy_cleaned %>%
  dplyr::select(.model, RMSE, MAE, MASE, RMSSE, winkler, CRPS)  # Explicitly specify dplyr::select
kbl(fc_accuracy_table, booktabs = TRUE, escape = FALSE) %>%
  kable_styling(latex_options = c("hold_position"), full_width = FALSE)
```
Then we tested the models by adding different predictors which affects the sales of the pharmaceuticals like physical count, budget release and budget closure by the government, COVID 19, campaign in to the models as predictors which will affects the sales of the key pharmaceuticals products and the result showed that still arima have a better forecast accuracy with RMSSE of 0.850 (Table 2). 
```{r}
#| label: result3
#| tbl-cap: "forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after considering predictors"
fc_accuracy_predictor_overall %>% 
  kbl(booktabs = TRUE, escape = FALSE) |>
  kable_styling(latex_options = c("hold_position"),full_width = F)
```
```{r}
#| label: result2
#| fig-cap: "forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after considering predictors"
library(ggplot2)  # Load the ggplot2 package
library(forcats)  # Load the forcats package
fc_accuracy_overall %>%
  select(`.model`, item, RMSSE) %>% 
  ggplot(mapping = aes(x = RMSSE, y = fct_reorder(`.model`, RMSSE, .fun = median))) +
  geom_boxplot()
```
Individual item with different forecasting models 
We tested 10 models namely arima, arima with predictor, combination, exponential smoothing, mean, naive, regression, regression with predictor, snaive and stlf. Individually on the selected key pharmaceuticals and the result indicated that 11 (32.43%) of the selected pharmaceuticals have better forecast accuracy for arima predictor models, followed by naive 6 (16.21%), mean, regresstion, arima, and STLF methods each of them with 3 (8.10%) (Fig 2)

```{r}
# Sample data (replace with your actual data)
pharmaceuticals <- c("ARIMA Predictor", "Naive", "Mean", "Regression", "ARIMA", "STLF")
accuracy_percentages <- c(32.43, 16.21, 8.10, 8.10, 8.10, 8.10)

# Create a data frame
data <- data.frame(Pharmaceuticals = pharmaceuticals, Accuracy_Percentage = accuracy_percentages)
# Create a bar chart
library(ggplot2)

ggplot(data, aes(x = reorder(Pharmaceuticals, -Accuracy_Percentage), y = Accuracy_Percentage)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Pharmaceutical Forecasting Models", y = "Accuracy Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Those pharmaceuticals who had a better forecast accuracy with Arima predictors are Amoxacillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine, Atenolol, Atrovastatin 200mg, Ferrosu sulphate with follic acid, frusemide injection, Insuline Iosphane bipahsic, Lamivudine +Efaverienze + Tenofovir, Lidocaine + HCL and Propyl thiro uracil.The  six pharmaceuticasl which showed a better forecast accuracy with naive models were Ciprofloxacillin 500mg tablet, Magnesoum sulphate injection, Omeprazole capsule, Omeprazole injection, Rapid diagnostic test, Sulphamethaxone + Trimetoprine tablet.

Discussion:

The objective of this study was to evaluate the forecasting accuracy of different 
models for key selected pharmaceutical products from the Ethiopian Pharmaceutical 
Supply Services (EPSS). The analysis encompassed a total of 37 key pharmaceutical 
products, and the results highlight the performance of various forecasting models in 
predicting sales for these products. In the initial assessment of forecasting models, 
the ARIMA model emerged as the most accurate for the majority of the pharmaceutical 
products, exhibiting a Root Mean Squared Scaled Error (RMSSE) value of 0.850. This 
finding suggests that ARIMA has a strong predictive capability and outperforms other 
models, including the mean and regression models. The superiority of the ARIMA models 
over the Moving Average might be explained by  most of the data set have seasonal 
patern and MA  is one of the simplest prediction techniques for making projections 
about time-series without a noticeable seasonal pattern [@chopra2001supply].The 
superiority of the ARIMA model underscores its suitability for pharmaceutical sales 
forecasting within the context of EPSS. Similar findings were reported in forecasting 
ethanol demand in India where ARIMA model outperforms other models like linear and 
non-linear regression models [@dey2023forecasting].

To enhance the predictive power of the models, additional predictors were incorporated 
into the analysis. Predictors such as physical count, budget release and closure by the government, the impact of COVID-19, and promotional campaigns were considered as 
potential drivers of pharmaceutical sales. Despite the inclusion of these predictors, 
the ARIMA model maintained its superior forecasting accuracy with an RMSSE of 0.850. 
This resilience further validates the robustness of the ARIMA model and its ability to 
capture complex interactions between variables influencing pharmaceutical sales.

To delve deeper into the individual performance of the forecasting models, a 
comprehensive evaluation was conducted on each of the 37 pharmaceutical products. Among the ten models tested, ARIMA with predictors exhibited the highest forecast accuracy 
for 32.43% of the products. This was followed by the naive model, which performed well 
for 16.21% of the products. Notably, a few pharmaceuticals demonstrated a strong fit 
with specific models. For instance, Amoxicillin 500mg capsule, Anti RhD, Artehmether + 
Lumefantrine, and others exhibited superior forecast accuracy when utilizing the ARIMA 
model with predictors.

Similarly, other pharmaceuticals displayed better forecasting outcomes with alternative models. Notably, Ciprofloxacin 
500mg tablet, Magnesium sulfate injection, Omeprazole capsule, and others showed 
improved accuracy with the naive model. These findings underscore the importance of 
tailoring the choice of forecasting model based on the characteristics and 
dynamics of individual pharmaceutical products.This finding confirmed that There is not
a single best technique to solve 
time-series forecasting problems [@zhang2007quarterly]. In order to deal with 
time-series forecasting, each problem might be solved with a different approach [@ensafi2022time].

In conclusion, this study highlights the pivotal role of accurate forecasting in the 
pharmaceutical supply chain management. The ARIMA model, particularly when integrated 
with relevant predictors, emerges as a powerful tool for 
predicting pharmaceutical sales within the EPSS. However, the selection of the most 
appropriate model should consider the specific attributes of each pharmaceutical 
product. Further research could explore the optimization of model parameters and the 
incorporation of additional contextual variables to enhance forecasting accuracy and 
supply chain efficiency.

Limitation

One notable limitation of this study revolves around the availability and quality of 
the data used for analysis. While we employed a comprehensive dataset from the EPSS, it is essential to acknowledge that the accuracy of forecasting models is heavily 
contingent on the quality, granularity, and completeness of historical sales data. 
Variations in data recording practices, potential errors, missing values, or 
inconsistencies within the dataset could impact the precision of the forecasts 
generated by the employed models. Moreover, the inherent complexity of pharmaceutical 
demand, influenced by a myriad of external factors such as socioeconomic changes,stock 
outs, prolonged procurement lead time, rationing of avaiable products, pushing of 
pharmaceuticals to different health facility without demand, healthcare policies, and 
unforeseen events, may introduce an additional layer of uncertainty. It is important to recognize that despite our best efforts to address data-related challenges, the 
robustness and reliability of our forecasting outcomes may still be subject to the 
limitations inherent in the original data.

### 2024-10-09 manuscript/05-conclusion.qmd
Initial commit

# Conclusion {#sec-conclusion}
In this study, we embarked on a comprehensive analysis of pharmaceutical demand forecasting within the Ethiopian 
Pharmaceutical Supply Services (EPSS) context. By leveraging five  years of sales data for 37 key pharmaceutical 
products, we aimed to  enhance the accuracy of demand predictions and consequently  contribute to more effective supply 
chain management. Our findings  underscore the significance of accurate demand forecasting in  optimizing inventory 
levels, resource allocation, and overall supply chain operations.

The application of diverse forecasting models yielded valuable insights into their respective performance. Notably, the 
Autoregressive Integrated Moving Average (ARIMA) model emerged as a  robust choice, consistently exhibiting superior forecast accuracy  across a majority of the selected pharmaceutical products.  Additionally, the incorporation of predictor variables, such as  physical count, budget release and closure by the government,  COVID-19 impact, and campaign events, further underscored the  efficacy of the ARIMA model in accommodating external influences and delivering reliable forecasts.

The individualized assessment of forecasting models revealed specific strengths across pharmaceutical categories. ARIMA 
models with predictors demonstrated exceptional performance for pharmaceuticals including Amoxacillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine, Atenolol, and others, affirming their utility in capturing complex demand patterns. 
Moreover, the study shed light on pharmaceuticals best suited for alternative models like naive, mean, and regression 
methods, contributing to a comprehensive understanding of forecasting applicability.

While these findings contribute significantly to the discourse on pharmaceutical demand forecasting, it's essential to 
acknowledge limitations. The accuracy of our forecasts is intrinsically linked to the quality and completeness of 
historical sales data, and the intricate interplay of external variables adds complexity to the forecasting landscape.

In conclusion, this study serves as a pivotal step towards enhancing pharmaceutical supply chain management through 
refined demand forecasting. The demonstrated success of ARIMA models, particularly when augmented with predictor 
variables, holds promise for fostering resilient and adaptable supply chains. As the pharmaceutical landscape continues 
to evolve, our insights encourage further exploration of advanced forecasting techniques and continuous refinement of 
strategies to bolster supply chain efficiency and ensure consistent availability of essential pharmaceuticals.

In light of the study's findings, several actionable recommendations emerge. First and foremost, the integration of the 
Autoregressive Integrated Moving Average (ARIMA) model with predictor variables stands out as a viable strategy for 
enhancing pharmaceutical demand forecasting accuracy within Ethiopian Pharmaceutical Supply Services (EPSS). This entails 
harnessing the predictive power of external factors to refine forecasting precision. To further advance forecasting 
practices, the exploration of additional predictor variables such as public health campaigns, disease outbreaks, and 
economic indicators could offer valuable insights into demand dynamics. Moreover, refining the granularity of historical 
sales data, possibly through finer temporal intervals and geographic diversity, holds potential for more nuanced and 
accurate predictions. Consideration could also be given to hybrid forecasting models that amalgamate ARIMA's short-term 
accuracy with machine learning's predictive prowess for long-term trends. Regular model evaluation, sensitivity analysis, and collaborative stakeholder engagement are recommended to ensure ongoing accuracy and alignment with changing demand patterns.

Capacitating EPSS staff through training programs in forecasting methodologies and data interpretation can empower the 
workforce to make informed decisions. Replication of this study in diverse healthcare settings and comparative analyses 
can validate the findings' applicability and adaptability. By embracing these recommendations, EPSS can fortify its 
pharmaceutical supply chain with resilient, efficient, and responsive operations, ultimately ensuring the consistent 
availability of vital pharmaceuticals.
  a

### 2024-10-09 manuscript/main.qmd
Initial commit

---
title: "Pharmaceutical Supply Chain Forecasting by using Machine Learning in Ethiopian Pharmaceutical Supply Services"
author:
  - name: Arebu Issa Bilal
    email: arebu.issa@aau.edu.et
    affiliations:
        - id: Addis 
          name: Addis Ababa  University
          department: Department of Pharmaceutics and Social Pharmacy at School of Pharmacy
          country: Ethiopia
          postal-code: 9086
    attributes:
        corresponding: true
  - name: Arebu Issa Bilal
    email: arebu.issa@aau.edu.et
  - name: Bahman Rostami-Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliations:
        - id: carbs
          name: University
          department: Cardiff Business School 
          country: UK
          postal-code: CF5
  - name: Umit Sezer Bititci 
    email: U.S.Bititci@hw.ac.uk
    affiliations:
        - id: heriot- watt 
          name: University
          department: Edinburgh Business School 
          country: UK
          postal-code: EH14 4AS
  - name: Teferi Gedif Fenta
    email: tgedif@gmail.com
    affiliations:
        - id: Addis 
          name: Addis Ababa  University
          department: Department of Pharmaceutics and Social Pharmacy at School of Pharmacy
          country: Ethiopia
          postal-code: 9086

abstract: |
  This study delves into the realm of pharmaceutical demand forecasting within the Ethiopian Pharmaceuticals Supply Services(EPSS) context. It encompasses a comprehensive analysis of 37 key pharmaceutical products,assessing their forecast accuracy across variation models. The initial assessment reveals that the Autoregressive Integrated Moving Average (ARIMA) model emerges as a frotrunner, boasting a Root Mean Squared Error (RMSSE) value of 0.850. ARIMA's prowess in capturing complex demand patterns suggests its suitability for pharmaceutical sales forecasting in the EPSS landscape. To argument forecasting accuarcy, predictor varaibles such as physical count, budget release and closure by the goverment, the impact of COVID-19, and promotional campaigns were incorpoarted. Remarkably, the ARIMA model mantainsits superiority even in the presences of these external factors, reiterating its roustness. Delving further into individual pharmaceutical products, ARIMA with predictors demonstrates exceptional performance for a substantial portion (32.43%) of the products, including Amoxicillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine, and others. Conversely,alternative models like naive, mean, and regression methods showcase strengths for different pharmaceutical categories, underscoring the importance of tailoring forecasting approaches to product-specific dynamics. While this study significantly advances the discourse on pharmaceutical demand forecasting, limitations stemming from data quality and external variables underscore the intricacies of the forecasting landscape. In conclusion, it advocates the integration of the ARIMA model with predictor variables as a strategic approach for refining pharmaceutical demand forecasting within EPSS. It also encourages further exploration of advanced forecasting techniques and ongoing refinement to bolster supply chain efficiency and ensure consistent pharmaceutical availability. Capacity building through staff training and replication of the study in diverse healthcare settings are recommended steps toward resilient and efficient pharmaceutical supply chains.These findings hold promise for fostering supply chains that are adaptable, responsive, and capable of ensuring the continuous availability of essential pharmaceuticals in Ethiopia.
keywords:
  - Pharmaceutical Demand Forecasting
  - Ethiopian Pharmaceutical Supply Services
  - ARIMA Model
  - Forecasting Variables
  - Forecast Accuracy
bibliography: references.bib
execute:
  echo: false
  message: false
  warning: false
  pdf-engine: pdflatex
  prefer-html: true
  cache: true
format:
  html:
    number-sections: true
  docx:
    number-sections: true
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
        \mathtoolsset{showonlyrefs}
        \setlength{\parindent}{0cm}
    journal:
      name: Journal of 
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

```{r}
#| label: setup
#| include: false
library(fpp3)
library(kableExtra)
library(tidyverse)
library(kableExtra)
library(ggthemes)
library(patchwork)
library(readr)
library(forcats)
# Color blind color palettes
options(
  ggplot2.discrete.colour = c("#D55E00", "#0072B2","#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442"),
  ggplot2.discrete.fill = c("#D55E00", "#0072B2","#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442")
)
# Few theme
ggplot2::theme_set(ggthemes::theme_few())
```

```{r intro, child="01-introduction.qmd"}

```

```{r literature, child="02-literature.qmd"}

```

```{r experiment, child="02-experiment.qmd"}

```

```{r results, child="03-results.qmd"}

```

```{r conclusion, child="04-conclusion.qmd"}

```

# Reproducibility {.unnumbered}

R code to produce all results in this paper is available at

# References {#references .unnumbered}

### 2024-10-09 rscript/analyse_accuracy_result_for_all_items.R

@@ -1,0 +1,5 @@
+# analyse_accuracy_result_for_all_items
+
+library(tidyverse)
+
+fc_accuracy_all_item <- read_csv("result/fc_accuracy_all_item.csv")

### 2024-10-09 rscript/data_preparation.R

@@ -1,0 +1,92 @@
+
+
+
+library(tidyverse)
+library(readxl)
+library(lubridate)
+
+# ----read data ----------
+
+issued_forecast_original <- read_excel("data/full_issued_data.xlsx") |> janitor::clean_names() 
+
+# full_issued <- read_excel("data/full_issued_data.xlsx") |> janitor::clean_names()
+
+write_rds(issued_forecast_original,"data/full_issued.rds") 
+
+full_issues_original <- read_rds("data/full_issued.rds")
+
+stock_out_days_original <- read_excel("data/number_of_stock_out_days.xlsx") |> janitor::clean_names()
+
+full_issues_original <- full_issues_original |> filter(item  !="Rifampicin + Isoniazid + Pyrazinamide + Ethambutol (RHZE)")
+
+HP_wastage_rate_original <- read_excel("data/stock_out_data.xlsx", sheet = "HP wastage rate")|>
+  janitor::clean_names()
+
+##--------------- preparation
+
+full_issues <- full_issues_original |> mutate(expiry_date = as_date(expiry_date)) |>
+select(transaction_date,item,expiry_date,unit_price,unit,quantity_unit_issued=quantity_issued,
+       quantity_product_issued=quantity_issued_new_each, amount_issued,environment,region,zone,woreda,institution_id,
+       institution_type) |> drop_na()
+
+full_issues_correct_items <- full_issues |> mutate(
+  item = case_when(
+    item == "Tetracycline - 0.01 - Eye Ointment" ~ "Tetracycline - 1% - Eye Ointment" ,
+    item == "Sulphamethoxazole + Trimethoprim - (200mg + 40mg)/5ml - Mixture" ~ "Sulphamethoxazole + Trimethoprim - (200mg + 40mg)/5ml - Suspension",
+    item == "Insulin Soluble Human - 100IU/ml in 10ml Vial - Injection(Solution)" ~ "Insulin Soluble Human" ,
+    item == "Insulin Soluble / Neutral (HPB) - 100 Units/ml  - Injection" ~ "Insulin Soluble Human" ,
+    item == "Propylthiouracil - 100mg - Tablet (Scored)"  ~ "Propylthiouracil - 100mg - Tablet",
+    item == "Insuline Isophane Biphasic (Soluble/Isophane Mixture) - (30 + 70)IU/ml in 10ml Vial - Injection (Suspension)"  ~ "Insuline Isophane Biphasic",
[…]

### 2024-10-09 rscript/data_visualisation.R

@@ -1,0 +1,111 @@
+library(tidyverse)
+library(lubridate)
+library(fpp3)
+library(ggthemes)
+#----------------
+
+# read data-------
+
+issues <- read_rds("data/full_issues_correct_amountissues.rds")
+
+# manipulate data---------------
+issues_selected <- issues |> select(transaction_date,item,quantity_product_issued,amount_issued)
+daily_issues <- issues_selected |> mutate(date=as_date(transaction_date)) |> group_by(date,item) |> 
+  summarise(quantity_product_issued=sum(quantity_product_issued), amount_issued=sum(amount_issued)) |> 
+  ungroup()
+
+## create a tsibble
+daily_issues_tsibble <- daily_issues |> as_tsibble(index = date, key=item) |> 
+  fill_gaps(quantity_product_issued = 0, amount_issued=0)
+
+
+# daily visualisation---------------------
+
+issue_plot <- function(my_item) {
+ggplot(data = daily_issues_tsibble |> filter(item == my_item), mapping = aes(x = date, y = quantity_product_issued))+
+  geom_point(size=1)+
+  geom_line()+
+  theme_few()
+}
+
+
+issue_plot("Sulphamethoxazole + Trimethoprim - (200mg + 40mg)/5ml - Suspension")
+
+daily_issues_tsibble_features <- daily_issues_tsibble |> 
+  features(quantity_product_issued, feature_set(pkgs = "feasts"))
+
+ggplot(data = daily_issues_tsibble_features, mapping = aes(x = trend_strength, y = seasonal_strength_week))+
+  geom_point()+
+  theme_few()
+
+  
+high_sesonal <- daily_issues_tsibble_features |> filter(seasonal_strength_week==max(seasonal_strength_week)) |> pull(item)
+
+gg_season(daily_issues_tsibble |> filter(item == high_sesonal), period = "week")
+
+## Monthly visualisation
+
+monthly_issues_tsibble <- daily_issues_tsibble |> index_by(month=yearmonth(date)) |> group_by_key() |> 
+  summarise(quantity_product_issued=sum(quantity_product_issued),
+            amount_issued=sum(amount_issued))
+
+write_rds(monthly_issues_tsibble,"data/monthly_issues_tsibble.rds")
+
+
+# MONTHLY visualisation--------
+items <- unique(monthly_issues_tsibble$item)
[…]

### 2024-10-09 rscript/epss_forecast_preparation.R

@@ -1,0 +1,73 @@
+library(tidyverse)
+library(readxl)
+library(lubridate)
+# ----read data ----------
+
+# issued_forecast_original <- read_excel("data/forecast_issued-data.xlsx") |> janitor::clean_names()
+# write_rds(issued_forecast_original,"data/issued_forecast_original.rds")
+
+full_issues_original <- read_rds("../data/full_issued.rds")
+
+##--------------- preparation
+
+full_issues <- full_issues_original |> mutate(expiry_date = as_date(expiry_date)) |>
+select(transaction_date,item,expiry_date,unit_price,unit,quantity_unit_issued=quantity_issued,
+       quantity_product_issued=quantity_issued_new_each, amount_issued,environment,region,zone,woreda,institution_id,
+       institution_type) |> drop_na()
+
+
+full_issues_correct_items <- full_issues |> mutate(
+  item = case_when(
+    item == "Tetracycline - 0.01 - Eye Ointment" ~ "Tetracycline - 1% - Eye Ointment" ,
+    item == "Sulphamethoxazole + Trimethoprim - (200mg + 40mg)/5ml - Mixture" ~ "Sulphamethoxazole + Trimethoprim - (200mg + 40mg)/5ml - Suspension",
+    item == "Insulin Soluble Human - 100IU/ml in 10ml Vial - Injection(Solution)" ~ "Insulin Soluble Human" ,
+    item == "Insulin Soluble / Neutral (HPB) - 100 Units/ml  - Injection" ~ "Insulin Soluble Human" ,
+    item == "Propylthiouracil - 100mg - Tablet (Scored)"  ~ "Propylthiouracil - 100mg - Tablet",
+    item == "Insuline Isophane Biphasic (Soluble/Isophane Mixture) - (30 + 70)IU/ml in 10ml Vial - Injection (Suspension)"  ~ "Insuline Isophane Biphasic",
+    item == "Insuline Isophane Biphasic (Soluble/Isophane Mixture) - (30 + 70)IU/ml in 10ml Vial - Injection(Suspension)"  ~ "Insuline Isophane Biphasic",
+    item == "Sulphamethoxazole + Trimethoprim - (200mg +40mg)/5ml - Suspension"  ~  "Sulphamethoxazole + Trimethoprim - (200mg + 40mg)/5ml - Suspension",
[…]

### 2024-10-09 rscript/forecasting_model_predictors.R

@@ -1,0 +1,103 @@
+
+library(distributional)
+library(MASS)
+library(writexl)
+library(openxlsx)
+library(fpp3)
+library(readr)
+library(readxl)
+
+
+
+write.xlsx(full_issues_correct_amountissues, "C:/Users/issaa/Documents/EPSS_research/epps_forecasting_inventory/data.xlsx")
+
+#write.xlsx(full_issues_correct_amountissues, "C:/Users/issaa/Documents/EPSS_research/epps_forecasting_inventory/data.xlsx")
+
+monthly_issues_tsibble <- read_rds("data/monthly_issues_tsibble.rds") |> filter_index("2017 Aug" ~ .)
+
+monthly_predictors_tsibble <- read_excel("data/predictors_Final_Tssible 3.xlsx") |> janitor::clean_names() |> 
+  mutate(month = yearmonth(month)) |> as_tsibble(index = month, key = item) |> 
+  filter_index("2017 Aug" ~ .) |> filter(item  !="Rifampicin + Isoniazid + Pyrazinamide + Ethambutol (RHZE)")
+
+monthly_issues_all <- monthly_issues_tsibble |> left_join(monthly_predictors_tsibble) |> 
+  mutate(fical_year_counting_lag = lag(fical_year_counting), fical_year_counting_laed = lead(fical_year_counting))
+#write_rds(monthly_issues_all,"data/monthly_issues_all.rds")
+
+# we split the data into train an test
+
+f_horizon <- 6 #  forecast horizon = 6 months
+percentage_test <- 0.3 #20% of time series for test set
+
+monthly_issues_all |> filter(item == "Artemether + Lumefanthrine") -> malaria_data
+
+train <- monthly_issues_all |>
+  filter_index(. ~ as.character(max(monthly_issues_all$month)-(round(percentage_test*length(unique(monthly_issues_all$month))))))
+
+test <- monthly_issues_all |> 
+  filter_index(as.character(max(monthly_issues_all$month)-round(percentage_test*length(unique(monthly_issues_all$month)))+1) ~ .)
+
+
+
+#Time series cross validation. Create different rolling origin series (i.e. id)
+train_tscv <- monthly_issues_all |> 
+  filter_index(. ~ as.character(max(monthly_issues_all$month)-(f_horizon))) |>  
+  stretch_tsibble(.init = length(unique(train$month)), .step = 1)
+
+
+test_tscv <- test |> 
[…]

### 2024-10-09 rscript/forecasting_model_without_predictor.R

@@ -1,0 +1,74 @@
+install.packages("urca") # installed packages 
+
+library(fpp3)
+library(readr)
+library(tictoc)
+library(distributional)
+
+# Read data
+monthly_issues_tsibble <- read_rds("data/monthly_issues_tsibble.rds") |> filter_index("2017 Aug" ~ .)
+
+# we split the data into train an test
+
+f_horizon <- 6 #  forecast horizon = 6 months
+percentage_test <- 0.3 #20% of time series for test set
+
+
+
+train <- monthly_issues_tsibble |>
+  filter_index(. ~ as.character(max(monthly_issues_tsibble$month)-(round(percentage_test*length(unique(monthly_issues_tsibble$month))))))
+
+test <- monthly_issues_tsibble |> 
+  filter_index(as.character(max(monthly_issues_tsibble$month)-round(percentage_test*length(unique(monthly_issues_tsibble$month)))+1) ~ .)
+
+
+
+#Time series cross validation. Create different rolling origin series (i.e. id)
+train_tscv <- monthly_issues_tsibble |> 
+  filter_index(. ~ as.character(max(monthly_issues_tsibble$month)-(f_horizon))) |>  
+  stretch_tsibble(.init = length(unique(train$month)), .step = 1)
+
+
+# Specify and train models
+
+model_tscv <- train_tscv |>
+  model(
+    mean = MEAN(quantity_product_issued),#total average
+    naive = NAIVE(quantity_product_issued),#naive
+    snaive = SNAIVE(quantity_product_issued),#seasonal naive
+    exponential_smoothing = ETS(quantity_product_issued),#exponential smoothing naive
+    arima = ARIMA(quantity_product_issued),#ARIMA,
+    regression = TSLM(quantity_product_issued ~ trend() + season()),#regression
+  )
+
+write_rds(model_tscv,"result/ae_model_tscv.rds")
+
+# ae_model_tscv <- read_rds("result/ae_model_tscv.rds")
+# Produce forecast
+
+ae_fcst_tscv <-  model_tscv |> forecast(h=f_horizon)
+
+forecast_truncated <-  ae_fcst_tscv |> 
+  mutate(quantity_product_issued= dist_truncated(quantity_product_issued, lower = 0))
+ae_fcst_tscv1 <- forecast_truncated |> hilo(level = 95) |> unpack_hilo(`95%`)
+write_rds(forecast_truncated,"result/forecast_truncated.rds")
+
+#Evaluate forecast accuracy
[…]

### 2024-10-09 rscript/item_level_forecast.R

@@ -1,0 +1,118 @@
+library(distributional)
+library(fpp3)
+library(tidyverse)
+library(dplyr)
+library(tsibble)
+
+
+f_horizon <- 12 #  forecast horizon = 12 months
+percentage_test <- 0.3 #20% of time series for test set
+monthly_issues_all <- read_rds("data/monthly_issues_all.rds")
+item_selected <- "Metronidazole - 250mg - Capsule" 
+# [1] "Adrenaline (Epinephrine) - 0.1% in 1ml Ampoule - Injection"        
+# [2] "Albendazole - 400mg - Tablet"                                      
+# [3] "Amlodipine - 5mg - Tablet"                                         
+# [4] "Amoxicillin - 500mg - Capsule"                                     
+# [5] "Anti-Rho (D)"                                                      
+# [6] "Artemether + Lumefanthrine"                                        
+# [7] "Atenolol - 50mg - Tablet"                                          
+# [8] "Atrovastatin - 20mg - Tablet"                                      
+# [9] "Ceftriaxone"                                                       
+# [10] "Ciprofloxacin - 500mg - Tablet"                                    
+# [11] "Dextrose"                                                          
+# [12] "Ferrous + Folic Acid"                                              
+# [13] "Frusemide - 10mg/ml in 2ml Ampoule - Injection"                    
+# [14] "Gentamicin"                                                        
+# [15] "Hydralazine - 20mg/ml in 1ml Ampoule - Injection"                  
+# [16] "Insulin Isophane Human(Suspension)"                                
+# [17] "Insulin Soluble Human"                                             
+# [18] "Insuline Isophane Biphasic"                                        
+# [19] "Lamivudine  + Efavirenz +Tenofovir"                                
+# [20] "Lamivudine + Zidovudine"                                           
+# [21] "Lidocaine HCL"                                                     
+# [22] "Magnesium Sulphate"                                                
+# [23] "Medroxyprogesterone"                                               
+# [24] "Metformin - 500mg - Tablet"                                        
+# [25] "Metronidazole - 250mg - Capsule"                                   
[…]

### 2024-10-09 manuscript/02-literature.qmd
Revised sections

@@
 # Research background {#sec-lit}
-# focus on forecasting on pharmaceuticals products 
-# 2 pages 
-# how forecasting is done 
-# what other researches had done 
-# what are the limitation of the other papers
-# 15 relevant papaers
-
-# 02- literature
 
@@
 Table 1. Summary of Some Studies on Forecasting in pharmaceuticals products
-
----
-title: "Table 1. Summary of Some Studies on Forecasting in pharmaceuticals products"
-tbl-cap-location:top
-
-```{r}
-
-library(knitr)
-data <- data.frame(
-  Header1 = c("Row 1", "Row 2", "Row 3"),
-  Header2 = c("Data 1", "Data 2", "Data 3"),
-  Header3 = c("Data 1", "Data 2", "Data 3")
-)
-kable(data, format = "html")
-
-```
-
- |                       |           |               |             |          | References       |  | Year      | Variable      |             |
- | :-----------           |-----------| ------------- |-----------  |
- | Siddiqui et al            |           |               |             | 
- |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               | |
- |            |           |               |
- |            |           |               |         
- |            |           |               |
- |            |           |               |
- |            |           |               |
- |            |           |               |
- |            |           |               |
- |            |           |               |
- |            |           |               |
- |            |           |               |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |          
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           ||            |           |
- | -----------|-----------|
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |          
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |          
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |           |
- |            |
- |            |
- |            |
- |            |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |
- |

### 2024-10-09 manuscript/03-experiment.qmd
Revised sections

@@
 
-# you need to talk about the data set 
-# some visualization 
-# make reduce to paragaphs minimize it 
-# forecast accuracy measures provid ethe formulas
+## Data {#sec-data}
 
-## Data {#sec-data}
-# 2.1. Dataset
+### Dataset
+
 For this study, we utilized a dataset spanning five years (2018-2022) of sales data obtained from EPSS.
@@
 
-# Visualising monthly time series
-
-## Title
+## Visualising monthly time series
 

### 2024-10-09 manuscript/04-results.qmd
Revised sections

@@
 # Results and discussion {#sec-results}
+
 ```{r}
@@
 ```
+
 from the total of 37 key pharmaceuticals selected and tested with the  different forecasting models arima showed that a better forecast accuracy for most of the pharmaceuticals followed by mean and  regresstion models with RMSSE value of 0.850, 0.942 and 0.945 respectively (Table 1).
@@
 ```
+
 Then we tested the models by adding different predictors which affects the sales of the pharmaceuticals like physical count, budget release and budget closure by the government, COVID 19, campaign in to the models as predictors which will affects the sales of the key pharmaceuticals products and the result showed that still arima have a better forecast accuracy with RMSSE of 0.850 (Table 2).
+
 ```{r}
@@
 ```
+
 Individual item with different forecasting models 
@@
 ```
+
 Those pharmaceuticals who had a better forecast accuracy with Arima predictors are Amoxacillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine, Atenolol, Atrovastatin 200mg, Ferrosu sulphate with follic acid, frusemide injection, Insuline Iosphane bipahsic, Lamivudine +Efaverienze + Tenofovir, Lidocaine + HCL and Propyl thiro uracil.The  six pharmaceuticasl which showed a better forecast accuracy with naive models were Ciprofloxacillin 500mg tablet, Magnesoum sulphate injection, Omeprazole capsule, Omeprazole injection, Rapid diagnostic test, Sulphamethaxone + Trimetoprine tablet.

### 2024-10-09 manuscript/05-conclusion.qmd
modify

@@
 # Conclusion {#sec-conclusion}
+
 In this study, we embarked on a comprehensive analysis of pharmaceutical demand forecasting within the Ethiopian 

### 2024-10-09 manuscript/main.tex
modify

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
\mathtoolsset{showonlyrefs}
\setlength{\parindent}{0cm}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Journal of}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Pharmaceutical Supply Chain Forecasting by using Machine Learning in Ethiopian Pharmaceutical Supply Services},
  pdfauthor={Arebu Issa Bilal; Arebu Issa Bilal; Bahman Rostami-Tabar; Umit Sezer Bititci; Teferi Gedif Fenta},
  pdfkeywords={Pharmaceutical Demand Forecasting, Ethiopian
Pharmaceutical Supply Services, ARIMA Model, Forecasting
Variables, Forecast Accuracy},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\begin{document}

\begin{frontmatter}
\title{Pharmaceutical Supply Chain Forecasting by using Machine Learning
in Ethiopian Pharmaceutical Supply Services}
\author[1]{Arebu Issa Bilal%
\corref{cor1}%
}
 \ead{arebu.issa@aau.edu.et} 
\author[]{Arebu Issa Bilal%
%
}
 \ead{arebu.issa@aau.edu.et} 
\author[2]{Bahman Rostami-Tabar%
%
}
 \ead{rostami-tabarb@cardiff.ac.uk} 
\author[3]{Umit Sezer Bititci%
%
}
 \ead{U.S.Bititci@hw.ac.uk} 
\author[1]{Teferi Gedif Fenta%
%
}
 \ead{tgedif@gmail.com}

\affiliation[1]{organization={Addis Ababa University, Department of
Pharmaceutics and Social Pharmacy at School of
Pharmacy},country={Ethiopia},countrysep={,},postcode={9086},postcodesep={}}
\affiliation[2]{organization={University, Cardiff Business
School},country={UK},countrysep={,},postcode={CF5},postcodesep={}}
\affiliation[3]{organization={University, Edinburgh Business
School},country={UK},countrysep={,},postcode={EH14 4AS},postcodesep={}}

\cortext[cor1]{Corresponding author}

\begin{abstract}
This study delves into the realm of pharmaceutical demand forecasting
within the Ethiopian Pharmaceuticals Supply Services(EPSS) context. It
encompasses a comprehensive analysis of 37 key pharmaceutical
products,assessing their forecast accuracy across variation models. The
initial assessment reveals that the Autoregressive Integrated Moving
Average (ARIMA) model emerges as a frotrunner, boasting a Root Mean
Squared Error (RMSSE) value of 0.850. ARIMA's prowess in capturing
complex demand patterns suggests its suitability for pharmaceutical
sales forecasting in the EPSS landscape. To argument forecasting
accuarcy, predictor varaibles such as physical count, budget release and
closure by the goverment, the impact of COVID-19, and promotional
campaigns were incorpoarted. Remarkably, the ARIMA model mantainsits
superiority even in the presences of these external factors, reiterating
its roustness. Delving further into individual pharmaceutical products,
ARIMA with predictors demonstrates exceptional performance for a
substantial portion (32.43\%) of the products, including Amoxicillin
500mg capsule, Anti RhD, Artehmether + Lumefantrine, and others.
Conversely,alternative models like naive, mean, and regression methods
showcase strengths for different pharmaceutical categories, underscoring
the importance of tailoring forecasting approaches to product-specific
dynamics. While this study significantly advances the discourse on
pharmaceutical demand forecasting, limitations stemming from data
quality and external variables underscore the intricacies of the
forecasting landscape. In conclusion, it advocates the integration of
the ARIMA model with predictor variables as a strategic approach for
refining pharmaceutical demand forecasting within EPSS. It also
encourages further exploration of advanced forecasting techniques and
ongoing refinement to bolster supply chain efficiency and ensure
consistent pharmaceutical availability. Capacity building through staff
training and replication of the study in diverse healthcare settings are
recommended steps toward resilient and efficient pharmaceutical supply
chains.These findings hold promise for fostering supply chains that are
adaptable, responsive, and capable of ensuring the continuous
availability of essential pharmaceuticals in Ethiopia.
\end{abstract}

\begin{keyword}
    Pharmaceutical Demand Forecasting \sep Ethiopian Pharmaceutical
Supply Services \sep ARIMA Model \sep Forecasting Variables \sep 
    Forecast Accuracy
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

Universal access essential medicines is a cornerstone of effective
healthcare systems, yet ensuring equitable availability of these vital
medications remains an urgent global
challenge\citep{quick2003essential, world2004annual}. Across continents,
especially in Africa and Asia, many nations grapple with restricted
access due to factors like unaffordability and suboptimal supply chain
management \citep{world2004medicines}. A survey conducted across eight
sub-Saharan African countries indicated that there were unacceptably low
availability of essential medicines.The mean availability of 12 priority
essential medicines for women ranged from 22\% to 40\% and from 28\% to
57\% for children \citep{droti2019poor}. However, the issue of drug
shortages extends beyond developing countries, with instances reported
in the United States and Europe, adversely impacting healthcare systems
and compromising patient care quality
\citetext{\citealp[\citet{kaakeh2011impact}]{fox2003managing}; \citealp{johnson2011drug}; \citealp{huys2013european}; \citealp{le2011prevalence}}.

The consequences of drug shortages are felt both financially within the
healthcare sector and in terms of the quality of patient care
\citep{alspach2012drug, kaakeh2011impact, kaakeh2011impact, baumer2004national}.The
consequences extend across patients, providers, healthcare institutions,
and research programs. Patients face delays or cancellations of
treatments, substandard care due to the unavailability of optimal
therapies, and increased costs in a secondary ``grey market.''
Compromised patient safety, complications, prolonged hospital stays, and
even deaths have been reported {[}\citet{alspach2012drug};{]}.Addressing
these shortages is crucial to preserving the quality of patient care and
ensuring the effectiveness of healthcare delivery. Thus, a smoothly
operating pharmaceutical supply chain is essential to ensure timely
access to essential medicines, making accurate demand forecasting an
indispensable component.

An efficient pharmaceutical supply chain is vital for timely access to
essential health commodities. One of the most crucial components of such
a supply chain is a well-defined demand forecasting process. Forecasting
accurately the future demand for essential medicines provides valuable
insights for anticipating future healthcare requirements including
procurement and replenishment of
inventory\citep{subramanian2021effective}. Despite the significance of
forecasting for supply chain management, pharmaceutical product
forecasting remains a complex endeavor influenced by a multitude of
factors. Key considerations include the availability of data, the
specific position within the supply chain, and the nuanced understanding
of contributing elements. The position within the supply chain
introduces unique challenges and dynamics, requiring tailored
forecasting approaches. Additionally, a nuanced understanding of
contributing elements involves grasping intricate factors that impact
forecasting accuracy, such as market trends, regulatory changes, and
external influences. Schneider, Wilson, and Rosenbeck (2010) and Hyndman
and Athanasopoulos (2018) highlight the intricate nature of
pharmaceutical forecasting, emphasizing the need for a comprehensive
approach that incorporates these multifaceted aspects to enhance supply
chain management
effectiveness\citep{schneider2010pharmaceutical, hyndman2018forecasting}.

The Ethiopian pharmaceutical sector's distinct characteristics, marked
by a diverse array of medicines and unique data challenges stemming from
service delivery points, staff capacity limitations, communication
hurdles, prolonged procurement and distribution lead times, poor
forecast accuracy, and policy intricacies, necessitates tailored
forecasting strategies that cater to the nation's specific context
\citep{boche2022procurement}. In this context, a comprehensive
exploration of effective forecasting models that align with Ethiopia's
pharmaceutical landscape is paramount to mitigating stock outs and
enhancing healthcare delivery. The implications of pharmaceutical demand
forecasting extend beyond financial considerations to societal
well-being \citep{rostami2022forecasting}. In Ethiopia, a country
undergoing demographic and healthcare evolution, precise demand
forecasting assumes immense importance for resource allocation, supply
chain optimization, and policy formulation. With mounting interest in
refining demand forecasting in Ethiopia's pharmaceutical sector, this
study seeks to contribute to existing knowledge by investigating and
recommending proficient demand forecasting methods. Our investigation
centers on the application of mean, naive, exponential smoothing,
regression, and ARIMA models to bolster forecast accuracy and
effectiveness. This paper aims to forecast the demand for selected
pharmaceuticals in EPSS over the next 12 months by utilizing various
forecasting models and integrating predictors. The integration of
predictors with the models will be examined to determine whether they
enhance the forecasting accuracy. These predictors are derived from a
historical analysis of events in Ethiopia over the last 5 years that
have impacted pharmaceutical sales in EPSS. These events include
COVID-19, conflicts in the country, disease campaigns, budget closures,
budget releases, and physical inventory, all of which, in varying
degrees, influence the demand for pharmaceuticals in Ethiopia.

The current paper offers the following contributions we aim at
forecasting pharmaceutical sales over a 12-month period these models aim
to enhance the accuracy and reliability of demand predictions within the
context of EPSS. Moreover an essential contribution lies in the
identification of key factors influencing pharmaceutical sales. The
study develops a model that effectively incorporates these driving
factors, providing a more nuanced and comprehensive approach to demand
prediction. We generate forecasts and evaluate model performance using
the Root Mean Squared Error (RMSE), Root Mean Squared Scaled Error
(RMSSE) and Mean Absolute Error (MAE), Mean Absolute Scaled Error
(MASE), Crips and Winkler are used to assess probabilistic forecast
accuracy. \#how you know the forecasting accutacy problem and provide
eveidence about the forecasting problems \# most forecast focus on point
forecast and we want to focus on probablistic forecast \# i will write
the contribution of the reearch rather than these objectives

This research is guided by several objectives:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Evaluate the robustness of various forecasting models for
  pharmaceutical demand prediction within EPSS.
\item
  Identify pharmaceuticals that exhibit superior forecast accuracy under
  specific models.
\item
  Assess the performance of forecast accuracy metrics to gauge the
  efficacy of selected models.
\item
  Test the incorporation of diverse predictors affecting forecasts and
  evaluate resulting forecasting accuracy. In pursuing these objectives,
  this study endeavors to elevate the precision and reliability of
  pharmaceutical demand forecasting, thereby contributing to optimized
  supply chain management and enhanced healthcare outcomes in Ethiopia.
\end{enumerate}

\hypertarget{sec-lit}{%
\section{Research background}\label{sec-lit}}

Precise sales prediction is an essential and inexpensive way for each
company to augment their profits, decrease their costs, and achieve
greater flexibility to changes. In other words, exact sales forecasting
is utilized for capturing the tradeoff between customer demand
satisfaction and inventory costs \citep{gupta2000mid}. Especially, for
the pharmaceutical industry, successful sales forecasting systems can be
very beneficial, due to the short shelf-life of many pharmaceutical
products and the importance of the product quality which is closely
related to the human health \citep{doganis2006time}.There are many
decisions that depend on the quality of forecasts in the health care
system, from capacity planning to layout decisions to the daily
schedules. In general,the role of forecasting in health care is to
inform both clinical and non-clinical decisions. While the former
concerns decisions related to patients and their treatments
\citep{makridakis2019forecasting}, the latter involves
policy/management, and supply chain decisions that support the delivery
of high-quality care for patients.In the coming paragarphs we will
review what other researchers had done in the area of pharmaceuticals
forecasting.Morover today,manufacturing companies are organized
following the principles of just-in-time (JIT) delivery which focuses on
efficiency and eliminating the waste of excessive inventory
\citep{van2021using}. In JIT systems, accurate Stock Keeping Unit (SKU)
short-term demand forecasts, typically on a daily, weekly or monthly
basis, are crucial as they are the fundamental inputs to drive the
supply chain planning system.Therefore, forecasting performance can
strongly affect a company's financial results, since both under- and
overestimating short-term demand are expensive as they result in supply
shortages and consequently poor customer service, and excess inventories
and product obsolescence, respectively (Sanders and Graman, 2009).In
this section, we provide a brief review of studies on forecasting
harmaceuticals products.In conclusion, the research findings indicated
that the hybrid neural network approach, which combines linear and
nonlinear modeling techniques, outperformed traditional methods like
ARIMA and standalone neural networks in forecasting pharmaceutical
sales. The hybrid approach demonstrated the ability to capture both
linear and nonlinear patterns in sales data accurately, especially in
scenarios with limited historical records for individual drugs.

A study by Khalil Zadeh et al.~developed an intelligent sales prediction
model for Pharmaceutical Distribution Companies (PDCs) by combining
network analysis tools and time series forecasting methods. This
approach addresses the challenge of limited historical sales data for
pharmaceutical products, making traditional forecasting methods
inadequate\citep{zhang2004neural}.The researchers employed Mean Squared
Error (MSE) and Mean Absolute Error (MAE) to evaluate the performance of
various forecasting models, including ARIMA, neural networks, and a
hybrid approach combining both. The hybrid model, inspired by Zhang's
work\citep{zhang2004neural}, used ARIMA for linear components and neural
networks for nonlinear components.Three methods were used: (1) ARIMA,
(2) a hybrid neural network with each drug's past records, and (3) a
hybrid neural network with both each drug's and its group members' past
records. Grouping products and using group members' records
significantly improved forecasting accuracy by addressing the issue of
limited data and leveraging both linear and nonlinear modeling
capabilities.The study concluded that the hybrid approach, incorporating
both individual and group past records, performed better than using
either method alone \citep{khalil2014intelligent}.

A study by Van Bell et al.~used various forecasting models for point of
sale data to assess the impact of including sell-through data on
forecast accuracy. They compared information sharing (IS) methods with
non-information sharing (NIS) methods using the following approaches.
Substitution Approach (ETS-W and ARIMA-W): These models use downstream
wholesaler sales time series as input instead of prior manufacturer
shipments, suitable when all goods pass through wholesalers. Integration
Approach (Extrapolative Time Series Methods with Exogenous Variables):
This includes methods that integrate downstream information using
extrapolative time series techniques and machine learning (ML)
approaches. Extrapolative Time Series Methods (ETS and ARIMA):
Traditional forecasting methods based on modeling past time series
structures and extrapolating into the future. The study evaluated these
methods using Mean Squared Error (MSE) and Mean Absolute Error (MAE).
Additionally, the researchers used naive methods, IS and NIS with ETS
and ARIMA models, and machine learning techniques like LASSO, Support
Vector Regression (SVR), and Random Forest (RF).LASSO (Least Absolute
Shrinkage and Selection Operator) was identified as the best-performing
method for improving forecast accuracy, particularly for one-step ahead
forecasts. It outperformed other methods for longer horizons (h
\textgreater{} 1) by effectively selecting the most relevant variables
and reducing overfitting. ETS was found to be the most effective
univariate baseline method, and SVR showed the best performance for h =
1. The study concluded that IS methods generally outperformed NIS
methods, indicating that incorporating downstream information, such as
sell-through data, can significantly enhance short-term forecast
accuracy in supply chain management. Machine learning techniques like
LASSO, SVR, and RF faced challenges related to overfitting and model
complexity, especially with high-dimensional data. Balancing model
complexity and interpretability was crucial for robust forecasting.
Overall, the research suggests that LASSO, along with ETS and SVR, are
effective for improving demand prediction accuracy in multi-echelon
supply chains by leveraging downstream information and collaborative
approaches\citep{van2021using}.

A study by Nikolopoulos et al.~forecasted UK pharmaceutical time series
before and after patent expiry, a critical point when a generic version
enters the market. The study used various models, including naïve and
moving averages, the Bass diffusion model, repeat purchase diffusion
models, three regression models, and two exponential smoothing (ES)
approaches for non-trended (simple ES) and trended data (Holt ES)
\citep{nikolopoulos2016forecasting}.Holt Exponential Smoothing: This
model was the most accurate for forecasting both branded and generic
drugs across short-term, mid-term, and long-term horizons.Naïve Model:
Used as a benchmark, the relative absolute error (RAE) was calculated to
compare forecasting accuracy.Bass Diffusion Model: Although popular in
marketing research, it had limitations in accurately forecasting
pharmaceutical time series due to the complex dynamics between branded
and generic drugs. Comparative Analysis: The study compared various
forecasting models to evaluate their performance, highlighting the
importance of selecting appropriate models based on the data and
context.The research found that Holt's exponential smoothing model
outperformed others, including naïve models, moving averages, and ARIMA
models, especially for short-term forecasts. However, forecasting
accuracy varied across different horizons. Despite their limitations,
diffusion models were included to provide additional insights.Overall,
the study emphasized the strengths and limitations of different
forecasting approaches in the context of pharmaceutical time series,
with Holt's exponential smoothing model demonstrating superior
performance\citep{nikolopoulos2016forecasting}.

Astudy by Siddiqui et al used to develop a combined forecasting
model.The combination is developed using two frequent, but separately
used models, i.e.~Autoregressive Integrated Moving Average (ARIMA) model
and Holts Winter model and a combined model were proposed which is known
as the ARHOW model.The ARIMA model is commonly used for time series
forecasting, while Holt's Winter model is effective for handling
seasonality in data \citep[markovska2016comparative]{inproceedings}.The
ARHOW model aims to enhance forecasting accuracy by leveraging the
strengths of both ARIMA and Holt's Winter models. The model equation for
obtaining a consolidated forecast is as follows: ARHOW = β *
ArimaForecast + β1 * HoltWinterForecast In this equation: ARHOW
represents the Combined Weighted Forecast. ArimaForecast and
HoltWinterForecast are the forecasts obtained by applying the ARIMA and
Holt's Winter models over the training set.β and β1 are the weights
assigned to the ARIMA and Holt's Winter forecasts, respectively. By
combining the forecasts generated by these two models, the ARHOW model
seeks to mitigate the impact of volatility and external factors on
forecasting accuracy, ultimately providing more reliable predictions for
pharmaceutical sales data \citep{siddiqui2022hybrid} . Combining
multiple forecasting techniques like ARIMA and Holt's Winter to create
the ARHOW model introduces complexity. Managing and interpreting the
results of a hybrid model may pose challenges in practical
implementation and decision-making processes .Morover External factors
such as regulatory changes, technological advancements, and market
dynamics can introduce forecasting errors. While the ARHOW model aims to
improve accuracy, it may still be susceptible to errors resulting from
unforeseen external influences \citep{siddiqui2022hybrid}.The key
forecasting metrics employed in the study include:

Mean Absolute Percentage Error (MAPE), Mean, Mean Absolute Deviation
(MAD), Mean Square Error (MSE), Root Mean Square Error (RMSE).The Mean
Absolute Percentage Error (MAPE) and other forecasting metrics suggest
that the ARHOW model provides more accurate predictions compared to
traditional forecasting techniques like ARIMA, Holt's Winter, ETS,
Theta, and Naïve models. The model's efficiency in forecasting
pharmaceutical demand can help manufacturers plan inventory and prevent
stockouts or overstock situations .In conclusion, the study underscores
the effectiveness of the ARHOW model in improving demand forecasting
accuracy in the pharmaceutical industry and provides insights for
enhancing supply chain efficiency through advanced forecasting
techniques .

A study by de Oliveira et al.~compared five machine learning regression
approaches for lead time forecasting in the pharmaceutical supply chain.
The models used were k-nearest neighbors (KNN), support vector machines
(SVM), random forests (RF), linear regression (LR), and multi-layer
perceptrons (MLP). These models aimed to predict supplier delivery
times, enhancing planning for production and logistics services in the
pharmaceutical field \citep{de2021lead}. The study focused on using
machine learning techniques to forecast lead times for purchasing
products in the pharmaceutical industry. The comparison aimed to
identify the most effective approach for predicting purchasing lead
times, demonstrating high accuracy potential, especially with linear
support vector regression. Key limitations noted included computational
complexity, sensitivity to outliers, difficulty in interpretability, and
overfitting. Using mean squared error (MSE) as the forecast accuracy
metric, the study found that the SVM approach achieved the best
performance, with an average error of less than two days. This indicates
that SVM was the most effective model among those tested for forecasting
purchasing lead times in the pharmaceutical supply
chain\citep[de2021lead]{inproceedings}.

A study by Rathipriya et al.~focused on demand forecasting for
time-series pharmaceutical data using both shallow and deep neural
network models\citep{rathipriya2023demand}. The researchers employed
several models, including the Autoregressive Integrated Moving Average
(ARIMA) model, known for its simplicity and ability to handle
non-stationary series.The shallow neural network models used were,
Radial Basis Function Neural Network (RBF\_NN),Generalized Regression
Neural Network (GR\_NN),Probabilistic Neural Network (P\_NN),For deep
neural networks, the study investigated:Long Short-Term Memory
(LSTM),Stacked LSTM,Additionally, the study explored hybrid models,
combining ARIMA with the Adaptive Neural Fuzzy Inference System (ANFIS)
to forecast Iran's annual energy usage, demonstrating higher accuracy
compared to individual models. The researchers also employed methods
like Genetic Algorithms (GA), Bayesian structures, and Reinforcing
Artificial Neural Networks (RANN) for forecasting short-run demand.The
performance of the demand forecast models (DFMs) was evaluated using
Root Mean Squared Error (RMSE). The study found that shallow neural
network models, such as GR\_NN, RBF\_NN, and P\_NN, had almost identical
mean RMSE values across all eight drug categories, indicating their
effectiveness in predicting pharmaceutical demand. Conversely, deep
neural network models like LSTM and Stacked LSTM did not perform as well
due to the limited number of samples available for training. The
research concluded that shallow neural network-based DFMs performed well
for five out of eight drug categories, while the ARIMA model was best
for the remaining three. This suggests that no single forecasting model
is optimal for all drug categories\citep[rathipriya2023demand]{article}.

A study by Candan et al.~on demand forecasting in the pharmaceutical
industry utilized a neuro-fuzzy approach as the forecasting
method\citep[hrupelj2017demand]{inproceedings}. Neuro-fuzzy modeling
combines information from empirical models, heuristics, and data,
providing effective models that explain solutions better than black-box
models like neural networks\citep{abraham2001neuro, babuvska2003neuro}.
The study also explored the emotional learning fuzzy inference system
(ELFIS) and locally linear neuro-fuzzy (LLNF) methods proposed by
Abdollahzade et al.~(2012) for predicting the stock prices of a
pharmaceutical
company\citep{abdollahzade2012application, hrupelj2017demand}. While the
neuro-fuzzy approach combines fuzzy logic and neural networks to improve
forecasting, it may still struggle with extremely complex or non-linear
relationships within the data. The study used paired t-tests and mean
difference analysis to evaluate forecast accuracy. Overall, the metrics
supported the conclusion that the neuro-fuzzy approach is a robust
method for demand forecasting in the pharmaceutical industry, providing
accurate predictions closely aligned with actual demand quantities.

A study by Burinskiene discusses various forecasting techniques in the
pharmaceutical retail industry. These include: Single Exponential
Smoothing: Applies weighted averages with exponentially decreasing
weights for older observations. Moving Average: Used for trend-based
estimation to predict demand and standard deviation. Naive Method:
Assumes all predictions are equal to the last observation, serving as a
benchmark. Holt's Linear Method: An extension of single exponential
smoothing for trend prediction. Holt-Winters No Trend Method: A
variation without a trend component\citep{burinskiene2022forecasting}.
The study focuses on short-term forecasting, excluding long-term
forecasting and safety stock considerations, which are crucial in the
pharmaceutical industry due to varying demand
patterns\citep{nikolopoulos2016forecasting}. The naive method fails to
capture trends or seasonality, while the moving average method struggles
with sudden changes and irregular patterns\citep{restyana2021analysis}.
Holt's linear method is limited in handling complex seasonal patterns,
and the Holt-Winters method requires careful parameter tuning and
historical data analysis\citep{koehler2001forecasting}. Single
exponential smoothing is not effective for sudden changes or
outliers\citep[zahra2019forecasting]{inproceedings}. The study used
MAPE, MSE, and Theil's U2 test to evaluate forecast accuracy, with
Theil's U2 test comparing results to the naive forecast. A U2 value
greater than 1 indicates better performance than the naive method
\citep{burinskiene2022forecasting}. The Holt-Winters Multiplicative
method was found to improve weekly forecasts and outperform other
methods, especially in addressing product shortages in pharmacies.

A study conducted by Anusha et al.~in the Indian pharmaceutical retail
sector utilized forecasting models such as Simple Moving Average,
Exponential Smoothing, and Winter's Exponential
Smoothing\citep{anusha2014demand}. These models were employed to predict
demand for both seasonal pharmaceutical items like Okacet 10mg tablets
and random pharmaceutical items like Stamlo Beta tablets. Quantitative
indicators like Mean Absolute Deviation (MAD), Mean Squared Error (MSE),
and Mean Absolute Percentage Error (MAPE) were used to assess forecast
error. The Winter's Exponential Smoothing (WES) model showed the least
sales forecast error for the seasonal pharmaceutical item Okacet 10mg
tablets, with a MAPE of 27.50, indicating high forecast accuracy.
Conversely, for the non-seasonal pharmaceutical item Stamlo Beta
tablets, the Six-month Moving Average model performed best with a MAPE
of 45.62\citep{anusha2014demand}. However, the study did not account for
unexplained seasonal variations in staff, shifts in the number of
potential patients, physician turnover, preferences for pharmaceuticals,
and the influence of marketing representatives on surrounding clinics.
These factors were not factored into the analysis, potentially affecting
the accuracy of demand forecasting in the pharmaceutical retail setting.
These limitations suggest that the study may not have considered all
relevant variables impacting demand forecasting accuracy.

Zhu et al.~conducted a study comparing various forecasting methods in
the pharmaceutical industry, including Moving Average (MA), Linear
Regression (LR), Exponential Smoothing (ES), Exponential State Space
Model (ETS), Vector Autoregressive with exogenous variables (VARX),
Symbolic regression, Graph-based analysis, Artificial Neural Networks
(ANN), Clustering, and Recurrent Neural Networks
(RNN)\citep{zhu2021demand}. Additionally, machine learning algorithms
such as Linear Regression (LR), Support Vector Regression (SVR), Random
Forest (RF), and RNN with cross-series training were assessed.The study
evaluated forecast accuracy using metrics like Normalized Mean Squared
Error (NMSE), Normalized Mean Absolute Error (NMAE), and Bias. Results
indicated that the cross-series training models outperformed baseline
models (Moving Average, Exponential Smoothing, Linear Regression) in
terms of NMSE, NMAE, and Bias. RNN models, particularly when implemented
with cross-series training, showed superiority and generalizability
across datasets compared to other machine learning algorithms (LR, SVR,
RF), confirming the effectiveness of the proposed forecasting framework
in the pharmaceutical industry.

Ribeiro et al.~conducted a study aimed at detecting outliers and
predicting sales for a pharmaceutical distribution company. They
employed the dampened Pegels method, a time-series data mining
technique, to forecast future quantities of medicines to be ordered by
the company\citep[ribeiro2017improving]{article}. The accuracy of the
forecasting model was evaluated using the Symmetric Mean Absolute
Percentage Error (SMAPE) metric, which indicated that the Pegels method
provided more accurate sales forecasts compared to the company's
previous method. Furthermore, these forecasts effectively met demand,
aiding in efficient stock management and ensuring customer satisfaction
while minimizing inventory costs.

The study by Bon and Ng at the University Health Centre aimed to enhance
inventory demand forecasting for Panadol 650mg through the utilization
of various forecasting techniques. Their research design encompassed
both qualitative and quantitative data collection methods within a
descriptive framework. Historical data spanning from January 2009 to
August 2014 were analyzed employing methods like single moving average,
exponential smoothing, and regression analysis. Forecast accuracy was
assessed using metrics including RMSE, MSE, MAD, and MAPE to identify
the most precise forecasting technique. The objective was to optimize
inventory management by selecting the most accurate forecasting method
based on these measurements\citep{bon2017optimization}. Results
highlighted regression analysis as the superior technique among the ten
methods tested, exhibiting the least forecasting error for optimizing
Panadol 650mg inventory demand at the University Health
Centre\citep[bon2017optimization]{inproceedings}.

The study conducted by Newberne and MIKE focused on utilizing
Holt-Winters Forecasting in Healthcare, particularly in the context of
healthcare management involving Pseudoephedrine prescriptions and Upper
Respiratory Infection (URI) visits \citep{newberne2006holt}. Data from
the Military Healthcare System's M2 database were analyzed using the
Holt-Winters method to demonstrate its applicability in healthcare
forecasting. Forecasting accuracy was assessed using the Mean Absolute
Percentage Error (MAPE) metric. The study employed various forecasting
methods including Naïve, Seasonal Naïve, ARIMA, and Deep Learning Model:
Long Short-Term Memory (LSTM) Recurrent Neural Network. Additionally,
techniques like Simple Exponential Smoothing, Holt with linear and
damped trends, Holt-Winters, ES state space model, and Theta were
explored to forecast drug quantities for distribution in the Brazilian
Public Health System. Among these methods, the LSTM model often
exhibited the best performance based on Mean Absolute Error (MAE) as a
measure of forecast accuracy. However, deep learning models like LSTM
may pose challenges due to their complexity and computational intensity,
requiring specialized expertise for tuning and implementation in
practical settings.

The study conducted by Zahra and Putra focused on using ARIMA and Single
Exponential Smoothing methods to predict and calculate EOQ (Economic
Order Quantity). To measure forecast accuracy, the researchers utilized
RMSE, MAPE, MAD, and MSE metrics. The findings revealed that the ARIMA
(1,0,0) method exhibited the smallest error values, with a MAPE of 13\%
and MSE of 283,715, indicating its superior accuracy compared to the
Exponential Smoothing method. Based on these results, the researchers
predicted a need for 222 boxes of medicine for the next period and
recommended procuring 223 boxes to maintain a service level of 95\%. The
study underscored the significance of accurate forecasting in optimizing
medicine procurement processes and ensuring adequate stock levels to
efficiently meet demand. By employing MAPE and MSE as forecast accuracy
metrics, the researchers could compare the performance of different
forecasting methods and determine the most effective approach for
predicting medicine needs in hospital
settings\citep{zahra2019forecasting}.

The study conducted by Restyana et al.~focused on analyzing drug
forecasting within the Jombang Regency from 2017 to 2019. They employed
the Single Moving Average and Single Exponential Smoothing methods to
predict drug demand across 35 public health centers. The objective was
to identify the method with the lowest forecasting error by assessing
metrics like Mean Absolute Deviation (MAD) and Mean Square Error (MSE).
By analyzing annual primary data and testing errors, the researchers
aimed to evaluate the accuracy of the forecasting methods. The findings
revealed that the Single Exponential Smoothing (SES) approach exhibited
superior forecast accuracy, as evidenced by lower MAD and MSE values.The
study by Kim et al.~used Vector Autoregressive with Exogenous Variables
(VARX) models for demand forecasting of medicines. Two models were
constructed to test the effect of exogenous variables: Model 1, a pure
autoregressive model without exogenous variables, and Model 2, the
suggested VARX model incorporating topic trends from social media
analysis. The study used sales data as endogenous variables and moving
averages of topic trends as exogenous variables in the VARX models.The
results showed that incorporating topic trends into the VARX models
enhanced the performance of demand forecasting, particularly for
medicines sensitive to environmental and seasonal changes. By analyzing
topic trends related to certain diseases and incorporating them into
VARX models, the researchers found significant improvement in forecast
accuracy.Prediction error rates were used as the forecast accuracy
measure, calculated for four different medicines for each model (Model 1
and Model 2). The comparison of error rates between the two models
demonstrated that adding topic trends as exogenous variables in the VARX
model significantly improved the accuracy of demand forecasts,
highlighting the importance of integrating social media data into
forecasting processes\citep{kim2015demand}.

Table 1 provides a summary of some studies in the literature on
forecasting in pharmaceuticals products. We have identified several
limitations in the existing literature on pharmaceutical product
forecasting that have motivated us to pursue this research. These
limitations can be summarized as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Almost all of the studies have primarily concentrated on point
  forecasts at a single level.
\item
  A majority of the research has not explored longer time series data
  periods.
\item
  Only a few studies have integrated domain knowledge with time series
  data to enhance pharmaceutical forecasting accuracy.
\item
  Additionally, no studies have been conducted in African settings,
  where challenges related to forecasting and pharmaceutical
  availability are prevalent.
\end{enumerate}

Table 1. Summary of Some Studies on Forecasting in pharmaceuticals
products

\hypertarget{sec-experiment}{%
\section{Experiment setup}\label{sec-experiment}}

\hypertarget{sec-data}{%
\subsection{Data}\label{sec-data}}

\hypertarget{dataset}{%
\subsubsection{Dataset}\label{dataset}}

For this study, we utilized a dataset spanning five years (2018-2022) of
sales data obtained from EPSS. Rigorous checks were conducted to ensure
the consistency and completeness of the collected data. From the
extensive pool of pharmaceutical products within EPSS, we strategically
selected a set of 37 key pharmaceuticals, encompassing various programs
and representing different classes of drugs. This deliberate selection
enriches the research by providing a comprehensive representation across
the pharmaceutical landscape. We are presenting the graph for some of
the pharmaceuticals sales for the last five years by using visualization
for the readers to undesrtand the sales of the pharmaceuticals.

\hypertarget{visualising-monthly-time-series}{%
\subsection{Visualising monthly time
series}\label{visualising-monthly-time-series}}

\includegraphics{main_files/figure-pdf/unnamed-chunk-10-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-11-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-12-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-13-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-14-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-15-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-16-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-17-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-18-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-19-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-20-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-21-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-22-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-23-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-24-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-25-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-26-1.pdf}

\includegraphics{main_files/figure-pdf/unnamed-chunk-27-1.pdf}

\hypertarget{data-splitting-and-preparation}{%
\section{2. 2. Data Splitting and
Preparation}\label{data-splitting-and-preparation}}

The primary objective of this study is to forecast pharmaceutical demand
within EPSS for the year 2023. To achieve this, the historical sales
data for the chosen pharmaceuticals in EPSS was partitioned into
distinct training and testing subsets. Employing the widely recognized
70:30 split method, 70\% of the available data was designated as the
training set, while the remaining 30\% was reserved for testing and
validation purposes \citep{chan2019comparison}. As a result, the dataset
was segregated into a training set consisting of 42 months and a testing
set spanning 18 months

\hypertarget{forecasting-methods}{%
\section{2.3.1. Forecasting Methods}\label{forecasting-methods}}

In pursuit of accurate pharmaceutical demand forecasts, we explored a
spectrum of forecasting techniques. The following methods were applied:

Naive Method: A simple approach relying on the assumption that the
future value will be the same as the most recent observed value.This
Methods uses when there is no seasonality
\citetext{\citealp[burinskiene2022forecasting]{article}; \citealp[nikolopoulos2016forecasting]{article}}
Mean Method: A straightforward technique forecasting future values based
on the mean of past observations. Seasonal Naive Method: Similar to the
naive method, this approach predicts future values using the most recent
observed value from the same season. Exponential Smoothing: A method
that assigns exponentially decreasing weights to past observations to
forecast future values. Regression: Employing statistical regression
models to establish relationships between variables for prediction.
Autoregressive Integrated Moving Average Model (ARIMA): A time series
model incorporating autoregressive, differencing, and moving average
components for accurate forecasting. ARIMA with Predictors: Extending
the ARIMA model with additional predictor variables to enhance
forecasting accuracy. Combination of Models: Exploring ensemble methods
that combine multiple models for improved predictive performance. By
applying these diverse forecasting methodologies, we aimed to
comprehensively assess and compare their efficacy in predicting
pharmaceutical demand within EPSS. Each method was evaluated against the
testing dataset to ascertain its forecasting accuracy and suitability
for the specific

\hypertarget{sec-method}{%
\subsection{Forecasting methods}\label{sec-method}}

Model building

\hypertarget{naive}{%
\section{Naive}\label{naive}}

The naive forecast method is a simple approach that assumes future
values will be the same as the most recent observed value, making it
useful for stable or slowly changing time series. While it doesn't
consider underlying patterns, it serves as a baseline for evaluating
more advanced techniques\citep{hyndman2018forecasting}. It's
particularly suitable for products with stable demand or limited
seasonality, providing a straightforward reference for forecasting and
inventory management.

\hypertarget{mean}{%
\section{Mean}\label{mean}}

The mean forecast method predicts future values by averaging historical
data, making it suitable for stable time series. While it overlooks
trends and seasonality, it serves as a baseline for evaluating advanced
techniques \citep{hyndman2018forecasting}. Its simplicity makes it
valuable in certain forecasting applications
\citep{hyndman2018forecasting}.

\hypertarget{seasonal-naive}{%
\section{Seasonal naive}\label{seasonal-naive}}

The seasonal naive forecast method predicts future values based on the
most recent observed value from the corresponding season in the previous
year, making it useful for data with strong seasonal patterns. It serves
as a baseline for evaluating seasonal forecasting models, capturing
repetitive patterns effectively. Its simplicity and ability to capture
seasonality make it valuable in seasonal forecasting
applications\citep{hyndman2018forecasting}.

\hypertarget{exponential-smoothing}{%
\section{Exponential smoothing}\label{exponential-smoothing}}

Exponential Smoothing is regarded as one of the most accurate and robust
methods for time series forecasting. Not only does it exhibit high
accuracy, but it also offers versatility and robustness. What makes
Exponential Smoothing particularly appealing is its intuitive nature,
making it easy to understand and interpret. Furthermore, this method
boasts minimal data storage and computational requirements, making it
highly efficient for real-time applications. In the business realm,
Exponential Smoothing finds extensive use for inventory demand
forecasting, showcasing its practicality \citep{gardner1985exponential}.
Surprisingly, even when pitted against more sophisticated techniques,
Exponential Smoothing has demonstrated impressive performance in
forecasting competitions
\citep[\citet{makridakis1982accuracy}]{makridakis2000m3}. Its simplicity
and reliability make it a widely trusted approach in the field of
forecasting.

\hypertarget{regression}{%
\section{Regression}\label{regression}}

Regression methods play a crucial role in forecasting and optimizing the
health supply chain, ensuring the availability and efficient
distribution of essential medical resources. By analyzing historical
data, regression models can identify patterns and relationships between
variables such as patient demand, disease prevalence, and inventory
levels. The basic concept is that we forecast the time series of
interest assuming that it has a linear relationship with other time
series. For example, we might wish to forecast monthly sales y using
total advertising spend as a predictor X. Or we might forecast daily
electricity y demand using temperature x1 and the day of week x2 as
predictors. The forecast variable y is sometimes also called the regress
and, dependent or explained variable. The predictor variables x are
sometimes also called the regressors, independent or explanatory
variables.

Simple linear regression In the simplest case, the regression model
allows for a linear relationship between the forecast variable y and a
single predictor variable x: yt = β0 + β1xt + ε Example of data from
such a model is shown in Figure 5.1. The coefficients β0 and β1 denote
the intercept and the slope of the line respectively. The intercept β0
represents the predicted value of y when x=0 . The slope represents the
average predicted change in y resulting from a one unit increase in x.

\hypertarget{figure-an-example-of-data-from-a-simple-linear-regression-model}{%
\section{Figure : An example of data from a simple linear regression
model}\label{figure-an-example-of-data-from-a-simple-linear-regression-model}}

Notice that the observations do not lie on the straight line but are
scattered around it. We can think of each observation yt as consisting
of the systematic or explained part of the model, = β0 + β1xt , and the
random ``error'', εt. The ``error'' term does not imply a mistake, but a
deviation from the underlying straight line model. It captures anything
that may affect yt other than xt

Multiple linear regressions When there are two or more predictor
variables, the model is called a multiple regression model. The general
form of a multiple regression model is

yt = β0 + β1x1,t + β2x2,t + ⋯ + βkxk,t + εt, (1)

where is the variable to be forecast and are the predictor variables.
Each of the predictor variables must be numerical. The coefficients
β1,\ldots.. βk measure the effect of each predictor after taking into
account the effects of all the other predictors in the model. Thus, the
coefficients measure the marginal effects of the predictor variables.
Assumptions When we use a linear regression model, we are implicitly
making some assumptions about the variables in Equation (1).

First, we assume that the model is a reasonable approximation to
reality; that is, the relationship between the forecast variable and the
predictor variables satisfies this linear equation.

Second, we make the following assumptions about the errors (ε1, \ldots{}
, εT ) :

they have mean zero; otherwise the forecasts will be systematically
biased.

they are not autocorrelated; otherwise the forecasts will be
inefficient, as

there is more information in the data that can be exploited.

they are unrelated to the predictor variables; otherwise there would be
more information that should be included in the systematic part of the
model.

It is also useful to have the errors being normally distributed with a
constant variance σ2 in order to easily produce prediction intervals.
Another important assumption in the linear regression model is that each
predictor x is not a random variable. If we were performing a controlled
experiment in a laboratory, we could control the values of each x (so
they would not be random) and observe the resulting values of y. With
observational data (including most data in business and economics), it
is not possible to control the value of x, we simply observe it. Hence
we make this an assumption.

\hypertarget{auto-regressive-integrated-moving-average-model-arima}{%
\section{Auto regressive Integrated Moving Average Model
(ARIMA)}\label{auto-regressive-integrated-moving-average-model-arima}}

The Autoregressive Integrated Moving Average (ARIMA) model is a widely
utilized approach for forecasting and analyzing time series data. It
combines three key components: autoregression (AR), differencing (I),
and moving average (MA). ARIMA models have been extensively studied and
applied in various domains due to their flexibility and effectiveness in
capturing both short-term and long-term dependencies in time series.Box
and Jenkins (1970) {[}1{]} introduced the ARIMA model as a powerful tool
for time series analysis. The authors provided a comprehensive framework
for identifying, estimating, and diagnosing ARIMA models. They
emphasized the importance of differencing to achieve stationarity in
non-stationary time series data, allowing the model to capture trends
and patterns effectively.

\citep{chase2013demand} describes the application of the ARIMA model. It
is the forecasting technique that integrates vital components of
time-series and methods of regression. In 1970, George Box and Gwilym
Jenkins introduced the ARIMA forecasting method, and they developed a
comprehensive approach for forecasting \citep{chase2013demand}.ARIMA
(p,d,q) models gather relevant factors from the historical data by using
auto-correlation between them to distinguish those slacked request
verifiable qualities that best anticipate future demand; whereas p is
the order of Autoregressive (AR) term, q is the order of MA term

ARIMA can demonstrate cycle and regularity, and is mainly used for
mid-term aggregate forecasts and pre-sent far superior estimations than
time series models or even casual models; explained by
\citep{stadtler2014supply} explicitly considers dependent demands.
Additionally, different components, for example, informative factors
that impact request. It is also considered a good fit for the Mean
Absolute Percentage Errors (MAPE) calculation.

ARIMA model helps analyse the time series's probabilistic properties and
develop sin-gle or multi-equation models \citep{moritz2015comparison}.
Analysing a time series data is that there is a straightforward yet
extensive arrangement of useful models that can speak to numerous
conceivable examples of information found in time series. For a
stationary time-series, we can envision the information creating process
as a weighted blend of earlier perceptions in addition to an irregular
request term. Besides, identifying the ARIMA model parameters (i.e.~p, q
and d) is often challenging. Thereby, we use R software's statistical
package; it is widely used in the relevant literature
\citep{bokde2016psf, dhamo2010using}. Moreover, `R' generated the best
combination of p, d and q values because it estimates the value of the
parameter based on AIC (Akaike Information Criterion) and BIC (Bayesian
Information Criterion).

\#Performance evaluation \{\#sec-performance\}

\hypertarget{forecast-accuracy-measures}{%
\section{2.4. Forecast accuracy
measures}\label{forecast-accuracy-measures}}

To evaluate the forecast accuracy and bias, several measures including
Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), Mean Absolute
Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE),and Winkler
are used. We only present the results for RMSSE due to the space limit
of the journal and also because it is the recommended error metric in M5
competition \citep{makridakis2022m5}.

Point forecast accuracy is measured via the Mean Squared Scaled Error
(MSSE) and the Mean Absolute Scaled Error (MASE). The Mean Absolute
Scaled Error (MASE)
\citep[hyndman2018forecasting,\citet{article}hyndman2006another,]{book}
is calculated as ::: \{.cell
hash=`main\_cache/pdf/unnamed-chunk-28\_a96c898b6cc2c723121cc3be6c515a81'\}

:::

\hypertarget{sec-results}{%
\section{Results and discussion}\label{sec-results}}

from the total of 37 key pharmaceuticals selected and tested with the
different forecasting models arima showed that a better forecast
accuracy for most of the pharmaceuticals followed by mean and
regresstion models with RMSSE value of 0.850, 0.942 and 0.945
respectively (Table 1).

\begin{table}[!h]
\centering
\begin{tabular}[t]{lrrrrrr}
\toprule
.model & RMSE & MAE & MASE & RMSSE & winkler & CRPS\\
\midrule
arima & 109214.1 & 91630.74 & 0.925905 & 0.8503067 & 683471.4 & 64771.83\\
exponential_smoothing & 117985.6 & 98588.74 & 1.048426 & 0.9579627 & 695575.3 & 68638.44\\
mean & 121312.5 & 107616.64 & 1.087621 & 0.9422224 & 582790.3 & 73334.54\\
naive & 135300.5 & 116398.62 & 1.268535 & 1.0841642 & 570090.5 & 74691.10\\
regression & 131416.4 & 114329.68 & 1.065268 & 0.9455063 & 899543.3 & 83315.32\\
\addlinespace
snaive & 143861.2 & 122377.80 & 1.165142 & 1.0463095 & 732046.7 & 85266.80\\
\bottomrule
\end{tabular}
\end{table}

Then we tested the models by adding different predictors which affects
the sales of the pharmaceuticals like physical count, budget release and
budget closure by the government, COVID 19, campaign in to the models as
predictors which will affects the sales of the key pharmaceuticals
products and the result showed that still arima have a better forecast
accuracy with RMSSE of 0.850 (Table 2).

\begin{table}[!h]
\caption{forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after
considering predictors}\tabularnewline

\centering
\begin{tabular}[t]{lrrrrrr}
\toprule
.model & RMSE & MAE & MASE & RMSSE & winkler & CRPS\\
\midrule
arima_predictor & 114064.7 & 95753.88 & 0.9192677 & 0.851753 & 807586.4 & 68631.70\\
regression_predictor & 132117.6 & 113194.12 & 1.0474410 & 0.941347 & 1000434.4 & 83390.61\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}

{\centering \includegraphics{main_files/figure-pdf/result2-1.pdf}

}

\caption{forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022
after considering predictors}

\end{figure}

Individual item with different forecasting models We tested 10 models
namely arima, arima with predictor, combination, exponential smoothing,
mean, naive, regression, regression with predictor, snaive and stlf.
Individually on the selected key pharmaceuticals and the result
indicated that 11 (32.43\%) of the selected pharmaceuticals have better
forecast accuracy for arima predictor models, followed by naive 6
(16.21\%), mean, regresstion, arima, and STLF methods each of them with
3 (8.10\%) (Fig 2)

\includegraphics{main_files/figure-pdf/unnamed-chunk-43-1.pdf}

Those pharmaceuticals who had a better forecast accuracy with Arima
predictors are Amoxacillin 500mg capsule, Anti RhD, Artehmether +
Lumefantrine, Atenolol, Atrovastatin 200mg, Ferrosu sulphate with follic
acid, frusemide injection, Insuline Iosphane bipahsic, Lamivudine
+Efaverienze + Tenofovir, Lidocaine + HCL and Propyl thiro uracil.The
six pharmaceuticasl which showed a better forecast accuracy with naive
models were Ciprofloxacillin 500mg tablet, Magnesoum sulphate injection,
Omeprazole capsule, Omeprazole injection, Rapid diagnostic test,
Sulphamethaxone + Trimetoprine tablet.

Discussion:

The objective of this study was to evaluate the forecasting accuracy of
different models for key selected pharmaceutical products from the
Ethiopian Pharmaceutical Supply Services (EPSS). The analysis
encompassed a total of 37 key pharmaceutical products, and the results
highlight the performance of various forecasting models in predicting
sales for these products. In the initial assessment of forecasting
models, the ARIMA model emerged as the most accurate for the majority of
the pharmaceutical products, exhibiting a Root Mean Squared Scaled Error
(RMSSE) value of 0.850. This finding suggests that ARIMA has a strong
predictive capability and outperforms other models, including the mean
and regression models. The superiority of the ARIMA models over the
Moving Average might be explained by most of the data set have seasonal
patern and MA is one of the simplest prediction techniques for making
projections about time-series without a noticeable seasonal pattern
\citep{chopra2001supply}.The superiority of the ARIMA model underscores
its suitability for pharmaceutical sales forecasting within the context
of EPSS. Similar findings were reported in forecasting ethanol demand in
India where ARIMA model outperforms other models like linear and
non-linear regression models \citep{dey2023forecasting}.

To enhance the predictive power of the models, additional predictors
were incorporated into the analysis. Predictors such as physical count,
budget release and closure by the government, the impact of COVID-19,
and promotional campaigns were considered as potential drivers of
pharmaceutical sales. Despite the inclusion of these predictors, the
ARIMA model maintained its superior forecasting accuracy with an RMSSE
of 0.850. This resilience further validates the robustness of the ARIMA
model and its ability to capture complex interactions between variables
influencing pharmaceutical sales.

To delve deeper into the individual performance of the forecasting
models, a comprehensive evaluation was conducted on each of the 37
pharmaceutical products. Among the ten models tested, ARIMA with
predictors exhibited the highest forecast accuracy for 32.43\% of the
products. This was followed by the naive model, which performed well for
16.21\% of the products. Notably, a few pharmaceuticals demonstrated a
strong fit with specific models. For instance, Amoxicillin 500mg
capsule, Anti RhD, Artehmether + Lumefantrine, and others exhibited
superior forecast accuracy when utilizing the ARIMA model with
predictors.

Similarly, other pharmaceuticals displayed better forecasting outcomes
with alternative models. Notably, Ciprofloxacin 500mg tablet, Magnesium
sulfate injection, Omeprazole capsule, and others showed improved
accuracy with the naive model. These findings underscore the importance
of tailoring the choice of forecasting model based on the
characteristics and dynamics of individual pharmaceutical products.This
finding confirmed that There is not a single best technique to solve
time-series forecasting problems \citep{zhang2007quarterly}. In order to
deal with time-series forecasting, each problem might be solved with a
different approach \citep{ensafi2022time}.

In conclusion, this study highlights the pivotal role of accurate
forecasting in the pharmaceutical supply chain management. The ARIMA
model, particularly when integrated with relevant predictors, emerges as
a powerful tool for predicting pharmaceutical sales within the EPSS.
However, the selection of the most appropriate model should consider the
specific attributes of each pharmaceutical product. Further research
could explore the optimization of model parameters and the incorporation
of additional contextual variables to enhance forecasting accuracy and
supply chain efficiency.

Limitation

One notable limitation of this study revolves around the availability
and quality of the data used for analysis. While we employed a
comprehensive dataset from the EPSS, it is essential to acknowledge that
the accuracy of forecasting models is heavily contingent on the quality,
granularity, and completeness of historical sales data. Variations in
data recording practices, potential errors, missing values, or
inconsistencies within the dataset could impact the precision of the
forecasts generated by the employed models. Moreover, the inherent
complexity of pharmaceutical demand, influenced by a myriad of external
factors such as socioeconomic changes,stock outs, prolonged procurement
lead time, rationing of avaiable products, pushing of pharmaceuticals to
different health facility without demand, healthcare policies, and
unforeseen events, may introduce an additional layer of uncertainty. It
is important to recognize that despite our best efforts to address
data-related challenges, the robustness and reliability of our
forecasting outcomes may still be subject to the limitations inherent in
the original data.

\hypertarget{sec-conclusion}{%
\section{Conclusion}\label{sec-conclusion}}

In this study, we embarked on a comprehensive analysis of pharmaceutical
demand forecasting within the Ethiopian Pharmaceutical Supply Services
(EPSS) context. By leveraging five years of sales data for 37 key
pharmaceutical products, we aimed to enhance the accuracy of demand
predictions and consequently contribute to more effective supply chain
management. Our findings underscore the significance of accurate demand
forecasting in optimizing inventory levels, resource allocation, and
overall supply chain operations.

The application of diverse forecasting models yielded valuable insights
into their respective performance. Notably, the Autoregressive
Integrated Moving Average (ARIMA) model emerged as a robust choice,
consistently exhibiting superior forecast accuracy across a majority of
the selected pharmaceutical products. Additionally, the incorporation of
predictor variables, such as physical count, budget release and closure
by the government, COVID-19 impact, and campaign events, further
underscored the efficacy of the ARIMA model in accommodating external
influences and delivering reliable forecasts.

The individualized assessment of forecasting models revealed specific
strengths across pharmaceutical categories. ARIMA models with predictors
demonstrated exceptional performance for pharmaceuticals including
Amoxacillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine,
Atenolol, and others, affirming their utility in capturing complex
demand patterns. Moreover, the study shed light on pharmaceuticals best
suited for alternative models like naive, mean, and regression methods,
contributing to a comprehensive understanding of forecasting
applicability.

While these findings contribute significantly to the discourse on
pharmaceutical demand forecasting, it's essential to acknowledge
limitations. The accuracy of our forecasts is intrinsically linked to
the quality and completeness of historical sales data, and the intricate
interplay of external variables adds complexity to the forecasting
landscape.

In conclusion, this study serves as a pivotal step towards enhancing
pharmaceutical supply chain management through refined demand
forecasting. The demonstrated success of ARIMA models, particularly when
augmented with predictor variables, holds promise for fostering
resilient and adaptable supply chains. As the pharmaceutical landscape
continues to evolve, our insights encourage further exploration of
advanced forecasting techniques and continuous refinement of strategies
to bolster supply chain efficiency and ensure consistent availability of
essential pharmaceuticals.

In light of the study's findings, several actionable recommendations
emerge. First and foremost, the integration of the Autoregressive
Integrated Moving Average (ARIMA) model with predictor variables stands
out as a viable strategy for enhancing pharmaceutical demand forecasting
accuracy within Ethiopian Pharmaceutical Supply Services (EPSS). This
entails harnessing the predictive power of external factors to refine
forecasting precision. To further advance forecasting practices, the
exploration of additional predictor variables such as public health
campaigns, disease outbreaks, and economic indicators could offer
valuable insights into demand dynamics. Moreover, refining the
granularity of historical sales data, possibly through finer temporal
intervals and geographic diversity, holds potential for more nuanced and
accurate predictions. Consideration could also be given to hybrid
forecasting models that amalgamate ARIMA's short-term accuracy with
machine learning's predictive prowess for long-term trends. Regular
model evaluation, sensitivity analysis, and collaborative stakeholder
engagement are recommended to ensure ongoing accuracy and alignment with
changing demand patterns.

Capacitating EPSS staff through training programs in forecasting
methodologies and data interpretation can empower the workforce to make
informed decisions. Replication of this study in diverse healthcare
settings and comparative analyses can validate the findings'
applicability and adaptability. By embracing these recommendations, EPSS
can fortify its pharmaceutical supply chain with resilient, efficient,
and responsive operations, ultimately ensuring the consistent
availability of vital pharmaceuticals. a

\hypertarget{reproducibility}{%
\section*{Reproducibility}\label{reproducibility}}
\addcontentsline{toc}{section}{Reproducibility}

R code to produce all results in this paper is available at

\renewcommand\refname{References}
  \bibliography{references.bib}

\end{document}

### 2024-10-09 manuscript/03-experiment.qmd

@@
 We are presenting the graph for some of the pharmaceuticals sales for the last five years by using visualization for the readers to undesrtand the sales of the pharmaceuticals.
+
 ```{r}
@@
 # 2. 2. Data Splitting and Preparation
-The primary objective of this study is to forecast pharmaceutical demand within EPSS for the year 2023.
-To achieve this, the historical sales data for the chosen pharmaceuticals in EPSS was partitioned into distinct training and testing subsets.
-Employing the widely recognized 70:30 split method, 70% of the available data was designated as the training set, while the remaining 30% was reserved for testing and validation purposes [@chan2019comparison].
-As a result, the dataset was segregated into a training set consisting of 42 months and a testing set spanning 18 months
 
-# 2.3.1. Forecasting Methods
-In pursuit of accurate pharmaceutical demand forecasts, we explored a spectrum of forecasting techniques. The following methods were applied:
-
-Naive Method: A simple approach relying on the assumption that the future value will be the same as the most recent observed value.This Methods uses when there is no
-seasonality [@article{burinskiene2022forecasting; @article{nikolopoulos2016forecasting]
-Mean Method: A straightforward technique forecasting future values based on the mean of past observations.
-Seasonal Naive Method: Similar to the naive method, this approach predicts future values using the most recent observed value from the same season.
-Exponential Smoothing: A method that assigns exponentially decreasing weights to past observations to forecast future values.
-Regression: Employing statistical regression models to establish relationships between variables for prediction.
-Autoregressive Integrated Moving Average Model (ARIMA): A time series model incorporating autoregressive, differencing, and moving average components for accurate forecasting.
-ARIMA with Predictors: Extending the ARIMA model with additional predictor variables to enhance forecasting accuracy.
-Combination of Models: Exploring ensemble methods that combine multiple models for improved predictive performance.
-By applying these diverse forecasting methodologies, we aimed to comprehensively assess and compare their efficacy in predicting pharmaceutical demand within EPSS.
-Each method was evaluated against the testing dataset to ascertain its forecasting accuracy and suitability for the specific
-
-## Forecasting methods {#sec-method}
-Model building
-
-# Naive 
-The naive forecast method is a simple approach that assumes future values will be the same as the most recent observed value, making it useful for stable or slowly changing time series.
-While it doesn't consider underlying patterns, it serves as a baseline for evaluating more advanced techniques[@hyndman2018forecasting].
-It's particularly suitable for products with stable demand or limited seasonality, providing a straightforward reference for forecasting and inventory management.
-
-# Mean 
-The mean forecast method predicts future values by averaging historical data, making it suitable for stable time series.
-While it overlooks trends and seasonality, it serves as a baseline for evaluating advanced techniques [@hyndman2018forecasting].
-Its simplicity makes it valuable in certain forecasting applications [@hyndman2018forecasting].
-
-# Seasonal naive
-
-The seasonal naive forecast method predicts future values based on the most recent observed value from the corresponding season in the previous year, making it useful for data with strong seasonal patterns.
-It serves as a baseline for evaluating seasonal forecasting models, capturing repetitive patterns effectively.
-Its simplicity and ability to capture seasonality make it valuable in seasonal forecasting applications[@hyndman2018forecasting].
-
-# Exponential smoothing
-
-Exponential Smoothing is regarded as one of the most accurate and robust methods for time series forecasting.
-Not only does it exhibit high accuracy, but it also offers versatility and robustness.
-What makes Exponential Smoothing particularly appealing is its intuitive nature, making it easy to understand and interpret.
-Furthermore, this method boasts minimal data storage and computational requirements, making it highly efficient for real-time applications.
-In the business realm, Exponential Smoothing finds extensive use for inventory demand forecasting, showcasing its practicality [@gardner1985exponential].
-Surprisingly, even when pitted against more sophisticated techniques, Exponential Smoothing has demonstrated impressive performance in forecasting competitions [@makridakis2000m3,@makridakis1982accuracy].
-Its simplicity and reliability make it a widely trusted approach in the field of forecasting.
-
-# Regression
-
-Regression methods play a crucial role in forecasting and optimizing the health supply chain, ensuring the availability and efficient distribution of essential medical resources.
-By analyzing historical data, regression models can identify patterns and relationships between variables such as patient demand, disease prevalence, and inventory levels.
-The basic concept is that we forecast the time series of interest assuming that it has a linear relationship with other time series.
-For example, we might wish to forecast monthly sales y using total advertising spend as a predictor X.
-Or we might forecast daily electricity y demand using temperature x1   and the day of week x2 as predictors.
-The forecast variable y  is sometimes also called the regress and, dependent or explained variable.
-The predictor variables x  are sometimes also called the regressors, independent or explanatory variables.
-
-Simple linear regression 
-In the simplest case, the regression model allows for a linear relationship between the forecast variable y and a single predictor variable x: yt = β0 + β1xt + ε 
-Example of data from such a model is shown in Figure 5.1.
-The coefficients β0 and β1 denote the intercept and the slope of the line respectively.
-The intercept β0 represents the predicted value of  y when x=0 .
-The slope represents the average predicted change in y resulting from a one unit increase in x.
-
-# Figure : An example of data from a simple linear regression model
-
-Notice that the observations do not lie on the straight line but are scattered around it.
-We can think of each observation yt  as consisting of the systematic or explained part of the model, = β0 + β1xt , and the random “error”, εt.
-The “error” term does not imply a mistake, but a deviation from the underlying straight line model.
-It captures anything that may affect yt other than xt
-
-Multiple linear regressions 
-When there are two or more predictor variables, the model is called a multiple regression model. The general form of a multiple regression model is
-
-yt = β0 + β1x1,t + β2x2,t + ⋯ + βkxk,t + εt, (1)
-
-where is the variable to be forecast and are the predictor variables.
-Each of the predictor variables must be numerical.
-The coefficients β1,…..
-βk measure the effect  of each predictor after taking into account the effects  of all the other predictors in the model.
-Thus, the coefficients measure the marginal effects of the predictor variables.
-Assumptions 
-When we use a linear regression model, we are implicitly making some assumptions about the variables in Equation (1).
-
-First, we assume that the model is a reasonable approximation to reality; that is, the relationship between the forecast variable and the predictor variables satisfies this linear equation.
-
-Second, we make the following assumptions about the errors (ε1, … , εT ) :
-
-they have mean zero; otherwise the forecasts will be systematically biased.
-
-they are not autocorrelated; otherwise the forecasts will be inefficient, as
-
-there is more information in the data that can be exploited.
-
-they are unrelated to the predictor variables; otherwise there would be more information that should be included in the systematic part of the model.
-
-It is also useful to have the errors being normally distributed with a constant variance σ2  in order to easily produce prediction intervals. 
-Another important assumption in the linear regression model is that each predictor x is not a random variable.
-If we were performing a controlled experiment in a laboratory, we could control the values of each x (so they would not be random) and observe the resulting values of y.
-With observational data (including most data in business and economics), it is not possible to control the
-value of x, we simply observe it. Hence we make this an assumption.
-
-# Auto regressive Integrated Moving Average Model (ARIMA)
-
-The Autoregressive Integrated Moving Average (ARIMA) model is a widely utilized approach for forecasting and analyzing time series data.
-It combines three key components: autoregression (AR), differencing (I), and moving average (MA).
-ARIMA models have been extensively studied and applied in various domains due to their flexibility and effectiveness in capturing both short-term and long-term dependencies in time series.Box and Jenkins (1970) [1] introduced the ARIMA model as a powerful tool for time series analysis.
-The authors provided a comprehensive framework for identifying, estimating, and diagnosing ARIMA models.
-They emphasized the importance of differencing to achieve stationarity in non-stationary time series data, allowing the model to capture trends and patterns effectively.
-
-[@chase2013demand] describes the application of the ARIMA model.
-It is the forecasting technique that integrates vital components of time-series and methods of regression.
-In 1970, George Box and Gwilym Jenkins introduced the ARIMA forecasting method, and they developed a comprehensive approach for forecasting [@chase2013demand].ARIMA (p,d,q) models gather relevant factors from the historical data by using auto-correlation between them to distinguish those slacked request verifiable qualities that best anticipate future demand; whereas p is the order of Autoregressive (AR) term, q is the order of MA term
-
-ARIMA can demonstrate cycle and regularity, and is mainly used for mid-term aggregate forecasts and pre-sent far superior estimations than time series models or even casual models; explained by [@stadtler2014supply] explicitly considers dependent demands.
-Additionally, different components, for example, informative factors that impact request.
-It is also considered a good fit for the Mean Absolute Percentage Errors (MAPE) calculation.
-
-ARIMA model helps analyse the time series’s probabilistic properties and develop sin-gle or multi-equation models [@moritz2015comparison].
-Analysing a time series data is that there is a straightforward yet extensive arrangement of useful models that can speak to numerous conceivable examples of information found in time series.
-For a stationary time-series, we can envision the information creating process as a weighted blend of earlier perceptions in addition to an irregular request term.
-Besides, identifying the ARIMA model parameters (i.e.
-p, q and d) is often challenging.
-Thereby, we use R software’s statistical package; it is widely used in the relevant literature [@bokde2016psf; @dhamo2010using].
-Moreover, ‘R’ generated the best combination of p, d and q values because it estimates the value of the parameter based on AIC (Akaike Information Criterion) and BIC (Bayesian Information Criterion).
-
-#Performance evaluation {#sec-performance}
-
-# 2.4. Forecast accuracy measures
-To evaluate the forecast accuracy and bias, several measures including Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), Mean Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE),and Winkler are used.
-We only present the results for RMSSE due to the space limit of the journal and also because it is the recommended error metric in M5 competition [@makridakis2022m5].
-
-Point forecast accuracy is measured via the Mean Squared Scaled Error (MSSE) and the Mean Absolute Scaled Error (MASE).
-The Mean Absolute Scaled Error (MASE) [@book{hyndman2018forecasting,@article{hyndman2006another,] is calculated as
-```{r}
-
-```
+The primary objective of this study is to forecast pharmaceutical demand within EPSS for the year 2023. To achieve this, the historical sales data for the chosen pharmaceuticals

### 2024-10-09 manuscript/04-results.qmd

@@
 ```{r}
-library(kableExtra)
-library(dplyr)  # Load the dplyr package
-fc_accuracy_cleaned <- na.omit(fc_accuracy_overall__without_predictor)
-fc_accuracy_table <- fc_accuracy_cleaned %>%
-  dplyr::select(.model, RMSE, MAE, MASE, RMSSE, winkler, CRPS)  # Explicitly specify dplyr::select
-kbl(fc_accuracy_table, booktabs = TRUE, escape = FALSE) %>%
-  kable_styling(latex_options = c("hold_position"), full_width = FALSE)
+# library(kableExtra)
+# library(dplyr)  # Load the dplyr package
+# fc_accuracy_cleaned <- na.omit(fc_accuracy_overall__without_predictor)
+# fc_accuracy_table <- fc_accuracy_cleaned %>%
+#   dplyr::select(.model, RMSE, MAE, MASE, RMSSE, winkler, CRPS)  # Explicitly specify dplyr::select
+# kbl(fc_accuracy_table, booktabs = TRUE, escape = FALSE) %>%
+#   kable_styling(latex_options = c("hold_position"), full_width = FALSE)
 ```
@@
 ```{r}
-#| label: result3
-#| tbl-cap: "forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after considering predictors"
-fc_accuracy_predictor_overall %>% 
-  kbl(booktabs = TRUE, escape = FALSE) |>
-  kable_styling(latex_options = c("hold_position"),full_width = F)
+# #| label: result3
+# #| tbl-cap: "forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after considering predictors"
+# fc_accuracy_predictor_overall %>% 
+#   kbl(booktabs = TRUE, escape = FALSE) |>
+#   kable_styling(latex_options = c("hold_position"),full_width = F)
 ```
+
 ```{r}
-#| label: result2
-#| fig-cap: "forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after considering predictors"
-library(ggplot2)  # Load the ggplot2 package
-library(forcats)  # Load the forcats package
-fc_accuracy_overall %>%
-  select(`.model`, item, RMSSE) %>% 
-  ggplot(mapping = aes(x = RMSSE, y = fct_reorder(`.model`, RMSSE, .fun = median))) +
-  geom_boxplot()
+# #| label: result2
+# #| fig-cap: "forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after considering predictors"
+# library(ggplot2)  # Load the ggplot2 package
+# library(forcats)  # Load the forcats package
+# fc_accuracy_overall %>%
+#   select(`.model`, item, RMSSE) %>% 
+#   ggplot(mapping = aes(x = RMSSE, y = fct_reorder(`.model`, RMSSE, .fun = median))) +
+#   geom_boxplot()
 ```
@@
 # Sample data (replace with your actual data)
-pharmaceuticals <- c("ARIMA Predictor", "Naive", "Mean", "Regression", "ARIMA", "STLF")
-accuracy_percentages <- c(32.43, 16.21, 8.10, 8.10, 8.10, 8.10)
-
-# Create a data frame
-data <- data.frame(Pharmaceuticals = pharmaceuticals, Accuracy_Percentage = accuracy_percentages)
-# Create a bar chart
-library(ggplot2)
-
-ggplot(data, aes(x = reorder(Pharmaceuticals, -Accuracy_Percentage), y = Accuracy_Percentage)) +
-  geom_bar(stat = "identity", fill = "blue") +
-  labs(x = "Pharmaceutical Forecasting Models", y = "Accuracy Percentage") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# pharmaceuticals <- c("ARIMA Predictor", "Naive", "Mean", "Regression", "ARIMA", "STLF")
+# accuracy_percentages <- c(32.43, 16.21, 8.10, 8.10, 8.10, 8.10)
+# 
+# # Create a data frame
+# data <- data.frame(Pharmaceuticals = pharmaceuticals, Accuracy_Percentage = accuracy_percentages)
+# # Create a bar chart
+# library(ggplot2)
+# 
+# ggplot(data, aes(x = reorder(Pharmaceuticals, -Accuracy_Percentage), y = Accuracy_Percentage)) +
+#   geom_bar(stat = "identity", fill = "blue") +
+#   labs(x = "Pharmaceutical Forecasting Models", y = "Accuracy Percentage") +
+#   theme_minimal() +
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 

### 2024-10-10 manuscript/02-literature.qmd
Revised literature review

@@
 
-Precise sales prediction is an essential and inexpensive way for each company to augment their profits, decrease their costs, and achieve greater flexibility to changes.
-In other words, exact sales forecasting is utilized for capturing the tradeoff between customer demand satisfaction and inventory costs [@gupta2000mid].
-Especially, for the pharmaceutical industry, successful sales forecasting systems can be very beneficial, due to the short shelf-life of many pharmaceutical products
-and the importance of the product quality which is closely related to the human health [@doganis2006time].There are many decisions that depend on the quality of forecasts in the health care system, from capacity planning to layout decisions to the daily schedules.
-In general,the role of forecasting in health care is to inform both clinical and non-clinical decisions.
-While the former concerns decisions related to patients and their treatments [@makridakis2019forecasting], the latter involves policy/management, and supply chain decisions that support the delivery of high-quality care for patients.In the coming paragarphs we will review what other researchers had done in the area of pharmaceuticals forecasting.Morover today,manufacturing companies are organized following the principles of just-in-time (JIT) delivery which focuses on efficiency and eliminating the waste of excessive inventory [@van2021using].
-In JIT systems, accurate Stock Keeping Unit (SKU) short-term demand forecasts, typically on a daily, weekly or monthly basis, are crucial as they are the fundamental inputs to drive the supply chain planning system.Therefore, forecasting performance can strongly affect a company’s financial results, since both under- and overestimating short-term demand are expensive as they result in supply shortages and consequently poor customer service, and excess inventories and product obsolescence, respectively (Sanders and Graman, 2009).In this section, we provide a brief review of studies on forecasting harmaceuticals products.In conclusion, the research findings indicated that the hybrid neural network approach, which combines linear and nonlinear modeling techniques, outperformed traditional methods like ARIMA and standalone neural networks in forecasting pharmaceutical sales.
-The hybrid approach demonstrated the ability to capture both linear and nonlinear patterns in sales data accurately, especially in scenarios with limited historical records for individual drugs.
+Accurate sales forecasting is crucial in the pharmaceutical industry, where it directly impacts profit maximization, cost minimization, and the ability to respond to market changes.
+Forecasting in healthcare not only influences clinical decisions but also plays a pivotal role in supply chain management, ensuring that the right drugs are available when needed.
+The challenge lies in balancing customer demand with inventory costs, which is particularly complex in pharmaceuticals due to factors like short shelf life and quality constraints [@gupta2000mid, @makridakis2020forecasting].
 
-A study by Khalil Zadeh et al.
-developed an intelligent sales prediction model for Pharmaceutical Distribution Companies (PDCs) by combining network analysis tools and time series forecasting methods.
-This approach addresses the challenge of limited historical sales data for pharmaceutical products, making traditional forecasting methods inadequate[@zhang2004neural].The researchers employed Mean Squared Error (MSE) and Mean Absolute Error (MAE) to evaluate the performance of various forecasting models, including ARIMA, neural networks, and a hybrid approach combining both.
-The hybrid model, inspired by Zhang's work[@zhang2004neural], used ARIMA for linear components and neural networks for nonlinear components.Three methods were used: (1) ARIMA, (2) a hybrid neural network with each drug’s past records, and (3) a hybrid neural network with both each drug’s and its group members’ past records.
-Grouping products and using group members' records significantly improved forecasting accuracy by addressing the issue of limited data and leveraging both linear and nonlinear modeling capabilities.The study concluded that the hybrid approach, incorporating both individual and group past records, performed better than using either method alone [@khalil2014intelligent].
+In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies utilizing a combination of simple and advanced methods to enhance accuracy and reliability[@rathipriya2023demand,@merkuryeva2019demand,@nikolopoulos2016forecasting,@van2021using,@khalil2014intelligent,@zhu2021demand,@anusha2014demand,@burinskiene2022forecasting,@bon2017optimization,@sousa2019statistical,@zahra2019forecasting].
+Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods like the Holt-Winters method, Single Moving Average (SMA), and Single Exponential Smoothing (SES), which are simpler but effective for certain applications.
+In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand,@candan2014demand, and @ribeiro2017improving].
 
-A study by Van Bell et al.
-used various forecasting models for point of sale data to assess the impact of including sell-through data on forecast accuracy.
-They compared information sharing (IS) methods with non-information sharing (NIS) methods using the following approaches.
-Substitution Approach (ETS-W and ARIMA-W): These models use downstream wholesaler sales time series as input instead of prior manufacturer shipments, suitable when all goods pass through wholesalers.
-Integration Approach (Extrapolative Time Series Methods with Exogenous Variables): This includes methods that integrate downstream information using extrapolative time series techniques and machine learning (ML) approaches.
-Extrapolative Time Series Methods (ETS and ARIMA): Traditional forecasting methods based on modeling past time series structures and extrapolating into the future.
-The study evaluated these methods using Mean Squared Error (MSE) and Mean Absolute Error (MAE).
-Additionally, the researchers used naive methods, IS and NIS with ETS and ARIMA models, and machine learning techniques like LASSO, Support Vector Regression (SVR), and Random Forest (RF).LASSO (Least Absolute Shrinkage and Selection Operator) was identified as the best-performing method for improving forecast accuracy, particularly for one-step ahead forecasts.
-It outperformed other methods for longer horizons (h > 1) by effectively selecting the most relevant variables and reducing overfitting.
-ETS was found to be the most effective univariate baseline method, and SVR showed the best performance for h = 1.
-The study concluded that IS methods generally outperformed NIS methods, indicating that incorporating downstream information, such as sell-through data, can significantly enhance short-term forecast accuracy in supply chain management.
-Machine learning techniques like LASSO, SVR, and RF faced challenges related to overfitting and model complexity, especially with high-dimensional data.
-Balancing model complexity and interpretability was crucial for robust forecasting.
-Overall, the research suggests that LASSO, along with ETS and SVR, are effective for improving demand prediction accuracy in multi-echelon supply chains by leveraging downstream information and collaborative approaches[@van2021using].
+Early research in pharmaceutical forecasting primarily focused on traditional statistical methods.
+These methods, including ARIMA, exponential smoothing, and moving averages, were widely used due to their simplicity and effectiveness in relatively stable environments [@zahra2019forecasting].
+These models typically assess forecast accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean Squared Error (MSE), and Root Mean Squared Error (RMSE).
+For example, @newberne2006holt applied the Holt-Winters method to forecast healthcare data, focusing on prescription trends.
+Their findings highlighted the method's utility in short-term planning and resource management, validated through metrics like MAPE, MAD, and MSD.
+Similarly, @restyana2021analysis compared Single Moving Average (SMA) and Single Exponential Smoothing (SES) in drug demand forecasting, concluding that SES provided more accurate results, as indicated by lower MAD and MSE values.
 
-A study by Nikolopoulos et al.
-forecasted UK pharmaceutical time series before and after patent expiry, a critical point when a generic version enters the market.
-The study used various models, including naïve and moving averages, the Bass diffusion model, repeat purchase diffusion models, three regression models, and two exponential smoothing (ES) approaches for non-trended (simple ES) and trended data (Holt ES) [@nikolopoulos2016forecasting].Holt Exponential Smoothing: This model was the most accurate for forecasting both branded and generic drugs across short-term, mid-term, and long-term horizons.Naïve Model: Used as a benchmark, the relative absolute error (RAE) was calculated to compare forecasting accuracy.Bass Diffusion Model: Although popular in marketing research, it had limitations in accurately forecasting pharmaceutical time series due to the complex dynamics between branded and generic drugs.
-Comparative Analysis: The study compared various forecasting models to evaluate their performance, highlighting the importance of selecting appropriate models based on the data and context.The research found that Holt’s exponential smoothing model outperformed others, including naïve models, moving averages, and ARIMA models, especially for short-term forecasts.
-However, forecasting accuracy varied across different horizons.
-Despite their limitations, diffusion models were included to provide additional insights.Overall, the study emphasized the strengths and limitations of different forecasting approaches in the context of pharmaceutical time series, with Holt’s exponential smoothing model demonstrating superior performance[@nikolopoulos2016forecasting].
+Recognizing the limitations of traditional models, particularly in capturing the complexities of pharmaceutical demand (e.g., seasonality, external influences), researchers began developing hybrid models.
+These models combine linear methods like ARIMA with nonlinear techniques such as neural networks, aiming to improve forecasting accuracy by addressing diverse data patterns.
+@khalil2014intelligent introduced a hybrid model that integrated ARIMA with neural networks.
+Their study demonstrated that this approach significantly outperformed traditional models, especially in scenarios with limited historical data.
+The hybrid model's effectiveness was confirmed by improvements in key performance indicators like MAPE and RMSE.
 
-Astudy by Siddiqui et al used to develop a combined forecasting model.The combination is developed using two frequent, but separately used models, i.e.
-Autoregressive Integrated Moving Average (ARIMA)  model and Holts Winter model and a combined model were proposed which is known as the ARHOW model.The ARIMA model is commonly used for time series forecasting, while Holt’s Winter model is effective for handling seasonality in data [@inproceedings{markovska2016comparative].The ARHOW model aims to enhance forecasting accuracy by leveraging the strengths of both ARIMA and Holt’s Winter models.
-The model equation for obtaining a consolidated forecast is as follows: ARHOW = β * ArimaForecast + β1 * HoltWinterForecast
-In this equation:
-ARHOW represents the Combined Weighted Forecast.
-ArimaForecast and HoltWinterForecast are the forecasts obtained by applying the ARIMA and Holt’s Winter models over the training set.β and β1 are the weights assigned to the ARIMA and Holt’s Winter forecasts, respectively.
-By combining the forecasts generated by these two models, the ARHOW model seeks to mitigate the impact of volatility and external factors on forecasting accuracy, ultimately providing more reliable predictions for pharmaceutical sales data [@siddiqui2022hybrid] .
-Combining multiple forecasting techniques like ARIMA and Holt’s Winter to create the ARHOW model introduces complexity.
-Managing and interpreting the results of a hybrid model may pose challenges in practical implementation and decision-making processes .Morover External factors such as regulatory changes, technological advancements, and market dynamics can introduce forecasting errors.
-While the ARHOW model aims to improve accuracy, it may still be susceptible to errors resulting from unforeseen external influences [@siddiqui2022hybrid].The key forecasting metrics employed in the study include:
+The pharmaceutical industry's increasing complexity and the availability of large datasets have led to the adoption of advanced machine learning techniques.
+These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power.@van2021using, showed that incorporating downstream information into machine learning models, like LASSO and SVR, greatly enhances forecast accuracy.
+Their study on multi-echelon supply chains revealed that advanced models, particularly those integrating external variables, consistently produced lower forecast errors, as measured by AvgRelRMSE.
+Similarly, @rathipriya2023demand compared shallow neural networks and deep learning models for drug demand forecasting.
+The study found that shallow models outperformed deep learning models for most drug categories, while ARIMA was more effective for the remaining categories.
+This suggests that no single model is universally optimal, and model selection should be context-specific.
 
-Mean Absolute Percentage Error (MAPE), Mean, Mean Absolute Deviation (MAD), Mean Square Error (MSE), Root Mean Square Error (RMSE).The Mean Absolute Percentage Error (MAPE) and other forecasting metrics suggest that the ARHOW model provides more accurate predictions compared to traditional forecasting techniques like ARIMA, Holt’s Winter, ETS, Theta, and Naïve models.
-The model's efficiency in forecasting pharmaceutical demand can help manufacturers plan inventory and prevent stockouts or overstock situations .In conclusion, the study underscores the effectiveness of the ARHOW model in improving demand forecasting accuracy in the pharmaceutical industry and provides insights for enhancing supply chain efficiency through advanced forecasting techniques .
+More sophisticated models, such as Neuro-Fuzzy Systems, have emerged as powerful tools for pharmaceutical forecasting.
+These systems combine the learning capabilities of neural networks with the reasoning abilities of fuzzy logic, offering a balanced approach that incorporates both empirical data and expert knowledge.
+Deep learning models, while powerful, present challenges such as the need for large datasets and substantial computational resources.
+Despite these challenges, models like LSTM have shown potential in capturing long-term dependencies in time series data, as demonstrated by @sousa2019statistical in their study on drug distribution in the Brazilian Public Health System.
+@candan2014demand employed an Adaptive Neuro-Fuzzy Inference System (ANFIS) to forecast pharmaceutical demand.
+Their study highlighted the system's ability to capture complex patterns in the data, resulting in highly accurate forecasts.
+The effectiveness of this approach was further validated through statistical tests, such as paired T-tests and mean difference analysis, which confirmed its superiority over traditional methods.
 
-A study by de Oliveira et al.
-compared five machine learning regression approaches for lead time forecasting in the pharmaceutical supply chain.
-The models used were k-nearest neighbors (KNN), support vector machines (SVM), random forests (RF), linear regression (LR), and multi-layer perceptrons (MLP).
-These models aimed to predict supplier delivery times, enhancing planning for production and logistics services in the pharmaceutical field [@de2021lead].
-The study focused on using machine learning techniques to forecast lead times for purchasing products in the pharmaceutical industry.
-The comparison aimed to identify the most effective approach for predicting purchasing lead times, demonstrating high accuracy potential, especially with linear support vector regression.
-Key limitations noted included computational complexity, sensitivity to outliers, difficulty in interpretability, and overfitting.
-Using mean squared error (MSE) as the forecast accuracy metric, the study found that the SVM approach achieved the best performance, with an average error of less than two days.
-This indicates that SVM was the most effective model among those tested for forecasting purchasing lead times in the pharmaceutical supply chain[@inproceedings{de2021lead].
+Despite the advancements in forecasting methods, several gaps remain, particularly in addressing uncertainty and incorporating domain knowledge into the modeling process.
+Most existing studies focus on point forecasts, which may not fully capture the range of possible outcomes in pharmaceutical demand.
+This limitation is critical, given the high stakes of inventory management in this industry.
+Moreover, there is a lack of research on the impact of integrating domain knowledge—such as insights from experienced professionals—into forecasting models.
+This integration could potentially improve model accuracy and reliability, especially in complex and rapidly changing environments.
+In summary, the literature on pharmaceutical forecasting has evolved from traditional statistical models to advanced machine learning techniques, each offering unique insights and capabilities.
+However, there is still room for improvement, particularly in addressing uncertainty and incorporating domain knowledge.
+Future research should focus on developing probabilistic forecasting models and exploring ways to integrate expert insights into the forecasting process.
+These advancements could significantly enhance decision-making and operational efficiency in the pharmaceutical supply chain.
 
-A study by Rathipriya et al.
-focused on demand forecasting for time-series pharmaceutical data using both shallow and deep neural network models[@rathipriya2023demand].
-The researchers employed several models, including the Autoregressive Integrated Moving Average (ARIMA) model, known for its simplicity and ability to handle non-stationary series.The shallow neural network models used were, Radial Basis Function Neural Network (RBF_NN),Generalized Regression Neural Network (GR_NN),Probabilistic Neural Network (P_NN),For deep neural networks, the study investigated:Long Short-Term Memory (LSTM),Stacked LSTM,Additionally, the study explored hybrid models, combining ARIMA with the Adaptive Neural Fuzzy Inference System (ANFIS) to forecast Iran's annual energy usage, demonstrating higher accuracy compared to individual models.
-The researchers also employed methods like Genetic Algorithms (GA), Bayesian structures, and Reinforcing Artificial Neural Networks (RANN) for forecasting short-run demand.The performance of the demand forecast models (DFMs) was evaluated using Root Mean Squared Error (RMSE).
-The study found that shallow neural network models, such as GR_NN, RBF_NN, and P_NN, had almost identical mean RMSE values across all eight drug categories, indicating their effectiveness in predicting pharmaceutical demand.
-Conversely, deep neural network models like LSTM and Stacked LSTM did not perform as well due to the limited number of samples available for training.
-The research concluded that shallow neural network-based DFMs performed well for five out of eight drug categories, while the ARIMA model was best for the remaining three.
-This suggests that no single forecasting model is optimal for all drug categories[@article{rathipriya2023demand].
+# Table 1: Summary of some studies on forecasting in pharmaceutical supply chain
 
-A study by Candan et al.
-on demand forecasting in the pharmaceutical industry utilized a neuro-fuzzy approach as the forecasting method[@inproceedings{hrupelj2017demand].
-Neuro-fuzzy modeling combines information from empirical models, heuristics, and data, providing effective models that explain solutions better than black-box models like neural networks[@abraham2001neuro; @babuvska2003neuro].
-The study also explored the emotional learning fuzzy inference system (ELFIS) and locally linear neuro-fuzzy (LLNF) methods proposed by Abdollahzade et al.
-(2012) for predicting the stock prices of a pharmaceutical company[@abdollahzade2012application; @hrupelj2017demand].
-While the neuro-fuzzy approach combines fuzzy logic and neural networks to improve forecasting, it may still struggle with extremely complex or non-linear relationships within the data.
-The study used paired t-tests and mean difference analysis to evaluate forecast accuracy.
-Overall, the metrics supported the conclusion that the neuro-fuzzy approach is a robust method for demand forecasting in the pharmaceutical industry, providing accurate predictions closely aligned with actual demand quantities.
+# How the domain knowledge were collected
 
-A study by Burinskiene discusses various forecasting techniques in the pharmaceutical retail industry.
-These include: Single Exponential Smoothing: Applies weighted averages with exponentially decreasing weights for older observations.
-Moving Average: Used for trend-based estimation to predict demand and standard deviation.
-Naive Method: Assumes all predictions are equal to the last observation, serving as a benchmark.
-Holt's Linear Method: An extension of single exponential smoothing for trend prediction.
-Holt-Winters No Trend Method: A variation without a trend component[@burinskiene2022forecasting].
-The study focuses on short-term forecasting, excluding long-term forecasting and safety stock considerations, which are crucial in the pharmaceutical industry due to varying demand patterns[@nikolopoulos2016forecasting].
-The naive method fails to capture trends or seasonality, while the moving average method struggles with sudden changes and irregular patterns[@restyana2021analysis].
-Holt's linear method is limited in handling complex seasonal patterns, and the Holt-Winters method requires careful parameter tuning and historical data analysis[@koehler2001forecasting].
-Single exponential smoothing is not effective for sudden changes or outliers[@inproceedings{zahra2019forecasting].
-The study used MAPE, MSE, and Theil's U2 test to evaluate forecast accuracy, with Theil's U2 test comparing results to the naive forecast.
-A U2 value greater than 1 indicates better performance than the naive method [@burinskiene2022forecasting].
-The Holt-Winters Multiplicative method was found to improve weekly forecasts and outperform other methods, especially in addressing product shortages in pharmacies.
+To gather domain knowledge, we contacted experts from the Ethiopian Pharmaceuticals Supply Service (EPSS) who work in the distribution unit.
+Five experts with over 10 years of experience in managing the distribution of the pharmaceuticals in EPSS agreed to participate.
+The researchers, along with these experts, examined the electronic bin cards of selected pharmaceuticals.
+A bin card is a physical or electronic record used to track the inventory levels of individual items in a storage location, such as a warehouse or stockroom.
+The term 'bin' refers to the specific storage location for an item.
+Information typically recorded on a bin card includes the Item ID, unit of measure, product description, account details, date, references, beginning balance, manufacturer, batch number, expiry date, quantity received, quantity issued, and balance.
+In summary, a bin card is a fundamental tool in inventory management, especially in environments where physical stock tracking is essential.
+It provides a straightforward and immediate way to monitor item quantities, helping to ensure effective inventory management.
 
-A study conducted by Anusha et al.
-in the Indian pharmaceutical retail sector utilized forecasting models such as Simple Moving Average, Exponential Smoothing, and Winter’s Exponential Smoothing[@anusha2014demand].
-These models were employed to predict demand for both seasonal pharmaceutical items like Okacet 10mg tablets and random pharmaceutical items like Stamlo Beta tablets.
-Quantitative indicators like Mean Absolute Deviation (MAD), Mean Squared Error (MSE), and Mean Absolute Percentage Error (MAPE) were used to assess forecast error.
-The Winter’s Exponential Smoothing (WES) model showed the least sales forecast error for the seasonal pharmaceutical item Okacet 10mg tablets, with a MAPE of 27.50, indicating high forecast accuracy.
-Conversely, for the non-seasonal pharmaceutical item Stamlo Beta tablets, the Six-month Moving Average model performed best with a MAPE of 45.62[@anusha2014demand].
-However, the study did not account for unexplained seasonal variations in staff, shifts in the number of potential patients, physician turnover, preferences for pharmaceuticals, and the influence of marketing representatives on surrounding clinics.
-These factors were not factored into the analysis, potentially affecting the accuracy of demand forecasting in the pharmaceutical retail setting.
-These limitations suggest that the study may not have considered all relevant variables impacting demand forecasting accuracy.
+We meticulously traced sales data for the past five years, triangulating graphs with bin card records to identify predictors influencing demand for each product.
+The expert uses their domain knowledge and experience to identify about the distribution charachterstics of the specific products and sometimes when they assume that they do not have enough history or do not remember about that specific products they call to other experts who remember specific issues and decide whetrere the specific predictos affect the distribution and if all of them agree the include it as predcitors.
+The data collection process took more than a month.
+We downloaded and analyzed over 100 bin cards per product.
+Dates on the bin cards were converted from the Ethiopian calendar to the Gregorian calendar for accuracy.
+Transactions due to storage or warehouse changes within EPSS were excluded.
+After cleaning the data,and discussing with the experts  we identified the following key predictors affecting distribution and sales: stock replenishment, stock outs, COVID-19, fiscal year counting, conflict, and malaria seasonality.
 
-Zhu et al.
-conducted a study comparing various forecasting methods in the pharmaceutical industry, including Moving Average (MA), Linear Regression (LR), Exponential Smoothing (ES), Exponential State Space Model (ETS), Vector Autoregressive with exogenous variables (VARX), Symbolic regression, Graph-based analysis, Artificial Neural Networks (ANN), Clustering, and Recurrent Neural Networks (RNN)[@zhu2021demand].
-Additionally, machine learning algorithms such as Linear Regression (LR), Support Vector Regression (SVR), Random Forest (RF), and RNN with cross-series training were assessed.The study evaluated forecast accuracy using metrics like Normalized Mean Squared Error (NMSE), Normalized Mean Absolute Error (NMAE), and Bias.
-Results indicated that the cross-series training models outperformed baseline models (Moving Average, Exponential Smoothing, Linear Regression) in terms of NMSE, NMAE, and Bias.
-RNN models, particularly when implemented with cross-series training, showed superiority and generalizability across datasets compared to other machine learning algorithms (LR, SVR, RF), confirming the effectiveness of the proposed forecasting framework in the pharmaceutical industry.
+•	Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
+Whenever there was stock replenishment at the central EPSS, sales and distribution to hubs and health facilities increased.
+This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.
+•	Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand.
+This can happen when demand exceeds the available supply, or when there is a delay or failure in the replenishment process.
+Resulted in no transactions and a decline in sales.
 
-Ribeiro et al.
-conducted a study aimed at detecting outliers and predicting sales for a pharmaceutical distribution company.
-They employed the dampened Pegels method, a time-series data mining technique, to forecast future quantities of medicines to be ordered by the company[@article{ribeiro2017improving].
-The accuracy of the forecasting model was evaluated using the Symmetric Mean Absolute Percentage Error (SMAPE) metric, which indicated that the Pegels method provided more accurate sales forecasts compared to the company's previous method.
-Furthermore, these forecasts effectively met demand, aiding in efficient stock management and ensuring customer satisfaction while minimizing inventory costs.
+•	COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain.
+The pandemic disrupted global manufacturing and supply, leading to shortages of essential medicines and raw materials, Ethiopia, like many other countries, relies heavily on imported pharmaceutical products.
+With the pandemic causing lockdowns and restrictions in manufacturing hubs like India and China, the supply of essential medicines and raw materials to Ethiopia was significantly disrupted.
+The impact of COVID-19 on sales was significant.
+After the announcement of COVID-19 in Ethiopia, the central EPSS pushed a six-month supply of products to hubs to prevent stock outs.
+This push was from March to August 2020.
+During this period, if products were not stocked out, sales continued to hubs and health facilities.
 
-The study by Bon and Ng at the University Health Centre aimed to enhance inventory demand forecasting for Panadol 650mg through the utilization of various forecasting techniques.
-Their research design encompassed both qualitative and quantitative data collection methods within a descriptive framework.
-Historical data spanning from January 2009 to August 2014 were analyzed employing methods like single moving average, exponential smoothing, and regression analysis.
-Forecast accuracy was assessed using metrics including RMSE, MSE, MAD, and MAPE to identify the most precise forecasting technique.
-The objective was to optimize inventory management by selecting the most accurate forecasting method based on these measurements[@bon2017optimization].
-Results highlighted regression analysis as the superior technique among the ten methods tested, exhibiting the least forecasting error for optimizing Panadol 650mg inventory demand at the University Health Centre[@inproceedings{bon2017optimization].
+•	Physical inventory: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
+The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
+Physical inventory counting periods also influenced sales.
+Stores closed during these periods, halting transactions.
+We observed increased sales before inventory counting periods, as hubs and facilities stocked up.
+July and August were identified as physical counting periods each year.
 
-The study conducted by Newberne and MIKE focused on utilizing Holt-Winters Forecasting in Healthcare, particularly in the context of healthcare management involving Pseudoephedrine prescriptions and Upper Respiratory Infection (URI) visits [@newberne2006holt].
-Data from the Military Healthcare System's M2 database were analyzed using the Holt-Winters method to demonstrate its applicability in healthcare forecasting.
-Forecasting accuracy was assessed using the Mean Absolute Percentage Error (MAPE) metric.
-The study employed various forecasting methods including Naïve, Seasonal Naïve, ARIMA, and Deep Learning Model: Long Short-Term Memory (LSTM) Recurrent Neural Network.
-Additionally, techniques like Simple Exponential Smoothing, Holt with linear and damped trends, Holt-Winters, ES state space model, and Theta were explored to forecast drug quantities for distribution in the Brazilian Public Health System.
-Among these methods, the LSTM model often exhibited the best performance based on Mean Absolute Error (MAE) as a measure of forecast accuracy.
-However, deep learning models like LSTM may pose challenges due to their complexity and computational intensity, requiring specialized expertise for tuning and implementation in practical settings.
+•	Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters.
+Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises.
+Conflicts often lead to the destruction of infrastructure, including roads, bridges, and warehouses, making it difficult to transport pharmaceutical products to affected areas.
+This disruption can cause significant delays or prevent the delivery of essential medicines altogether.
+Internal displacements and conflicts in Ethiopia over the past five years also affected sales and distribution.
+We traced periods of instability and included them as predictors.
+The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.
 
-The study conducted by Zahra and Putra focused on using ARIMA and Single Exponential Smoothing methods to predict and calculate EOQ (Economic Order Quantity).
-To measure forecast accuracy, the researchers utilized RMSE, MAPE, MAD, and MSE metrics.
-The findings revealed that the ARIMA (1,0,0) method exhibited the smallest error values, with a MAPE of 13% and MSE of 283,715, indicating its superior accuracy compared to the Exponential Smoothing method.
-Based on these results, the researchers predicted a need for 222 boxes of medicine for the next period and recommended procuring 223 boxes to maintain a service level of 95%.
-The study underscored the significance of accurate forecasting in optimizing medicine procurement processes and ensuring adequate stock levels to efficiently meet demand.
-By employing MAPE and MSE as forecast accuracy metrics, the researchers could compare the performance of different forecasting methods and determine the most effective approach for predicting medicine needs in hospital settings[@zahra2019forecasting].
+•	Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
+In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease.
+Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced.
+During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments.
+Malaria seasonality was another significant predictor.
+Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks.
+We identified epidemic periods affecting sales: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.
 
-The study conducted by Restyana et al.
-focused on analyzing drug forecasting within the Jombang Regency from 2017 to 2019.
-They employed the Single Moving Average and Single Exponential Smoothing methods to predict drug demand across 35 public health centers.
-The objective was to identify the method with the lowest forecasting error by assessing metrics like Mean Absolute Deviation (MAD) and Mean Square Error (MSE).
-By analyzing annual primary data and testing errors, the researchers aimed to evaluate the accuracy of the forecasting methods.
-The findings revealed that the Single Exponential Smoothing (SES) approach exhibited superior forecast accuracy, as evidenced by lower MAD and MSE values.The study by Kim et al.
-used Vector Autoregressive with Exogenous Variables (VARX) models for demand forecasting of medicines.
-Two models were constructed to test the effect of exogenous variables: Model 1, a pure autoregressive model without exogenous variables, and Model 2, the suggested VARX model incorporating topic trends from social media analysis.
-The study used sales data as endogenous variables and moving averages of topic trends as exogenous variables in the VARX models.The results showed that incorporating topic trends into the VARX models enhanced the performance of demand forecasting, particularly for medicines sensitive to environmental and seasonal changes.
-By analyzing topic trends related to certain diseases and incorporating them into VARX models, the researchers found significant improvement in forecast accuracy.Prediction error rates were used as the forecast accuracy measure, calculated for four different medicines for each model (Model 1 and Model 2).
-The comparison of error rates between the two models demonstrated that adding topic trends as exogenous variables in the VARX model significantly improved the accuracy of demand forecasts, highlighting the importance of integrating social media data into forecasting processes[@kim2015demand].
-
-Table 1 provides a summary of some studies in the literature on forecasting in pharmaceuticals products.
-We have identified several limitations in the existing literature on pharmaceutical product forecasting that have motivated us to pursue this research.
-These limitations can be summarized as follows:
-
-1. Almost all of the studies have primarily concentrated on point forecasts at a single level.
-2. A majority of the research has not explored longer time series data periods.
-3. Only a few studies have integrated domain knowledge with time series data to enhance pharmaceutical forecasting accuracy.
-4. Additionally, no studies have been conducted in African settings, where challenges related to forecasting and pharmaceutical availability are prevalent.
-
-Table 1. Summary of Some Studies on Forecasting in pharmaceuticals products
+The comprehensive analysis and inclusion of these predictors ensured a robust model for forecasting pharmaceutical sales and distribution, leveraging expert domain knowledge and extensive historical data analysis.

### 2024-10-12 manuscript/01-introduction.qmd
Introduction revised according to Prof Bahman comments

@@
 
-Universal access  essential medicines is a cornerstone of effective healthcare systems, yet ensuring equitable availability of these vital medications remains an urgent global challenge[@quick2003essential;@world2004annual].
-Across continents, especially in Africa and Asia, many nations grapple with restricted access due to factors like unaffordability and suboptimal supply chain management [@world2004medicines].
-A survey conducted across eight sub-Saharan African countries indicated that there were  unacceptably low availability of essential medicines.The mean availability of 12 priority essential medicines for women ranged from 22% to 40% and from 28% to 57% for children [@droti2019poor].
-However, the issue of drug shortages extends beyond developing countries, with instances reported in the United States and Europe, adversely impacting healthcare systems and compromising patient care quality [@fox2003managing, @kaakeh2011impact; @johnson2011drug; @huys2013european;@le2011prevalence].
+Universal access to essential medicines is a fundamental goal of effective healthcare systems, yet ensuring equitable availability continues to be a critical global challenge [@quick2003essential; @world2004annual].
+Many countries, particularly in Africa and Asia, face significant barriers to accessing these medications, often due to high costs and inefficient supply chain management [@world2004medicines].
+A survey in eight sub-Saharan African countries revealed concerningly low availability of essential medicines, with mean availability ranging from 22% to 40% for women’s priority medicines and from 28% to 57% for children’s medicines [@droti2019poor].
+However, drug shortages are not limited to low-income regions; they also occur in developed areas like the United States and Europe, where they disrupt healthcare delivery and compromise the quality of patient care [@fox2003managing; @kaakeh2011impact; @johnson2011drug; @huys2013european; @le2011prevalence].
 
-The consequences of drug shortages are felt both financially within the healthcare sector and in terms of the quality of patient care [@alspach2012drug; @kaakeh2011impact;@kaakeh2011impact; @baumer2004national].The consequences extend across patients, providers, healthcare institutions, and research programs.
-Patients face delays or cancellations of treatments, substandard care due to the unavailability of optimal therapies, and increased costs in a secondary "grey market." Compromised patient safety, complications, prolonged hospital stays, and even deaths have been reported [@alspach2012drug;].Addressing these shortages is crucial to preserving the quality of patient care and ensuring the effectiveness of healthcare delivery.
-Thus, a smoothly operating pharmaceutical supply chain is essential to ensure timely access to essential medicines, making accurate demand forecasting an indispensable component.
+The consequences of drug shortages extend beyond immediate healthcare delivery, impacting healthcare costs, treatment quality, and patient safety [@alspach2012drug; @kaakeh2011impact; @baumer2004national].
+Such shortages can result in treatment delays, suboptimal care due to the unavailability of preferred therapies, increased reliance on costly secondary markets, and serious patient outcomes, including complications, longer hospital stays, and even deaths [@alspach2012drug].
+Addressing these challenges requires a resilient pharmaceutical supply chain that can ensure the timely availability of medicines.
+In this context, accurate demand forecasting becomes critical for effective procurement and inventory management [@subramanian2021effective].
 
-An efficient pharmaceutical supply chain is vital for timely access to essential health commodities.
-One of the most crucial components of such a supply chain is a well-defined demand forecasting process.
-Forecasting accurately the future demand for essential medicines provides valuable insights for anticipating future healthcare requirements including procurement and replenishment of inventory[@subramanian2021effective].
-Despite the significance of forecasting for supply chain management, pharmaceutical product forecasting remains a complex endeavor influenced by a multitude of factors.
-Key considerations include the availability of data, the specific position within the supply chain, and the nuanced understanding of contributing elements.
-The position within the supply chain introduces unique challenges and dynamics, requiring tailored forecasting approaches.
-Additionally, a nuanced understanding of contributing elements involves grasping intricate factors that impact forecasting accuracy, such as market trends, regulatory changes, and external influences.
-Schneider, Wilson, and Rosenbeck (2010) and Hyndman and Athanasopoulos (2018) highlight the intricate nature of pharmaceutical forecasting, emphasizing the need for a comprehensive approach that incorporates these multifaceted aspects to enhance supply chain management effectiveness[@schneider2010pharmaceutical; @hyndman2018forecasting].
+Demand forecasting in pharmaceutical supply chains is a complex endeavor, influenced by factors such as data quality, supply chain positioning, and external drivers, including market trends and regulatory changes [@schneider2010pharmaceutical; @hyndman2018forecasting].
+While vast amounts of data are often collected for transactional purposes—such as sales—many forecasting approaches overlook the significance of domain knowledge, which is as crucial as quantitative data for understanding the context and making more informed predictions.
+Domain expertise helps identify the impact of nuanced factors, such as clinical guidelines, disease outbreaks, and local policies, that are not always evident from sales data alone.
 
-The Ethiopian pharmaceutical sector's distinct characteristics, marked by a diverse array of medicines and unique data challenges stemming from service delivery points, staff capacity limitations, communication hurdles, prolonged procurement and distribution lead times, poor forecast accuracy, and policy intricacies, necessitates tailored forecasting strategies that cater to the nation's specific context [@boche2022procurement].
-In this context, a comprehensive exploration of effective forecasting models that align with Ethiopia's pharmaceutical landscape is paramount to mitigating stock outs and enhancing healthcare delivery.
-The implications of pharmaceutical demand forecasting extend beyond financial considerations to societal well-being [@rostami2022forecasting].
-In Ethiopia, a country undergoing demographic and healthcare evolution, precise demand forecasting assumes immense importance for resource allocation, supply chain optimization, and policy formulation.
-With mounting interest in refining demand forecasting in Ethiopia's pharmaceutical sector, this study seeks to contribute to existing knowledge by investigating and recommending proficient demand forecasting methods.
-Our investigation centers on the application of mean, naive, exponential smoothing, regression, and ARIMA models to bolster forecast accuracy and effectiveness.
-This paper aims to forecast the demand for selected pharmaceuticals in EPSS over the next 12 months by utilizing various forecasting models and integrating predictors.
-The integration of predictors with the models will be examined to determine whether they enhance the forecasting accuracy.
-These predictors are derived from a historical analysis of events in Ethiopia over the last 5 years that have impacted pharmaceutical sales in EPSS.
-These events include COVID-19, conflicts in the country, disease campaigns, budget closures, budget releases, and physical inventory, all of which, in varying degrees, influence the demand for pharmaceuticals in Ethiopia.
+Combining statistical models with domain insights offers a more refined forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview].
+Domain knowledge enables analysts to incorporate relevant factors, such as market conditions, policy changes, and supply chain disruptions, into the forecasting model.
+It aids in guiding feature selection, parameter tuning, and the interpretation of complex data patterns, ultimately leading to more accurate and actionable forecasts [@sinha2008incorporating; @dash2022review].
+This integration is particularly beneficial in fields like healthcare, where expert insights can align machine learning algorithms with real-world applications and ethical considerations [@dash2022review].
 
-The current paper offers the following contributions we aim at forecasting pharmaceutical sales over a 12-month period these models aim to enhance the accuracy and reliability of demand predictions within the context of EPSS.
-Moreover an essential contribution lies in the identification of key factors influencing pharmaceutical sales.
-The study develops a model that effectively incorporates these driving factors, providing a more nuanced and comprehensive approach to demand prediction.
-We generate forecasts and evaluate model performance using the Root Mean Squared Error (RMSE), Root Mean Squared Scaled Error (RMSSE) and Mean Absolute Error (MAE), Mean Absolute Scaled Error (MASE), Crips and Winkler are used to assess probabilistic forecast accuracy.
-#how you know the forecasting accutacy problem and provide eveidence about the forecasting problems 
-# most forecast focus on point forecast and we want to focus on probablistic forecast
-# i will write the contribution of the reearch rather than these objectives
+Despite the considerable body of research on integrating domain knowledge with statistical techniques, relatively few studies focus specifically on the unique requirements of pharmaceutical forecasting.
+This study aims to bridge that gap by employing advanced machine learning models combined with domain expertise to address the distinctive characteristics of Ethiopia’s pharmaceutical sector.
+The challenges in this sector, such as diverse product categories, limited data from service delivery points, communication issues, long lead times, forecast inaccuracies, and complex policies, necessitate tailored forecasting approaches [@bilal2024challenges; @boche2022procurement].
+Accurate demand forecasting in Ethiopia is essential for optimizing resource allocation, streamlining supply chain operations, and shaping effective healthcare policies [@rostami2022forecasting].
 
-This research is guided by several objectives:
+This research will forecast the demand for selected pharmaceuticals in the Ethiopian Pharmaceutical Supply Service (EPSS) over the next six months using various methods, including dynamic regression ARIMA, multiple regression, neural networks, machine learning, and time-series GPT models.
+We will explore whether incorporating domain knowledge can improve forecast accuracy by integrating predictors derived from historical events over the past five years, such as stock replenishment cycles, stockouts, fiscal calendar variations, COVID-19, conflicts, and seasonal trends like malaria outbreaks.
 
-1. Evaluate the robustness of various forecasting models for pharmaceutical demand prediction within EPSS.
-2. Identify pharmaceuticals that exhibit superior forecast accuracy under specific models.
-3. Assess the performance of forecast accuracy metrics to gauge the efficacy of selected models.
-4. Test the incorporation of diverse predictors affecting forecasts and evaluate resulting forecasting accuracy.
-In pursuing these objectives, this study endeavors to elevate the precision and reliability of pharmaceutical demand forecasting, thereby contributing to optimized supply chain management and enhanced healthcare outcomes in Ethiopia.
+Our contributions include not only generating point forecasts but also employing probabilistic forecasting techniques that capture the range of potential outcomes with associated probabilities, offering a more comprehensive view of demand uncertainty than traditional methods.
+Additionally, we identify key factors influencing pharmaceutical demand and incorporate these into our forecasting models to better reflect the complexities of the Ethiopian context.
+Model performance will be evaluated using metrics such as Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), and probabilistic measures like the Continuous Ranked Probability Score (CRPS) and Winkler score.
+
+The remainder of this paper is structured as follows: Section 2 reviews the literature and identifies gaps to position our work; Section 3 outlines the experimental design, including data sources, forecasting methods, and evaluation criteria; Section 4 presents the results and analysis; and Section 5 concludes with a summary of findings and future research directions.

### 2024-10-18 manuscript/03-experiment.qmd
Experiment revised according to Prof Bahman

@@
 Rigorous checks were conducted to ensure the consistency and completeness of the collected data.
-From the extensive pool of pharmaceutical products within EPSS, we strategically selected a set of 37 key pharmaceuticals, encompassing various programs and representing different classes of drugs.
+From the extensive pool of pharmaceutical products within EPSS, we strategically selected a set of 33 key pharmaceuticals, encompassing various programs and representing different classes of drugs.
 This deliberate selection enriches the research by providing a comprehensive representation across the pharmaceutical landscape.
 We are presenting the graph for some of the pharmaceuticals sales for the last five years by using visualization for the readers to undesrtand the sales of the pharmaceuticals.
+
+# How the domain knowledge were collected (Key informant interview)
+
+To gather domain knowledge, we contacted experts from the Ethiopian Pharmaceuticals Supply Service (EPSS) who work in the distribution unit.
+Five experts with over 10 years of experience in managing the distribution of the pharmaceuticals in EPSS agreed to participate.
+The researchers, along with these experts, examined the electronic bin cards of selected pharmaceuticals.
+A bin card is a physical or electronic record used to track the inventory levels of individual items in a storage location, such as a warehouse or stockroom.
+The term 'bin' refers to the specific storage location for an item.
+Information typically recorded on a bin card includes the Item ID, unit of measure, product description, account details, date, references, beginning balance, manufacturer, batch number, expiry date, quantity received, quantity issued, and balance.
+In summary, a bin card is a fundamental tool in inventory management, especially in environments where physical stock tracking is essential.
+It provides a straightforward and immediate way to monitor item quantities, helping to ensure effective inventory management.
+
+We meticulously traced sales data for the past five years, triangulating graphs with bin card records to identify predictors influencing demand for each product.
+The expert uses their domain knowledge and experience to identify about the distribution charachterstics of the specific products and sometimes when they assume that they do not have enough history or do not remember about that specific products they call to other experts who remember specific issues and decide whetrere the specific predictos affect the distribution and if all of them agree the include it as predcitors.
+The data collection process took more than a month.
+We downloaded and analyzed over 100 bin cards per product.
+Dates on the bin cards were converted from the Ethiopian calendar to the Gregorian calendar for accuracy.
+Transactions due to storage or warehouse changes within EPSS were excluded.
+After cleaning the data,and discussing with the experts  we identified the following key predictors affecting distribution and sales: stock replenishment, stock outs, COVID-19, fiscal year counting, conflict, and malaria seasonality.
+
+•	Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
+Whenever there was stock replenishment at the central EPSS, sales and distribution to hubs and health facilities increased.
+This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.
+•	Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand.
+This can happen when demand exceeds the available supply, or when there is a delay or failure in the replenishment process.
+Resulted in no transactions and a decline in sales.
+
+•	COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain.
+The pandemic disrupted global manufacturing and supply, leading to shortages of essential medicines and raw materials, Ethiopia, like many other countries, relies heavily on imported pharmaceutical products.
+With the pandemic causing lockdowns and restrictions in manufacturing hubs like India and China, the supply of essential medicines and raw materials to Ethiopia was significantly disrupted.
+The impact of COVID-19 on sales was significant.
+After the announcement of COVID-19 in Ethiopia, the central EPSS pushed a six-month supply of products to hubs to prevent stock outs.
+This push was from March to August 2020.
+During this period, if products were not stocked out, sales continued to hubs and health facilities.
+
+•	Physical inventory: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
+The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
+Physical inventory counting periods also influenced sales.
+Stores closed during these periods, halting transactions.
+We observed increased sales before inventory counting periods, as hubs and facilities stocked up.
+July and August were identified as physical counting periods each year.
+
+•	Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters.
+Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises.
+Conflicts often lead to the destruction of infrastructure, including roads, bridges, and warehouses, making it difficult to transport pharmaceutical products to affected areas.
+This disruption can cause significant delays or prevent the delivery of essential medicines altogether.
+Internal displacements and conflicts in Ethiopia over the past five years also affected sales and distribution.
+We traced periods of instability and included them as predictors.
+The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.
+
+•	Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
+In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease.
+Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced.
+During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments.
+Malaria seasonality was another significant predictor.
+Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks.
+We identified epidemic periods affecting sales: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.
 

### 2024-10-24 manuscript/01-introduction.qmd
my first draft

@@
 Many countries, particularly in Africa and Asia, face significant barriers to accessing these medications, often due to high costs and inefficient supply chain management [@world2004medicines].
-A survey in eight sub-Saharan African countries revealed concerningly low availability of essential medicines, with mean availability ranging from 22% to 40% for women’s priority medicines and from 28% to 57% for children’s medicines [@droti2019poor].
+A survey conducted in eight sub-Saharan African countries revealed alarmingly low availability of essential medicines.
+The average availability of women's priority medicines ranged from just 22% to 40%, while children's medicines were only slightly more accessible, with availability ranging from 28% to 57% [@droti2019poor].
 However, drug shortages are not limited to low-income regions; they also occur in developed areas like the United States and Europe, where they disrupt healthcare delivery and compromise the quality of patient care [@fox2003managing; @kaakeh2011impact; @johnson2011drug; @huys2013european; @le2011prevalence].
@@
 
-Demand forecasting in pharmaceutical supply chains is a complex endeavor, influenced by factors such as data quality, supply chain positioning, and external drivers, including market trends and regulatory changes [@schneider2010pharmaceutical; @hyndman2018forecasting].
-While vast amounts of data are often collected for transactional purposes—such as sales—many forecasting approaches overlook the significance of domain knowledge, which is as crucial as quantitative data for understanding the context and making more informed predictions.
-Domain expertise helps identify the impact of nuanced factors, such as clinical guidelines, disease outbreaks, and local policies, that are not always evident from sales data alone.
+Demand forecasting in pharmaceutical supply chains is a highly complex process, shaped by several factors.
+These include the quality of available data, the specific position within the supply chain, and external influences such as market trends, regulatory changes, and disruptive events such as pandemics, conflicts, flooding.
+[@schneider2010pharmaceutical; @hyndman2018forecasting].
+Recently, major donors  such as Bill & Melinda Gates Foundation have invested in data collection technologies in sub-Saharan Africa, recognizing the potential of data to provide valuable insights.
+This investment is essential and offers significant benefits for improving supply chain delivery.
+However, much of the data being collected focuses on transactional data, such as consumption and sale.
+Almost all platforms collecting data tend to overlook the critical contextual factors and events that influence these transactions—factors that are just as crucial for accurate forecasting.
+This presents a major barrier, as understanding the broader context, including administration procedures, local policies, disruptions, conflicts, and so on is vital for building reliable forecasting models.
+This gap—known as business context and domain knowledge—requires urgent attention in Pharmaceutical supply chain forecasting.
+Without incorporating this knowledge, any forecasting model is unlikely to produce reliable results.
+We must prioritize collecting and integrating this contextual information to improve the accuracy of demand forecasts in pharmaceutical supply chains.
+Combining statistical models with domain insights offers a more refined forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview].
+Domain knowledge enables analysts to incorporate relevant events, such as supply chain disruptions, into the forecasting model.
+This integration is particularly beneficial in fields like healthcare, where expert insights can align algorithms with real-world applications [@dash2022review].
 
-Combining statistical models with domain insights offers a more refined forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview].
-Domain knowledge enables analysts to incorporate relevant factors, such as market conditions, policy changes, and supply chain disruptions, into the forecasting model.
-It aids in guiding feature selection, parameter tuning, and the interpretation of complex data patterns, ultimately leading to more accurate and actionable forecasts [@sinha2008incorporating; @dash2022review].
-This integration is particularly beneficial in fields like healthcare, where expert insights can align machine learning algorithms with real-world applications and ethical considerations [@dash2022review].
-
-Despite the considerable body of research on integrating domain knowledge with statistical techniques, relatively few studies focus specifically on the unique requirements of pharmaceutical forecasting.
 This study aims to bridge that gap by employing advanced machine learning models combined with domain expertise to address the distinctive characteristics of Ethiopia’s pharmaceutical sector.
@@
 
-This research will forecast the demand for selected pharmaceuticals in the Ethiopian Pharmaceutical Supply Service (EPSS) over the next six months using various methods, including dynamic regression ARIMA, multiple regression, neural networks, machine learning, and time-series GPT models.
-We will explore whether incorporating domain knowledge can improve forecast accuracy by integrating predictors derived from historical events over the past five years, such as stock replenishment cycles, stockouts, fiscal calendar variations, COVID-19, conflicts, and seasonal trends like malaria outbreaks.
+This research uses consumption data from 33 health commodities, covering the period from December 2017 to July 2022, provided by the Ethiopian Pharmaceutical Supply Service (EPSS), to develop models for six-month-ahead forecasting.
+Initially, we implement univariate models that rely solely on consumption data.
+Following this, we conduct exploratory data analysis and engage in interviews with EPSS experts to gather domain knowledge on events influencing consumption patterns.
+Through these interviews, the experts identified five key factors collectively impacting consumption: stock replenishment period, stockouts, fiscal calendar, COVID-19, conflicts, and the seasonal malaria.
+We then develop forecasting models that incorporate the data gathered through the interview process and compare their performance with univariate models that rely solely on consumption data.
 
-Our contributions include not only generating point forecasts but also employing probabilistic forecasting techniques that capture the range of potential outcomes with associated probabilities, offering a more comprehensive view of demand uncertainty than traditional methods.
-Additionally, we identify key factors influencing pharmaceutical demand and incorporate these into our forecasting models to better reflect the complexities of the Ethiopian context.
-Model performance will be evaluated using metrics such as Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), and probabilistic measures like the Continuous Ranked Probability Score (CRPS) and Winkler score.
+Our primary aims is to assess whether incorporating domain knowledge improves forecast accuracy.
+Moreover, we also fill a gap in the pharmaceutical demand forecasting literature by not only generating point forecasts but also producing probabilistic forecasting techniques that capture the range of potential outcomes with associated probabilities, offering a more comprehensive view of demand uncertainty than traditional methods.
+Model performance will be evaluated using point accuracy measure including Mean Absolute Scaled Error(MASE) and Root Mean Squared Scaled Error (RMSSE), and probabilistic measures like the Continuous Ranked Probability Score (CRPS) and Winkler score.
 

### 2024-10-24 manuscript/02-literature.qmd
my first draft

@@
 
-In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies utilizing a combination of simple and advanced methods to enhance accuracy and reliability[@rathipriya2023demand,@merkuryeva2019demand,@nikolopoulos2016forecasting,@van2021using,@khalil2014intelligent,@zhu2021demand,@anusha2014demand,@burinskiene2022forecasting,@bon2017optimization,@sousa2019statistical,@zahra2019forecasting].
+In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies using a combination of simple and advanced methods to enhance accuracy [@nikolopoulos2016forecasting,@zhu2021demand,@anusha2014demand].
 Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods like the Holt-Winters method, Single Moving Average (SMA), and Single Exponential Smoothing (SES), which are simpler but effective for certain applications.
@@
 
-# Table 1: Summary of some studies on forecasting in pharmaceutical supply chain
+@tbl-literature provides a summary of key studies in the literature on forecasting for pharmaceutical products.
+We identify several limitations both in the existing research and in practice, which highlight important gaps that motivate our current study.
+These gaps are summarized as follows:
 
-# How the domain knowledge were collected
+1.
+Despite significant investment in data collection technologies and logistics management systems, these systems primarily collect transactional data, such as consumption.
+However, no data is collected on events that influence consumption variability—information that would be crucial for building reliable forecasting models.
+No studies to date have explored how domain expertise could enhance pharmaceutical demand forecasting.
 
-To gather domain knowledge, we contacted experts from the Ethiopian Pharmaceuticals Supply Service (EPSS) who work in the distribution unit.
-Five experts with over 10 years of experience in managing the distribution of the pharmaceuticals in EPSS agreed to participate.
-The researchers, along with these experts, examined the electronic bin cards of selected pharmaceuticals.
-A bin card is a physical or electronic record used to track the inventory levels of individual items in a storage location, such as a warehouse or stockroom.
-The term 'bin' refers to the specific storage location for an item.
-Information typically recorded on a bin card includes the Item ID, unit of measure, product description, account details, date, references, beginning balance, manufacturer, batch number, expiry date, quantity received, quantity issued, and balance.
-In summary, a bin card is a fundamental tool in inventory management, especially in environments where physical stock tracking is essential.
-It provides a straightforward and immediate way to monitor item quantities, helping to ensure effective inventory management.
+2.
+Current research on forecasting for pharmaceutical supply chain predominantly focuses on generating point forecasts.
+There is a lack of studies that consider the entire forecast distribution of monthly consumption, which would better capture the uncertainty of future demand and provide a valuable risk management tool for decision-makers.
 
-We meticulously traced sales data for the past five years, triangulating graphs with bin card records to identify predictors influencing demand for each product.
-The expert uses their domain knowledge and experience to identify about the distribution charachterstics of the specific products and sometimes when they assume that they do not have enough history or do not remember about that specific products they call to other experts who remember specific issues and decide whetrere the specific predictos affect the distribution and if all of them agree the include it as predcitors.
-The data collection process took more than a month.
-We downloaded and analyzed over 100 bin cards per product.
-Dates on the bin cards were converted from the Ethiopian calendar to the Gregorian calendar for accuracy.
-Transactions due to storage or warehouse changes within EPSS were excluded.
-After cleaning the data,and discussing with the experts  we identified the following key predictors affecting distribution and sales: stock replenishment, stock outs, COVID-19, fiscal year counting, conflict, and malaria seasonality.
+3.
+Reproducibility remains a major challenge in this field.
+It is often difficult for readers to reproduce previous studies without direct assistance from the authors, limiting the practical application and validation of existing research.
 
-•	Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
-Whenever there was stock replenishment at the central EPSS, sales and distribution to hubs and health facilities increased.
-This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.
-•	Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand.
-This can happen when demand exceeds the available supply, or when there is a delay or failure in the replenishment process.
-Resulted in no transactions and a decline in sales.
-
-•	COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain.
-The pandemic disrupted global manufacturing and supply, leading to shortages of essential medicines and raw materials, Ethiopia, like many other countries, relies heavily on imported pharmaceutical products.
-With the pandemic causing lockdowns and restrictions in manufacturing hubs like India and China, the supply of essential medicines and raw materials to Ethiopia was significantly disrupted.
-The impact of COVID-19 on sales was significant.
-After the announcement of COVID-19 in Ethiopia, the central EPSS pushed a six-month supply of products to hubs to prevent stock outs.
-This push was from March to August 2020.
-During this period, if products were not stocked out, sales continued to hubs and health facilities.
-
-•	Physical inventory: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
-The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
-Physical inventory counting periods also influenced sales.
-Stores closed during these periods, halting transactions.
-We observed increased sales before inventory counting periods, as hubs and facilities stocked up.
-July and August were identified as physical counting periods each year.
-
-•	Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters.
-Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises.
-Conflicts often lead to the destruction of infrastructure, including roads, bridges, and warehouses, making it difficult to transport pharmaceutical products to affected areas.
-This disruption can cause significant delays or prevent the delivery of essential medicines altogether.
-Internal displacements and conflicts in Ethiopia over the past five years also affected sales and distribution.
-We traced periods of instability and included them as predictors.
-The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.
-
-•	Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
-In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease.
-Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced.
-During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments.
-Malaria seasonality was another significant predictor.
-Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks.
-We identified epidemic periods affecting sales: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.
-
-The comprehensive analysis and inclusion of these predictors ensured a robust model for forecasting pharmaceutical sales and distribution, leveraging expert domain knowledge and extensive historical data analysis.
+```{r}
+#| label: tbl-literature
+#| echo: false
+#| cache: false
+#| tbl-cap: "Summary of some studies on forecasting in pharmaceutical supply chain"
+library(kableExtra)
+readxl::read_xlsx("Literature_review_table.xlsx") |>
+  knitr::kable(booktabs = T, linesep = "") |>
+  kable_styling(latex_options = "scale_down", font_size = 11) |>
+  column_spec(3, "6em") |>
+  column_spec(5, "4em") |>
+  column_spec(6, "10em") |>
+  column_spec(7, "8em") |>
+  landscape()
+```

### 2024-10-24 manuscript/03-experiment.qmd
my first draft

@@
 
-### Dataset
-
-For this study, we utilized a dataset spanning five years (2018-2022) of sales data obtained from EPSS.
+For this study, we use a dataset spanning five years (Dec.
+2017- July 2022) of consumption data obtained from EPSS.
 Rigorous checks were conducted to ensure the consistency and completeness of the collected data.
@@
 This deliberate selection enriches the research by providing a comprehensive representation across the pharmaceutical landscape.
+
 We are presenting the graph for some of the pharmaceuticals sales for the last five years by using visualization for the readers to undesrtand the sales of the pharmaceuticals.
 
-# How the domain knowledge were collected (Key informant interview)
+### Explaoratory Data Analysis with consumption data
+
+Given the number of produts in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understadn data.
+@fig-feature shows the strength of trend versus seasonality.
+Each point represents one time series with the strength of trend in x-axis and the strength of seasonality in y-axis.
+Both measures are on a scale of [0,1].
+the strength of trend and seasonality were calculated using the "STL" (Seasonal and Trend decomposition using Loess) decomposition method, as described by @mstl.
+
+```{r}
+#| label: fig-feature
+#| cache: true
+#| out.width: "70%"
+#| fig.align: center
+#| fig-cap: "The strength of the trend and seasonality in the time series of pharmaceutical product consumption. The scatter plot shows 33 data points, with each point corresponding to a product."
+read_csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
+  janitor::clean_names()
+med_qty_tsb <- med_qty |> 
+  mutate(month = yearmonth(month)) |> 
+  as_tsibble(index = month, key = item)
+
+med_qty_tsb %>%
+  features(quantity, feat_stl) %>%
+  ggplot(aes(x = trend_strength, y = seasonal_strength_week)) +
+  geom_point(alpha = 0.25) +
+  labs(x = "Strength of trend", y = "Strength of weekly seasonality")
+```
+
+It is clear that there are some series showing strong trends and/or seasonality, corresponding to series at the higher levels of the hierarchy.
+The majority of series show low trend and seasonality.
+These are time series belonging to the bottom series, series related to the nature of incidents for a given control, health board and priority level.
+Bottom series are dominated by noise with little or no systematic patterns.
+
+In addition to displaying the strength of trend and seasonality, we also visualize all time series, revealing various patterns such as trends, seasonality, and instances of consumption with chaotic behavior during certain months.
+For instance, some series exhibit both seasonality and trend, while others show a low volume of attended incidents and higher entropy, making them more volatile and challenging to forecast.
+At the incident level, combined with categories from other levels, many series contain zeros or low counts.
+This highlights the diversity of daily time series patterns in the data set.
+Figure @fig-dataviz2 illustrates the time plots for a few selected products.
+
+```{r}
+#| label: fig-dataviz2
+#| cache: true
+#| dependson: "fig-feature"
+#| out.width: "100%"
+#| fig-width: 8
+#| fig-height: 10
+#| fig-cap: "Monthly time plot of consumption.
+X-axis shows the month, consisting of 60 data points (months) and y-axis shows the comsumption.
+The panels display data from four products to give a glimpse of the consumption patterns."
+
+(my_plot[[1]] | my_plot[[1]]) /
+(my_plot[[1]] | my_plot[[1]])
+
+```
+
+Visualizing the data revealed that various events significantly impacted the consumption of different products, but these were not reflected in the available data.
+Expert insights were crucial for understanding the nature and timing of these events, filling gaps in the system, and incorporating qualitative factors that EPSS logistics system miss.
+Therefore, we conducted interviews with domain experts to collect information on external factors that influence consumption.
+These interviews allowed us to account for unusual patterns and customize our forecasting models to more accurately reflect the unique consumption behaviors of each product.
+
+### Domain knowledge colelction though interview with EPSS experts
 
@@
 
-•	Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
+- Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
 Whenever there was stock replenishment at the central EPSS, sales and distribution to hubs and health facilities increased.
 This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.
-•	Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand.
+
+- Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand.
 This can happen when demand exceeds the available supply, or when there is a delay or failure in the replenishment process.
@@
 
-•	COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain.
+- COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain.
 The pandemic disrupted global manufacturing and supply, leading to shortages of essential medicines and raw materials, Ethiopia, like many other countries, relies heavily on imported pharmaceutical products.
@@
 
-•	Physical inventory: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
+- Physical inventory: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
 The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
@@
 
-•	Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters.
+- Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters.
 Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises.
@@
 
-•	Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
+- Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
 In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease.
@@
 
-```{r}
-#| include: false
-library(tidyverse)
-library(lubridate)
-library(fpp3)
-library(ggthemes)
-monthly_issues_tsibble <- read_rds("../data/monthly_issues_tsibble.rds")
-monthly_issues_tsibble <- monthly_issues_tsibble |> filter_index("2017 Aug" ~ "2022 Jun")
+## Forecasting models
 
-knitr::opts_chunk$set(fig.align = 'center', out.width = "90%")
+**Multiple Linear Regression (MLR)**:
 
-monthly_issues_tsibble |> as_tibble() |> select(month) -> predictors
+**ARIMA and ARIMAX (LSTM)**:
 
-write_csv(predictors, "../data/predictors.csv")
-getwd()
-```
+**Long Short-Term Memory neural network (LSTM)**: The LSTM model is a specialised form of recurrent neural network (RNN) used to model sequential data by capturing long-term dependencies \[\@<https://doi.org/10.1162/neco.1997.9.8.1735>\].
+Unlike traditional RNNs, LSTMs can learn to retain information for longer time periods due to their unique architecture, which consists of several gates that control the flow of information.
+This makes LSTMs particularly effective for time series forecasting.
+In our implementation, we used a sequential model architecture, comprising one LSTM layer with 50 units, followed by dense layers, with the final output being a single linear unit.
+The Adam optimizer was employed to minimize the mean squared error, and the model was trained for 100 epochs using the keras_model_sequential() function from the Keras package in R.
 
-```{r}
-#| include: false
-items <- unique(monthly_issues_tsibble$item)
-issue_plot_monthly <- function(my_item) {
-  ggplot(data = monthly_issues_tsibble |> filter(item == my_item), mapping = aes(x = month, y = quantity_product_issued))+
-    geom_point(size=1)+
-    geom_line()+
-    theme_few()+
-    labs(title=my_item)+
-    scale_x_yearmonth(date_breaks = "2 month")+
-    theme(axis.text.x = element_text(angle=90))
-}
+**TimeGPT**: TimeGPT is the first pre-trained foundational model designed specifically for time series forecasting, developed by Nixtla [@[https://doi.org/10.48550/arXiv.2310.0358910.03589](https://doi.org/10.48550/arXiv.2310.03589)].
+It utilizes a transformer-based architecture with an encoder-decoder configuration but differs from other models in that it is not based on large language models.
+Instead, it is built from the ground up to handle time series data.
+TimeGPT was trained on more than 100 billion data points, drawing from publicly available time series across various sectors, such as retail, healthcare, transportation, demographics, energy, banking, and web traffic.
+This wide range of data sources, each with unique temporal patterns, enables the model to manage diverse time series characteristics effectively.
+Furthermore, TimeGPT supports the inclusion of external regressors in its forecasts and can generate quantile forecasts, providing reliable uncertainty estimation.
 
-my_plot <- list()
-for (i in (1:40)) {
-  my_plot[[i]] <- issue_plot_monthly(items[i])
-}
-```
+## Performance evaluation
 
-## Visualising monthly time series
+To evaluate the performance of the various forecasting approaches, we split the data into a series of ten training and test sets.
+We use a time series cross-validation approach [@hyndman2021forecasting], with a forecast horizon of 84 days, and each training set expanding in 42-day steps.
+The first training set uses all data up to 2018-04-25, and the first test set uses the 84 days beginning 2018-04-26.
+The second training set uses all data up to 2018-06-06, with the second test set using the following 84 days.
+The largest training set ends on 2019-05-09, with the test set ending on 2019-07-31.
+Model development and hyper-parameter tuning is performed using the training data and the errors are assess using the corresponding test set.
+While we compute forecast errors for the entire 12 weeks, we are most interested in the last 42 days of each test set, because that corresponds to how forecasts are generated for planning in practice.
+Forecasting performance is evaluated using both point and probabilistic error measures.
 
-```{r}
-my_plot[[1]]
-```
+The error metrics provided below consider a forecasting horizon denoted by $j$, representing the number of time periods ahead we are predicting.
+In our study, this forecasting horizon ranges from 1 to 84 days, $j= 1,2,\dots, 84$.
 
-```{r}
-my_plot[[2]]
-```
+Point forecast accuracy is measured via the Mean Squared Scaled Error (MSSE) and the Mean Absolute Scaled Error (MASE).
+The Mean Absolute Scaled Error (MASE) [@HK06; @hyndman2021forecasting] is calculated as:
+$$
+  \text{MASE} = \text{mean}(|q_{j}|),
+$$
+where
+$$
+  q_{j} = \frac{ e_{j}}
+ {\displaystyle\frac{1}{T-m}\sum_{t=m+1}^T |y_{t}-y_{t-m}|},
+$$
+and $e_{j}$ is the point forecast error for forecast horizon $j$, $m = 7$ (as we have daily seasonal series), $y_t$ is the observation for period $t$, and $T$ is the sample size (the number of observations used for training the forecasting model).
+The denominator is the mean absolute error of the seasonal naive method in the fitting sample of $T$ observations and is used to scale the error.
+Smaller MASE values suggest more accurate forecasts.
+Note that the measure is scale-independent, thus allowing us to average the results across series.
 
-```{r}
-my_plot[[4]]
-```
+A related measure is MSSE [@hyndman2021forecasting;@makridakis2022m5], which uses squared errors rather than absolute errors:
+$$
+  \text{MSSE} = \text{mean}(q_{j}^2),
+$$ {#eq-RMSSE}
+where,
+$$
+  q^2_{j} = \frac{ e^2_{j}}
+ {\displaystyle\frac{1}{T-m}\sum_{t=m+1}^T (y_{t}-y_{t-m})^2},
+$$
+Again, this is scale-independent, and smaller MSSE values suggest more accurate forecasts.
 
-```{r}
-my_plot[[6]]
-```
-```{r}
-my_plot[[9]]
-```
+Using scale-independent measures, such as MASE and MSSE, enables more appropriate comparisons between time series at different levels and scales, as these measures are not influenced by the magnitude of the data.
+This is of particular importance in our study, as we work with time series at various levels of hierarchy, with varying scales, resulting in different magnitudes of error.
+By employing scale-independent measures, we can meaningfully assess the forecast accuracy across the entire hierarchy, ensuring a more robust comparison.
 
-```{r}
-my_plot[[11]]
-```
-```{r}
-my_plot[[12]]
-```
-```{r}
-my_plot[[15]]
-```
-```{r}
-my_plot[[17]]
-```
-```{r}
-my_plot[[19]]
-```
-```{r}
-my_plot[[21]]
-```
-```{r}
-my_plot[[25]]
-```
-```{r}
-my_plot[[26]]
-```
-```{r}
-my_plot[[29]]
-```
-```{r}
-my_plot[[30]]
-```
-```{r}
-my_plot[[32]]
-```
-```{r}
-my_plot[[36]]
-```
-```{r}
-my_plot[[37]]
-```
+To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@gneiting2014probabilistic;@hyndman2021forecasting].
+It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution.
+$$
+  \text{CRPS} = \text{mean}(p_j),
+$$ {#eq-CRPS}
+where
+$$
+  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
+$$
+where $G_j(x)$ is the forecasted probability distribution function for forecast horizon $j$, and $F_j(x)$ is the true probability distribution function for the same period.
 
-# 2. 2. Data Splitting and Preparation
+Calibration refers to the statistical consistency between the distributional forecasts and the observations.
+It measures how well the predicted probabilities match the observations.
+On the other hand, sharpness refers to the concentration of the forecast distributions --- a sharp forecast distribution results in narrow prediction intervals, indicating high confidence in the forecast.
+A model is well-calibrated if the predicted probabilities match the distribution of the observations, and it is sharp if it is confident in its predictions.
+The CRPS rewards sharpness and calibration by assigning lower scores to forecasts with sharper distributions, and to forecasts that are well-calibrated.
+Thus, it is a metric that combines both sharpness and miscalibration into a single score, making it a useful tool for evaluating the performance of probabilistic forecasts.
 
-The primary objective of this study is to forecast pharmaceutical demand within EPSS for the year 2023. To achieve this, the historical sales data for the chosen pharmaceuticals
+CRPS can be considered an average of all possible Winkler scores [@winkler1972decision;@hyndman2021forecasting, Section 5.9] or percentile scores [@hyndman2021forecasting, Section 5.9], and thus provides an evaluation of all possible prediction intervals or quantiles.
+A specific prediction interval could be evaluated using a Winkler score.
+Certain situations may also require assessing accuracy for a particular quantile, such as lower (e.g 5\%) or higher (e.g.
+95\%) quantiles.
+In such cases, a percentile score becomes useful in meeting this specific requirement.

### 2024-10-24 manuscript/04-results.qmd
my first draft

@@
 data_visualisation <- read.csv("../rscript/data_visualisation.R")
+```
+
+# Results and discussion {#sec-results}
+
+```{r}
+#| include: false
+rmsse <- readr::read_rds(here::here("results/rmsse.rds")) |>
+  mutate(msse = rmsse^2)
+mase <- readr::read_rds(here::here("results/mase.rds"))
+crps <- readr::read_rds(here::here("results/crps.rds")) |>
+  pivot_wider(names_from = model, values_from = crps) |>
+  group_by(method, h, series) |>
+  mutate(across(where(is.numeric), ~ .x / naiveecdf)) |>
+  ungroup()
+```
+
+In this section, we compare the forecasting performance of the Stationary, ETS, GLM, and TSGLM models along with the ensemble, using base forecast and Minimum Trace (MinT) reconciliation methods.
+We have also computed the forecast accuracy for Ordinary Least Square (OLS) and Weighted Least Square (WLS) approaches, along with bottom-up forecasting.
+However, they are not reported here because their accuracy is outperformed by MinT.
+We should also note that forecasts, and consequently their corresponding errors, are generated for the entire hierarchy and they could be reported at any level, if required.
+But to save space, we have reported only the top level (Total), the bottom level, and the levels corresponding to Control areas and Health boards.
+The latters are chosen because this is where decision-making takes place, so these forecasts are the most important.
+
+The overall forecasting performance is reported in @tbl-result, in which the average forecast accuracy over horizons 43--84 days (corresponding to the planning horizon) is presented per model, method, and the hierarchical level.
+Reported forecast accuracy is averaged across all forecast horizons, rolling origins, and series at each level.
+@tbl-result presents both point and probabilistic forecast accuracy at total, control area, health board and bottom-level series.
+Point forecast performance is reported using MASE and MSSE, while probabilistic forecast accuracy is reported using CRPS.
+The bold entries in each table identify a combination of method and model that performs best for the corresponding level (i.e.
+each column), based on the smallest values of accuracy measures.
+
+@tbl-result shows that forecast reconciliation (i.e.
+MinT) improves point forecast accuracy at the higher levels of the hierarchy including total, control area and health board.
+However, it does not result in accuracy improvement at the bottom-level series, for which base forecasts are more accurate.
+This might be due to the noisy structure of time series at the bottom level, and possibly due to very different patterns in the aggregated series.
+It is also clear from @tbl-result that the ensemble method improves forecast accuracy at total, control area and health board.
+However, this does not remain valid for bottom series where different individual methods perform best, depending on the accuracy measure.
+While the forecast reconciliation approach aims to enhance forecast accuracy, its effectiveness is not guaranteed, especially if the bottom-level series exhibit excessive noise and lack systematic patterns.
+Despite this, reconciling forecasts at the bottom level can offer advantages by generating coherent forecasts that facilitate alignment in planning across various teams within an organization, promote better coordination, and prevent conflicting decisions.
+Moreover, even when dealing with noisy and irregular bottom-level series, reconciliation can still improve forecast accuracy at higher levels of the hierarchy by leveraging the information available across the hierarchy.
+Therefore, although the bottom-level forecasts may not be highly accurate on their own, reconciling them with higher-level forecasts can still provide a more consistent view of future demand and potentially yield more accurate forecasts at other levels.
+
+```{r}
+#| label: tbl-result
+#| tbl-cap: "Average forecast performance calculated on the test sets at forecast horizons $h=43,\\dots,84$ days, with time series cross validation applied to attended incident data.
+The test set consists of 462 days.
+The best approach is highlighted in bold.
+Point forecast accuracy is measured using MASE and MSSE, while probabilistic forecast accuracy is measured using CRPS."
+# Set minimum in column to bold
+set_to_bold <- function(table) {
+  for (i in 3:6) {
+    best <- which.min(table[[i]])
+    table[[i]] <- sprintf(table[[i]], fmt = "%1.3f")
+    table[[i]][best] <- paste0("\\textbf{", table[[i]][best], "}")
+  }
+  return(table)
+}
+
+msse1 <- rmsse |>
+  filter(
+    method %in% c("mint", "base"),
+    model != "qcomb",
+    h > 42,
+    model != "ensemble2"
+  ) |>
+  mutate(
+    model = factor(model,
+      levels = c("naiveecdf", "ets", "iglm", "tscount", "ensemble", "ensemble2"),
+      labels = c("Stationary", "ETS", "GLM", "TSGLM", "Ensemble", "Ensemble2")
+    ),
+    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT")),
+  ) |>
+  group_by(method, model, series) |>
+  summarise(msse = mean(msse), .groups = "drop") |>
+  pivot_wider(names_from = series, values_from = msse) |>
+  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
+  set_to_bold()
+mase1 <- mase |>
+  filter(
+    method %in% c("mint", "base"),
+    model != "qcomb",
+    h > 42,
+    model != "ensemble2"
+  ) |>
+  mutate(
+    model = factor(model,
+      levels = c("naiveecdf", "ets", "iglm", "tscount", "ensemble", "ensemble2"),
+      labels = c("Stationary", "ETS", "GLM", "TSGLM", "Ensemble", "Ensemble2")
+    ),
+    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT"))
+  ) |>
+  group_by(method, model, series) |>
+  summarise(mase = mean(mase), .groups = "drop") |>
+  pivot_wider(names_from = series, values_from = mase) |>
+  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
+  set_to_bold()
+crps1 <- readr::read_rds(here::here("results/crps.rds")) |>
+  filter(
+    method %in% c("mint", "base"),
+    model != "qcomb",
+    h > 42,
+    model != "ensemble2"
+  ) |>
+  mutate(
+    model = factor(model,
+      levels = c("naiveecdf", "ets", "iglm", "tscount", "ensemble", "ensemble2"),
+      labels = c("Stationary", "ETS", "GLM", "TSGLM", "Ensemble", "Ensemble2")
+    ),
+    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT"))
+  ) |>
+  group_by(method, model, series) |>
+  summarise(crps = mean(crps), .groups = "drop") |>
+  pivot_wider(names_from = series, values_from = crps) |>
+  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
+  set_to_bold()
+bind_rows(mase1, msse1, crps1) |>
+  kable(format = "latex", booktabs = TRUE, escape = FALSE, align = "llrrrr") |>
+  pack_rows("MSSE", 1, 10) |>
+  pack_rows("MASE", 11, 20, hline_before = TRUE) |>
+  pack_rows("CRPS", 21, 30, hline_before = TRUE)
+```
+
+@tbl-result presents the accuracy of the forecast distributions measures by CRPS, which considers both forecasting reliability and interval sharpness.
+The smaller the value of CRPS, the better the comprehensive performance.
+We observe that forecast reconciliation results in forecast improvement for the total and health board level.
+CRPS is almost identical at the control area and botom levels.
+Base forecasts are slightly better at the control area level, while reconciliation is marginally accurate than base at the bottom level.
+The ensemble method is also more accurate for higher levels, but ETS performs well at the bottom level.
+@tbl-result also indicates that reconciliation using Mint  generates accurate *distributional* forecasts.
+The marginal improvement in the average probabilistic forecast accuracy at the bottom level might be due to the reconciliation method giving improved forecast accuracy in the tails of the forecast distribution, which are critical for managing risks.
+
+Overall, our results indicate that forecast reconciliation using the MinT method provides reliable forecasts and improves upon the base (unreconciled) forecasts at all levels except the bottom-level series.
+But even there, forecast reconciliation using MinT improves accuracy in the tails of the distribution.
+
+```{r}
+#| label: fig-accuracy
+#| fig-width: 8
+#| fig-height: 5
+#| out.width: "100%"
+#| fig-cap: "Average accuracy by week for 12 weeks using MinT reconciliation.
+The total number of days used to calculate the accuracy in the test set is 462.
+Forecasts are generateled every 42 days, therefore we use 11 samples to calculate the average accuracy.
+CRPS is relative to a stationary Empirical Cumulative Distribution Function (ECDF).
+MASE and MSSE are relative to the corresponding values for the training set."
+#| fig-pos: "H"
+crps <- crps |>
+  pivot_longer(ensemble:tscount, names_to = "model", values_to = "crps")
+
+accuracy <- bind_rows(
+  rmsse |> mutate(measure = "msse", accuracy = rmsse^2),
+  mase |> mutate(measure = "mase", accuracy = mase),
+  crps |> mutate(measure = "crps", accuracy = crps),
+) |>
+  select(-rmsse, -mase, -crps)
+
+# Plot of average accuracy vs week for each method for Total
+acc_summary <- accuracy |>
+  filter(
+    series %in% c("Total", "Control areas", "Health boards"),
+    method == "mint", model != "qcomb"
+  ) |>
+  mutate(model = case_when(
+    model == "naiveecdf" ~ "Stationary",
+    model == "ets" ~ "ETS",
+    model == "tscount" ~ "TSGLM",
+    model == "iglm" ~ "GLM",
+    model == "ensemble" ~ "Ensemble",
+    model == "ensemble2" ~ "Ensemble2"
+  )) |>
+  mutate(
+    measure = factor(measure, levels = c("mase", "msse", "crps"), labels = c("MASE", "MSSE", "CRPS")),
+    series = factor(series, levels = c("Total", "Control areas", "Health boards")),
+    model = factor(model, levels = c("Stationary", "ETS", "TSGLM", "GLM", "Ensemble", "Ensemble2")),
+    week = factor(trunc((h - 1) / 7) + 1)
+  ) |>
+  group_by(week, model, measure, series) |>
+  summarise(accuracy = mean(accuracy), .groups = "drop")
+
+acc_summary |>
+  filter(
+    model != "Ensemble2"
+  ) |>
+  ggplot(aes(x = week, y = accuracy, group = model, col = model)) +
+  geom_line() +
+  geom_point(size = .5) +
+  facet_grid(measure ~ series, scales = "free_y") +
+  labs(y = "Average accuracy", x = "Week ahead") +
+  ggthemes::scale_color_colorblind() +
+  ggthemes::theme_few()
+```
+
+In addition to the overall forecast accuracy presented in @tbl-result, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-accuracy.
+The figure focuses on the hierarchical levels important for decision-making including total, control area, and health board; however, the accuracy could be calculated for any level.
+We only illustrate the results of the MinT method, given its strong performance described in @tbl-result.
+For illustration purposes, we report the average weekly forecast accuracy instead of the daily forecast horizon, as this reduces the visual noise in the figure.
+Thus, the x-axis shows horizons from week 1 ($h= 1,\dots,7$) to week 12 ($h= 78,\dots,84$).
+The forecast horizon from week 7 to week 12 corresponds to the upcoming planning horizon, which is used by planners and decision-makers.
+For both the point forecast and distributional accuracy  we can see that the ensemble approach performs best across almost all horizons, with the biggest differences at the highest levels of aggregation.
+It is important to highlight that, all forecasting models outperform the stationary empirical distribution that is used as a benchmark for both point and probabilistic forecasts.
+
+Despite using Poisson regression models to create count distributions of attended incidents for the base forecasts, it is important to note that the reconciled forecast distributions do not maintain a count format.
+In practical scenarios, there might be a need to use integer forecasts.
+Count forecast reconciliation is an active area of research, and it would be interesting to explore how our approach could be adapted to generate count-reconciled probabilistic forecasts in future studies.
+Rounding the forecasts is one possible solution to this problem.
+However, the impact of rounding on forecast accuracy varies depending on the level of hierarchy and the scale of the data.
+In situations with high volume demand, the effects of rounding may be negligible, and forecast accuracy calculations can overlook integer effects.
+On the other hand, in low-volume demand settings, such as forecasts at the bottom level of the hierarchy, integer (rounding) effects may have a more noticeable influence on forecast accuracy.
+
+## An illustration of probabilistic forecast for EMS demand
+
+In this section, we provide an illustrative example of a probabilistic forecast for future demand, of the total attended incidents in the SB health board.
+Due to the complexity of including such plots in the manuscript for the entire hierarchy and 84 days ahead, only one example is presented here.
+However, it is feasible to generate these plots for the entire hierarchy and for any forecast horizon if necessary.}
+
+In practice, point forecasts are commonly used, but they have limitations as they ignore the uncertainty associated with the forecast.
+The future is inherently characterized by an irreducible level of uncertainty.
+Being prepared entails considering alternate courses of action.
+Probabilistic forecasts offer an alternative approach to anticipate future demand.
+Rather than providing a single value, they assign likelihoods to all possible demand outcomes, acknowledging that different numbers of attended incidents are possible, but with varying likelihoods.
+
+The purpose of probabilistic forecasting, as demonstrated in @fig-forecast-density-hstep and @fig-forecast-density, is to quantify uncertainty.
+@fig-forecast-density-hstep depicts the forecast distribution of total incidents in one health board over a 7-day period.
+It also gives the point forecast as well as the 80\% and 90\% prediction intervals.
+@fig-forecast-density zooms in on the first day to show the histogram more clearly, illustrating the range of possible outcomes and their likelihood.
+
+Decisions based on these forecasts could focus on the tails of the distribution: unexpectedly high demand leading to crowding and inefficiency, or unexpectedly low demand resulting in wasted resources.
+Such forecasts are valuable tools for decision-makers and planners, especially when dealing with low-probability, high-cost situations.
+Different EMS managements may have varying risk attitudes depending on resource availability, making it crucial to consider the entire distribution when making decisions.
+For instance, these forecasts enable management to calculate the probability of demand exceeding a certain threshold of available resources (e.g., 90\%), which can serve as an informative early warning measure for overcrowding.
+
+It is important to note that while point forecasts and prediction intervals can be obtained from the probabilistic forecasts, the reverse is not possible.
+A single number cannot be used to directly derive a probabilistic forecast.
+Prediction intervals, although helpful in indicating possible ranges, do not provide information on the probabilities of low or high demand.
+
+In EMS planning, future demand is just one aspect to consider.
+Other inputs, such as capacity, should also be treated as probability distributions to adopt a probabilistic approach to planning.
+To extract valuable insights and make informed decisions from probabilistic forecasts, specialized numerical tools are required, as the forecasts themselves are typically represented as explicit probability density functions or Monte Carlo generators.
+
+```{r}
+#| label: fig-forecast-density-hstep
+#| out.width: "80%"
+#| fig-cap: "A graphical illustration of the forecast distribution of ambulance demand (i.e.
+total incidence attended) for the SB health board for a horizon of seven days.
+For each day, we display the point forecast (black point), the histogram, and 80% (thick line) and 90% (thin line) prediction intervals.
+It also shows a portion of a historical time series as well as its fitted values."
+#| fig-pos: "H"
+incident_gthf <- readr::read_rds(here::here("data/incidents_gt.rds"))
+
+incident_cv <- incident_gthf |> filter(lhb == "SB", is_aggregated(category), is_aggregated(nature))
+fit <- incident_cv %>%
+  model(
+    ets = ETS(sqrt(incident))
+  )
+set.seed(2023)
+fcst <- fit %>% generate(h = 7, times = 5000, bootstrap = TRUE)
+
+fitted_ets <- fit |> augment()
+fcst_point <- fcst |> index_by(date) |> summarise(.mean=mean(.sim))
+p1 <- ggplot(data = fcst, mapping = aes(x = date, y = .sim))+
+  stat_histinterval(point_interval = "mean_qi",.width = c(0.80, 0.95),breaks=15)+
+  geom_line(data=fcst_point,aes(y=.mean, shape ="Point Forecast"))+
+  geom_line(aes(y = .fitted, shape ="Fitted"), data = filter_index(fitted_ets, "2019-07-17" ~ .))+
+  geom_point(aes(y = .fitted, shape ="Fitted"), data = filter_index(fitted_ets, "2019-07-17" ~ .))+
+  geom_line(aes(y = incident, shape ="Actual"),data = filter_index(incident_cv, "2019-07-17" ~ .))+
+  geom_point(aes(y = incident, shape ="Actual"),data = filter_index(incident_cv, "2019-07-17" ~ .))+
+  scale_shape_manual(name=NULL,
+                   breaks=c('Actual','Fitted','Point Forecast'),
+                   values=c('Actual'=17,'Fitted'=15, 'Point Forecast'=19))+
+  labs(x= "Date", y="Incidents")+
+  theme_few()+
+  theme(legend.position = "top")
+p1
+```
+
+```{r}
+#| label: fig-forecast-density
+#| out.width: 80%
+#| fig-cap: "An illustrative example of the forecast distribution of ambulance demand (i.e.
+total incidence attended) for the SB health board for one day ahead.
+This corresponds to the first forecast distribution in Figure 5.
+The horizontal axis shows all possible outcomes that may occur, with their likelihood shown on the vertical axis.
+The point in the middle shows the point forecast.
+Two lines at the bottom of the distribution highlights 80% (thick line) and 90% (thin line) prediction intervals."
+#| fig-pos: "H"
+
+fcst1 <- fcst %>% filter_index("2019-08-01")
+
+p2 <- ggplot(data = fcst1, mapping = aes(x = .sim))+
+  stat_histinterval(point_interval = "mean_qi",.width = c(0.80, 0.95),breaks=40)+
+  scale_x_continuous(breaks = seq(60, 140, 5))+
+  scale_y_continuous(labels = percent)+
+  labs(y = "Density", x = "Incidents")+
+  theme_few()
+p2
 ```

### 2024-11-02 manuscript/02-literature.qmd
first draft ready

@@
 
-Despite the advancements in forecasting methods, several gaps remain, particularly in addressing uncertainty and incorporating domain knowledge into the modeling process.
+@nguyen2023managing explore how sentiment analysis of news media can be utilized to improve demand forecasting for pharmaceutical products during disruptive events, such as pandemics and scandals.
+The authors employed a VARX (Vector Autoregressive with Exogenous Variables) model to forecast demand volatility.
+@papanagnou2018coping investigate the impact of big data analytics, specifically through text mining, on improving forecasting accuracy for pharmaceuticals in retail pharmacies, using HMX pharmacy stores in Nigeria as a case study.
+They use multivariate time series analysis technique known as VARX, which incorporates customer-generated content from sources like Google, YouTube, and online newspapers to predict demand for medicines.
+@fourkiotis2024applying focus on enhancing pharmaceutical sales forecasting through the application of both traditional statistical methods and advanced machine learning techniques.
+Various forecasting approaches, including traditional models like Naïve and ARIMA,Facebook Prophet, and machine learning methods such as XGBoost and LSTM neural networks are compared.
+The results demonstrated that machine learning models, particularly XGBoost, significantly outperformed traditional methods.
+@belghith2024new present a rolling forecasting framework employing moving average and three exponential smoothing methods (Brown’s, Holt’s, and Holt-Winters) implemented using Microsoft Power BI for enhanced data visualization.
+
+<!-- Despite the advancements in forecasting methods, several gaps remain, particularly in addressing uncertainty and incorporating domain knowledge into the modeling process.
 Most existing studies focus on point forecasts, which may not fully capture the range of possible outcomes in pharmaceutical demand.
@@
 However, there is still room for improvement, particularly in addressing uncertainty and incorporating domain knowledge.
-Future research should focus on developing probabilistic forecasting models and exploring ways to integrate expert insights into the forecasting process.
-These advancements could significantly enhance decision-making and operational efficiency in the pharmaceutical supply chain.
+-->
 
@@
 #| tbl-cap: "Summary of some studies on forecasting in pharmaceutical supply chain"
+#| tbl-pos: "H"
 library(kableExtra)
-readxl::read_xlsx("Literature_review_table.xlsx") |>
+readxl::read_xlsx("Literature_review.xlsx") |>
   knitr::kable(booktabs = T, linesep = "") |>
-  kable_styling(latex_options = "scale_down", font_size = 11) |>
-  column_spec(3, "6em") |>
-  column_spec(5, "4em") |>
-  column_spec(6, "10em") |>
-  column_spec(7, "8em") |>
-  landscape()
+  kable_styling(latex_options = c("scale_down"), font_size = 11) |>
+  column_spec(4, "10em") |>
+  column_spec(7, "10em") |>
+  column_spec(8, "10em") |>
+  column_spec(9, "2em") |>
+  column_spec(10, "2em") |>
+  column_spec(11, "2em") |>
+  kableExtra::landscape()
 ```

### 2024-11-02 manuscript/03-experiment.qmd
first draft ready

@@
 Rigorous checks were conducted to ensure the consistency and completeness of the collected data.
-From the extensive pool of pharmaceutical products within EPSS, we strategically selected a set of 33 key pharmaceuticals, encompassing various programs and representing different classes of drugs.
-This deliberate selection enriches the research by providing a comprehensive representation across the pharmaceutical landscape.
-
-We are presenting the graph for some of the pharmaceuticals sales for the last five years by using visualization for the readers to undesrtand the sales of the pharmaceuticals.
+From the extensive pool of pharmaceutical products within EPSS, we selected a set of 33 key pharmaceuticals, including various programs and representing different classes of drugs.
 
@@
 
-Given the number of produts in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understadn data.
+Given the number of products in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understadn data.
 @fig-feature shows the strength of trend versus seasonality.
@@
 #| cache: true
-#| out.width: "70%"
+#| out.width: "60%"
 #| fig.align: center
 #| fig-cap: "The strength of the trend and seasonality in the time series of pharmaceutical product consumption. The scatter plot shows 33 data points, with each point corresponding to a product."
-read_csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
-  janitor::clean_names()
+
+med_qty <- read_csv("../data/predicotrs_final_tssible_July_04_2024.csv") |> 
+  janitor::clean_names() |> mutate(month = yearmonth(month))
+
 med_qty_tsb <- med_qty |> 
-  mutate(month = yearmonth(month)) |> 
   as_tsibble(index = month, key = item)
@@
   features(quantity, feat_stl) %>%
-  ggplot(aes(x = trend_strength, y = seasonal_strength_week)) +
+  ggplot(aes(x = trend_strength, y = seasonal_strength_year)) +
   geom_point(alpha = 0.25) +
-  labs(x = "Strength of trend", y = "Strength of weekly seasonality")
+  labs(x = "Strength of trend", y = "Strength of seasonality")+
+  ggthemes::theme_few()
 ```
 
-It is clear that there are some series showing strong trends and/or seasonality, corresponding to series at the higher levels of the hierarchy.
-The majority of series show low trend and seasonality.
-These are time series belonging to the bottom series, series related to the nature of incidents for a given control, health board and priority level.
-Bottom series are dominated by noise with little or no systematic patterns.
-
-In addition to displaying the strength of trend and seasonality, we also visualize all time series, revealing various patterns such as trends, seasonality, and instances of consumption with chaotic behavior during certain months.
-For instance, some series exhibit both seasonality and trend, while others show a low volume of attended incidents and higher entropy, making them more volatile and challenging to forecast.
-At the incident level, combined with categories from other levels, many series contain zeros or low counts.
-This highlights the diversity of daily time series patterns in the data set.
+It is evident that some time series display strong trends and/or seasonality, while the majority exhibit low trends and minimal seasonality.
+A number of products show pronounced trends, and only a few demonstrate clear seasonal patterns.
+Beyond assessing the strength of trends and seasonality, we also visualized all time series to understand data and various patterns, including trends and instances of erratic consumption behavior during certain months.
+For example, some series show low consumption volumes over consecutive months, followed by peak consumption in specific months, making them more volatile and difficult to forecast.
+This underscores the diversity of monthly pharmaceutical time series patterns within the dataset and highlights the importance of understanding the factors driving these consumption behaviors.
 Figure @fig-dataviz2 illustrates the time plots for a few selected products.
@@
 #| dependson: "fig-feature"
-#| out.width: "100%"
-#| fig-width: 8
-#| fig-height: 10
+#| fig-width: 11
+#| fig-height: 6
 #| fig-cap: "Monthly time plot of consumption.
@@
 
-(my_plot[[1]] | my_plot[[1]]) /
-(my_plot[[1]] | my_plot[[1]])
+p1 <- ggplot(med_qty |> filter(item == "Oxytocin") |> select(-item), 
+       aes(x=month, y=quantity))+
+  geom_point()+
+  geom_line(aes(group =1))+
+  scale_x_yearmonth(date_breaks = "4 month")+
+  labs(x = "Month", y = "Consumption", title = "Oxytocin")+
+  ggthemes::theme_few()+
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
 
+p2 <- ggplot(med_qty |> filter(item == "Amoxicillin - 500mg - Capsule") |> select(-item), 
+       aes(x=month, y=quantity))+
+  geom_point()+
+  geom_line(aes(group =1))+
+  scale_x_yearmonth(date_breaks = "4 month")+
+    labs(x = "Month", y = "Consumption",title = "Amoxicillin - 500mg - Capsule")+
+  ggthemes::theme_few()+
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+
+p3 <- ggplot(med_qty |> filter(item == "Ceftriaxone") |> select(-item), 
+       aes(x=month, y=quantity))+
+  geom_point()+
+  geom_line(aes(group =1))+
+  scale_x_yearmonth(date_breaks = "4 month")+
+    labs(x = "Month", y = "Consumption",title = "Ceftriaxone")+
+  ggthemes::theme_few()+
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+
+p4 <- ggplot(med_qty |> filter(item == "Atenolol - 50mg - Tablet") |> select(-item), 
+       aes(x=month, y=quantity))+
+  geom_point()+
+  geom_line(aes(group =1))+
+  scale_x_yearmonth(date_breaks = "4 month")+
+    labs(x = "Month", y = "Consumption", title = "Atenolol - 50mg - Tablet")+
+  ggthemes::theme_few()+
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+
+patch <- (p1 | p2) /
+(p3 | p4) 
+patch+ggthemes::theme_few()
 ```
@@
 
-### Domain knowledge colelction though interview with EPSS experts
+### Collaborative expert review to identify factors affecting consumption
 
-To gather domain knowledge, we contacted experts from the Ethiopian Pharmaceuticals Supply Service (EPSS) who work in the distribution unit.
-Five experts with over 10 years of experience in managing the distribution of the pharmaceuticals in EPSS agreed to participate.
-The researchers, along with these experts, examined the electronic bin cards of selected pharmaceuticals.
+To understand the external factors influencing product consumption, we began by visualizing the time series data for consumption as highlighted in the previous section.
+This initial step helped identify specific instances within the data that showed unusual patterns, such as peaks, unusually high or low observations, and consecutive periods of low consumption.
+These observations indicated potential events or factors affecting consumption levels.
+
+To gather domain knowledge about factors affected the consumption of products, we contacted experts from the Ethiopian Pharmaceuticals Supply Service (EPSS) who work in the distribution unit.
+Next, we collaborated with six experts from the Ethiopian Pharmaceuticals Supply Service (EPSS), each with over 10 years of experience managing pharmaceutical distribution.
+In these sessions, we reviewed the time series of all products together to understand the potential drivers behind these anomalies.
+The experts provided insights based on their extensive experience and knowledge of distribution practices.
+When there was uncertainty or gaps in the data for certain products, additional experts were consulted to gather further context.
+If all experts reached a consensus, the influencing factors were confirmed and included in the analysis.
+
+As part of this process, we examined the electronic bin cards for 33 pharmaceutical products and conducted detailed data exploration.
 A bin card is a physical or electronic record used to track the inventory levels of individual items in a storage location, such as a warehouse or stockroom.
@@
 It provides a straightforward and immediate way to monitor item quantities, helping to ensure effective inventory management.
-
-We meticulously traced sales data for the past five years, triangulating graphs with bin card records to identify predictors influencing demand for each product.
-The expert uses their domain knowledge and experience to identify about the distribution charachterstics of the specific products and sometimes when they assume that they do not have enough history or do not remember about that specific products they call to other experts who remember specific issues and decide whetrere the specific predictos affect the distribution and if all of them agree the include it as predcitors.
-The data collection process took more than a month.
-We downloaded and analyzed over 100 bin cards per product.
-Dates on the bin cards were converted from the Ethiopian calendar to the Gregorian calendar for accuracy.
-Transactions due to storage or warehouse changes within EPSS were excluded.
-After cleaning the data,and discussing with the experts  we identified the following key predictors affecting distribution and sales: stock replenishment, stock outs, COVID-19, fiscal year counting, conflict, and malaria seasonality.
+Over a period of two weeks, we meticulously traced consumption data through these bin cards.
+Moreover, We ensured accuracy by converting dates from the Ethiopian calendar to the Gregorian calendar and excluded transactions related to internal storage or warehouse relocation within EPSS.
+Through this collaborative approach, we were able to identify several predictors affecting distribution and consumption that are summarised as followings:
 
 - Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
-Whenever there was stock replenishment at the central EPSS, sales and distribution to hubs and health facilities increased.
+Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased.
 This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.
 
-- Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand.
+<!-- - Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand.
 This can happen when demand exceeds the available supply, or when there is a delay or failure in the replenishment process.
-Resulted in no transactions and a decline in sales.
+Resulted in no transactions and a decline in consumption.
+-->
 
-- COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain.
+<!-- - COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain.
 The pandemic disrupted global manufacturing and supply, leading to shortages of essential medicines and raw materials, Ethiopia, like many other countries, relies heavily on imported pharmaceutical products.
 With the pandemic causing lockdowns and restrictions in manufacturing hubs like India and China, the supply of essential medicines and raw materials to Ethiopia was significantly disrupted.
-The impact of COVID-19 on sales was significant.
+The impact of COVID-19 on consumption was significant.
 After the announcement of COVID-19 in Ethiopia, the central EPSS pushed a six-month supply of products to hubs to prevent stock outs.
 This push was from March to August 2020.
-During this period, if products were not stocked out, sales continued to hubs and health facilities.
+During this period, if products were not stocked out, consumption continued to hubs and health facilities.
+-->
 
-- Physical inventory: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
+- Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
 The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
-Physical inventory counting periods also influenced sales.
+Physical inventory counting periods also influenced consumption.
 Stores closed during these periods, halting transactions.
-We observed increased sales before inventory counting periods, as hubs and facilities stocked up.
+We observed increased consumption before inventory counting periods, as hubs and facilities stocked up.
 July and August were identified as physical counting periods each year.
@@
 This disruption can cause significant delays or prevent the delivery of essential medicines altogether.
-Internal displacements and conflicts in Ethiopia over the past five years also affected sales and distribution.
+Internal displacements and conflicts in Ethiopia over the past five years also affected consumption and distribution.
 We traced periods of instability and included them as predictors.
@@
 Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks.
-We identified epidemic periods affecting sales: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.
+We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.
+
+The key predictors identified are deterministic values, since both past and future values of these predictors are known in advance.
+We then integrated them into the consumption data, creating the complete dataset required for model building and running the experiment.
 
@@
 
-**Multiple Linear Regression (MLR)**:
+We evaluate a range of univariate models and their counterparts that include predictors, spanning from simpler methods like regression, exponential smoothing, and ARIMA to more complex models such as long short-term memory (LSTM) networks and advanced foundational time series forecasting models.
+Below, we provide a brief overview of these approaches.
+Detailed implementation codes in R and Python are available in a GitHub repository and accessible for public.
 
-**ARIMA and ARIMAX (LSTM)**:
+**Exponential Smoothing State Space model (ETS):** ETS models, as described by @hyndman2021forecasting, combine trend, seasonality, and error components in time series using different configurations that can be additive, multiplicative, or mixed.
+The trend component can be specified as none ("N"), additive ("A"), or damped additive ("Ad"); the seasonality can be none ("N"), additive ("A"), or multiplicative ("M"); and the error term can be additive ("A") or multiplicative ("M").
+To forecast consumption, we use the ETS() function from the fable package in R, which automatically selects the optimal model for each time series based on the corrected Akaike’s Information Criterion (AICc).
+In our study, an automated algorithm determines the best configuration for trend, seasonality, and error components across each time series, leveraging the ets() function’s use of AIC to identify the optimal model.
+Given the high volume of time series (1530), manual selection of components is impractical, so the algorithm customizes model forms for each series based on its unique characteristics.
+This results in a tailored combination of additive or multiplicative components, adapting to the specific patterns of each time series.
 
-**Long Short-Term Memory neural network (LSTM)**: The LSTM model is a specialised form of recurrent neural network (RNN) used to model sequential data by capturing long-term dependencies \[\@<https://doi.org/10.1162/neco.1997.9.8.1735>\].
+**Multiple Linear Regression (MLR)**: We use Multiple linear regression to model the relationship between a consumption and potential variables influencing its variation.
+In our first model, we use multiple linear regression with a trend component to capture the underlying progression over time, i.e., _regression_.
+We also incorporate dummy variables for each month to account for seasonal fluctuations, without including any additional predictors, i.e., _regression_reg_.
+This approach helps us establish a baseline model focused solely on temporal trends and seasonal patterns.
+We then extend this model by introducing additional predictors that include variables such as replenishment cycles, fiscal year indicators, and periods with malaria prevalence.
+These additional predictors allow the model to see if capturing external factors can provide a better understanding of the factors influencing the consumption and result at enhanced accuracy.
+We produce forecasts using TLSM() function from the fable package in R.
+
+**ARIMA and ARIMA with regressors**: ARIMA (AutoRegressive Integrated Moving Average) is a powerful statistical model designed to forecast time series by capturing temporal dynamics and are widely used in time series forecasting due to their ability to model complex trends and patterns over time without relying on external predictors.
+ARIMAX (AutoRegressive Integrated Moving Average with eXogenous variables) extends ARIMA by incorporating external variables, or exogenous predictors, into the model.
+This modification allows to include relevant information from external factors such as malaria season, and fiscal year period that may explain variations in the series beyond its internal time dynamics, we refer to this in the result as _arima_reg_.
+By adding these predictors, ARIMAX combines the strengths of ARIMA’s time-series structure with the flexibility of regression models.
+In our study, we use an automated algorithm to determine the optimal configuration for ARIMA components, following the approach outlined by @hyndman2021forecasting and We use the ARIMA() function from the fable package in R.
+
+**Long Short-Term Memory neural network (LSTM)**: The LSTM model is a specialised form of recurrent neural network (RNN) used to model sequential data by capturing long-term dependencies [@graves2012long].
 Unlike traditional RNNs, LSTMs can learn to retain information for longer time periods due to their unique architecture, which consists of several gates that control the flow of information.
@@
 The Adam optimizer was employed to minimize the mean squared error, and the model was trained for 100 epochs using the keras_model_sequential() function from the Keras package in R.
+We use LSTM models both with and without predictors, referring to the LSTM model with predictors as _lstm_reg_.
 
-**TimeGPT**: TimeGPT is the first pre-trained foundational model designed specifically for time series forecasting, developed by Nixtla [@[https://doi.org/10.48550/arXiv.2310.0358910.03589](https://doi.org/10.48550/arXiv.2310.03589)].
-It utilizes a transformer-based architecture with an encoder-decoder configuration but differs from other models in that it is not based on large language models.
+**TimeGPT**: TimeGPT is the first pre-trained foundational model designed specifically for time series forecasting, developed by Nixtla [@garza2023timegpt].
+It uses a transformer-based architecture with an encoder-decoder configuration but differs from other models in that it is not based on large language models.
 Instead, it is built from the ground up to handle time series data.
@@
 Furthermore, TimeGPT supports the inclusion of external regressors in its forecasts and can generate quantile forecasts, providing reliable uncertainty estimation.
+We use TimeGpt models both with and without predictors, referring to the model with predictors as _timegpt_reg_.
 
@@
 
-To evaluate the performance of the various forecasting approaches, we split the data into a series of ten training and test sets.
-We use a time series cross-validation approach [@hyndman2021forecasting], with a forecast horizon of 84 days, and each training set expanding in 42-day steps.
-The first training set uses all data up to 2018-04-25, and the first test set uses the 84 days beginning 2018-04-26.
-The second training set uses all data up to 2018-06-06, with the second test set using the following 84 days.
-The largest training set ends on 2019-05-09, with the test set ending on 2019-07-31.
-Model development and hyper-parameter tuning is performed using the training data and the errors are assess using the corresponding test set.
-While we compute forecast errors for the entire 12 weeks, we are most interested in the last 42 days of each test set, because that corresponds to how forecasts are generated for planning in practice.
-Forecasting performance is evaluated using both point and probabilistic error measures.
+To assess the performance of our forecasting methods, we split the data into a series of training and test sets and apply time series cross-validation with a forecast horizon of 6 months.
+Each training set is expanded in monthly increments, allowing model development and hyperparameter tuning to be performed on the training data, with errors evaluated on the corresponding test sets.
+We assess forecasting performance using both point and probabilistic error measures.
 
-The error metrics provided below consider a forecasting horizon denoted by $j$, representing the number of time periods ahead we are predicting.
-In our study, this forecasting horizon ranges from 1 to 84 days, $j= 1,2,\dots, 84$.
+The error metrics presented here consider a forecasting horizon denoted by  by $j$, which represents the number of time periods ahead we are predicting, with $j$ ranging from 1 to 6 months in our study.
+Point forecast accuracy is measured using the Mean Squared Scaled Error (MSSE) and the Mean Absolute Scaled Error (MASE).
+The Mean Absolute Scaled Error (MASE) [@HK06; @hyndman2021forecasting] is calculated as follows:
 
-Point forecast accuracy is measured via the Mean Squared Scaled Error (MSSE) and the Mean Absolute Scaled Error (MASE).
-The Mean Absolute Scaled Error (MASE) [@HK06; @hyndman2021forecasting] is calculated as:
 $$
@@
 $$
-and $e_{j}$ is the point forecast error for forecast horizon $j$, $m = 7$ (as we have daily seasonal series), $y_t$ is the observation for period $t$, and $T$ is the sample size (the number of observations used for training the forecasting model).
+and $e_{j}$ is the point forecast error for forecast horizon $j$, $m = 12$ (as we have monthly seasonal series), $y_t$ is the observation for period $t$, and $T$ is the sample size (the number of observations used for training the forecasting model).
 The denominator is the mean absolute error of the seasonal naive method in the fitting sample of $T$ observations and is used to scale the error.
@@
 Note that the measure is scale-independent, thus allowing us to average the results across series.
+
+Here, $e_{j}$ represents the point forecast error for forecast horizon $j$, with $m = 12$ (since we are dealing with monthly seasonal data).
+The term $y_t$ denotes the observation at time $t$, and $T$ is the sample size, or the number of observations used for training the forecasting model.
+The denominator in the MASE formula is the mean absolute error of the seasonal naive method over the training sample of $T$ observations, providing a basis for scaling the forecast error.
+Lower MASE values indicate more accurate forecasts.
+Notably, this measure is scale-independent, allowing us to average results across different series for broader performance comparison.
 
@@
 Again, this is scale-independent, and smaller MSSE values suggest more accurate forecasts.
-
-Using scale-independent measures, such as MASE and MSSE, enables more appropriate comparisons between time series at different levels and scales, as these measures are not influenced by the magnitude of the data.
-This is of particular importance in our study, as we work with time series at various levels of hierarchy, with varying scales, resulting in different magnitudes of error.
-By employing scale-independent measures, we can meaningfully assess the forecast accuracy across the entire hierarchy, ensuring a more robust comparison.
 
@@
 
-CRPS can be considered an average of all possible Winkler scores [@winkler1972decision;@hyndman2021forecasting, Section 5.9] or percentile scores [@hyndman2021forecasting, Section 5.9], and thus provides an evaluation of all possible prediction intervals or quantiles.
-A specific prediction interval could be evaluated using a Winkler score.
-Certain situations may also require assessing accuracy for a particular quantile, such as lower (e.g 5\%) or higher (e.g.
-95\%) quantiles.
-In such cases, a percentile score becomes useful in meeting this specific requirement.
+CRPS can be considered an average of all possible qiantiles [@hyndman2021forecasting, Section 5.9], and thus provides an evaluation of all possible prediction intervals or quantiles.
+A specific prediction interval could be evaluated using a Winkler score, if required.
+
+<!-- If the  $100(1−\alpha)\%$ prediction interval at time  t is given by $[\ell_{\alpha,t}, u_{\alpha,t}]$, then the Winkler score is defined as the length of the interval plus a penalty if the observation is outside the interval -->
+
+<!-- $$ -->
+<!-- W_{\alpha,t} = \begin{cases} -->
+<!--   (u_{\alpha,t} - \ell_{\alpha,t}) + \frac{2}{\alpha} (\ell_{\alpha,t} - y_t) & \text{if } y_t < \ell_{\alpha,t} \\ -->
+<!--   (u_{\alpha,t} - \ell_{\alpha,t})   & \text{if }  \ell_{\alpha,t} \le y_t \le u_{\alpha,t} \\ -->
+<!--   (u_{\alpha,t} - \ell_{\alpha,t}) + \frac{2}{\alpha} (y_t - u_{\alpha,t}) & \text{if } y_t > u_{\alpha,t}. -->
+<!--   \end{cases} -->
+<!-- $$ -->
+
+<!-- In this study, we report skill scores for winkler aand CRPS, with skill scores, we compute a forecast accuracy measure relative to some benchmark method.
+For example, if we use the naïve method as a benchmark, and also compute forecasts using the ETS model, we can compute the CRPS and Winkler skill score of the ETS model relative to the naïve method.
+If the skill scores are negative, it indicates that the proposed method performs worse than the naïve method; positive scores suggest it performs better.
+In this sense, skill scores are not entirely independent of data scale, similar to the concepts behind MASE and RMSSE for point forecasts.
+-->
+
+<!-- $$ -->
+<!-- \frac{\text{CRPS}_{\text{Naïve}} - \text{CRPS}_{\text{ETS}}}{\text{CRPS}_{\text{Naïve}}}. -->
+<!-- $$ -->
+
+<!-- $$ -->
+<!-- \frac{\text{Winkler}_{\text{Naïve}} - \text{Winkler}_{\text{ETS}}}{\text{Winkler}_{\text{Naïve}}}. -->
+<!-- $$ -->

### 2024-11-02 manuscript/04-results.qmd
first draft ready

--- 2024-10-09 manuscript/04-results.qmd
@@
 library(readr)
-fc_accuracy_overall__without_predictor <- read_csv("../result/fc_accuracy_overall__without_predictor.csv")
-fc_accuracy_predictor_overall <- read_csv("../result/fc_accuracy_predictor_overall.csv")
-fc_accuracy_overall <- read_csv("../result/fc_accuracy_all_item.csv")
-data_visualisation <- read.csv("../rscript/data_visualisation.R")
-```
+point_accuracy_h <- read_rds("../result/accuracy_h.rds") |> janitor::clean_names() |> select(item, h,model,mase,rmsse)
+prob_accuracy_h <- read_rds("../result/accuracy_h.rds") |> janitor::clean_names() |> select(item, h,model,crps_skill,winkler_skill=skill,crps,winkler)
 
-from the total of 37 key pharmaceuticals selected and tested with the  different forecasting models arima showed that a better forecast accuracy for most of the pharmaceuticals followed by mean and  regresstion models with RMSSE value of 0.850, 0.942 and 0.945 respectively (Table 1).
+point_accuracy_id <- read_rds("../result/accuracy_id.rds")|> janitor::clean_names() |> select(item, id,model,mase,rmsse)
+prob_accuracy_id <- read_rds("../result/accuracy_id.rds")|> janitor::clean_names() |> select(item, id,model,crps_skill,winkler_skill=skill,crps,winkler)
 
-```{r}
-library(kableExtra)
-library(dplyr)  # Load the dplyr package
-fc_accuracy_cleaned <- na.omit(fc_accuracy_overall__without_predictor)
-fc_accuracy_table <- fc_accuracy_cleaned %>%
-  dplyr::select(.model, RMSE, MAE, MASE, RMSSE, winkler, CRPS)  # Explicitly specify dplyr::select
-kbl(fc_accuracy_table, booktabs = TRUE, escape = FALSE) %>%
-  kable_styling(latex_options = c("hold_position"), full_width = FALSE)
-```
+point_accuracy_horizon <- read_rds("../result/point_accuracy_horizon.rds") |> select(item, h=horizon,model,mase,rmsse)
+prob_accuracy_horizon <- read_rds("../result/prob_accuracy_horizon.rds") |>  janitor::clean_names() |> select(item, h=horizon,model,crps_skill=crps_skill_score,winkler_skill=winkler_skill_score,crps,winkler=winkler_score)
 
-Then we tested the models by adding different predictors which affects the sales of the pharmaceuticals like physical count, budget release and budget closure by the government, COVID 19, campaign in to the models as predictors which will affects the sales of the key pharmaceuticals products and the result showed that still arima have a better forecast accuracy with RMSSE of 0.850 (Table 2).
+point_accuracy_origin <- read_rds("../result/point_accuracy_origin.rds") |> janitor::clean_names() |> select(item, id=origin,model,mase,rmsse)
+prob_accuracy_origin <- read_rds("../result/prob_accuracy_origin.rds") |> janitor::clean_names() |> select(item, id=origin,model,crps_skill=crps_skill_score,winkler_skill=winkler_skill_score,crps,winkler=winkler_score)
 
-```{r}
-#| label: result3
-#| tbl-cap: "forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after considering predictors"
-fc_accuracy_predictor_overall %>% 
-  kbl(booktabs = TRUE, escape = FALSE) |>
-  kable_styling(latex_options = c("hold_position"),full_width = F)
-```
-```{r}
-#| label: result2
-#| fig-cap: "forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after considering predictors"
-library(ggplot2)  # Load the ggplot2 package
-library(forcats)  # Load the forcats package
-fc_accuracy_overall %>%
-  select(`.model`, item, RMSSE) %>% 
-  ggplot(mapping = aes(x = RMSSE, y = fct_reorder(`.model`, RMSSE, .fun = median))) +
-  geom_boxplot()
-```
+point_accuracy_h_all <- point_accuracy_h |> bind_rows(point_accuracy_horizon |> mutate(h= as.integer(h)))
+prob_accuracy_h_all <- prob_accuracy_h |> bind_rows(prob_accuracy_horizon |> mutate(h= as.integer(h)))
 
-Individual item with different forecasting models 
-We tested 10 models namely arima, arima with predictor, combination, exponential smoothing, mean, naive, regression, regression with predictor, snaive and stlf.
-Individually on the selected key pharmaceuticals and the result indicated that 11 (32.43%) of the selected pharmaceuticals have better forecast accuracy for arima predictor models, followed by naive 6 (16.21%), mean, regresstion, arima, and STLF methods each of them with 3 (8.10%) (Fig 2)
-
-```{r}
-# Sample data (replace with your actual data)
-pharmaceuticals <- c("ARIMA Predictor", "Naive", "Mean", "Regression", "ARIMA", "STLF")
-accuracy_percentages <- c(32.43, 16.21, 8.10, 8.10, 8.10, 8.10)
-
-# Create a data frame
-data <- data.frame(Pharmaceuticals = pharmaceuticals, Accuracy_Percentage = accuracy_percentages)
-# Create a bar chart
-library(ggplot2)
-
-ggplot(data, aes(x = reorder(Pharmaceuticals, -Accuracy_Percentage), y = Accuracy_Percentage)) +
-  geom_bar(stat = "identity", fill = "blue") +
-  labs(x = "Pharmaceutical Forecasting Models", y = "Accuracy Percentage") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+point_accuracy_origin_all <- point_accuracy_id |> bind_rows(point_accuracy_origin |> mutate(id= as.integer(id)))
+prob_accuracy_origin_all <- prob_accuracy_id |> bind_rows(prob_accuracy_origin |> mutate(id= as.integer(id)))
 
@@
 
-Those pharmaceuticals who had a better forecast accuracy with Arima predictors are Amoxacillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine, Atenolol, Atrovastatin 200mg, Ferrosu sulphate with follic acid, frusemide injection, Insuline Iosphane bipahsic, Lamivudine +Efaverienze + Tenofovir, Lidocaine + HCL and Propyl thiro uracil.The  six pharmaceuticasl which showed a better forecast accuracy with naive models were Ciprofloxacillin 500mg tablet, Magnesoum sulphate injection, Omeprazole capsule, Omeprazole injection, Rapid diagnostic test, Sulphamethaxone + Trimetoprine tablet.
+In this section, we compare the forecasting performance of various approaches, examining models that incorporate expert-identified business context predictors versus those that rely solely on historical consumption data.
+Point forecast performance is reported using MASE and MSSE, while probabilistic forecast accuracy is reported using CRPS.
 
-Discussion:
+The forecasting performance is reported in @fig-point, in which the average forecast accuracy over forecast horizon and across all products is calculated for each origin.
+we also report the ditribution of accurcy metrics  across all rollog origins, average across forecast horizon and all products.
+this shows how mdoels varies in roviding accuracy across different origins.
+further rserach would be required to understand how error analysis across forecast hrizon and across differne origins could be used as afiagnos!
+y-axis shows the models sorted based on ther values of mase and rmsse, when the one i the bottom has lowest error. this modesis TimeGpt, which is the foundational time sei models.
 
-The objective of this study was to evaluate the forecasting accuracy of different 
-models for key selected pharmaceutical products from the Ethiopian Pharmaceutical 
-Supply Services (EPSS). The analysis encompassed a total of 37 key pharmaceutical 
-products, and the results highlight the performance of various forecasting models in 
-predicting sales for these products. In the initial assessment of forecasting models, 
-the ARIMA model emerged as the most accurate for the majority of the pharmaceutical 
-products, exhibiting a Root Mean Squared Scaled Error (RMSSE) value of 0.850. This 
-finding suggests that ARIMA has a strong predictive capability and outperforms other 
-models, including the mean and regression models. The superiority of the ARIMA models 
-over the Moving Average might be explained by  most of the data set have seasonal 
-patern and MA  is one of the simplest prediction techniques for making projections 
-about time-series without a noticeable seasonal pattern [@chopra2001supply].The 
-superiority of the ARIMA model underscores its suitability for pharmaceutical sales 
-forecasting within the context of EPSS. Similar findings were reported in forecasting 
-ethanol demand in India where ARIMA model outperforms other models like linear and 
-non-linear regression models [@dey2023forecasting].
+@fig-point also indicate that predictors collected through interacting with experts with domaine knoeldge improves point forecast accuracy for all models, highlighting the essential needs to systematically colelct such information alognside the tarsactional data about consuption.
 
-To enhance the predictive power of the models, additional predictors were incorporated 
-into the analysis. Predictors such as physical count, budget release and closure by the government, the impact of COVID-19, and promotional campaigns were considered as 
-potential drivers of pharmaceutical sales. Despite the inclusion of these predictors, 
-the ARIMA model maintained its superior forecasting accuracy with an RMSSE of 0.850. 
-This resilience further validates the robustness of the ARIMA model and its ability to 
-capture complex interactions between variables influencing pharmaceutical sales.
+We observed that the LSTM model had the highest error compared to all other models.
+While LSTMs are powerful for capturing complex, non-linear temporal patterns, they may underperform in certain scenarios, especially when compared to simpler time series forecasting models.
+This underperformance can be attributed to overfitting, as LSTMs have a large number of parameters and require substantial amounts of data to generalize effectively.
+In contrast, simpler models like ARIMA with exogenous variables often handle straightforward trends or seasonality more effectively and may outperform LSTMs when long-term dependencies or non-linearity in the data are not significant.
 
-To delve deeper into the individual performance of the forecasting models, a 
-comprehensive evaluation was conducted on each of the 37 pharmaceutical products. Among the ten models tested, ARIMA with predictors exhibited the highest forecast accuracy 
-for 32.43% of the products. This was followed by the naive model, which performed well 
-for 16.21% of the products. Notably, a few pharmaceuticals demonstrated a strong fit 
-with specific models. For instance, Amoxicillin 500mg capsule, Anti RhD, Artehmether + 
-Lumefantrine, and others exhibited superior forecast accuracy when utilizing the ARIMA 
-model with predictors.
+```{r}
+#| label: fig-point
+#| fig-width: 7
+#| fig-height: 5
+#| fig-cap: "Distribution of point forecast accuracy across different origins, averaged across the forecat horizon and all products.
+The total number of months used to calculate the accuracy in the test set is 12 months for each product.
+MASE and MSSE are relative to the corresponding values for the training set."
+#| fig-pos: "H"
 
-Similarly, other pharmaceuticals displayed better forecasting outcomes with alternative models. Notably, Ciprofloxacin 
-500mg tablet, Magnesium sulfate injection, Omeprazole capsule, and others showed 
-improved accuracy with the naive model. These findings underscore the importance of 
-tailoring the choice of forecasting model based on the characteristics and 
-dynamics of individual pharmaceutical products.This finding confirmed that There is not
-a single best technique to solve 
-time-series forecasting problems [@zhang2007quarterly]. In order to deal with 
-time-series forecasting, each problem might be solved with a different approach [@ensafi2022time].
+p_mase <- point_accuracy_origin_all |> ggplot(aes(x= mase, y =  fct_reorder(model, mase)))+
+  geom_boxplot()+
+  labs(y = "Models")+
+  ggthemes::theme_few()
 
-In conclusion, this study highlights the pivotal role of accurate forecasting in the 
-pharmaceutical supply chain management. The ARIMA model, particularly when integrated 
-with relevant predictors, emerges as a powerful tool for 
-predicting pharmaceutical sales within the EPSS. However, the selection of the most 
-appropriate model should consider the specific attributes of each pharmaceutical 
-product. Further research could explore the optimization of model parameters and the 
-incorporation of additional contextual variables to enhance forecasting accuracy and 
-supply chain efficiency.
+p_rmsse <- point_accuracy_origin_all |> ggplot(aes(x= rmsse, y =  fct_reorder(model, mase)))+
+  geom_boxplot()+
+  labs(y = "Models")+
+  ggthemes::theme_few()
 
-Limitation
+point_patch <- p_mase/p_rmsse
 
-One notable limitation of this study revolves around the availability and quality of 
-the data used for analysis. While we employed a comprehensive dataset from the EPSS, it is essential to acknowledge that the accuracy of forecasting models is heavily 
-contingent on the quality, granularity, and completeness of historical sales data. 
-Variations in data recording practices, potential errors, missing values, or 
-inconsistencies within the dataset could impact the precision of the forecasts 
-generated by the employed models. Moreover, the inherent complexity of pharmaceutical 
-demand, influenced by a myriad of external factors such as socioeconomic changes,stock 
-outs, prolonged procurement lead time, rationing of avaiable products, pushing of 
-pharmaceuticals to different health facility without demand, healthcare policies, and 
-unforeseen events, may introduce an additional layer of uncertainty. It is important to recognize that despite our best efforts to address data-related challenges, the 
-robustness and reliability of our forecasting outcomes may still be subject to the 
-limitations inherent in the original data.
+point_patch+ggthemes::theme_few()
+
+```
+
+```{r}
+#| label: fig-prob
+#| fig-width: 9
+#| fig-height: 3
+#| fig-cap: "Distribution of probabilistic forecast accuracy across different origins, averaged across the forecat horizon and all products.
+The total number of months used to calculate the accuracy in the test set is 12 months for each product."
+#| fig-pos: "H"
+
+prob_accuracy_origin_all |> ggplot(aes(x= crps, y =  fct_reorder(model, crps)))+
+  geom_boxplot()+
+  labs(y = "Models")+
+  ggthemes::theme_few()
+
+```
+
+```{r}
+#| label: fig-horizon
+#| fig-width: 11
+#| fig-height: 5
+#| fig-cap: "Average accuracy by month for 6 month. The total number of months used to calculate the accuracy in the test set is 12 for each product."
+#| fig-pos: "H"
+
+p1h <- point_accuracy_h_all |> group_by(model,h) |> summarise(mase=mean(mase) , .groups = "drop")|> filter(model != "lstm", model != "lstm_reg") |>  ggplot(aes(x = h, y = mase, colour =model))+
+  geom_point()+
+  geom_line()+
+  ggthemes::theme_few()+
+  ggthemes::scale_color_colorblind()+
+  labs(x= " Forecast horizon")+
+  theme(legend.position="none")
+
+p2h <- point_accuracy_h_all |> group_by(model,h) |> summarise(rmsse=mean(rmsse),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |> 
+  ggplot(aes(x = h, y = rmsse, colour =model))+
+  geom_point()+
+  geom_line()+
+  ggthemes::theme_few()+
+  ggthemes::scale_color_colorblind()+
+  labs(x= " Forecast horizon")+
+  theme(legend.position="bottom")
+
+p3h <- prob_accuracy_h_all |> group_by(model,h) |> summarise(crps=mean(crps),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |> 
+  ggplot(aes(x = h, y = crps, colour =model))+
+  geom_point()+
+  geom_line()+
+  ggthemes::theme_few()+
+  ggthemes::scale_color_colorblind()+
+  labs(x= " Forecast horizon")+
+  theme(legend.position="none")
+
+prob_patch <- (p1h|p2h|p3h)
+prob_patch
+```
+
+@fig-prob illustrates the forecast distribution accuracy measured by the Continuous Ranked Probability Score (CRPS), which evaluates both forecasting calibration and interval sharpness.
+A smaller CRPS value indicates better overall performance.
+We observed that incorporating domain knowledge improved forecast accuracy for most models, enhancing not only point forecasts but also probabilistic forecasts.
+Notably, TimeGpt achieved the highest probabilistic forecast accuracy, while LSTM models had the lowest.
+This aligns with the findings related to point forecast accuracy, reinforcing the earlier explanations for these results.
+
+In addition to reporting the forecast accuracy presented across different rolling origins, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-horizon.
+this is the accuracy reorted as the average across rolling origins and all products fro each given horizon.
+For both the point forecast and distributional accuracy  we can see that models incoporating information collected by domain experts improves compared to counterparts that that uses only the historical comsuptions.
+
+Overall, our results indicate that forecast produces using models with infromation colelcted via the domain expert engagement provides reliable forecasts and improves upon the models relying solely on historcial consuption.
+
+Our analysis highlights the value of using foundational time series forecasting models like TimeGpt, particularly in developing countries and the global health and humanitarian sectors.
+In these contexts, the lack of advanced analytical skills and systematic data collection can hinder effective forecasting and decision-making.
+
+The findings also emphasize the importance of incorporating relevant domain knowledge into forecasting models.
+In pharmaceutical supply logistics administration systems, data often consists solely of transactional records on consumption and distribution.
+Including exogenous variables such as administrative procedures, seasonal patterns, or conflicts, or any other relevant factor that could be identified by those with domain knowledge proved essential for reducing forecast errors.
+This approach improved both point and probabilistic forecast accuracy, enabling a more confident assessment of uncertainty.
+Therefore, the systematic collection of information about significant events and their impact on consumption is vital.
+Recording details of events such as policy changes, administration procedures, conflicts, and incorporating this information into forecasting models creates a comprehensive understanding of consumption.
+This practice enhances modeling reliability and allows institutions in developing countries and humanitarian organizations to better forecast demand, allocate resources effectively, and respond proactively.
+Establishing robust data collection systems is thus a critical step for strengthening forecasting capabilities and operational resilience.
+
+## An illustration of probabilistic forecast for Pharmacuitical product consumption
+
+In this section, we present an illustrative example of a probabilistic forecast for future consumption of _Sodium Chloride (Normal Saline)_ product.
+Due to the complexity of including such visualizations for all products, only one example is shown here.
+However, it is feasible to generate these plots for all products if needed.
+
+In practice, point forecasts are commonly used despite their limitations, but they do not account for the inherent uncertainty associated with forecasts.
+The future is inherently uncertain, and effective planning requires considering alternative scenarios.
+Probabilistic forecasts offer a comprehensive approach by assigning likelihoods to a range of possible outcomes, recognizing that different consumption levels may occur with varying probabilities.
+The primary purpose of probabilistic forecasting, as illustrated in @fig-forecast-density-hstep, is to quantify and communicate uncertainty.
+This figure displays the forecast distribution of consumption over a 6-month horizon using a density plot.
+For each month within the forecast period, a separate distribution is generated.
+The plot also includes the point forecast alongside 80% and 90% prediction intervals to illustrate potential variability.
+
+It is important to note that while point forecasts and prediction intervals can be derived from probabilistic forecasts, the reverse is not true.
+A single-point forecast cannot inherently provide the probabilistic context needed to capture the range of possible outcomes.
+Although prediction intervals can indicate a range of potential values, they do not convey the detailed probabilities of low or high consumption.
+This distinction highlights the value of probabilistic forecasting in supporting informed decision-making by offering a clearer view of future uncertainties.
+
+```{r}
+#| label: fig-forecast-density-hstep
+#| fig-width: 9
+#| fig-height: 4
+#| fig-cap: "A graphical illustration of the forecast distribution of a pharmacutical product (i.e.
+total incidence attended) for the SB health board for a horizon of six month.
+For each month, we display the point forecast (black point), the histogram, and 80% (thick line) and 90% (thin line) prediction intervals.
+It also shows a portion of a historical time series as well as its fitted values."
+#| fig-pos: "H"
+
+amoxicillin <- med_qty |> filter(item == "Sodium Chloride (Normal Saline)") |> select(month,quantity) |> 
+  as_tsibble(index = month)
+
+fit <- amoxicillin |> filter_index(. ~ "2022 Jan") |>
+  model(ets = ARIMA(quantity))
+
+fcst <- fit |> forecast(h = 6)
+fitted_ets <- fit |> augment()
+
+ggplot(data = fcst, mapping = aes(x = month, ydist = quantity))+
+  ggdist::stat_halfeye(alpha = .4)+
+  geom_line(aes(y=.mean, colour ="Point Forecast"))+
+  geom_line(aes(y = .fitted, colour ="Fitted"), data = filter_index(fitted_ets, "2021 Jan" ~ .))+
+  geom_point(aes(y = .fitted, colour ="Fitted"), data = filter_index(fitted_ets, "2021 Jan" ~ .))+
+  geom_line(aes(y = quantity, colour ="Actual"),data = filter_index(amoxicillin, "2021 Jan" ~ .))+
+  geom_point(aes(y = quantity, colour ="Actual"),data = filter_index(amoxicillin, "2021 Jan" ~ .))+
+  scale_color_manual(name=NULL,
+                     breaks=c('Fitted', 'Actual',"Point Forecast"),
+                     values=c('Fitted'='#E69F00', 'Actual'='#0072B2',"Point Forecast"="#000000"))+
+  ggthemes::theme_few()
+```

### 2024-11-02 manuscript/05-conclusion.qmd
first draft ready

@@
-# Conclusion {#sec-conclusion}
+# Conclusion and future reserach {#sec-conclusion}
 
-In this study, we embarked on a comprehensive analysis of pharmaceutical demand forecasting within the Ethiopian 
-Pharmaceutical Supply Services (EPSS) context. By leveraging five  years of sales data for 37 key pharmaceutical 
-products, we aimed to  enhance the accuracy of demand predictions and consequently  contribute to more effective supply 
-chain management. Our findings  underscore the significance of accurate demand forecasting in  optimizing inventory 
-levels, resource allocation, and overall supply chain operations.
+In this study, we conducted an extensive analysis of pharmaceutical demand forecasting within the Ethiopian Pharmaceutical Supply Service (EPSS).
+Using five years of consumption data for 33 key pharmaceutical products, along with additional information gathered through collaboration with domain experts at EPSS, our goal is to contribute to enhancing the demand forecasting process and emphasize the importance of integrating domain knowledge into model building.
+This step is critical, as the phase of understanding the data and incorporating valuable contextual information is often overlooked.
+Too frequently, forecasting efforts rely solely on consumption data, leading to building models with little relevance to reality.
+Our approach highlights the necessity of including domain insights to construct more accurate and effective forecasting models.
 
-The application of diverse forecasting models yielded valuable insights into their respective performance. Notably, the 
-Autoregressive Integrated Moving Average (ARIMA) model emerged as a  robust choice, consistently exhibiting superior forecast accuracy  across a majority of the selected pharmaceutical products.
-Additionally, the incorporation of predictor variables, such as  physical count, budget release and closure by the government,  COVID-19 impact, and campaign events, further underscored the  efficacy of the ARIMA model in accommodating external influences and delivering reliable forecasts.
+Our findings highlight the importance of collecting and incorporating domain knowledge when building forecasting models.
+We evaluated both point and probabilistic forecasts using a range of models, from simple univariate approaches like ARIMA to complex models such as LSTM.
+Notably, we demonstrated that employing complex models does not necessarily lead to more accurate forecasts in the pharmaceutical supply context.
+Additionally, our research revealed that newly developed foundational time series forecasting models have potential applicability, particularly in environments where advanced analytical skills and knowledge are limited, as often seen in pharmaceutical supply chains in developing countries.
 
-The individualized assessment of forecasting models revealed specific strengths across pharmaceutical categories. ARIMA 
-models with predictors demonstrated exceptional performance for pharmaceuticals including Amoxacillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine, Atenolol, and others, affirming their utility in capturing complex demand patterns.
-Moreover, the study shed light on pharmaceuticals best suited for alternative models like naive, mean, and regression 
-methods, contributing to a comprehensive understanding of forecasting applicability.
+To advance forecasting practices further, exploring additional predictor variables—such as public health campaigns, disease outbreaks, and economic indicators—could provide valuable insights into demand dynamics.
+Improving the granularity of historical consumption data, through finer temporal intervals and geographic differentiation, also holds potential for more precise predictions.
+Additionally, hybrid forecasting models that combine judgmental point forecasts with probabilistic machine learning models could enhance predictive performance.
+Empowering EPSS staff through training in foundational time series models, data interpretation, and collection will enable informed decision-making and strengthen workforce capabilities.
+Replicating this study across diverse healthcare settings and conducting comparative analyses would help validate the adaptability and applicability of the findings.
+Additionally, it is recommended that pharmaceutical supply services systematically collect and maintain detailed records of events that may influence consumption, alongside consumption data.
+This practice is essential for refining demand models and enhancing forecast accuracy.
 
-While these findings contribute significantly to the discourse on pharmaceutical demand forecasting, it's essential to 
-acknowledge limitations. The accuracy of our forecasts is intrinsically linked to the quality and completeness of 
-historical sales data, and the intricate interplay of external variables adds complexity to the forecasting landscape.
+## Practical challenges and limitations
 
-In conclusion, this study serves as a pivotal step towards enhancing pharmaceutical supply chain management through 
-refined demand forecasting. The demonstrated success of ARIMA models, particularly when augmented with predictor 
-variables, holds promise for fostering resilient and adaptable supply chains. As the pharmaceutical landscape continues 
-to evolve, our insights encourage further exploration of advanced forecasting techniques and continuous refinement of 
-strategies to bolster supply chain efficiency and ensure consistent availability of essential pharmaceuticals.
+Despite dedicated efforts to engage with domain experts to better understand the data—and spending significant time over two weeks collaborating with them—we found that the process of interpreting data through domain knowledge is complex and presents unique challenges.
+These challenges include defining what constitutes domain knowledge and determining how best to incorporate it to enhance model reliability and accuracy.
+Below, we summarize some of these key challenges.
+One issue involved capturing accurate information from bin cards and expert review, particularly during periods of disruption like COVID-19 and conflicts.
+For instance, during the COVID-19 pandemic, experts noted a decrease in demand due to travel restrictions.
+However, despite the reduced demand, there were still significant distributions of pharmaceuticals from the EPSS central to various hubs and health facilities.
+This was because, in response to anticipated shortages, a political decision was made to push products to these facilities before the travel restrictions took full effect.
+The rationale was that if the facilities remained closed, patients would have no access to medications, necessitating the preemptive stockpiling of essential pharmaceuticals.
+This scenario illustrates how policy decisions can significantly impact supply chain data, making it challenging to model distribution patterns accurately.
+Similarly, experts indicated that conflicts would disrupt the distribution of pharmaceuticals.
+While conflict zones did indeed hinder transportation, distribution still occurred whenever roads were temporarily opened, even amidst on-going conflict.
+This created inconsistencies in the data, as periods of halted distribution were followed by rapid replenishment once access was restored.
+Such fluctuations add complexity to modelling the supply chain from the central EPSS to hubs and health facilities.
+Other challenges involved seasonal or periodic activities, such as the fiscal year-end stock counting.
+During this time, EPSS temporarily halts transactions to conduct a full inventory count, which can take anywhere from one to two months, depending on the circumstances.
+This inconsistency in the duration of inventory counts adds a layer of unpredictability to supply chain modelling.
 
-In light of the study's findings, several actionable recommendations emerge. First and foremost, the integration of the 
-Autoregressive Integrated Moving Average (ARIMA) model with predictor variables stands out as a viable strategy for 
-enhancing pharmaceutical demand forecasting accuracy within Ethiopian Pharmaceutical Supply Services (EPSS). This entails 
-harnessing the predictive power of external factors to refine forecasting precision. To further advance forecasting 
-practices, the exploration of additional predictor variables such as public health campaigns, disease outbreaks, and 
-economic indicators could offer valuable insights into demand dynamics. Moreover, refining the granularity of historical 
-sales data, possibly through finer temporal intervals and geographic diversity, holds potential for more nuanced and 
-accurate predictions. Consideration could also be given to hybrid forecasting models that amalgamate ARIMA's short-term 
-accuracy with machine learning's predictive prowess for long-term trends.
-Regular model evaluation, sensitivity analysis, and collaborative stakeholder engagement are recommended to ensure ongoing accuracy and alignment with changing demand patterns.
+## Future reserach
 
-Capacitating EPSS staff through training programs in forecasting methodologies and data interpretation can empower the 
-workforce to make informed decisions. Replication of this study in diverse healthcare settings and comparative analyses 
-can validate the findings' applicability and adaptability. By embracing these recommendations, EPSS can fortify its 
-pharmaceutical supply chain with resilient, efficient, and responsive operations, ultimately ensuring the consistent 
-availability of vital pharmaceuticals.
-  a
+Following the current research, several promising areas could further advance knowledge and practical applications in this area:
+
+- Future research could focus on developing intelligent systems that incorporate domain-specific knowledge into model building.
+This would involve creating algorithms capable of identifying relevant domain insights and integrating them effectively into model training processes.
+Such tools would bridge the gap between purely data-driven approaches and expert-knowledge-enhanced modeling, leading to more robust and contextually informed forecasts.
+
+- The use of distribution-based representation of error metrics across different rolling origins or forecast horizons, such as boxplots, to showcase forecast accuracy provides essential insights into variability in accuracy at different point of time in data.
+Building on this, future work could aim at designing methodologies that identify diagnostic patterns within the distribution of errors.
+These approaches could suggest model improvements by analyzing error distributions and help practitioners to refine models based on error behavior.
+
+- While current study focus on monthly aggregate consumption at the organizational level, there is potential in exploring more granular forecasts.
+Future research could investigate daily or weekly forecasting and examine the impacts at hierarchical levels, such as specific sites or health facilities.
+Investigating how fine-grained temporal and hierarchical forecasts influence inventory management and operational efficiency could yield valuable insights for reducing stockouts, minimizing costs and reducing waste.

### 2024-11-02 manuscript/main.qmd
Revised sections
first draft ready

@@
 ---
-title: "Pharmaceutical Supply Chain Forecasting by using Machine Learning in Ethiopian Pharmaceutical Supply Services"
+title: "The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting"
 author:
@@
           name: Addis Ababa University
-          department: Department of Pharmaceutics and Social Pharmacy at School of Pharmacy
-          country: Ethiopia
-          postal-code: 9086
+          department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
+          country: Ethiopia.
     attributes:
         corresponding: true
-  - name: Arebu Issa Bilal
-    email: arebu.issa@aau.edu.et
   - name: Bahman Rostami-Tabar
@@
         - id: carbs
-          name: University
-          department: Cardiff Business School 
-          country: UK
-          postal-code: CF5
+          name: Cardiff University
+          department: Data Lab for Social Good Reserach Group, Cardiff Business School 
+          country: UK.
+  - name: Harsha Halgamuwe Hewage
+    email: HalgamuweHewageHR@cardiff.ac.uk
+    affiliations:
+        - ref: carbs
   - name: Umit Sezer Bititci 
@@
     affiliations:
-        - id: heriot- watt 
-          name: University
-          department: Edinburgh Business School 
-          country: UK
-          postal-code: EH14 4AS
+        - id: herio 
+          name: Heriot-Watt University
+          department: Edinburgh Business School, Heriot-Watt University
+          country: UK.
   - name: Teferi Gedif Fenta
@@
     affiliations:
-        - id: Addis 
-          name: Addis Ababa  University
-          department: Department of Pharmaceutics and Social Pharmacy at School of Pharmacy
-          country: Ethiopia
-          postal-code: 9086
+        - ref: Addis
 
 abstract: |
-  This study delves into the realm of pharmaceutical demand forecasting within the Ethiopian Pharmaceuticals Supply Services(EPSS) context.
-It encompasses a comprehensive analysis of 37 key pharmaceutical products,assessing their forecast accuracy across variation models.
-The initial assessment reveals that the Autoregressive Integrated Moving Average (ARIMA) model emerges as a frotrunner, boasting a Root Mean Squared Error (RMSSE) value of 0.850.
-ARIMA's prowess in capturing complex demand patterns suggests its suitability for pharmaceutical sales forecasting in the EPSS landscape.
-To argument forecasting accuarcy, predictor varaibles such as physical count, budget release and closure by the goverment, the impact of COVID-19, and promotional campaigns were incorpoarted.
-Remarkably, the ARIMA model mantainsits superiority even in the presences of these external factors, reiterating its roustness.
-Delving further into individual pharmaceutical products, ARIMA with predictors demonstrates exceptional performance for a substantial portion (32.43%) of the products, including Amoxicillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine, and others.
-Conversely,alternative models like naive, mean, and regression methods showcase strengths for different pharmaceutical categories, underscoring the importance of tailoring forecasting approaches to product-specific dynamics.
-While this study significantly advances the discourse on pharmaceutical demand forecasting, limitations stemming from data quality and external variables underscore the intricacies of the forecasting landscape.
-In conclusion, it advocates the integration of the ARIMA model with predictor variables as a strategic approach for refining pharmaceutical demand forecasting within EPSS.
-It also encourages further exploration of advanced forecasting techniques and ongoing refinement to bolster supply chain efficiency and ensure consistent pharmaceutical availability.
-Capacity building through staff training and replication of the study in diverse healthcare settings are recommended steps toward resilient and efficient pharmaceutical supply chains.These findings hold promise for fostering supply chains that are adaptable, responsive, and capable of ensuring the continuous availability of essential pharmaceuticals in Ethiopia.
+  Accurate forecasting in pharmaceutical supply chains is critical for ensuring the continuous availability of essential medicines, particularly in developing countries where resource constraints and logistical challenges are more pronounced.
+Effective demand forecasting supports procurement and inventory management, which in turn could help to prevent stockouts, reduce wastage due to overstocking, and enhance overall healthcare delivery.
+In such settings, reliable forecasts, that acknowledge uncertainties, are essential to ensure that limited resources are used efficiently to meet health needs.
+  Despite advancements in forecasting, significant gaps remain in effective forecasting in poor resource countries.
+Most existing literature on pharmaceutical supply chain forecasting focuses primarily on point estimation of consumption while neglecting the inherent uncertainty of these forecasts Additionally, these studies often rely solely on historical consumption data without considering the broader context that influences consumption Furthermore, many works lack rigorous methodological design, transparency, and reproducibility.
+This study addresses these gaps by integrating domain-specific knowledge, gathered through expert interviews and engagement with key members of the Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models.
+Using a dataset spanning five years (December 2017 to July 2022) from EPSS, we developed forecasting models that integrate expert-identified variables such as stock replenishment schedules, fiscal inventory counts, conflicts, and disease outbreaks.
+Evaluation metrics such as Mean Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE), and Continuous Ranked Probability Score (CRPS) are used to report forecast accuracy.
+The findings underscore the significance of contextual data in developing robust forecasting models that are adaptable to complex, real-world conditions.
+Our results also highlight the effectiveness of foundational time series methods, such as LLM models, for forecasting.
+These methods are particularly appealing for resource-constrained countries that may lack advanced analytical expertise.
+To promote usability, generalizability, and reproducibility, we share the complete dataset and code in R and Python, and the entire paper is written in Quarto via a GitHub repository to facilitate these practices.
+
 keywords:
-  - Pharmaceutical Demand Forecasting
-  - Ethiopian Pharmaceutical Supply Services
-  - ARIMA Model
-  - Forecasting Variables
-  - Forecast Accuracy
+  - Forecasting
+  - Pharmaceutical supply chain
+  - Domain knowlege
+  - Forecast accuracy
+  - Developing countries
 bibliography: references.bib
@@
 ggplot2::theme_set(ggthemes::theme_few())
+
 ```
@@
 
-```{r experiment, child="02-experiment.qmd"}
+```{r experiment, child="03-experiment.qmd"}
 
@@
 
-```{r results, child="03-results.qmd"}
+```{r results, child="04-results.qmd"}
 
@@
 
-```{r conclusion, child="04-conclusion.qmd"}
+```{r conclusion, child="05-conclusion.qmd"}
 
@@
 
-R code to produce all results in this paper is available at
+We use R and Python for model development and accuracy evaluation. The paper is written in Quarto, and all code, data, and files will be shared in a GitHub repository upon acceptance for publication.
 

### 2024-11-02 rscript/00_data_preperation.R

@@ -1,0 +1,88 @@
+# cross validation data preperation for LLMs and DL models
+
+library(tidyverse)
+library(fable)
+library(tsibble)
+
+
+# read data
+
+med_qty <- read.csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
+  janitor::clean_names()
+
+# make the tsibble
+
+med_qty_tsb <- med_qty |> 
+  mutate(month = yearmonth(month)) |> 
+  group_by(item) |> 
+  mutate(item_id = sprintf("id_%02d", cur_group_id())) |>  #creating a item id to handle the data easy
+  ungroup() |>
+  select(-stock_out, -covid_19, -conflict) |> 
+  as_tsibble(index = month, key = c(item_id, item))
+
+
+# create product master
+
+product_master <- med_qty_tsb |> 
+  as_tibble() |> 
+  select(item, item_id) |>
+  unique() |> 
+  write_rds('data/product_master.rds')
+
+
+# check for temporal gaps
+
+med_qty_tsb |> count_gaps()  # seems there are no temporal gaps
+
+
+# create tscv train data
+
+f_horizon <- 6
+
+test_length <- 12
+
+med_tsb_train_tscv <- med_qty_tsb |> 
+  filter_index(. ~ as.character(max(med_qty_tsb$month)-(f_horizon))) |>
+  stretch_tsibble(.init = (length(unique(med_qty_tsb$month)) - test_length), .step = 1) |> 
+  mutate(unique_id = paste0(item_id, '-', .id))  |>  
+  rename(horizon = .id)
+
+
+# create tscv test data
+
+med_tsb_test_tscv <- med_qty_tsb |> 
+  filter_index(as.character(max(med_qty_tsb$month)-(test_length) + 1) ~ .) |>
+  slide_tsibble(.size = 6, .step = 1) %>% 
+  mutate(unique_id = paste0(item_id, '-', .id)) %>% 
+  rename(horizon = .id)
+
+
+# filter malaria products and save data
+
+med_tsb_train_tscv_no_malaria <- med_tsb_train_tscv |> 
+  filter(item != 'Artemether + Lumefanthrine' , item != 'Rapid Diagnostic Test') |> 
+  as_tibble() |> 
+  select(-item_id, -item) |> 
+  write.csv('data/train_tscv_no_malaria.csv', row.names = F)
+
[…]

### 2024-11-02 rscript/01_timegpt_dist.R

@@ -1,0 +1,157 @@
+# generate distributions for TimeGPT models
+
+# load libraries
+
+library(tidyverse)
+
+
+# data preparation --------------------------------------------------------
+
+timegpt_malaria <- read.csv('results/timegpt_malaria.csv')
+timegpt_no_malaria <- read.csv('results/timegpt_no_malaria.csv')
+timegpt_malaria_reg <- read.csv('results/timegpt_malaria_reg.csv')
+timegpt_no_malaria_reg <- read.csv('results/timegpt_no_malaria_reg.csv')
+
+product_master <- read_rds('data/product_master.rds')
+
+
+# Interpolate points ------------------------------------------------------
+
+# Function to interpolate 1000 points from quantiles (timegpt only generates quantiles)
+
+interpolate_points <- function(row_quantiles, quantiles, n_points = 1022) {
+  approx(x = quantiles, y = row_quantiles, xout = seq(0, 1, length.out = n_points))$y
+}
+
+# Define quantile levels
+
+quantiles <- quantiles <- seq(0.01, 0.99, by = 0.01)
+
+
+# malaria_reg -------------------------------------------------------------
+
+# Apply the interpolation function row-wise
+
+distribution_points <- t(apply(timegpt_malaria_reg %>% select(-unique_id, -ds, -TimeGPT), 1, interpolate_points, quantiles = quantiles))
+
+
+# Create column names for the distribution points
+
+colnames(distribution_points) <- paste0('X', seq_len(ncol(distribution_points)))
+
+
+# Bind the distribution points to the original data
+
+timegpt_malaria_reg_updated <- timegpt_malaria_reg %>%
+  bind_cols(as.data.frame(distribution_points)) %>% 
+  select(where(function(x) !all(is.na(x)))) %>% 
+  select(-contains("Time")) %>% 
+  mutate(across(starts_with("X"), ~ if_else(. < 0, 0, .))) %>%
+  mutate(prediction = rowMeans(across(starts_with("X")))) %>% 
+  rename_with(~ paste0("X", as.integer(str_extract(., "\\d+")) - 12), starts_with("X")) %>% 
+  separate(unique_id, into = c("id", "horizon"), sep = "-") %>%
+  rename(month = ds, item_id = id) %>%  
+  mutate(horizon = as.integer(horizon),
+         month = yearmonth(month),
+         model = 'timegpt_reg') %>% 
+  as_tibble()
+
+
[…]

### 2024-11-02 rscript/02_timegpt_dist.R

@@ -1,0 +1,158 @@
+# generate distributions for TimeGPT models
+
+# load libraries
+
+library(tidyverse)
+library(fable)
+library(tsibble)
+
+
+# data preparation --------------------------------------------------------
+
+timegpt_malaria <- read.csv('results/timegpt_malaria.csv')
+timegpt_no_malaria <- read.csv('results/timegpt_no_malaria.csv')
+timegpt_malaria_reg <- read.csv('results/timegpt_malaria_reg.csv')
+timegpt_no_malaria_reg <- read.csv('results/timegpt_no_malaria_reg.csv')
+
+product_master <- read_rds('data/product_master.rds')
+
+
+# Interpolate points ------------------------------------------------------
+
+# Function to interpolate 1000 points from quantiles (timegpt only generates quantiles)
+
+interpolate_points <- function(row_quantiles, quantiles, n_points = 1022) {
+  approx(x = quantiles, y = row_quantiles, xout = seq(0, 1, length.out = n_points))$y
+}
+
+# Define quantile levels
+
+quantiles <- quantiles <- seq(0.01, 0.99, by = 0.01)
+
+
+# malaria_reg -------------------------------------------------------------
+
+# Apply the interpolation function row-wise
+
+distribution_points <- t(apply(timegpt_malaria_reg %>% select(-unique_id, -ds, -TimeGPT), 1, interpolate_points, quantiles = quantiles))
+
+
+# Create column names for the distribution points
+
+colnames(distribution_points) <- paste0('X', seq_len(ncol(distribution_points)))
+
+
+# Bind the distribution points to the original data
+
+timegpt_malaria_reg_updated <- timegpt_malaria_reg %>%
+  bind_cols(as.data.frame(distribution_points)) %>% 
+  select(where(function(x) !all(is.na(x)))) %>% 
+  select(-contains("Time")) %>% 
+  mutate(across(starts_with("X"), ~ if_else(. < 0, 0, .))) %>%
+  mutate(prediction = rowMeans(across(starts_with("X")))) %>% 
+  rename_with(~ paste0("X", as.integer(str_extract(., "\\d+")) - 12), starts_with("X")) %>% 
+  separate(unique_id, into = c("id", "horizon"), sep = "-") %>%
+  rename(month = ds, item_id = id) %>%  
+  mutate(horizon = as.integer(horizon),
+         month = yearmonth(month),
+         model = 'timegpt_reg') %>% 
[…]

### 2024-11-02 rscript/03_lstm.R

@@ -1,0 +1,605 @@
+# lstm model
+
+# Install necessary packages if not already installed
+
+# install.packages(c("keras", "tensorflow"))
+
+
+# Load libraries
+
+library(keras)
+library(tensorflow)
+library(tidyverse)
+library(tsibble)
+library(doParallel)
+library(foreach)
+
+
+# Prepare data ------------------------------------------------------------
+
+# read data
+
+med_qty <- read.csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
+  janitor::clean_names()
+
+
+# make the tsibble
+
+med_qty_tsb <- med_qty |> 
+  mutate(month = yearmonth(month)) |> 
+  group_by(item) |> 
+  mutate(item_id = sprintf("id_%02d", cur_group_id())) |>  #creating a item id to handle the data easy
+  ungroup() |>
+  as_tsibble(index = month, key = c(item_id, item)) |> 
+  select(-stock_out, -covid_19, -conflict)
+
+
+# create tscv train data
+
+f_horizon <- 6
+
+test_length <- 12
+
+med_tsb_train_tscv <- med_qty_tsb |> 
+  filter_index(. ~ as.character(max(med_qty_tsb$month)-(f_horizon))) |>
+  stretch_tsibble(.init = (length(unique(med_qty_tsb$month)) - test_length), .step = 1) |> 
+  mutate(unique_id = paste0(item_id, '-', .id))  |>  
+  rename(horizon = .id)
+
+
+# create tscv test data
+
+med_tsb_test_tscv <- med_qty_tsb |> 
+  filter_index(as.character(max(med_qty_tsb$month)-(test_length) + 1) ~ .) |>
+  slide_tsibble(.size = 6, .step = 1) |> 
+  mutate(unique_id = paste0(item_id, '-', .id)) |> 
+  rename(horizon = .id)
+
+# univariate model setup (without scaling) --------------------------------------------------
+
+
+# Define function to create sequences for LSTM
+
+create_sequences <- function(data, target_column, time_steps) {
+  X <- array(dim = c(0, time_steps, ncol(data)))
+  Y <- NULL
+  for (i in 1:(nrow(data) - time_steps)) {
+    # X: sequence of time_steps
+    sequence <- array(as.matrix(data[i:(i + time_steps - 1), ]), dim = c(1, time_steps, ncol(data)))
+    X <- abind::abind(X, sequence, along = 1)
[…]

### 2024-11-02 rscript/04_accuracy_evaluation.R

@@ -1,0 +1,268 @@
+# cross validation
+
+library(tidyverse)
+library(tsibble)
+library(Metrics)
+library(SpecsVerification)
+library(greybox)
+
+
+# data preperation --------------------------------------------------------
+
+# read data
+
+med_qty <- read.csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
+  janitor::clean_names()
+
+product_master <- read_rds('data/product_master.rds')
+
+snaive <- read_rds('results/snaive_prob.rds') |> 
+  rename(origin = horizon) |> # i have mistakenly named origin as horizon
+  mutate(unique_id = paste0(item_id,'-',origin))
+
+
+# make the tsibble
+
+med_qty_tsb <- med_qty |> 
+  mutate(month = yearmonth(month)) |> 
+  group_by(item) |> 
+  mutate(item_id = sprintf("id_%02d", cur_group_id())) |>  #creating a item id to handle the data easy. 
+  ungroup() |>
+  as_tsibble(index = month, key = c(item_id, item))
+
+
+# create tscv train data
+
+f_horizon <- 6
+
+test_length <- 12
+
+med_tsb_train_tscv <- med_qty_tsb |> 
+  filter_index(. ~ as.character(max(med_qty_tsb$month)-(f_horizon))) |>
+  stretch_tsibble(.init = (length(unique(med_qty_tsb$month)) - test_length), .step = 1) |> 
+  mutate(unique_id = paste0(item_id, '-', .id))  |>  
+  rename(origin = .id)
+
+
+# create tscv test data
+
+med_tsb_test_tscv <- med_qty_tsb |> 
+  filter_index(as.character(max(med_qty_tsb$month)-(test_length) + 1) ~ .) |>
+  slide_tsibble(.size = 6, .step = 1) |> 
+  mutate(unique_id = paste0(item_id, '-', .id)) |> 
+  rename(origin = .id)
+
+
+# Create prediction masters -----------------------------------------------
+
+
+pred_master <- read_rds('results/timegpt_fc.rds') |> 
+  bind_rows(read_rds('results/lstm_uni.rds') |> 
+              mutate(model = 'lstm')) |> 
+  bind_rows(read_rds('results/lstm_multi.rds') |> 
+              mutate(model = 'lstm_reg')) |> 
+  rename(origin = horizon) # i have mistakenly named origin as horizon
+
+  
+point_pred <- pred_master |> 
[…]

### 2024-11-02 rscript/04_accuracy_evaluation_origin.R

@@ -1,0 +1,272 @@
+# cross validation
+
+library(tidyverse)
+library(tsibble)
+library(Metrics)
+library(SpecsVerification)
+library(greybox)
+
+
+# data preperation --------------------------------------------------------
+
+# read data
+
+med_qty <- read.csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
+  janitor::clean_names()
+
+product_master <- read_rds('data/product_master.rds')
+
+snaive <- read_rds('results/snaive_prob.rds') |> 
+  rename(origin = horizon) |> # i have mistakenly named origin as horizon
+  mutate(unique_id = paste0(item_id,'-',origin))
+
+
+# make the tsibble
+
+med_qty_tsb <- med_qty |> 
+  mutate(month = yearmonth(month)) |> 
+  group_by(item) |> 
+  mutate(item_id = sprintf("id_%02d", cur_group_id())) |>  #creating a item id to handle the data easy. 
+  ungroup() |>
+  as_tsibble(index = month, key = c(item_id, item))
+
+
+# create tscv train data
+
+f_horizon <- 6
+
+test_length <- 12
+
+med_tsb_train_tscv <- med_qty_tsb |> 
+  filter_index(. ~ as.character(max(med_qty_tsb$month)-(f_horizon))) |>
+  stretch_tsibble(.init = (length(unique(med_qty_tsb$month)) - test_length), .step = 1) |> 
+  mutate(unique_id = paste0(item_id, '-', .id))  |>  
+  rename(origin = .id)
+
+
+# create tscv test data
+
+med_tsb_test_tscv <- med_qty_tsb |> 
+  filter_index(as.character(max(med_qty_tsb$month)-(test_length) + 1) ~ .) |>
+  slide_tsibble(.size = 6, .step = 1) |> 
+  mutate(unique_id = paste0(item_id, '-', .id)) |> 
+  rename(origin = .id)
+
+
+# Create prediction masters -----------------------------------------------
+
+
+pred_master <- read_rds('results/timegpt_fc.rds') |> 
+  bind_rows(read_rds('results/lstm_uni.rds') |> 
+              mutate(model = 'lstm')) |> 
+  bind_rows(read_rds('results/lstm_multi.rds') |> 
+              mutate(model = 'lstm_reg')) |> 
+  rename(origin = horizon) # i have mistakenly named origin as horizon
+
+  
+point_pred <- pred_master |> 
[…]

### 2024-11-02 rscript/05_snaive.R

@@ -1,0 +1,113 @@
+# sNaive
+
+library(tidyverse)
+library(tsibble)
+library(fable)
+library(doParallel)
+library(foreach)
+
+
+# data preperation --------------------------------------------------------
+
+# read data
+
+med_qty <- read.csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
+  janitor::clean_names()
+
+product_master <- read_rds('data/product_master.rds')
+
+
+# make the tsibble
+
+med_qty_tsb <- med_qty |> 
+  mutate(month = yearmonth(month)) |> 
+  group_by(item) |> 
+  mutate(item_id = sprintf("id_%02d", cur_group_id())) |>  #creating a item id to handle the data easy
+  ungroup() |>
+  as_tsibble(index = month, key = c(item_id, item))
+
+
+# create tscv train data
+
+f_horizon <- 6
+
+test_length <- 12
+
+med_tsb_train_tscv <- med_qty_tsb |> 
+  filter_index(. ~ as.character(max(med_qty_tsb$month)-(f_horizon))) |>
+  stretch_tsibble(.init = (length(unique(med_qty_tsb$month)) - test_length), .step = 1) |> 
+  mutate(unique_id = paste0(item_id, '-', .id))  |>  
+  rename(horizon = .id)
+
+
+# sNaive ------------------------------------------------------------------
+
+# id list
+
+id_list <- med_tsb_train_tscv |> 
+  as_tibble() |> 
+  select(unique_id) |> 
+  pull() |> 
+  unique()
+
+
+# Register outer parallel backend with 5 cores
+
+cl_outer <- makeCluster(5)
+registerDoParallel(cl_outer)
+
+
+# parallel loop
+
+system.time(snaive_prob_pred <- foreach(i = id_list,
+                                        .combine = 'rbind',
+                                        .packages=c("doParallel", "foreach", "tidyverse", "fpp3")) %dopar% {
+                                          
+                                          # i <- id_list[1]
+                                          
+                                          # train sets
+                                          
+                                          train_data <- med_tsb_train_tscv |> 
+                                            filter(unique_id == i)
+                                          
+                                          
+                                          # model fitting
+                                          
+                                          snaive_fit <- train_data |> 
+                                            model(snaive = SNAIVE(quantity))
+                                          
+                                          
+                                          # forecasting
+                                          
+                                          fc_horizon <- 6
+                                          
+                                          boot <- 1000
+                                          
[…]

### 2024-11-02 rscript/06_accuracy_evaluation_horizon.R

@@ -1,0 +1,292 @@
+# accuracy evaluation - horizons
+
+# cross validation
+
+library(tidyverse)
+library(tsibble)
+library(Metrics)
+library(SpecsVerification)
+library(greybox)
+
+
+# data preperation --------------------------------------------------------
+
+# read data
+
+med_qty <- read.csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
+  janitor::clean_names()
+
+product_master <- read_rds('data/product_master.rds')
+
+snaive <- read_rds('results/snaive_prob.rds') |> 
+  rename(origin = horizon) |> # i have mistakenly named origin as horizon
+  group_by(month, item_id) |> 
+  mutate(horizon = row_number()) |> # creating the horizon
+  ungroup() |> 
+  mutate(unique_id = paste0(item_id,'-',horizon))
+
+
+# make the tsibble
+
+med_qty_tsb <- med_qty |> 
+  mutate(month = yearmonth(month)) |> 
+  group_by(item) |> 
+  mutate(item_id = sprintf("id_%02d", cur_group_id())) |>  #creating a item id to handle the data easy. 
+  ungroup() |>
+  as_tsibble(index = month, key = c(item_id, item))
+
+
+# create tscv train data
+
+f_horizon <- 6
+
+test_length <- 12
+
+med_tsb_train_tscv <- med_qty_tsb |> 
+  filter_index(. ~ as.character(max(med_qty_tsb$month)-(f_horizon))) |>
+  stretch_tsibble(.init = (length(unique(med_qty_tsb$month)) - test_length), .step = 1) |> 
+  rename(origin = .id) |> 
+  as_tibble() |> 
+  group_by(month, item_id) |> 
+  mutate(horizon = row_number()) |> # creating the horizon
+  ungroup() |> 
+  mutate(unique_id = paste0(item_id, '-', horizon))
+
+
+# create tscv test data
+
+med_tsb_test_tscv <- med_qty_tsb |> 
+  filter_index(as.character(max(med_qty_tsb$month)-(test_length) + 1) ~ .) |>
+  slide_tsibble(.size = 6, .step = 1) |> 
+  rename(origin = .id) |> 
+  as_tibble() |> 
+  group_by(month, item_id) |> 
+  mutate(horizon = row_number()) |> # creating the horizon
+  ungroup() |> 
+  mutate(unique_id = paste0(item_id, '-', horizon))
+
+
[…]

### 2024-11-02 rscript/bahman.R

@@ -1,0 +1,133 @@
+library(tidyverse)
+library(fpp3)
+
+med_qty1 <- read_csv('data/full_issued_data.csv') |>
+  janitor::clean_names() |> mutate(month = dmy(month), month = yearmonth(month))
+  
+med_qty <- read_csv('data/predicotrs_final_tssible_July_04_2024.csv') |> 
+  janitor::clean_names() |> mutate(month = yearmonth(month))
+
+itemselec <- med_qty |> filter(item == "Oxytocin") |> select(-item)
+ggplot(itemselec, 
+       aes(x=month, y=quantity))+
+  geom_point()+
+  geom_line(aes(group =1))+
+  geom_point(data = itemselec |> filter(stock_out==1) |> select(month,quantity),
+             size=3, colour="orange")+
+  ggthemes::theme_clean()+
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
+
+med_qty_tsb <- med_qty |> 
+  mutate(month = yearmonth(month)) |> 
+  as_tsibble(index = month, key = item)
+
+f_horizon <- 6
+
+test_length <- 12
+
+med_tsb_train_tscv <- med_qty_tsb |> 
+  filter_index(. ~ as.character(max(med_qty_tsb$month)-(f_horizon))) |>
+  stretch_tsibble(.init = (length(unique(med_qty_tsb$month)) - test_length), .step = 1)
+
+# create tscv test data
+
+med_tsb_test_tscv <- med_qty_tsb |> 
+  filter_index(as.character(max(med_qty_tsb$month)-(test_length) + 1) ~ .) |>
+  slide_tsibble(.size = 6, .step = 1)
+
+# med_tsb_test_tscv_stockout <- med_tsb_test_tscv |> mutate(quantity = if_else(stock_out ==1, NA,quantity))
+
+# filter malaria products and save data
+
+med_tsb_train_tscv_no_malaria <- med_tsb_train_tscv |> 
+  filter(item != 'Artemether + Lumefanthrine' , item != 'Rapid Diagnostic Test')
+
+med_tsb_train_tscv_malaria <- med_tsb_train_tscv |> 
+  filter(item == 'Artemether + Lumefanthrine' | item == 'Rapid Diagnostic Test')
+
+med_tsb_test_tscv_no_malaria <- med_tsb_test_tscv |> 
[…]

### 2024-12-13 manuscript/01-introduction.qmd
final draft for submission

@@
 Demand forecasting in pharmaceutical supply chains is a highly complex process, shaped by several factors.
-These include the quality of available data, the specific position within the supply chain, and external influences such as market trends, regulatory changes, and disruptive events such as pandemics, conflicts, flooding.
-[@schneider2010pharmaceutical; @hyndman2018forecasting].
+These include the quality of available data, the specific position within the supply chain, and external influences such as market trends, regulatory changes, and disruptive events such as pandemics, conflicts, flooding [@schneider2010pharmaceutical; @hyndman2018forecasting].
 Recently, major donors  such as Bill & Melinda Gates Foundation have invested in data collection technologies in sub-Saharan Africa, recognizing the potential of data to provide valuable insights.
@@
 Following this, we conduct exploratory data analysis and engage in interviews with EPSS experts to gather domain knowledge on events influencing consumption patterns.
-Through these interviews, the experts identified five key factors collectively impacting consumption: stock replenishment period, stockouts, fiscal calendar, COVID-19, conflicts, and the seasonal malaria.
+Through these interviews, the experts identified three key factors collectively impacting consumption: stock replenishment period, fiscal calendar, and the seasonal malaria.
 We then develop forecasting models that incorporate the data gathered through the interview process and compare their performance with univariate models that rely solely on consumption data.
@@
 Moreover, we also fill a gap in the pharmaceutical demand forecasting literature by not only generating point forecasts but also producing probabilistic forecasting techniques that capture the range of potential outcomes with associated probabilities, offering a more comprehensive view of demand uncertainty than traditional methods.
-Model performance will be evaluated using point accuracy measure including Mean Absolute Scaled Error(MASE) and Root Mean Squared Scaled Error (RMSSE), and probabilistic measures like the Continuous Ranked Probability Score (CRPS) and Winkler score.
+Model performance is evaluated using point accuracy measure including MASE, RMSSE, and probabilistic measure using CRPS.
 

### 2024-12-13 manuscript/02-literature.qmd
final draft for submission

@@
 Forecasting in healthcare not only influences clinical decisions but also plays a pivotal role in supply chain management, ensuring that the right drugs are available when needed.
-The challenge lies in balancing customer demand with inventory costs, which is particularly complex in pharmaceuticals due to factors like short shelf life and quality constraints [@gupta2000mid, @makridakis2020forecasting].
+The challenge lies in balancing customer demand with inventory costs, which is particularly complex in pharmaceuticals due to factors like short shelf life and quality constraints [@gupta2000mid; @makridakis2020forecasting].
 
-In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies using a combination of simple and advanced methods to enhance accuracy [@nikolopoulos2016forecasting,@zhu2021demand,@anusha2014demand].
+In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies using a combination of simple and advanced methods to enhance accuracy [@nikolopoulos2016forecasting;@zhu2021demand;@anusha2014demand].
 Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods like the Holt-Winters method, Single Moving Average (SMA), and Single Exponential Smoothing (SES), which are simpler but effective for certain applications.
-In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand,@candan2014demand, and @ribeiro2017improving].
+In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand, @candan2014demand, and @ribeiro2017improving.
 
@@
 The pharmaceutical industry's increasing complexity and the availability of large datasets have led to the adoption of advanced machine learning techniques.
-These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power.@van2021using, showed that incorporating downstream information into machine learning models, like LASSO and SVR, greatly enhances forecast accuracy.
+These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power.
+@van2021using showed that incorporating downstream information into machine learning models, like LASSO and SVR, greatly enhances forecast accuracy.
 Their study on multi-echelon supply chains revealed that advanced models, particularly those integrating external variables, consistently produced lower forecast errors, as measured by AvgRelRMSE.

### 2024-12-13 manuscript/03-experiment.qmd
final draft for submission

@@
 
-To gather domain knowledge about factors affected the consumption of products, we contacted experts from the Ethiopian Pharmaceuticals Supply Service (EPSS) who work in the distribution unit.
-Next, we collaborated with six experts from the Ethiopian Pharmaceuticals Supply Service (EPSS), each with over 10 years of experience managing pharmaceutical distribution.
+To gather domain knowledge about factors affected the consumption of products, we contacted experts from EPSS who work in the distribution unit.
+Next, we collaborated with six experts, each with over 10 years of experience managing pharmaceutical distribution.
 In these sessions, we reviewed the time series of all products together to understand the potential drivers behind these anomalies.
@@
 
-- Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters.
+<!-- - Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters.
 Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises.
@@
 The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.
+-->
 
@@
 ARIMAX (AutoRegressive Integrated Moving Average with eXogenous variables) extends ARIMA by incorporating external variables, or exogenous predictors, into the model.
-This modification allows to include relevant information from external factors such as malaria season, and fiscal year period that may explain variations in the series beyond its internal time dynamics, we refer to this in the result as _arima_reg_.
+This modification allows to include relevant information from external factors such as malaria season, and fiscal year period, and stock replenishement period that may explain variations in the series beyond its internal time dynamics, we refer to this in the result as _arima_reg_.
 By adding these predictors, ARIMAX combines the strengths of ARIMA’s time-series structure with the flexibility of regression models.
@@
 We use TimeGpt models both with and without predictors, referring to the model with predictors as _timegpt_reg_.
+
+In all models, probabilistic forecasts were generated using bootstrapping to create 1,000 possible future scenarios for each period within the forecast horizon [@hyndman2021forecasting].
 
@@
 
-To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@gneiting2014probabilistic;@hyndman2021forecasting].
+To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@hyndman2021forecasting].
 It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution.
@@
 
-CRPS can be considered an average of all possible qiantiles [@hyndman2021forecasting, Section 5.9], and thus provides an evaluation of all possible prediction intervals or quantiles.
+CRPS can be considered an average of all possible quantiles [@hyndman2021forecasting, Section 5.9], and thus provides an evaluation of all possible prediction intervals or quantiles.
 A specific prediction interval could be evaluated using a Winkler score, if required.

### 2024-12-13 manuscript/04-results.qmd
final draft for submission

@@
 In this section, we compare the forecasting performance of various approaches, examining models that incorporate expert-identified business context predictors versus those that rely solely on historical consumption data.
-Point forecast performance is reported using MASE and MSSE, while probabilistic forecast accuracy is reported using CRPS.
+Point forecast performance is reported using MASE and RMSSE, while probabilistic forecast accuracy is reported using CRPS.
 
 The forecasting performance is reported in @fig-point, in which the average forecast accuracy over forecast horizon and across all products is calculated for each origin.
-we also report the ditribution of accurcy metrics  across all rollog origins, average across forecast horizon and all products.
-this shows how mdoels varies in roviding accuracy across different origins.
-further rserach would be required to understand how error analysis across forecast hrizon and across differne origins could be used as afiagnos!
-y-axis shows the models sorted based on ther values of mase and rmsse, when the one i the bottom has lowest error. this modesis TimeGpt, which is the foundational time sei models.
-
-@fig-point also indicate that predictors collected through interacting with experts with domaine knoeldge improves point forecast accuracy for all models, highlighting the essential needs to systematically colelct such information alognside the tarsactional data about consuption.
-
-We observed that the LSTM model had the highest error compared to all other models.
-While LSTMs are powerful for capturing complex, non-linear temporal patterns, they may underperform in certain scenarios, especially when compared to simpler time series forecasting models.
-This underperformance can be attributed to overfitting, as LSTMs have a large number of parameters and require substantial amounts of data to generalize effectively.
-In contrast, simpler models like ARIMA with exogenous variables often handle straightforward trends or seasonality more effectively and may outperform LSTMs when long-term dependencies or non-linearity in the data are not significant.
+We report the distribution of accuracy metrics across all rolling origins.
+This shows how models varies in providing accuracy across different origins.
+The y-axis displays models sorted by their MASE and RMSSE values, with the model exhibiting the lowest error positioned at the bottom.
+This model is the Ariam model, which incorporates exogenous variables.
+Additionally, @fig-point indicates that predictors obtained through interactions with domain experts enhance point forecast accuracy across most models.
+This underscores the critical importance of systematically collecting such expert-informed data alongside transactional consumption data.
 
@@
 
-```{r}
-#| label: fig-horizon
-#| fig-width: 11
-#| fig-height: 5
-#| fig-cap: "Average accuracy by month for 6 month. The total number of months used to calculate the accuracy in the test set is 12 for each product."
-#| fig-pos: "H"
+<!-- ```{r} -->
+<!-- #| label: fig-horizon -->
+<!-- #| fig-width: 11 -->
+<!-- #| fig-height: 5 -->
+<!-- #| fig-cap: "Average accuracy by month for 6 month. The total number of months used to calculate the accuracy in the test set is 12 for each product." -->
+<!-- #| fig-pos: "H" -->
 
-p1h <- point_accuracy_h_all |> group_by(model,h) |> summarise(mase=mean(mase) , .groups = "drop")|> filter(model != "lstm", model != "lstm_reg") |>  ggplot(aes(x = h, y = mase, colour =model))+
-  geom_point()+
-  geom_line()+
-  ggthemes::theme_few()+
-  ggthemes::scale_color_colorblind()+
-  labs(x= " Forecast horizon")+
-  theme(legend.position="none")
+<!-- p1h <- point_accuracy_h_all |> group_by(model,h) |> summarise(mase=mean(mase) , .groups = "drop")|> filter(model != "lstm", model != "lstm_reg") |>  ggplot(aes(x = h, y = mase, colour =model))+ -->
+<!--   geom_point()+ -->
+<!--   geom_line()+ -->
+<!--   ggthemes::theme_few()+ -->
+<!--   ggthemes::scale_color_colorblind()+ -->
+<!--   labs(x= " Forecast horizon")+ -->
+<!--   theme(legend.position="none") -->
 
-p2h <- point_accuracy_h_all |> group_by(model,h) |> summarise(rmsse=mean(rmsse),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |> 
-  ggplot(aes(x = h, y = rmsse, colour =model))+
-  geom_point()+
-  geom_line()+
-  ggthemes::theme_few()+
-  ggthemes::scale_color_colorblind()+
-  labs(x= " Forecast horizon")+
-  theme(legend.position="bottom")
+<!-- p2h <- point_accuracy_h_all |> group_by(model,h) |> summarise(rmsse=mean(rmsse),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |>  -->
+<!--   ggplot(aes(x = h, y = rmsse, colour =model))+ -->
+<!--   geom_point()+ -->
+<!--   geom_line()+ -->
+<!--   ggthemes::theme_few()+ -->
+<!--   ggthemes::scale_color_colorblind()+ -->
+<!--   labs(x= " Forecast horizon")+ -->
+<!--   theme(legend.position="bottom") -->
 
-p3h <- prob_accuracy_h_all |> group_by(model,h) |> summarise(crps=mean(crps),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |> 
-  ggplot(aes(x = h, y = crps, colour =model))+
-  geom_point()+
-  geom_line()+
-  ggthemes::theme_few()+
-  ggthemes::scale_color_colorblind()+
-  labs(x= " Forecast horizon")+
-  theme(legend.position="none")
+<!-- p3h <- prob_accuracy_h_all |> group_by(model,h) |> summarise(crps=mean(crps),.groups = "drop") |> filter(model != "lstm", model != "lstm_reg") |>  -->
+<!--   ggplot(aes(x = h, y = crps, colour =model))+ -->
+<!--   geom_point()+ -->
+<!--   geom_line()+ -->
+<!--   ggthemes::theme_few()+ -->
+<!--   ggthemes::scale_color_colorblind()+ -->
+<!--   labs(x= " Forecast horizon")+ -->
+<!--   theme(legend.position="none")  -->
 
-prob_patch <- (p1h|p2h|p3h)
-prob_patch
-```
+<!-- prob_patch <- (p1h|p2h|p3h) -->
+<!-- prob_patch -->
+<!-- ``` -->
 
-@fig-prob illustrates the forecast distribution accuracy measured by the Continuous Ranked Probability Score (CRPS), which evaluates both forecasting calibration and interval sharpness.
+@fig-prob illustrates the forecast distribution accuracy measured by CRPS, which evaluates both forecasting calibration and interval sharpness.
 A smaller CRPS value indicates better overall performance.
 We observed that incorporating domain knowledge improved forecast accuracy for most models, enhancing not only point forecasts but also probabilistic forecasts.
-Notably, TimeGpt achieved the highest probabilistic forecast accuracy, while LSTM models had the lowest.
+Notably, ARIMA achieved the highest probabilistic forecast accuracy, while LSTM models had the lowest.
 This aligns with the findings related to point forecast accuracy, reinforcing the earlier explanations for these results.
 
-In addition to reporting the forecast accuracy presented across different rolling origins, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-horizon.
+We observe that LSTM models exhibit a widespread error distribution compared to other models.
+This is attributed to their performance variability, which can significantly depend on the characteristics of demand patterns across different products.
+For example, the product "Amlodipine - 5mg - Tablet" demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero demand.
+Such patterns align well with the strengths of LSTM models, which excel at capturing long-term dependencies and managing complex temporal fluctuations.
+In contrast, the demand for "Anti-Rho (D)" is erratic and sporadic, characterized by frequent, random fluctuations without any discernible long-term trends or seasonal patterns.
+This chaotic nature, combined with inconsistent and often low volumes, poses significant challenges for LSTM models.
+The lack of clear structure or stable patterns limits the models' ability to generalize effectively, resulting in poorer forecast performance.
+Although LSTM models can handle irregular data to some extent, they require sufficient and well-distributed training samples to achieve optimal performance.
+In our study, the majority of products exhibit erratic and sporadic demand patterns similar to "Anti-Rho (D)", resulting in this widespread error distribution.
+
+<!-- In addition to reporting the forecast accuracy presented across different rolling origins, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-horizon.
 this is the accuracy reorted as the average across rolling origins and all products fro each given horizon.
 For both the point forecast and distributional accuracy  we can see that models incoporating information collected by domain experts improves compared to counterparts that that uses only the historical comsuptions.
+-->
 
-Overall, our results indicate that forecast produces using models with infromation colelcted via the domain expert engagement provides reliable forecasts and improves upon the models relying solely on historcial consuption.
+<!-- Overall, our results indicate that forecast produces using models with infromation colelcted via the domain expert engagement provides reliable forecasts and improves upon the models relying solely on historcial consuption.
+-->
 
-Our analysis highlights the value of using foundational time series forecasting models like TimeGpt, particularly in developing countries and the global health and humanitarian sectors.
-In these contexts, the lack of advanced analytical skills and systematic data collection can hinder effective forecasting and decision-making.
-
-The findings also emphasize the importance of incorporating relevant domain knowledge into forecasting models.
+The findings emphasize the importance of incorporating relevant domain knowledge into forecasting models.
 In pharmaceutical supply logistics administration systems, data often consists solely of transactional records on consumption and distribution.
@@
 Establishing robust data collection systems is thus a critical step for strengthening forecasting capabilities and operational resilience.
+
+Our analysis also highlights the value of using foundational time series forecasting models like TimeGpt, particularly in developing countries and the global health and humanitarian sectors.
+In these contexts, the lack of advanced analytical skills and systematic data collection can hinder effective forecasting and decision-making.
 
@@
 
-fcst <- fit |> forecast(h = 6)
+fcst <- fit |> forecast(h = 6) |> mutate(quantity = distributional::dist_truncated(quantity, lower = 0), .mean=mean(quantity))
+
 fitted_ets <- fit |> augment()

### 2024-12-13 manuscript/05-conclusion.qmd
final draft for submission

@@
 
-In this study, we conducted an extensive analysis of pharmaceutical demand forecasting within the Ethiopian Pharmaceutical Supply Service (EPSS).
+In this study, we conducted an extensive analysis of pharmaceutical demand forecasting within the EPSS.
 Using five years of consumption data for 33 key pharmaceutical products, along with additional information gathered through collaboration with domain experts at EPSS, our goal is to contribute to enhancing the demand forecasting process and emphasize the importance of integrating domain knowledge into model building.
@@
 
+- Large Language Models (LLMs) for time series forecasting, like the one included in this study, offer transformative opportunities for developing countries by making advanced forecasting methods accessible, even in contexts with limited expertise or infrastructure.
+However, future research should explore their applicability, advantages, limitations, and performance relative to traditional and hybrid forecasting methods.
+
 - The use of distribution-based representation of error metrics across different rolling origins or forecast horizons, such as boxplots, to showcase forecast accuracy provides essential insights into variability in accuracy at different point of time in data.

### 2024-12-13 manuscript/main.qmd
final draft for submission

@@
 title: "The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting"
-author:
-  - name: Arebu Issa Bilal
-    email: arebu.issa@aau.edu.et
-    affiliations:
-        - id: Addis 
-          name: Addis Ababa University
-          department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
-          country: Ethiopia.
-    attributes:
-        corresponding: true
-  - name: Bahman Rostami-Tabar
-    email: rostami-tabarb@cardiff.ac.uk
-    affiliations:
-        - id: carbs
-          name: Cardiff University
-          department: Data Lab for Social Good Reserach Group, Cardiff Business School 
-          country: UK.
-  - name: Harsha Halgamuwe Hewage
-    email: HalgamuweHewageHR@cardiff.ac.uk
-    affiliations:
-        - ref: carbs
-  - name: Umit Sezer Bititci 
-    email: U.S.Bititci@hw.ac.uk
-    affiliations:
-        - id: herio 
-          name: Heriot-Watt University
-          department: Edinburgh Business School, Heriot-Watt University
-          country: UK.
-  - name: Teferi Gedif Fenta
-    email: tgedif@gmail.com
-    affiliations:
-        - ref: Addis
+# author:
+#   - name: Arebu Issa Bilal
+#     email: arebu.issa@aau.edu.et
+#     affiliations:
+#         - id: Addis 
+#           name: Addis Ababa University
+#           department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
+#           country: Ethiopia.
+#     attributes:
+#         corresponding: true
+#   - name: Bahman Rostami-Tabar
+#     email: rostami-tabarb@cardiff.ac.uk
+#     affiliations:
+#         - id: carbs
+#           name: Cardiff University
+#           department: Data Lab for Social Good Reserach Group, Cardiff Business School 
+#           country: UK.
+#   - name: Harsha Halgamuwe Hewage
+#     email: HalgamuweHewageHR@cardiff.ac.uk
+#     affiliations:
+#         - ref: carbs
+#   - name: Umit Sezer Bititci 
+#     email: U.S.Bititci@hw.ac.uk
+#     affiliations:
+#         - id: herio 
+#           name: Heriot-Watt University
+#           department: Edinburgh Business School, Heriot-Watt University
+#           country: UK.
+#   - name: Teferi Gedif Fenta
+#     email: tgedif@gmail.com
+#     affiliations:
+#         - ref: Addis
 
@@
   Despite advancements in forecasting, significant gaps remain in effective forecasting in poor resource countries.
-Most existing literature on pharmaceutical supply chain forecasting focuses primarily on point estimation of consumption while neglecting the inherent uncertainty of these forecasts Additionally, these studies often rely solely on historical consumption data without considering the broader context that influences consumption Furthermore, many works lack rigorous methodological design, transparency, and reproducibility.
+Most existing literature on pharmaceutical supply chain forecasting focuses primarily on point estimation of consumption while neglecting the inherent uncertainty of these forecasts.
+Additionally, these studies often rely solely on historical consumption data without considering the broader context that influences consumption.
+Furthermore, many works lack rigorous methodological design, transparency, and reproducibility.
 This study addresses these gaps by integrating domain-specific knowledge, gathered through expert interviews and engagement with key members of the Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models.
-Using a dataset spanning five years (December 2017 to July 2022) from EPSS, we developed forecasting models that integrate expert-identified variables such as stock replenishment schedules, fiscal inventory counts, conflicts, and disease outbreaks.
-Evaluation metrics such as Mean Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE), and Continuous Ranked Probability Score (CRPS) are used to report forecast accuracy.
+Using a dataset spanning five years (December 2017 to July 2022) from EPSS, we developed forecasting models that integrate expert-identified variables such as stock replenishment schedules, fiscal inventory counts, and disease outbreaks.
+Evaluation metrics including Mean Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE), and Continuous Ranked Probability Score (CRPS) are used to report forecast accuracy.
 The findings underscore the significance of contextual data in developing robust forecasting models that are adaptable to complex, real-world conditions.
-Our results also highlight the effectiveness of foundational time series methods, such as LLM models, for forecasting.
-These methods are particularly appealing for resource-constrained countries that may lack advanced analytical expertise.
+Our results also highlight the effectiveness of foundational time series models for forecasting.
+These models are particularly appealing for resource-constrained countries that may lack advanced analytical expertise.
 To promote usability, generalizability, and reproducibility, we share the complete dataset and code in R and Python, and the entire paper is written in Quarto via a GitHub repository to facilitate these practices.
@@
     journal:
-      name: Journal of 
+      name: International Journal of Production Research
       formatting: preprint

### 2024-12-13 manuscript/main.tex
final draft for submission

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
 \fi
@@
 \setlength{\parindent}{0cm}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\journal{Journal of}
+\journal{International Journal of Production Research}
 \ifLuaTeX
@@
 \bibliographystyle{elsarticle-harv}
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \hypersetup{
-  pdftitle={Pharmaceutical Supply Chain Forecasting by using Machine Learning in Ethiopian Pharmaceutical Supply Services},
-  pdfauthor={Arebu Issa Bilal; Arebu Issa Bilal; Bahman Rostami-Tabar; Umit Sezer Bititci; Teferi Gedif Fenta},
-  pdfkeywords={Pharmaceutical Demand Forecasting, Ethiopian
-Pharmaceutical Supply Services, ARIMA Model, Forecasting
-Variables, Forecast Accuracy},
+  pdftitle={The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting},
+  pdfkeywords={Forecasting, Pharmaceutical supply chain, Domain
+knowlege, Forecast accuracy, Developing countries},
   colorlinks=true,
@@
 \begin{frontmatter}
-\title{Pharmaceutical Supply Chain Forecasting by using Machine Learning
-in Ethiopian Pharmaceutical Supply Services}
-\author[1]{Arebu Issa Bilal%
-\corref{cor1}%
-}
- \ead{arebu.issa@aau.edu.et} 
-\author[]{Arebu Issa Bilal%
-%
-}
- \ead{arebu.issa@aau.edu.et} 
-\author[2]{Bahman Rostami-Tabar%
-%
-}
- \ead{rostami-tabarb@cardiff.ac.uk} 
-\author[3]{Umit Sezer Bititci%
-%
-}
- \ead{U.S.Bititci@hw.ac.uk} 
-\author[1]{Teferi Gedif Fenta%
-%
-}
- \ead{tgedif@gmail.com}
-
-\affiliation[1]{organization={Addis Ababa University, Department of
-Pharmaceutics and Social Pharmacy at School of
-Pharmacy},country={Ethiopia},countrysep={,},postcode={9086},postcodesep={}}
-\affiliation[2]{organization={University, Cardiff Business
-School},country={UK},countrysep={,},postcode={CF5},postcodesep={}}
-\affiliation[3]{organization={University, Edinburgh Business
-School},country={UK},countrysep={,},postcode={EH14 4AS},postcodesep={}}
+\title{The missing puzzle piece: Contextual insights for enhanced
+pharmaceutical supply chain forecasting}
 
@@
 \begin{abstract}
-This study delves into the realm of pharmaceutical demand forecasting
-within the Ethiopian Pharmaceuticals Supply Services(EPSS) context. It
-encompasses a comprehensive analysis of 37 key pharmaceutical
-products,assessing their forecast accuracy across variation models. The
-initial assessment reveals that the Autoregressive Integrated Moving
-Average (ARIMA) model emerges as a frotrunner, boasting a Root Mean
-Squared Error (RMSSE) value of 0.850. ARIMA's prowess in capturing
-complex demand patterns suggests its suitability for pharmaceutical
-sales forecasting in the EPSS landscape. To argument forecasting
-accuarcy, predictor varaibles such as physical count, budget release and
-closure by the goverment, the impact of COVID-19, and promotional
-campaigns were incorpoarted. Remarkably, the ARIMA model mantainsits
-superiority even in the presences of these external factors, reiterating
-its roustness. Delving further into individual pharmaceutical products,
-ARIMA with predictors demonstrates exceptional performance for a
-substantial portion (32.43\%) of the products, including Amoxicillin
-500mg capsule, Anti RhD, Artehmether + Lumefantrine, and others.
-Conversely,alternative models like naive, mean, and regression methods
-showcase strengths for different pharmaceutical categories, underscoring
-the importance of tailoring forecasting approaches to product-specific
-dynamics. While this study significantly advances the discourse on
-pharmaceutical demand forecasting, limitations stemming from data
-quality and external variables underscore the intricacies of the
-forecasting landscape. In conclusion, it advocates the integration of
-the ARIMA model with predictor variables as a strategic approach for
-refining pharmaceutical demand forecasting within EPSS. It also
-encourages further exploration of advanced forecasting techniques and
-ongoing refinement to bolster supply chain efficiency and ensure
-consistent pharmaceutical availability. Capacity building through staff
-training and replication of the study in diverse healthcare settings are
-recommended steps toward resilient and efficient pharmaceutical supply
-chains.These findings hold promise for fostering supply chains that are
-adaptable, responsive, and capable of ensuring the continuous
-availability of essential pharmaceuticals in Ethiopia.
+Accurate forecasting in pharmaceutical supply chains is critical for
+ensuring the continuous availability of essential medicines,
+particularly in developing countries where resource constraints and
+logistical challenges are more pronounced. Effective demand forecasting
+supports procurement and inventory management, which in turn could help
+to prevent stockouts, reduce wastage due to overstocking, and enhance
+overall healthcare delivery. In such settings, reliable forecasts, that
+acknowledge uncertainties, are essential to ensure that limited
+resources are used efficiently to meet health needs. Despite
+advancements in forecasting, significant gaps remain in effective
+forecasting in poor resource countries. Most existing literature on
+pharmaceutical supply chain forecasting focuses primarily on point
+estimation of consumption while neglecting the inherent uncertainty of
+these forecasts. Additionally, these studies often rely solely on
+historical consumption data without considering the broader context that
+influences consumption. Furthermore, many works lack rigorous
+methodological design, transparency, and reproducibility. This study
+addresses these gaps by integrating domain-specific knowledge, gathered
+through expert interviews and engagement with key members of the
+Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models.
+Using a dataset spanning five years (December 2017 to July 2022) from
+EPSS, we developed forecasting models that integrate expert-identified
+variables such as stock replenishment schedules, fiscal inventory
+counts, and disease outbreaks. Evaluation metrics including Mean
+Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE),
+and Continuous Ranked Probability Score (CRPS) are used to report
+forecast accuracy. The findings underscore the significance of
+contextual data in developing robust forecasting models that are
+adaptable to complex, real-world conditions. Our results also highlight
+the effectiveness of foundational time series models for forecasting.
+These models are particularly appealing for resource-constrained
+countries that may lack advanced analytical expertise. To promote
+usability, generalizability, and reproducibility, we share the complete
+dataset and code in R and Python, and the entire paper is written in
+Quarto via a GitHub repository to facilitate these practices.
 \end{abstract}
@@
 \begin{keyword}
-    Pharmaceutical Demand Forecasting \sep Ethiopian Pharmaceutical
-Supply Services \sep ARIMA Model \sep Forecasting Variables \sep 
-    Forecast Accuracy
+    Forecasting \sep Pharmaceutical supply chain \sep Domain
+knowlege \sep Forecast accuracy \sep 
+    Developing countries
 \end{keyword}
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
-
-\hypertarget{sec-intro}{%
-\section{Introduction}\label{sec-intro}}
-
-Universal access essential medicines is a cornerstone of effective
-healthcare systems, yet ensuring equitable availability of these vital
-medications remains an urgent global
-challenge\citep{quick2003essential, world2004annual}. Across continents,
-especially in Africa and Asia, many nations grapple with restricted
-access due to factors like unaffordability and suboptimal supply chain
-management \citep{world2004medicines}. A survey conducted across eight
-sub-Saharan African countries indicated that there were unacceptably low
-availability of essential medicines.The mean availability of 12 priority
-essential medicines for women ranged from 22\% to 40\% and from 28\% to
-57\% for children \citep{droti2019poor}. However, the issue of drug
-shortages extends beyond developing countries, with instances reported
-in the United States and Europe, adversely impacting healthcare systems
-and compromising patient care quality
-\citetext{\citealp[\citet{kaakeh2011impact}]{fox2003managing}; \citealp{johnson2011drug}; \citealp{huys2013european}; \citealp{le2011prevalence}}.
-
-The consequences of drug shortages are felt both financially within the
-healthcare sector and in terms of the quality of patient care
-\citep{alspach2012drug, kaakeh2011impact, kaakeh2011impact, baumer2004national}.The
-consequences extend across patients, providers, healthcare institutions,
-and research programs. Patients face delays or cancellations of
-treatments, substandard care due to the unavailability of optimal
-therapies, and increased costs in a secondary ``grey market.''
-Compromised patient safety, complications, prolonged hospital stays, and
-even deaths have been reported {[}\citet{alspach2012drug};{]}.Addressing
-these shortages is crucial to preserving the quality of patient care and
-ensuring the effectiveness of healthcare delivery. Thus, a smoothly
-operating pharmaceutical supply chain is essential to ensure timely
-access to essential medicines, making accurate demand forecasting an
-indispensable component.
-
-An efficient pharmaceutical supply chain is vital for timely access to
-essential health commodities. One of the most crucial components of such
-a supply chain is a well-defined demand forecasting process. Forecasting
-accurately the future demand for essential medicines provides valuable
-insights for anticipating future healthcare requirements including
-procurement and replenishment of
-inventory\citep{subramanian2021effective}. Despite the significance of
-forecasting for supply chain management, pharmaceutical product
-forecasting remains a complex endeavor influenced by a multitude of
-factors. Key considerations include the availability of data, the
-specific position within the supply chain, and the nuanced understanding
-of contributing elements. The position within the supply chain
-introduces unique challenges and dynamics, requiring tailored
-forecasting approaches. Additionally, a nuanced understanding of
-contributing elements involves grasping intricate factors that impact
-forecasting accuracy, such as market trends, regulatory changes, and
-external influences. Schneider, Wilson, and Rosenbeck (2010) and Hyndman
-and Athanasopoulos (2018) highlight the intricate nature of
-pharmaceutical forecasting, emphasizing the need for a comprehensive
-approach that incorporates these multifaceted aspects to enhance supply
-chain management
-effectiveness\citep{schneider2010pharmaceutical, hyndman2018forecasting}.
-
-The Ethiopian pharmaceutical sector's distinct characteristics, marked
-by a diverse array of medicines and unique data challenges stemming from
-service delivery points, staff capacity limitations, communication
-hurdles, prolonged procurement and distribution lead times, poor
-forecast accuracy, and policy intricacies, necessitates tailored
-forecasting strategies that cater to the nation's specific context
-\citep{boche2022procurement}. In this context, a comprehensive
-exploration of effective forecasting models that align with Ethiopia's
-pharmaceutical landscape is paramount to mitigating stock outs and
-enhancing healthcare delivery. The implications of pharmaceutical demand
-forecasting extend beyond financial considerations to societal
-well-being \citep{rostami2022forecasting}. In Ethiopia, a country
-undergoing demographic and healthcare evolution, precise demand
-forecasting assumes immense importance for resource allocation, supply
-chain optimization, and policy formulation. With mounting interest in
-refining demand forecasting in Ethiopia's pharmaceutical sector, this
-study seeks to contribute to existing knowledge by investigating and
-recommending proficient demand forecasting methods. Our investigation
-centers on the application of mean, naive, exponential smoothing,
-regression, and ARIMA models to bolster forecast accuracy and
-effectiveness. This paper aims to forecast the demand for selected
-pharmaceuticals in EPSS over the next 12 months by utilizing various
-forecasting models and integrating predictors. The integration of
-predictors with the models will be examined to determine whether they
-enhance the forecasting accuracy. These predictors are derived from a
-historical analysis of events in Ethiopia over the last 5 years that
-have impacted pharmaceutical sales in EPSS. These events include
-COVID-19, conflicts in the country, disease campaigns, budget closures,
-budget releases, and physical inventory, all of which, in varying
-degrees, influence the demand for pharmaceuticals in Ethiopia.
-
-The current paper offers the following contributions we aim at
-forecasting pharmaceutical sales over a 12-month period these models aim
-to enhance the accuracy and reliability of demand predictions within the
-context of EPSS. Moreover an essential contribution lies in the
-identification of key factors influencing pharmaceutical sales. The
-study develops a model that effectively incorporates these driving
-factors, providing a more nuanced and comprehensive approach to demand
-prediction. We generate forecasts and evaluate model performance using
-the Root Mean Squared Error (RMSE), Root Mean Squared Scaled Error
-(RMSSE) and Mean Absolute Error (MAE), Mean Absolute Scaled Error
-(MASE), Crips and Winkler are used to assess probabilistic forecast
-accuracy. \#how you know the forecasting accutacy problem and provide
-eveidence about the forecasting problems \# most forecast focus on point
-forecast and we want to focus on probablistic forecast \# i will write
-the contribution of the reearch rather than these objectives
-
-This research is guided by several objectives:
+
+\section{Introduction}\label{sec-intro}
+
+Universal access to essential medicines is a fundamental goal of
+effective healthcare systems, yet ensuring equitable availability
+continues to be a critical global challenge
+\citep{quick2003essential, world2004annual}. Many countries,
+particularly in Africa and Asia, face significant barriers to accessing
+these medications, often due to high costs and inefficient supply chain
+management \citep{world2004medicines}. A survey conducted in eight
+sub-Saharan African countries revealed alarmingly low availability of
+essential medicines. The average availability of women's priority
+medicines ranged from just 22\% to 40\%, while children's medicines were
+only slightly more accessible, with availability ranging from 28\% to
+57\% \citep{droti2019poor}. However, drug shortages are not limited to
+low-income regions; they also occur in developed areas like the United
+States and Europe, where they disrupt healthcare delivery and compromise
+the quality of patient care
+\citep{fox2003managing, kaakeh2011impact, johnson2011drug, huys2013european, le2011prevalence}.
+
+The consequences of drug shortages extend beyond immediate healthcare
+delivery, impacting healthcare costs, treatment quality, and patient
+safety \citep{alspach2012drug, kaakeh2011impact, baumer2004national}.
+Such shortages can result in treatment delays, suboptimal care due to
+the unavailability of preferred therapies, increased reliance on costly
+secondary markets, and serious patient outcomes, including
+complications, longer hospital stays, and even deaths
+\citep{alspach2012drug}. Addressing these challenges requires a
+resilient pharmaceutical supply chain that can ensure the timely
+availability of medicines. In this context, accurate demand forecasting
+becomes critical for effective procurement and inventory management
+\citep{subramanian2021effective}.
+
+Demand forecasting in pharmaceutical supply chains is a highly complex
+process, shaped by several factors. These include the quality of
+available data, the specific position within the supply chain, and
+external influences such as market trends, regulatory changes, and
+disruptive events such as pandemics, conflicts, flooding
+\citep{schneider2010pharmaceutical, hyndman2018forecasting}. Recently,
+major donors such as Bill \& Melinda Gates Foundation have invested in
+data collection technologies in sub-Saharan Africa, recognizing the
+potential of data to provide valuable insights. This investment is
+essential and offers significant benefits for improving supply chain
+delivery. However, much of the data being collected focuses on
+transactional data, such as consumption and sale. Almost all platforms
+collecting data tend to overlook the critical contextual factors and
+events that influence these transactions---factors that are just as
+crucial for accurate forecasting. This presents a major barrier, as
+understanding the broader context, including administration procedures,
+local policies, disruptions, conflicts, and so on is vital for building
+reliable forecasting models. This gap---known as business context and
+domain knowledge---requires urgent attention in Pharmaceutical supply
+chain forecasting. Without incorporating this knowledge, any forecasting
+model is unlikely to produce reliable results. We must prioritize
+collecting and integrating this contextual information to improve the
+accuracy of demand forecasts in pharmaceutical supply chains. Combining
+statistical models with domain insights offers a more refined
+forecasting process, as evidenced by research that highlights the value
+of judgmental adjustments to enhance forecast accuracy
+\citep{fildes2007against, taylor2022artificial, soyiri2013overview}.
+Domain knowledge enables analysts to incorporate relevant events, such
+as supply chain disruptions, into the forecasting model. This
+integration is particularly beneficial in fields like healthcare, where
+expert insights can align algorithms with real-world applications
+\citep{dash2022review}.
+
+This study aims to bridge that gap by employing advanced machine
+learning models combined with domain expertise to address the
+distinctive characteristics of Ethiopia's pharmaceutical sector. The
+challenges in this sector, such as diverse product categories, limited
+data from service delivery points, communication issues, long lead
+times, forecast inaccuracies, and complex policies, necessitate tailored
+forecasting approaches
+\citep{bilal2024challenges, boche2022procurement}. Accurate demand
+forecasting in Ethiopia is essential for optimizing resource allocation,
+streamlining supply chain operations, and shaping effective healthcare
+policies \citep{rostami2022forecasting}.
+
+This research uses consumption data from 33 health commodities, covering
+the period from December 2017 to July 2022, provided by the Ethiopian
+Pharmaceutical Supply Service (EPSS), to develop models for
+six-month-ahead forecasting. Initially, we implement univariate models
+that rely solely on consumption data. Following this, we conduct
+exploratory data analysis and engage in interviews with EPSS experts to
+gather domain knowledge on events influencing consumption patterns.
+Through these interviews, the experts identified three key factors
+collectively impacting consumption: stock replenishment period, fiscal
+calendar, and the seasonal malaria. We then develop forecasting models
+that incorporate the data gathered through the interview process and
+compare their performance with univariate models that rely solely on
+consumption data.
+
+Our primary aims is to assess whether incorporating domain knowledge
+improves forecast accuracy. Moreover, we also fill a gap in the
+pharmaceutical demand forecasting literature by not only generating
+point forecasts but also producing probabilistic forecasting techniques
+that capture the range of potential outcomes with associated
+probabilities, offering a more comprehensive view of demand uncertainty
+than traditional methods. Model performance is evaluated using point
+accuracy measure including MASE, RMSSE, and probabilistic measure using
+CRPS.
+
+The remainder of this paper is structured as follows: Section 2 reviews
+the literature and identifies gaps to position our work; Section 3
+outlines the experimental design, including data sources, forecasting
+methods, and evaluation criteria; Section 4 presents the results and
+analysis; and Section 5 concludes with a summary of findings and future
+research directions.
+
+\section{Research background}\label{sec-lit}
+
+Accurate sales forecasting is crucial in the pharmaceutical industry,
+where it directly impacts profit maximization, cost minimization, and
+the ability to respond to market changes. Forecasting in healthcare not
+only influences clinical decisions but also plays a pivotal role in
+supply chain management, ensuring that the right drugs are available
+when needed. The challenge lies in balancing customer demand with
+inventory costs, which is particularly complex in pharmaceuticals due to
+factors like short shelf life and quality constraints
+\citep{gupta2000mid, makridakis2020forecasting}.
+
+In this review, a range of forecasting methods has been applied, from
+traditional statistical models to advanced machine learning techniques,
+with many studies using a combination of simple and advanced methods to
+enhance accuracy
+\citep{nikolopoulos2016forecasting, zhu2021demand, anusha2014demand}.
+Some studies, such as those by \citet{newberne2006holt}, and
+\citet{restyana2021analysis}, applied traditional forecasting methods
+like the Holt-Winters method, Single Moving Average (SMA), and Single
+Exponential Smoothing (SES), which are simpler but effective for certain
+applications. In contrast, other researchers explored more advanced
+approaches, including hybrid models and machine learning techniques, to
+address complex forecasting challenges, as seen in the studies by
+\citet{siddiqui2022hybrid}, \citet{de2021lead}, \citet{kim2015demand},
+\citet{candan2014demand}, and \citet{ribeiro2017improving}.
+
+Early research in pharmaceutical forecasting primarily focused on
+traditional statistical methods. These methods, including ARIMA,
+exponential smoothing, and moving averages, were widely used due to
+their simplicity and effectiveness in relatively stable environments
+\citep{zahra2019forecasting}. These models typically assess forecast
+accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean
+Squared Error (MSE), and Root Mean Squared Error (RMSE). For example,
+\citet{newberne2006holt} applied the Holt-Winters method to forecast
+healthcare data, focusing on prescription trends. Their findings
+highlighted the method's utility in short-term planning and resource
+management, validated through metrics like MAPE, MAD, and MSD.
+Similarly, \citet{restyana2021analysis} compared Single Moving Average
+(SMA) and Single Exponential Smoothing (SES) in drug demand forecasting,
+concluding that SES provided more accurate results, as indicated by
+lower MAD and MSE values.
+
+Recognizing the limitations of traditional models, particularly in
+capturing the complexities of pharmaceutical demand (e.g., seasonality,
+external influences), researchers began developing hybrid models. These
+models combine linear methods like ARIMA with nonlinear techniques such
+as neural networks, aiming to improve forecasting accuracy by addressing
+diverse data patterns. \citet{khalil2014intelligent} introduced a hybrid
+model that integrated ARIMA with neural networks. Their study
+demonstrated that this approach significantly outperformed traditional
+models, especially in scenarios with limited historical data. The hybrid
+model's effectiveness was confirmed by improvements in key performance
+indicators like MAPE and RMSE.
+
+The pharmaceutical industry's increasing complexity and the availability
+of large datasets have led to the adoption of advanced machine learning
+techniques. These methods, such as Support Vector Regression (SVR),
+Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer
+greater flexibility and predictive power. \citet{van2021using} showed
+that incorporating downstream information into machine learning models,
+like LASSO and SVR, greatly enhances forecast accuracy. Their study on
+multi-echelon supply chains revealed that advanced models, particularly
+those integrating external variables, consistently produced lower
+forecast errors, as measured by AvgRelRMSE. Similarly,
+\citet{rathipriya2023demand} compared shallow neural networks and deep
+learning models for drug demand forecasting. The study found that
+shallow models outperformed deep learning models for most drug
+categories, while ARIMA was more effective for the remaining categories.
+This suggests that no single model is universally optimal, and model
+selection should be context-specific.
+
+More sophisticated models, such as Neuro-Fuzzy Systems, have emerged as
+powerful tools for pharmaceutical forecasting. These systems combine the
+learning capabilities of neural networks with the reasoning abilities of
+fuzzy logic, offering a balanced approach that incorporates both
+empirical data and expert knowledge. Deep learning models, while
+powerful, present challenges such as the need for large datasets and
+substantial computational resources. Despite these challenges, models
+like LSTM have shown potential in capturing long-term dependencies in
+time series data, as demonstrated by \citet{sousa2019statistical} in
+their study on drug distribution in the Brazilian Public Health System.
+\citet{candan2014demand} employed an Adaptive Neuro-Fuzzy Inference
+System (ANFIS) to forecast pharmaceutical demand. Their study
+highlighted the system's ability to capture complex patterns in the
+data, resulting in highly accurate forecasts. The effectiveness of this
+approach was further validated through statistical tests, such as paired
+T-tests and mean difference analysis, which confirmed its superiority
+over traditional methods.
+
+\citet{nguyen2023managing} explore how sentiment analysis of news media
+can be utilized to improve demand forecasting for pharmaceutical
+products during disruptive events, such as pandemics and scandals. The
+authors employed a VARX (Vector Autoregressive with Exogenous Variables)
+model to forecast demand volatility. \citet{papanagnou2018coping}
+investigate the impact of big data analytics, specifically through text
+mining, on improving forecasting accuracy for pharmaceuticals in retail
+pharmacies, using HMX pharmacy stores in Nigeria as a case study. They
+use multivariate time series analysis technique known as VARX, which
+incorporates customer-generated content from sources like Google,
+YouTube, and online newspapers to predict demand for medicines.
+\citet{fourkiotis2024applying} focus on enhancing pharmaceutical sales
+forecasting through the application of both traditional statistical
+methods and advanced machine learning techniques. Various forecasting
+approaches, including traditional models like Naïve and ARIMA,Facebook
+Prophet, and machine learning methods such as XGBoost and LSTM neural
+networks are compared. The results demonstrated that machine learning
+models, particularly XGBoost, significantly outperformed traditional
+methods. \citet{belghith2024new} present a rolling forecasting framework
+employing moving average and three exponential smoothing methods
+(Brown's, Holt's, and Holt-Winters) implemented using Microsoft Power BI
+for enhanced data visualization.
+
+Table~\ref{tbl-literature} provides a summary of key studies in the
+literature on forecasting for pharmaceutical products. We identify
+several limitations both in the existing research and in practice, which
+highlight important gaps that motivate our current study. These gaps are
+summarized as follows:
 
@@
 \def\labelenumi{\arabic{enumi}.}
+\item
+  Despite significant investment in data collection technologies and
+  logistics management systems, these systems primarily collect
+  transactional data, such as consumption. However, no data is collected
+  on events that influence consumption variability---information that
+  would be crucial for building reliable forecasting models. No studies
+  to date have explored how domain expertise could enhance
+  pharmaceutical demand forecasting.
+\item
+  Current research on forecasting for pharmaceutical supply chain
+  predominantly focuses on generating point forecasts. There is a lack
+  of studies that consider the entire forecast distribution of monthly
+  consumption, which would better capture the uncertainty of future
+  demand and provide a valuable risk management tool for
+  decision-makers.
+\item
+  Reproducibility remains a major challenge in this field. It is often
+  difficult for readers to reproduce previous studies without direct
+  assistance from the authors, limiting the practical application and
+  validation of existing research.
+\end{enumerate}
+
+\begin{table}[H]
+
+\caption{\label{tbl-literature}Summary of some studies on forecasting in
+pharmaceutical supply chain}
+
+\centering{
+
+\centering\begingroup\fontsize{11}{13}\selectfont
+
+\resizebox{\linewidth}{!}{
+\begin{tabular}{lll>{\raggedright\arraybackslash}p{10em}ll>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{2em}}
+\toprule
+References & data granularity & Lenth of data & Forecast variable & Horizon & Probabilistic & Method & Metric & Items & Data & Code\\
+\midrule
+Current study & Monthly & 60 months & Consumption of pharmaceutical products & 6 months & Yes & TimeGpt (with and without predictors), LSTM (with and without predictors), Exponential Smoothing, ARIMA, Dynamic Regression (ARIMAX), Multiple Regression (with and without predictors) & RMSSE,CRIPS,MASE & 33 & Yes & Yes\\
+Siddiqui et al. (2022) & Monthly & 58 months & Sales of pharmaceutical products & Un known & No & ARIMA, HOLT WINTER, ETS,  Theta, ARHOW & MEAN, MAD, MSE, RMSE,  MAPE & 31 & No & No\\
+de Oliveira et al.(2021) & Daily & 365 days & Lead time & NA & No & k-NN, SVM, RF, LR, MLP & MSE & 11 & Yes & No\\
+Rathipriya et al. (2023) & Weekly & 260 weeks & Sales of pharmaceuticals product & Un known & No & SNN, ARIMA P\_NN GR\_NN RBF\_NN ARIMA P\_NN GR\_NN RBF\_NN & RMSE, Normalized RMSE & 57 & Yes & No\\
+Kim et al.(2015) & Monthly & 44 month & Medicines sales data & 1 month & No & VARX & Prediction error rate & 4 & No & No\\
+Merkuryeva et al. (2019) & Weekly & 20 weeks & Sales of pharmaceuticals & 1–week & No & MA, Symbolic Regression, LR & R², MAD & 1 & No & No\\
+Nikolopoulos et al. (2016) & Yearly & 21 year & Prescription records of pharmaceuticals & 1–5 & No & Diffusion models,  ARIMA, ES, LR & R², ME, MAE, MSE & 14 & No & No\\
+Van Belle et al.(2021) & Weekly & 253 weeks & Medcine sales & 1–5 weeks & No & ARIMA, ETS, MLP, SVR, RF, LR & RMSE & 50 & No & No\\
+Khalil Zadeh et al. (2014) & Monthly & 36 months & Sales  pharmaceutical products & 1 month & No & ARIMA, Graph-based analysis and ANN, & R2,MSE, MAE & 217 & No & No\\
+Zhu  X et al.(2021) & Weekly & 520 weeks & Medcine sales & 1–2 months & No & MA, LR, Clustering and RNN & NME, NMAE, NMSE & 245 & No & No\\
+Anusha et al.(2014) & Monthly & 36 months & Sales of pharmaceuticals & Un known & No & 6 Month MA, SES, WES & MAD, MSE, MAPE & 2 & Yes & No\\
+Burinskiene (2022) & Weekly & 13 weeks & Medcine sales & 36 days & No & SES, MA, Naïve, Holt & MAPE,MSE, U² & 8 & No & No\\
+Candan et al. (2014) & Yearly & 6 years & Medcine sales & 3 months & No & NF & SE Mean & 1 & No & No\\
+Ribeiro et al.(2017) & Monthly & 24 months & Medcine sales & 3 months & No & DPM & SMAPE & 357 & No & No\\
+Bon and NG (2017) & Monthly & 68 months & Medcine sales & 1 month & No & SMA, SES, DMA, DES, RA, HW-A, SA, HW-M, ARIMA-SM & RMSE, MSE, MAD,  MAPE  & 1 & No & No\\
+Newberne and Mike (2006)  & Monthly & 36 months & Pseudoephedrine prescriptions & 3 months & No & HW-M & MAPE & 1 & No & No\\
+Sousa et al.(2019) & Monthly & 79 months & drug distribution & quarterly & No & Naïve, Seasonal Naïve, ARIMA, DLM, LST, Recurrent Neural Network, SES, Holts linear and damped trends, HW, ES-SSM, Theta & MAE & 30 & Yes & Yes\\
+Zahra and Putra (2019) & Monthly & 36 months & Consumption & 12 months & No & ARIMA, ES & RMSE, MAPE, MAD, MSE & 1 & No & No\\
+Nguyen et  al. (2023) & Un known & Un known & Pharmaceutical products & Un known & No & VARX & ME, MAE,MAPE,RMSE & Unknown & No & No\\
+Christos et al.(2017) & Weekly & 129 weeks & Web-scraping  of analgesic medcines & Un known & Yes & VARX  & AIC, BIC, MSE, MAE & Unknown & No & No\\
+Belghith et al.(2024) & Daily & 1095 days & Sales of pharmaceuticals & 1 year & No & Brown’s method, Holt’s method, and Holt-Winters method, moving average & MAD, MAPE & 125 & No & No\\
+Konstantinos  et al.(2024) & Weekly & 260 weeks & Medcine sales & 1 year & No & Naïve, ARIMA, XGBoost , LSTM, Prophet & MAPE,MSE & 57 & No & No\\
+\bottomrule
+\end{tabular}}
+\endgroup{}
+
+}
+
+\end{table}%
+
+\section{Experiment setup}\label{sec-experiment}
+
+\subsection{Data}\label{sec-data}
+
+For this study, we use a dataset spanning five years (Dec.~2017- July
+2022) of consumption data obtained from EPSS. Rigorous checks were
+conducted to ensure the consistency and completeness of the collected
+data. From the extensive pool of pharmaceutical products within EPSS, we
+selected a set of 33 key pharmaceuticals, including various programs and
+representing different classes of drugs.
+
+\subsubsection{Explaoratory Data Analysis with consumption
+data}\label{explaoratory-data-analysis-with-consumption-data}
+
+Given the number of products in this study, we created several data plot
+and computed features of the time series, including the strength of
+trend and seasonality to better understadn data.
+Figure~\ref{fig-feature} shows the strength of trend versus seasonality.
+Each point represents one time series with the strength of trend in
+x-axis and the strength of seasonality in y-axis. Both measures are on a
+scale of {[}0,1{]}. the strength of trend and seasonality were
+calculated using the ``STL'' (Seasonal and Trend decomposition using
+Loess) decomposition method, as described by \citet{mstl}.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{main_files/figure-pdf/fig-feature-1.pdf}
+
+}
+
+\caption{\label{fig-feature}The strength of the trend and seasonality in
+the time series of pharmaceutical product consumption. The scatter plot
+shows 33 data points, with each point corresponding to a product.}
+
+\end{figure}%
+
+It is evident that some time series display strong trends and/or
+seasonality, while the majority exhibit low trends and minimal
+seasonality. A number of products show pronounced trends, and only a few
+demonstrate clear seasonal patterns. Beyond assessing the strength of
+trends and seasonality, we also visualized all time series to understand
+data and various patterns, including trends and instances of erratic
+consumption behavior during certain months. For example, some series
+show low consumption volumes over consecutive months, followed by peak
+consumption in specific months, making them more volatile and difficult
+to forecast. This underscores the diversity of monthly pharmaceutical
+time series patterns within the dataset and highlights the importance of
+understanding the factors driving these consumption behaviors. Figure
+Figure~\ref{fig-dataviz2} illustrates the time plots for a few selected
+products.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{main_files/figure-pdf/fig-dataviz2-1.pdf}
+
+}
+
+\caption{\label{fig-dataviz2}Monthly time plot of consumption. X-axis
+shows the month, consisting of 60 data points (months) and y-axis shows
+the comsumption. The panels display data from four products to give a
+glimpse of the consumption patterns.}
+
+\end{figure}%
+
+Visualizing the data revealed that various events significantly impacted
+the consumption of different products, but these were not reflected in
+the available data. Expert insights were crucial for understanding the
+nature and timing of these events, filling gaps in the system, and
+incorporating qualitative factors that EPSS logistics system miss.
+Therefore, we conducted interviews with domain experts to collect
+information on external factors that influence consumption. These
+interviews allowed us to account for unusual patterns and customize our
+forecasting models to more accurately reflect the unique consumption
+behaviors of each product.
+
+\subsubsection{Collaborative expert review to identify factors affecting
+consumption}\label{collaborative-expert-review-to-identify-factors-affecting-consumption}
+
+To understand the external factors influencing product consumption, we
+began by visualizing the time series data for consumption as highlighted
+in the previous section. This initial step helped identify specific
+instances within the data that showed unusual patterns, such as peaks,
+unusually high or low observations, and consecutive periods of low
+consumption. These observations indicated potential events or factors
+affecting consumption levels.
+
+To gather domain knowledge about factors affected the consumption of
+products, we contacted experts from EPSS who work in the distribution
+unit. Next, we collaborated with six experts, each with over 10 years of
+experience managing pharmaceutical distribution. In these sessions, we
+reviewed the time series of all products together to understand the
+potential drivers behind these anomalies. The experts provided insights
+based on their extensive experience and knowledge of distribution
+practices. When there was uncertainty or gaps in the data for certain
+products, additional experts were consulted to gather further context.
+If all experts reached a consensus, the influencing factors were
+confirmed and included in the analysis.
+
+As part of this process, we examined the electronic bin cards for 33
+pharmaceutical products and conducted detailed data exploration. A bin
+card is a physical or electronic record used to track the inventory
+levels of individual items in a storage location, such as a warehouse or
+stockroom. The term `bin' refers to the specific storage location for an
+item. Information typically recorded on a bin card includes the Item ID,
+unit of measure, product description, account details, date, references,
+beginning balance, manufacturer, batch number, expiry date, quantity
+received, quantity issued, and balance. In summary, a bin card is a
+fundamental tool in inventory management, especially in environments
+where physical stock tracking is essential. It provides a
+straightforward and immediate way to monitor item quantities, helping to
+ensure effective inventory management. Over a period of two weeks, we
+meticulously traced consumption data through these bin cards. Moreover,
+We ensured accuracy by converting dates from the Ethiopian calendar to
+the Gregorian calendar and excluded transactions related to internal
+storage or warehouse relocation within EPSS. Through this collaborative
+approach, we were able to identify several predictors affecting
+distribution and consumption that are summarised as followings:
+
+\begin{itemize}
 \tightlist
 \item
-  Evaluate the robustness of various forecasting models for
-  pharmaceutical demand prediction within EPSS.
-\item
-  Identify pharmaceuticals that exhibit superior forecast accuracy under
-  specific models.
-\item
-  Assess the performance of forecast accuracy metrics to gauge the
-  efficacy of selected models.
-\item
-  Test the incorporation of diverse predictors affecting forecasts and
-  evaluate resulting forecasting accuracy. In pursuing these objectives,
-  this study endeavors to elevate the precision and reliability of
-  pharmaceutical demand forecasting, thereby contributing to optimized
-  supply chain management and enhanced healthcare outcomes in Ethiopia.
-\end{enumerate}
-
-\hypertarget{sec-lit}{%
-\section{Research background}\label{sec-lit}}
-
-Precise sales prediction is an essential and inexpensive way for each
-company to augment their profits, decrease their costs, and achieve
-greater flexibility to changes. In other words, exact sales forecasting
-is utilized for capturing the tradeoff between customer demand
-satisfaction and inventory costs \citep{gupta2000mid}. Especially, for
-the pharmaceutical industry, successful sales forecasting systems can be
-very beneficial, due to the short shelf-life of many pharmaceutical
-products and the importance of the product quality which is closely
-related to the human health \citep{doganis2006time}.There are many
-decisions that depend on the quality of forecasts in the health care
-system, from capacity planning to layout decisions to the daily
-schedules. In general,the role of forecasting in health care is to
-inform both clinical and non-clinical decisions. While the former
-concerns decisions related to patients and their treatments
-\citep{makridakis2019forecasting}, the latter involves
-policy/management, and supply chain decisions that support the delivery
-of high-quality care for patients.In the coming paragarphs we will
-review what other researchers had done in the area of pharmaceuticals
-forecasting.Morover today,manufacturing companies are organized
-following the principles of just-in-time (JIT) delivery which focuses on
-efficiency and eliminating the waste of excessive inventory
-\citep{van2021using}. In JIT systems, accurate Stock Keeping Unit (SKU)
-short-term demand forecasts, typically on a daily, weekly or monthly
-basis, are crucial as they are the fundamental inputs to drive the
-supply chain planning system.Therefore, forecasting performance can
-strongly affect a company's financial results, since both under- and
-overestimating short-term demand are expensive as they result in supply
-shortages and consequently poor customer service, and excess inventories
-and product obsolescence, respectively (Sanders and Graman, 2009).In
-this section, we provide a brief review of studies on forecasting
-harmaceuticals products.In conclusion, the research findings indicated
-that the hybrid neural network approach, which combines linear and
-nonlinear modeling techniques, outperformed traditional methods like
-ARIMA and standalone neural networks in forecasting pharmaceutical
-sales. The hybrid approach demonstrated the ability to capture both
-linear and nonlinear patterns in sales data accurately, especially in
-scenarios with limited historical records for individual drugs.
-
-A study by Khalil Zadeh et al.~developed an intelligent sales prediction
-model for Pharmaceutical Distribution Companies (PDCs) by combining
-network analysis tools and time series forecasting methods. This
-approach addresses the challenge of limited historical sales data for
-pharmaceutical products, making traditional forecasting methods
-inadequate\citep{zhang2004neural}.The researchers employed Mean Squared
-Error (MSE) and Mean Absolute Error (MAE) to evaluate the performance of
-various forecasting models, including ARIMA, neural networks, and a
-hybrid approach combining both. The hybrid model, inspired by Zhang's
-work\citep{zhang2004neural}, used ARIMA for linear components and neural
-networks for nonlinear components.Three methods were used: (1) ARIMA,
-(2) a hybrid neural network with each drug's past records, and (3) a
-hybrid neural network with both each drug's and its group members' past
-records. Grouping products and using group members' records
-significantly improved forecasting accuracy by addressing the issue of
-limited data and leveraging both linear and nonlinear modeling
-capabilities.The study concluded that the hybrid approach, incorporating
-both individual and group past records, performed better than using
-either method alone \citep{khalil2014intelligent}.
-
-A study by Van Bell et al.~used various forecasting models for point of
-sale data to assess the impact of including sell-through data on
-forecast accuracy. They compared information sharing (IS) methods with
-non-information sharing (NIS) methods using the following approaches.
-Substitution Approach (ETS-W and ARIMA-W): These models use downstream
-wholesaler sales time series as input instead of prior manufacturer
-shipments, suitable when all goods pass through wholesalers. Integration
-Approach (Extrapolative Time Series Methods with Exogenous Variables):
-This includes methods that integrate downstream information using
-extrapolative time series techniques and machine learning (ML)
-approaches. Extrapolative Time Series Methods (ETS and ARIMA):
-Traditional forecasting methods based on modeling past time series
-structures and extrapolating into the future. The study evaluated these
-methods using Mean Squared Error (MSE) and Mean Absolute Error (MAE).
-Additionally, the researchers used naive methods, IS and NIS with ETS
-and ARIMA models, and machine learning techniques like LASSO, Support
-Vector Regression (SVR), and Random Forest (RF).LASSO (Least Absolute
-Shrinkage and Selection Operator) was identified as the best-performing
-method for improving forecast accuracy, particularly for one-step ahead
-forecasts. It outperformed other methods for longer horizons (h
-\textgreater{} 1) by effectively selecting the most relevant variables
-and reducing overfitting. ETS was found to be the most effective
-univariate baseline method, and SVR showed the best performance for h =
-1. The study concluded that IS methods generally outperformed NIS
-methods, indicating that incorporating downstream information, such as
-sell-through data, can significantly enhance short-term forecast
-accuracy in supply chain management. Machine learning techniques like
-LASSO, SVR, and RF faced challenges related to overfitting and model
-complexity, especially with high-dimensional data. Balancing model
-complexity and interpretability was crucial for robust forecasting.
-Overall, the research suggests that LASSO, along with ETS and SVR, are
-effective for improving demand prediction accuracy in multi-echelon
-supply chains by leveraging downstream information and collaborative
-approaches\citep{van2021using}.
-
-A study by Nikolopoulos et al.~forecasted UK pharmaceutical time series
-before and after patent expiry, a critical point when a generic version
-enters the market. The study used various models, including naïve and
-moving averages, the Bass diffusion model, repeat purchase diffusion
-models, three regression models, and two exponential smoothing (ES)
-approaches for non-trended (simple ES) and trended data (Holt ES)
-\citep{nikolopoulos2016forecasting}.Holt Exponential Smoothing: This
-model was the most accurate for forecasting both branded and generic
-drugs across short-term, mid-term, and long-term horizons.Naïve Model:
-Used as a benchmark, the relative absolute error (RAE) was calculated to
-compare forecasting accuracy.Bass Diffusion Model: Although popular in
-marketing research, it had limitations in accurately forecasting
-pharmaceutical time series due to the complex dynamics between branded
-and generic drugs. Comparative Analysis: The study compared various
-forecasting models to evaluate their performance, highlighting the
-importance of selecting appropriate models based on the data and
-context.The research found that Holt's exponential smoothing model
-outperformed others, including naïve models, moving averages, and ARIMA
-models, especially for short-term forecasts. However, forecasting
-accuracy varied across different horizons. Despite their limitations,
-diffusion models were included to provide additional insights.Overall,
-the study emphasized the strengths and limitations of different
-forecasting approaches in the context of pharmaceutical time series,
-with Holt's exponential smoothing model demonstrating superior
-performance\citep{nikolopoulos2016forecasting}.
-
-Astudy by Siddiqui et al used to develop a combined forecasting
-model.The combination is developed using two frequent, but separately
-used models, i.e.~Autoregressive Integrated Moving Average (ARIMA) model
-and Holts Winter model and a combined model were proposed which is known
-as the ARHOW model.The ARIMA model is commonly used for time series
-forecasting, while Holt's Winter model is effective for handling
-seasonality in data \citep[markovska2016comparative]{inproceedings}.The
-ARHOW model aims to enhance forecasting accuracy by leveraging the
-strengths of both ARIMA and Holt's Winter models. The model equation for
-obtaining a consolidated forecast is as follows: ARHOW = β *
-ArimaForecast + β1 * HoltWinterForecast In this equation: ARHOW
-represents the Combined Weighted Forecast. ArimaForecast and
-HoltWinterForecast are the forecasts obtained by applying the ARIMA and
-Holt's Winter models over the training set.β and β1 are the weights
-assigned to the ARIMA and Holt's Winter forecasts, respectively. By
-combining the forecasts generated by these two models, the ARHOW model
-seeks to mitigate the impact of volatility and external factors on
-forecasting accuracy, ultimately providing more reliable predictions for
-pharmaceutical sales data \citep{siddiqui2022hybrid} . Combining
-multiple forecasting techniques like ARIMA and Holt's Winter to create
-the ARHOW model introduces complexity. Managing and interpreting the
-results of a hybrid model may pose challenges in practical
-implementation and decision-making processes .Morover External factors
-such as regulatory changes, technological advancements, and market
-dynamics can introduce forecasting errors. While the ARHOW model aims to
-improve accuracy, it may still be susceptible to errors resulting from
-unforeseen external influences \citep{siddiqui2022hybrid}.The key
-forecasting metrics employed in the study include:
-
-Mean Absolute Percentage Error (MAPE), Mean, Mean Absolute Deviation
-(MAD), Mean Square Error (MSE), Root Mean Square Error (RMSE).The Mean
-Absolute Percentage Error (MAPE) and other forecasting metrics suggest
-that the ARHOW model provides more accurate predictions compared to
-traditional forecasting techniques like ARIMA, Holt's Winter, ETS,
-Theta, and Naïve models. The model's efficiency in forecasting
-pharmaceutical demand can help manufacturers plan inventory and prevent
-stockouts or overstock situations .In conclusion, the study underscores
-the effectiveness of the ARHOW model in improving demand forecasting
-accuracy in the pharmaceutical industry and provides insights for
-enhancing supply chain efficiency through advanced forecasting
-techniques .
-
-A study by de Oliveira et al.~compared five machine learning regression
-approaches for lead time forecasting in the pharmaceutical supply chain.
-The models used were k-nearest neighbors (KNN), support vector machines
-(SVM), random forests (RF), linear regression (LR), and multi-layer
-perceptrons (MLP). These models aimed to predict supplier delivery
-times, enhancing planning for production and logistics services in the
-pharmaceutical field \citep{de2021lead}. The study focused on using
-machine learning techniques to forecast lead times for purchasing
-products in the pharmaceutical industry. The comparison aimed to
-identify the most effective approach for predicting purchasing lead
-times, demonstrating high accuracy potential, especially with linear
-support vector regression. Key limitations noted included computational
-complexity, sensitivity to outliers, difficulty in interpretability, and
-overfitting. Using mean squared error (MSE) as the forecast accuracy
-metric, the study found that the SVM approach achieved the best
-performance, with an average error of less than two days. This indicates
-that SVM was the most effective model among those tested for forecasting
-purchasing lead times in the pharmaceutical supply
-chain\citep[de2021lead]{inproceedings}.
-
-A study by Rathipriya et al.~focused on demand forecasting for
-time-series pharmaceutical data using both shallow and deep neural
-network models\citep{rathipriya2023demand}. The researchers employed
-several models, including the Autoregressive Integrated Moving Average
-(ARIMA) model, known for its simplicity and ability to handle
-non-stationary series.The shallow neural network models used were,
-Radial Basis Function Neural Network (RBF\_NN),Generalized Regression
-Neural Network (GR\_NN),Probabilistic Neural Network (P\_NN),For deep
-neural networks, the study investigated:Long Short-Term Memory
-(LSTM),Stacked LSTM,Additionally, the study explored hybrid models,
-combining ARIMA with the Adaptive Neural Fuzzy Inference System (ANFIS)
-to forecast Iran's annual energy usage, demonstrating higher accuracy
-compared to individual models. The researchers also employed methods
-like Genetic Algorithms (GA), Bayesian structures, and Reinforcing
-Artificial Neural Networks (RANN) for forecasting short-run demand.The
-performance of the demand forecast models (DFMs) was evaluated using
-Root Mean Squared Error (RMSE). The study found that shallow neural
-network models, such as GR\_NN, RBF\_NN, and P\_NN, had almost identical
-mean RMSE values across all eight drug categories, indicating their
-effectiveness in predicting pharmaceutical demand. Conversely, deep
-neural network models like LSTM and Stacked LSTM did not perform as well
-due to the limited number of samples available for training. The
-research concluded that shallow neural network-based DFMs performed well
-for five out of eight drug categories, while the ARIMA model was best
-for the remaining three. This suggests that no single forecasting model
-is optimal for all drug categories\citep[rathipriya2023demand]{article}.
-
-A study by Candan et al.~on demand forecasting in the pharmaceutical
-industry utilized a neuro-fuzzy approach as the forecasting
-method\citep[hrupelj2017demand]{inproceedings}. Neuro-fuzzy modeling
-combines information from empirical models, heuristics, and data,
-providing effective models that explain solutions better than black-box
-models like neural networks\citep{abraham2001neuro, babuvska2003neuro}.
-The study also explored the emotional learning fuzzy inference system
-(ELFIS) and locally linear neuro-fuzzy (LLNF) methods proposed by
-Abdollahzade et al.~(2012) for predicting the stock prices of a
-pharmaceutical
-company\citep{abdollahzade2012application, hrupelj2017demand}. While the
-neuro-fuzzy approach combines fuzzy logic and neural networks to improve
-forecasting, it may still struggle with extremely complex or non-linear
-relationships within the data. The study used paired t-tests and mean
-difference analysis to evaluate forecast accuracy. Overall, the metrics
-supported the conclusion that the neuro-fuzzy approach is a robust
-method for demand forecasting in the pharmaceutical industry, providing
-accurate predictions closely aligned with actual demand quantities.
-
-A study by Burinskiene discusses various forecasting techniques in the
-pharmaceutical retail industry. These include: Single Exponential
-Smoothing: Applies weighted averages with exponentially decreasing
-weights for older observations. Moving Average: Used for trend-based
-estimation to predict demand and standard deviation. Naive Method:
-Assumes all predictions are equal to the last observation, serving as a
-benchmark. Holt's Linear Method: An extension of single exponential
-smoothing for trend prediction. Holt-Winters No Trend Method: A
-variation without a trend component\citep{burinskiene2022forecasting}.
-The study focuses on short-term forecasting, excluding long-term
-forecasting and safety stock considerations, which are crucial in the
-pharmaceutical industry due to varying demand
-patterns\citep{nikolopoulos2016forecasting}. The naive method fails to
-capture trends or seasonality, while the moving average method struggles
-with sudden changes and irregular patterns\citep{restyana2021analysis}.
-Holt's linear method is limited in handling complex seasonal patterns,
-and the Holt-Winters method requires careful parameter tuning and
-historical data analysis\citep{koehler2001forecasting}. Single
-exponential smoothing is not effective for sudden changes or
-outliers\citep[zahra2019forecasting]{inproceedings}. The study used
-MAPE, MSE, and Theil's U2 test to evaluate forecast accuracy, with
-Theil's U2 test comparing results to the naive forecast. A U2 value
-greater than 1 indicates better performance than the naive method
-\citep{burinskiene2022forecasting}. The Holt-Winters Multiplicative
-method was found to improve weekly forecasts and outperform other
-methods, especially in addressing product shortages in pharmacies.
-
-A study conducted by Anusha et al.~in the Indian pharmaceutical retail
-sector utilized forecasting models such as Simple Moving Average,
-Exponential Smoothing, and Winter's Exponential
-Smoothing\citep{anusha2014demand}. These models were employed to predict
-demand for both seasonal pharmaceutical items like Okacet 10mg tablets
-and random pharmaceutical items like Stamlo Beta tablets. Quantitative
-indicators like Mean Absolute Deviation (MAD), Mean Squared Error (MSE),
-and Mean Absolute Percentage Error (MAPE) were used to assess forecast
-error. The Winter's Exponential Smoothing (WES) model showed the least
-sales forecast error for the seasonal pharmaceutical item Okacet 10mg
-tablets, with a MAPE of 27.50, indicating high forecast accuracy.
-Conversely, for the non-seasonal pharmaceutical item Stamlo Beta
-tablets, the Six-month Moving Average model performed best with a MAPE
-of 45.62\citep{anusha2014demand}. However, the study did not account for
-unexplained seasonal variations in staff, shifts in the number of
-potential patients, physician turnover, preferences for pharmaceuticals,
-and the influence of marketing representatives on surrounding clinics.
-These factors were not factored into the analysis, potentially affecting
-the accuracy of demand forecasting in the pharmaceutical retail setting.
-These limitations suggest that the study may not have considered all
-relevant variables impacting demand forecasting accuracy.
-
-Zhu et al.~conducted a study comparing various forecasting methods in
-the pharmaceutical industry, including Moving Average (MA), Linear
-Regression (LR), Exponential Smoothing (ES), Exponential State Space
-Model (ETS), Vector Autoregressive with exogenous variables (VARX),
-Symbolic regression, Graph-based analysis, Artificial Neural Networks
-(ANN), Clustering, and Recurrent Neural Networks
-(RNN)\citep{zhu2021demand}. Additionally, machine learning algorithms
-such as Linear Regression (LR), Support Vector Regression (SVR), Random
-Forest (RF), and RNN with cross-series training were assessed.The study
-evaluated forecast accuracy using metrics like Normalized Mean Squared
-Error (NMSE), Normalized Mean Absolute Error (NMAE), and Bias. Results
-indicated that the cross-series training models outperformed baseline
-models (Moving Average, Exponential Smoothing, Linear Regression) in
-terms of NMSE, NMAE, and Bias. RNN models, particularly when implemented
-with cross-series training, showed superiority and generalizability
-across datasets compared to other machine learning algorithms (LR, SVR,
-RF), confirming the effectiveness of the proposed forecasting framework
-in the pharmaceutical industry.
-
-Ribeiro et al.~conducted a study aimed at detecting outliers and
-predicting sales for a pharmaceutical distribution company. They
-employed the dampened Pegels method, a time-series data mining
-technique, to forecast future quantities of medicines to be ordered by
-the company\citep[ribeiro2017improving]{article}. The accuracy of the
-forecasting model was evaluated using the Symmetric Mean Absolute
-Percentage Error (SMAPE) metric, which indicated that the Pegels method
-provided more accurate sales forecasts compared to the company's
-previous method. Furthermore, these forecasts effectively met demand,
-aiding in efficient stock management and ensuring customer satisfaction
-while minimizing inventory costs.
-
-The study by Bon and Ng at the University Health Centre aimed to enhance
-inventory demand forecasting for Panadol 650mg through the utilization
-of various forecasting techniques. Their research design encompassed
-both qualitative and quantitative data collection methods within a
-descriptive framework. Historical data spanning from January 2009 to
-August 2014 were analyzed employing methods like single moving average,
-exponential smoothing, and regression analysis. Forecast accuracy was
-assessed using metrics including RMSE, MSE, MAD, and MAPE to identify
-the most precise forecasting technique. The objective was to optimize
-inventory management by selecting the most accurate forecasting method
-based on these measurements\citep{bon2017optimization}. Results
-highlighted regression analysis as the superior technique among the ten
-methods tested, exhibiting the least forecasting error for optimizing
-Panadol 650mg inventory demand at the University Health
-Centre\citep[bon2017optimization]{inproceedings}.
-
-The study conducted by Newberne and MIKE focused on utilizing
-Holt-Winters Forecasting in Healthcare, particularly in the context of
-healthcare management involving Pseudoephedrine prescriptions and Upper
-Respiratory Infection (URI) visits \citep{newberne2006holt}. Data from
-the Military Healthcare System's M2 database were analyzed using the
-Holt-Winters method to demonstrate its applicability in healthcare
-forecasting. Forecasting accuracy was assessed using the Mean Absolute
-Percentage Error (MAPE) metric. The study employed various forecasting
-methods including Naïve, Seasonal Naïve, ARIMA, and Deep Learning Model:
-Long Short-Term Memory (LSTM) Recurrent Neural Network. Additionally,
-techniques like Simple Exponential Smoothing, Holt with linear and
-damped trends, Holt-Winters, ES state space model, and Theta were
-explored to forecast drug quantities for distribution in the Brazilian
-Public Health System. Among these methods, the LSTM model often
-exhibited the best performance based on Mean Absolute Error (MAE) as a
-measure of forecast accuracy. However, deep learning models like LSTM
-may pose challenges due to their complexity and computational intensity,
-requiring specialized expertise for tuning and implementation in
-practical settings.
-
-The study conducted by Zahra and Putra focused on using ARIMA and Single
-Exponential Smoothing methods to predict and calculate EOQ (Economic
-Order Quantity). To measure forecast accuracy, the researchers utilized
-RMSE, MAPE, MAD, and MSE metrics. The findings revealed that the ARIMA
-(1,0,0) method exhibited the smallest error values, with a MAPE of 13\%
-and MSE of 283,715, indicating its superior accuracy compared to the
-Exponential Smoothing method. Based on these results, the researchers
-predicted a need for 222 boxes of medicine for the next period and
-recommended procuring 223 boxes to maintain a service level of 95\%. The
-study underscored the significance of accurate forecasting in optimizing
-medicine procurement processes and ensuring adequate stock levels to
-efficiently meet demand. By employing MAPE and MSE as forecast accuracy
-metrics, the researchers could compare the performance of different
-forecasting methods and determine the most effective approach for
-predicting medicine needs in hospital
-settings\citep{zahra2019forecasting}.
-
-The study conducted by Restyana et al.~focused on analyzing drug
-forecasting within the Jombang Regency from 2017 to 2019. They employed
-the Single Moving Average and Single Exponential Smoothing methods to
-predict drug demand across 35 public health centers. The objective was
-to identify the method with the lowest forecasting error by assessing
-metrics like Mean Absolute Deviation (MAD) and Mean Square Error (MSE).
-By analyzing annual primary data and testing errors, the researchers
-aimed to evaluate the accuracy of the forecasting methods. The findings
-revealed that the Single Exponential Smoothing (SES) approach exhibited
-superior forecast accuracy, as evidenced by lower MAD and MSE values.The
-study by Kim et al.~used Vector Autoregressive with Exogenous Variables
-(VARX) models for demand forecasting of medicines. Two models were
-constructed to test the effect of exogenous variables: Model 1, a pure
-autoregressive model without exogenous variables, and Model 2, the
-suggested VARX model incorporating topic trends from social media
-analysis. The study used sales data as endogenous variables and moving
-averages of topic trends as exogenous variables in the VARX models.The
-results showed that incorporating topic trends into the VARX models
-enhanced the performance of demand forecasting, particularly for
-medicines sensitive to environmental and seasonal changes. By analyzing
-topic trends related to certain diseases and incorporating them into
-VARX models, the researchers found significant improvement in forecast
-accuracy.Prediction error rates were used as the forecast accuracy
-measure, calculated for four different medicines for each model (Model 1
-and Model 2). The comparison of error rates between the two models
-demonstrated that adding topic trends as exogenous variables in the VARX
-model significantly improved the accuracy of demand forecasts,
-highlighting the importance of integrating social media data into
-forecasting processes\citep{kim2015demand}.
-
-Table 1 provides a summary of some studies in the literature on
-forecasting in pharmaceuticals products. We have identified several
-limitations in the existing literature on pharmaceutical product
-forecasting that have motivated us to pursue this research. These
-limitations can be summarized as follows:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
+  Stock replenishment: refers to the process of restocking or refilling
+  inventory to ensure that there are sufficient quantities of products
+  or materials available to meet demand. Whenever there was stock
+  replenishment at the central EPSS, consumption and distribution to
+  hubs and health facilities increased. This increase was attributed to
+  the need to restock depleted inventories and the push from central
+  EPSS to manage space constraints.
+\end{itemize}
+
+\begin{itemize}
 \tightlist
 \item
-  Almost all of the studies have primarily concentrated on point
-  forecasts at a single level.
+  Physical fiscal year inventory counting: refers to the process of
+  manually counting and verifying the actual quantities of
+  pharmaceutical products available in stock at a specific location. The
+  process is critical for maintaining the accuracy of inventory records,
+  ensuring that medicines are available when needed, and preventing
+  stockouts or overstocking. Physical inventory counting periods also
+  influenced consumption. Stores closed during these periods, halting
+  transactions. We observed increased consumption before inventory
+  counting periods, as hubs and facilities stocked up. July and August
+  were identified as physical counting periods each year.
+\end{itemize}
+
+\begin{itemize}
+\tightlist
 \item
-  A majority of the research has not explored longer time series data
-  periods.
+  Malaria seasonality: Refers to the predictable patterns and
+  fluctuations in malaria incidence throughout the year, typically
+  influenced by climate and environmental conditions. In many regions,
+  malaria transmission peaks during and shortly after the rainy season,
+  when conditions such as stagnant water pools create ideal breeding
+  sites for the Anopheles mosquitoes that transmit the disease.
+  Conversely, malaria cases often decline during the dry season when
+  mosquito breeding sites are reduced. During peak malaria seasons,
+  there is a significant surge in the demand for antimalarial drugs and
+  other related treatments. Malaria seasonality was another significant
+  predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine
+  and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We
+  identified epidemic periods affecting consumption: September to
+  December 2017, March to May 2018, September to December 2018, March to
+  May 2019, September to December 2019, March to May 2020, September to
+  December 2020, March to May 2021, September to December 2021, and
+  March to May 2022.
+\end{itemize}
+
+The key predictors identified are deterministic values, since both past
+and future values of these predictors are known in advance. We then
+integrated them into the consumption data, creating the complete dataset
+required for model building and running the experiment.
+
+\subsection{Forecasting models}\label{forecasting-models}
+
+We evaluate a range of univariate models and their counterparts that
+include predictors, spanning from simpler methods like regression,
+exponential smoothing, and ARIMA to more complex models such as long
+short-term memory (LSTM) networks and advanced foundational time series
+forecasting models. Below, we provide a brief overview of these
+approaches. Detailed implementation codes in R and Python are available
+in a GitHub repository and accessible for public.
+
+\textbf{Exponential Smoothing State Space model (ETS):} ETS models, as
+described by \citet{hyndman2021forecasting}, combine trend, seasonality,
+and error components in time series using different configurations that
+can be additive, multiplicative, or mixed. The trend component can be
+specified as none (``N''), additive (``A''), or damped additive
+(``Ad''); the seasonality can be none (``N''), additive (``A''), or
+multiplicative (``M''); and the error term can be additive (``A'') or
+multiplicative (``M''). To forecast consumption, we use the ETS()
+function from the fable package in R, which automatically selects the
+optimal model for each time series based on the corrected Akaike's
+Information Criterion (AICc). In our study, an automated algorithm
+determines the best configuration for trend, seasonality, and error
+components across each time series, leveraging the ets() function's use
+of AIC to identify the optimal model. Given the high volume of time
+series (1530), manual selection of components is impractical, so the
+algorithm customizes model forms for each series based on its unique
+characteristics. This results in a tailored combination of additive or
+multiplicative components, adapting to the specific patterns of each
+time series.
+
+\textbf{Multiple Linear Regression (MLR)}: We use Multiple linear
+regression to model the relationship between a consumption and potential
+variables influencing its variation. In our first model, we use multiple
+linear regression with a trend component to capture the underlying
+progression over time, i.e., \emph{regression}. We also incorporate
+dummy variables for each month to account for seasonal fluctuations,
+without including any additional predictors, i.e.,
+\emph{regression\_reg}. This approach helps us establish a baseline
+model focused solely on temporal trends and seasonal patterns. We then
+extend this model by introducing additional predictors that include
+variables such as replenishment cycles, fiscal year indicators, and
+periods with malaria prevalence. These additional predictors allow the
+model to see if capturing external factors can provide a better
+understanding of the factors influencing the consumption and result at
+enhanced accuracy. We produce forecasts using TLSM() function from the
+fable package in R.
+
+\textbf{ARIMA and ARIMA with regressors}: ARIMA (AutoRegressive
+Integrated Moving Average) is a powerful statistical model designed to
+forecast time series by capturing temporal dynamics and are widely used
+in time series forecasting due to their ability to model complex trends
+and patterns over time without relying on external predictors. ARIMAX
+(AutoRegressive Integrated Moving Average with eXogenous variables)
+extends ARIMA by incorporating external variables, or exogenous
+predictors, into the model. This modification allows to include relevant
+information from external factors such as malaria season, and fiscal
+year period, and stock replenishement period that may explain variations
+in the series beyond its internal time dynamics, we refer to this in the
+result as \emph{arima\_reg}. By adding these predictors, ARIMAX combines
+the strengths of ARIMA's time-series structure with the flexibility of
+regression models. In our study, we use an automated algorithm to
+determine the optimal configuration for ARIMA components, following the
+approach outlined by \citet{hyndman2021forecasting} and We use the
+ARIMA() function from the fable package in R.
+
+\textbf{Long Short-Term Memory neural network (LSTM)}: The LSTM model is
+a specialised form of recurrent neural network (RNN) used to model
+sequential data by capturing long-term dependencies
+\citep{graves2012long}. Unlike traditional RNNs, LSTMs can learn to
+retain information for longer time periods due to their unique
+architecture, which consists of several gates that control the flow of
+information. This makes LSTMs particularly effective for time series
+forecasting. In our implementation, we used a sequential model
+architecture, comprising one LSTM layer with 50 units, followed by dense
+layers, with the final output being a single linear unit. The Adam
+optimizer was employed to minimize the mean squared error, and the model
+was trained for 100 epochs using the keras\_model\_sequential() function
+from the Keras package in R. We use LSTM models both with and without
+predictors, referring to the LSTM model with predictors as
+\emph{lstm\_reg}.
+
+\textbf{TimeGPT}: TimeGPT is the first pre-trained foundational model
+designed specifically for time series forecasting, developed by Nixtla
+\citep{garza2023timegpt}. It uses a transformer-based architecture with
+an encoder-decoder configuration but differs from other models in that
+it is not based on large language models. Instead, it is built from the
+ground up to handle time series data. TimeGPT was trained on more than
+100 billion data points, drawing from publicly available time series
+across various sectors, such as retail, healthcare, transportation,
+demographics, energy, banking, and web traffic. This wide range of data
+sources, each with unique temporal patterns, enables the model to manage
+diverse time series characteristics effectively. Furthermore, TimeGPT
+supports the inclusion of external regressors in its forecasts and can
+generate quantile forecasts, providing reliable uncertainty estimation.
+We use TimeGpt models both with and without predictors, referring to the
+model with predictors as \emph{timegpt\_reg}.
+
+In all models, probabilistic forecasts were generated using
+bootstrapping to create 1,000 possible future scenarios for each period
+within the forecast horizon \citep{hyndman2021forecasting}.
+
+\subsection{Performance evaluation}\label{performance-evaluation}
+
+To assess the performance of our forecasting methods, we split the data
+into a series of training and test sets and apply time series
+cross-validation with a forecast horizon of 6 months. Each training set
+is expanded in monthly increments, allowing model development and
+hyperparameter tuning to be performed on the training data, with errors
+evaluated on the corresponding test sets. We assess forecasting
+performance using both point and probabilistic error measures.
+
+The error metrics presented here consider a forecasting horizon denoted
+by by \(j\), which represents the number of time periods ahead we are
+predicting, with \(j\) ranging from 1 to 6 months in our study. Point
+forecast accuracy is measured using the Mean Squared Scaled Error (MSSE)
+and the Mean Absolute Scaled Error (MASE). The Mean Absolute Scaled
+Error (MASE) \citep{HK06, hyndman2021forecasting} is calculated as
+follows:
+
+\[
+  \text{MASE} = \text{mean}(|q_{j}|),
+\] where \[
+  q_{j} = \frac{ e_{j}}
+ {\displaystyle\frac{1}{T-m}\sum_{t=m+1}^T |y_{t}-y_{t-m}|},
+\] and \(e_{j}\) is the point forecast error for forecast horizon \(j\),
+\(m = 12\) (as we have monthly seasonal series), \(y_t\) is the
+observation for period \(t\), and \(T\) is the sample size (the number
+of observations used for training the forecasting model). The
+denominator is the mean absolute error of the seasonal naive method in
+the fitting sample of \(T\) observations and is used to scale the error.
+Smaller MASE values suggest more accurate forecasts. Note that the
+measure is scale-independent, thus allowing us to average the results
+across series.
+
+Here, \(e_{j}\) represents the point forecast error for forecast horizon
+\(j\), with \(m = 12\) (since we are dealing with monthly seasonal
+data). The term \(y_t\) denotes the observation at time \(t\), and \(T\)
+is the sample size, or the number of observations used for training the
+forecasting model. The denominator in the MASE formula is the mean
+absolute error of the seasonal naive method over the training sample of
+\(T\) observations, providing a basis for scaling the forecast error.
+Lower MASE values indicate more accurate forecasts. Notably, this
+measure is scale-independent, allowing us to average results across
+different series for broader performance comparison.
+
+A related measure is MSSE
+\citep{hyndman2021forecasting, makridakis2022m5}, which uses squared
+errors rather than absolute errors:
+\begin{equation}\phantomsection\label{eq-RMSSE}{
+  \text{MSSE} = \text{mean}(q_{j}^2),
+}\end{equation} where, \[
+  q^2_{j} = \frac{ e^2_{j}}
+ {\displaystyle\frac{1}{T-m}\sum_{t=m+1}^T (y_{t}-y_{t-m})^2},
+\] Again, this is scale-independent, and smaller MSSE values suggest
+more accurate forecasts.
+
+To measure the forecast distribution accuracy, we calculate the
+Continuous Rank Probability Score \citep{hyndman2021forecasting}. It
+rewards sharpness and penalizes miscalibration, so it measures overall
+performance of the forecast distribution.
+\begin{equation}\phantomsection\label{eq-CRPS}{
+  \text{CRPS} = \text{mean}(p_j),
+}\end{equation} where \[
+  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
+\] where \(G_j(x)\) is the forecasted probability distribution function
+for forecast horizon \(j\), and \(F_j(x)\) is the true probability
+distribution function for the same period.
+
+Calibration refers to the statistical consistency between the
+distributional forecasts and the observations. It measures how well the
+predicted probabilities match the observations. On the other hand,
+sharpness refers to the concentration of the forecast distributions ---
+a sharp forecast distribution results in narrow prediction intervals,
+indicating high confidence in the forecast. A model is well-calibrated
+if the predicted probabilities match the distribution of the
+observations, and it is sharp if it is confident in its predictions. The
+CRPS rewards sharpness and calibration by assigning lower scores to
+forecasts with sharper distributions, and to forecasts that are
+well-calibrated. Thus, it is a metric that combines both sharpness and
+miscalibration into a single score, making it a useful tool for
+evaluating the performance of probabilistic forecasts.
+
+CRPS can be considered an average of all possible quantiles
+\citep[Section 5.9]{hyndman2021forecasting}, and thus provides an
+evaluation of all possible prediction intervals or quantiles. A specific
+prediction interval could be evaluated using a Winkler score, if
+required.
+
+\section{Results and discussion}\label{sec-results}
+
+In this section, we compare the forecasting performance of various
+approaches, examining models that incorporate expert-identified business
+context predictors versus those that rely solely on historical
+consumption data. Point forecast performance is reported using MASE and
+RMSSE, while probabilistic forecast accuracy is reported using CRPS.
+
+The forecasting performance is reported in Figure~\ref{fig-point}, in
+which the average forecast accuracy over forecast horizon and across all
+products is calculated for each origin. We report the distribution of
+accuracy metrics across all rolling origins. This shows how models
+varies in providing accuracy across different origins. The y-axis
+displays models sorted by their MASE and RMSSE values, with the model
+exhibiting the lowest error positioned at the bottom. This model is the
+Ariam model, which incorporates exogenous variables. Additionally,
+Figure~\ref{fig-point} indicates that predictors obtained through
+interactions with domain experts enhance point forecast accuracy across
+most models. This underscores the critical importance of systematically
+collecting such expert-informed data alongside transactional consumption
+data.
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{main_files/figure-pdf/fig-point-1.pdf}
+
+}
+
+\caption{\label{fig-point}Distribution of point forecast accuracy across
+different origins, averaged across the forecat horizon and all products.
+The total number of months used to calculate the accuracy in the test
+set is 12 months for each product. MASE and MSSE are relative to the
+corresponding values for the training set.}
+
+\end{figure}%
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{main_files/figure-pdf/fig-prob-1.pdf}
+
+}
+
+\caption{\label{fig-prob}Distribution of probabilistic forecast accuracy
+across different origins, averaged across the forecat horizon and all
+products. The total number of months used to calculate the accuracy in
+the test set is 12 months for each product.}
+
+\end{figure}%
+
+Figure~\ref{fig-prob} illustrates the forecast distribution accuracy
+measured by CRPS, which evaluates both forecasting calibration and
+interval sharpness. A smaller CRPS value indicates better overall
+performance. We observed that incorporating domain knowledge improved
+forecast accuracy for most models, enhancing not only point forecasts
+but also probabilistic forecasts. Notably, ARIMA achieved the highest
+probabilistic forecast accuracy, while LSTM models had the lowest. This
+aligns with the findings related to point forecast accuracy, reinforcing
+the earlier explanations for these results.
+
+We observe that LSTM models exhibit a widespread error distribution
+compared to other models. This is attributed to their performance
+variability, which can significantly depend on the characteristics of
+demand patterns across different products. For example, the product
+``Amlodipine - 5mg - Tablet'' demonstrates periods of extreme
+variability, with spikes in demand followed by periods of very low or
+zero demand. Such patterns align well with the strengths of LSTM models,
+which excel at capturing long-term dependencies and managing complex
+temporal fluctuations. In contrast, the demand for ``Anti-Rho (D)'' is
+erratic and sporadic, characterized by frequent, random fluctuations
+without any discernible long-term trends or seasonal patterns. This
+chaotic nature, combined with inconsistent and often low volumes, poses
+significant challenges for LSTM models. The lack of clear structure or
+stable patterns limits the models' ability to generalize effectively,
+resulting in poorer forecast performance. Although LSTM models can
+handle irregular data to some extent, they require sufficient and
+well-distributed training samples to achieve optimal performance. In our
+study, the majority of products exhibit erratic and sporadic demand
+patterns similar to ``Anti-Rho (D)'', resulting in this widespread error
+distribution.
+
+The findings emphasize the importance of incorporating relevant domain
+knowledge into forecasting models. In pharmaceutical supply logistics
+administration systems, data often consists solely of transactional
+records on consumption and distribution. Including exogenous variables
+such as administrative procedures, seasonal patterns, or conflicts, or
+any other relevant factor that could be identified by those with domain
+knowledge proved essential for reducing forecast errors. This approach
+improved both point and probabilistic forecast accuracy, enabling a more
+confident assessment of uncertainty. Therefore, the systematic
+collection of information about significant events and their impact on
+consumption is vital. Recording details of events such as policy
+changes, administration procedures, conflicts, and incorporating this
+information into forecasting models creates a comprehensive
+understanding of consumption. This practice enhances modeling
+reliability and allows institutions in developing countries and
+humanitarian organizations to better forecast demand, allocate resources
+effectively, and respond proactively. Establishing robust data
+collection systems is thus a critical step for strengthening forecasting
+capabilities and operational resilience.
+
+Our analysis also highlights the value of using foundational time series
+forecasting models like TimeGpt, particularly in developing countries
+and the global health and humanitarian sectors. In these contexts, the
+lack of advanced analytical skills and systematic data collection can
+hinder effective forecasting and decision-making.
+
+\subsection{An illustration of probabilistic forecast for Pharmacuitical
+product
+consumption}\label{an-illustration-of-probabilistic-forecast-for-pharmacuitical-product-consumption}
+
+In this section, we present an illustrative example of a probabilistic
+forecast for future consumption of \emph{Sodium Chloride (Normal
+Saline)} product. Due to the complexity of including such visualizations
+for all products, only one example is shown here. However, it is
+feasible to generate these plots for all products if needed.
+
+In practice, point forecasts are commonly used despite their
+limitations, but they do not account for the inherent uncertainty
+associated with forecasts. The future is inherently uncertain, and
+effective planning requires considering alternative scenarios.
+Probabilistic forecasts offer a comprehensive approach by assigning
+likelihoods to a range of possible outcomes, recognizing that different
+consumption levels may occur with varying probabilities. The primary
+purpose of probabilistic forecasting, as illustrated in
+Figure~\ref{fig-forecast-density-hstep}, is to quantify and communicate
+uncertainty. This figure displays the forecast distribution of
+consumption over a 6-month horizon using a density plot. For each month
+within the forecast period, a separate distribution is generated. The
+plot also includes the point forecast alongside 80\% and 90\% prediction
+intervals to illustrate potential variability.
+
+It is important to note that while point forecasts and prediction
+intervals can be derived from probabilistic forecasts, the reverse is
+not true. A single-point forecast cannot inherently provide the
+probabilistic context needed to capture the range of possible outcomes.
+Although prediction intervals can indicate a range of potential values,
+they do not convey the detailed probabilities of low or high
+consumption. This distinction highlights the value of probabilistic
+forecasting in supporting informed decision-making by offering a clearer
+view of future uncertainties.
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{main_files/figure-pdf/fig-forecast-density-hstep-1.pdf}
+
+}
+
+\caption{\label{fig-forecast-density-hstep}A graphical illustration of
+the forecast distribution of a pharmacutical product (i.e.~total
+incidence attended) for the SB health board for a horizon of six month.
+For each month, we display the point forecast (black point), the
+histogram, and 80\% (thick line) and 90\% (thin line) prediction
+intervals. It also shows a portion of a historical time series as well
+as its fitted values.}
+
+\end{figure}%
+
+\section{Conclusion and future reserach}\label{sec-conclusion}
+
+In this study, we conducted an extensive analysis of pharmaceutical
+demand forecasting within the EPSS. Using five years of consumption data
+for 33 key pharmaceutical products, along with additional information
+gathered through collaboration with domain experts at EPSS, our goal is
+to contribute to enhancing the demand forecasting process and emphasize
+the importance of integrating domain knowledge into model building. This
+step is critical, as the phase of understanding the data and
+incorporating valuable contextual information is often overlooked. Too
+frequently, forecasting efforts rely solely on consumption data, leading
+to building models with little relevance to reality. Our approach
+highlights the necessity of including domain insights to construct more
+accurate and effective forecasting models.
+
+Our findings highlight the importance of collecting and incorporating
+domain knowledge when building forecasting models. We evaluated both
+point and probabilistic forecasts using a range of models, from simple
+univariate approaches like ARIMA to complex models such as LSTM.
+Notably, we demonstrated that employing complex models does not
+necessarily lead to more accurate forecasts in the pharmaceutical supply
+context. Additionally, our research revealed that newly developed
+foundational time series forecasting models have potential
+applicability, particularly in environments where advanced analytical
+skills and knowledge are limited, as often seen in pharmaceutical supply
+chains in developing countries.
+
+To advance forecasting practices further, exploring additional predictor
+variables---such as public health campaigns, disease outbreaks, and
+economic indicators---could provide valuable insights into demand
+dynamics. Improving the granularity of historical consumption data,
+through finer temporal intervals and geographic differentiation, also
+holds potential for more precise predictions. Additionally, hybrid
+forecasting models that combine judgmental point forecasts with
+probabilistic machine learning models could enhance predictive
+performance. Empowering EPSS staff through training in foundational time
+series models, data interpretation, and collection will enable informed
+decision-making and strengthen workforce capabilities. Replicating this
+study across diverse healthcare settings and conducting comparative
+analyses would help validate the adaptability and applicability of the
+findings. Additionally, it is recommended that pharmaceutical supply
+services systematically collect and maintain detailed records of events
+that may influence consumption, alongside consumption data. This
+practice is essential for refining demand models and enhancing forecast
+accuracy.
+
+\subsection{Practical challenges and
+limitations}\label{practical-challenges-and-limitations}
+
+Despite dedicated efforts to engage with domain experts to better
+understand the data---and spending significant time over two weeks
+collaborating with them---we found that the process of interpreting data
+through domain knowledge is complex and presents unique challenges.
+These challenges include defining what constitutes domain knowledge and
+determining how best to incorporate it to enhance model reliability and
+accuracy. Below, we summarize some of these key challenges. One issue
+involved capturing accurate information from bin cards and expert
+review, particularly during periods of disruption like COVID-19 and
+conflicts. For instance, during the COVID-19 pandemic, experts noted a
+decrease in demand due to travel restrictions. However, despite the
+reduced demand, there were still significant distributions of
+pharmaceuticals from the EPSS central to various hubs and health
+facilities. This was because, in response to anticipated shortages, a
+political decision was made to push products to these facilities before
+the travel restrictions took full effect. The rationale was that if the
+facilities remained closed, patients would have no access to
+medications, necessitating the preemptive stockpiling of essential
+pharmaceuticals. This scenario illustrates how policy decisions can
+significantly impact supply chain data, making it challenging to model
+distribution patterns accurately. Similarly, experts indicated that
+conflicts would disrupt the distribution of pharmaceuticals. While
+conflict zones did indeed hinder transportation, distribution still
+occurred whenever roads were temporarily opened, even amidst on-going
+conflict. This created inconsistencies in the data, as periods of halted
+distribution were followed by rapid replenishment once access was
+restored. Such fluctuations add complexity to modelling the supply chain
+from the central EPSS to hubs and health facilities. Other challenges
+involved seasonal or periodic activities, such as the fiscal year-end
+stock counting. During this time, EPSS temporarily halts transactions to
+conduct a full inventory count, which can take anywhere from one to two
+months, depending on the circumstances. This inconsistency in the
+duration of inventory counts adds a layer of unpredictability to supply
+chain modelling.
+
+\subsection{Future reserach}\label{future-reserach}
+
+Following the current research, several promising areas could further
+advance knowledge and practical applications in this area:
+
+\begin{itemize}
 \item
-  Only a few studies have integrated domain knowledge with time series
-  data to enhance pharmaceutical forecasting accuracy.
+  Future research could focus on developing intelligent systems that
+  incorporate domain-specific knowledge into model building. This would
+  involve creating algorithms capable of identifying relevant domain
+  insights and integrating them effectively into model training
+  processes. Such tools would bridge the gap between purely data-driven
+  approaches and expert-knowledge-enhanced modeling, leading to more
+  robust and contextually informed forecasts.
 \item
-  Additionally, no studies have been conducted in African settings,
-  where challenges related to forecasting and pharmaceutical
-  availability are prevalent.
-\end{enumerate}
-
-Table 1. Summary of Some Studies on Forecasting in pharmaceuticals
-products
-
-\hypertarget{sec-experiment}{%
-\section{Experiment setup}\label{sec-experiment}}
-
-\hypertarget{sec-data}{%
-\subsection{Data}\label{sec-data}}
-
-\hypertarget{dataset}{%
-\subsubsection{Dataset}\label{dataset}}
-
-For this study, we utilized a dataset spanning five years (2018-2022) of
-sales data obtained from EPSS. Rigorous checks were conducted to ensure
-the consistency and completeness of the collected data. From the
-extensive pool of pharmaceutical products within EPSS, we strategically
-selected a set of 37 key pharmaceuticals, encompassing various programs
-and representing different classes of drugs. This deliberate selection
-enriches the research by providing a comprehensive representation across
-the pharmaceutical landscape. We are presenting the graph for some of
-the pharmaceuticals sales for the last five years by using visualization
-for the readers to undesrtand the sales of the pharmaceuticals.
-
-\hypertarget{visualising-monthly-time-series}{%
-\subsection{Visualising monthly time
-series}\label{visualising-monthly-time-series}}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-10-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-11-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-12-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-13-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-14-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-15-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-16-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-17-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-18-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-19-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-20-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-21-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-22-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-23-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-24-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-25-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-26-1.pdf}
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-27-1.pdf}
-
-\hypertarget{data-splitting-and-preparation}{%
-\section{2. 2. Data Splitting and
-Preparation}\label{data-splitting-and-preparation}}
-
-The primary objective of this study is to forecast pharmaceutical demand
-within EPSS for the year 2023. To achieve this, the historical sales
-data for the chosen pharmaceuticals in EPSS was partitioned into
-distinct training and testing subsets. Employing the widely recognized
-70:30 split method, 70\% of the available data was designated as the
-training set, while the remaining 30\% was reserved for testing and
-validation purposes \citep{chan2019comparison}. As a result, the dataset
-was segregated into a training set consisting of 42 months and a testing
-set spanning 18 months
-
-\hypertarget{forecasting-methods}{%
-\section{2.3.1. Forecasting Methods}\label{forecasting-methods}}
-
-In pursuit of accurate pharmaceutical demand forecasts, we explored a
-spectrum of forecasting techniques. The following methods were applied:
-
-Naive Method: A simple approach relying on the assumption that the
-future value will be the same as the most recent observed value.This
-Methods uses when there is no seasonality
-\citetext{\citealp[burinskiene2022forecasting]{article}; \citealp[nikolopoulos2016forecasting]{article}}
-Mean Method: A straightforward technique forecasting future values based
-on the mean of past observations. Seasonal Naive Method: Similar to the
-naive method, this approach predicts future values using the most recent
-observed value from the same season. Exponential Smoothing: A method
-that assigns exponentially decreasing weights to past observations to
-forecast future values. Regression: Employing statistical regression
-models to establish relationships between variables for prediction.
-Autoregressive Integrated Moving Average Model (ARIMA): A time series
-model incorporating autoregressive, differencing, and moving average
-components for accurate forecasting. ARIMA with Predictors: Extending
-the ARIMA model with additional predictor variables to enhance
-forecasting accuracy. Combination of Models: Exploring ensemble methods
-that combine multiple models for improved predictive performance. By
-applying these diverse forecasting methodologies, we aimed to
-comprehensively assess and compare their efficacy in predicting
-pharmaceutical demand within EPSS. Each method was evaluated against the
-testing dataset to ascertain its forecasting accuracy and suitability
-for the specific
-
-\hypertarget{sec-method}{%
-\subsection{Forecasting methods}\label{sec-method}}
-
-Model building
-
-\hypertarget{naive}{%
-\section{Naive}\label{naive}}
-
-The naive forecast method is a simple approach that assumes future
-values will be the same as the most recent observed value, making it
-useful for stable or slowly changing time series. While it doesn't
-consider underlying patterns, it serves as a baseline for evaluating
-more advanced techniques\citep{hyndman2018forecasting}. It's
-particularly suitable for products with stable demand or limited
-seasonality, providing a straightforward reference for forecasting and
-inventory management.
-
-\hypertarget{mean}{%
-\section{Mean}\label{mean}}
-
-The mean forecast method predicts future values by averaging historical
-data, making it suitable for stable time series. While it overlooks
-trends and seasonality, it serves as a baseline for evaluating advanced
-techniques \citep{hyndman2018forecasting}. Its simplicity makes it
-valuable in certain forecasting applications
-\citep{hyndman2018forecasting}.
-
-\hypertarget{seasonal-naive}{%
-\section{Seasonal naive}\label{seasonal-naive}}
-
-The seasonal naive forecast method predicts future values based on the
-most recent observed value from the corresponding season in the previous
-year, making it useful for data with strong seasonal patterns. It serves
-as a baseline for evaluating seasonal forecasting models, capturing
-repetitive patterns effectively. Its simplicity and ability to capture
-seasonality make it valuable in seasonal forecasting
-applications\citep{hyndman2018forecasting}.
-
-\hypertarget{exponential-smoothing}{%
-\section{Exponential smoothing}\label{exponential-smoothing}}
-
-Exponential Smoothing is regarded as one of the most accurate and robust
-methods for time series forecasting. Not only does it exhibit high
-accuracy, but it also offers versatility and robustness. What makes
-Exponential Smoothing particularly appealing is its intuitive nature,
-making it easy to understand and interpret. Furthermore, this method
-boasts minimal data storage and computational requirements, making it
-highly efficient for real-time applications. In the business realm,
-Exponential Smoothing finds extensive use for inventory demand
-forecasting, showcasing its practicality \citep{gardner1985exponential}.
-Surprisingly, even when pitted against more sophisticated techniques,
-Exponential Smoothing has demonstrated impressive performance in
-forecasting competitions
-\citep[\citet{makridakis1982accuracy}]{makridakis2000m3}. Its simplicity
-and reliability make it a widely trusted approach in the field of
-forecasting.
-
-\hypertarget{regression}{%
-\section{Regression}\label{regression}}
-
-Regression methods play a crucial role in forecasting and optimizing the
-health supply chain, ensuring the availability and efficient
-distribution of essential medical resources. By analyzing historical
-data, regression models can identify patterns and relationships between
-variables such as patient demand, disease prevalence, and inventory
-levels. The basic concept is that we forecast the time series of
-interest assuming that it has a linear relationship with other time
-series. For example, we might wish to forecast monthly sales y using
-total advertising spend as a predictor X. Or we might forecast daily
-electricity y demand using temperature x1 and the day of week x2 as
-predictors. The forecast variable y is sometimes also called the regress
-and, dependent or explained variable. The predictor variables x are
-sometimes also called the regressors, independent or explanatory
-variables.
-
-Simple linear regression In the simplest case, the regression model
-allows for a linear relationship between the forecast variable y and a
-single predictor variable x: yt = β0 + β1xt + ε Example of data from
-such a model is shown in Figure 5.1. The coefficients β0 and β1 denote
-the intercept and the slope of the line respectively. The intercept β0
-represents the predicted value of y when x=0 . The slope represents the
-average predicted change in y resulting from a one unit increase in x.
-
-\hypertarget{figure-an-example-of-data-from-a-simple-linear-regression-model}{%
-\section{Figure : An example of data from a simple linear regression
-model}\label{figure-an-example-of-data-from-a-simple-linear-regression-model}}
-
-Notice that the observations do not lie on the straight line but are
-scattered around it. We can think of each observation yt as consisting
-of the systematic or explained part of the model, = β0 + β1xt , and the
-random ``error'', εt. The ``error'' term does not imply a mistake, but a
-deviation from the underlying straight line model. It captures anything
-that may affect yt other than xt
-
-Multiple linear regressions When there are two or more predictor
-variables, the model is called a multiple regression model. The general
-form of a multiple regression model is
-
-yt = β0 + β1x1,t + β2x2,t + ⋯ + βkxk,t + εt, (1)
-
-where is the variable to be forecast and are the predictor variables.
-Each of the predictor variables must be numerical. The coefficients
-β1,\ldots.. βk measure the effect of each predictor after taking into
-account the effects of all the other predictors in the model. Thus, the
-coefficients measure the marginal effects of the predictor variables.
-Assumptions When we use a linear regression model, we are implicitly
-making some assumptions about the variables in Equation (1).
-
-First, we assume that the model is a reasonable approximation to
-reality; that is, the relationship between the forecast variable and the
-predictor variables satisfies this linear equation.
-
-Second, we make the following assumptions about the errors (ε1, \ldots{}
-, εT ) :
-
-they have mean zero; otherwise the forecasts will be systematically
-biased.
-
-they are not autocorrelated; otherwise the forecasts will be
-inefficient, as
-
-there is more information in the data that can be exploited.
-
-they are unrelated to the predictor variables; otherwise there would be
-more information that should be included in the systematic part of the
-model.
-
-It is also useful to have the errors being normally distributed with a
-constant variance σ2 in order to easily produce prediction intervals.
-Another important assumption in the linear regression model is that each
-predictor x is not a random variable. If we were performing a controlled
-experiment in a laboratory, we could control the values of each x (so
-they would not be random) and observe the resulting values of y. With
-observational data (including most data in business and economics), it
-is not possible to control the value of x, we simply observe it. Hence
-we make this an assumption.
-
-\hypertarget{auto-regressive-integrated-moving-average-model-arima}{%
-\section{Auto regressive Integrated Moving Average Model
-(ARIMA)}\label{auto-regressive-integrated-moving-average-model-arima}}
-
-The Autoregressive Integrated Moving Average (ARIMA) model is a widely
-utilized approach for forecasting and analyzing time series data. It
-combines three key components: autoregression (AR), differencing (I),
-and moving average (MA). ARIMA models have been extensively studied and
-applied in various domains due to their flexibility and effectiveness in
-capturing both short-term and long-term dependencies in time series.Box
-and Jenkins (1970) {[}1{]} introduced the ARIMA model as a powerful tool
-for time series analysis. The authors provided a comprehensive framework
-for identifying, estimating, and diagnosing ARIMA models. They
-emphasized the importance of differencing to achieve stationarity in
-non-stationary time series data, allowing the model to capture trends
-and patterns effectively.
-
-\citep{chase2013demand} describes the application of the ARIMA model. It
-is the forecasting technique that integrates vital components of
-time-series and methods of regression. In 1970, George Box and Gwilym
-Jenkins introduced the ARIMA forecasting method, and they developed a
-comprehensive approach for forecasting \citep{chase2013demand}.ARIMA
-(p,d,q) models gather relevant factors from the historical data by using
-auto-correlation between them to distinguish those slacked request
-verifiable qualities that best anticipate future demand; whereas p is
-the order of Autoregressive (AR) term, q is the order of MA term
-
-ARIMA can demonstrate cycle and regularity, and is mainly used for
-mid-term aggregate forecasts and pre-sent far superior estimations than
-time series models or even casual models; explained by
-\citep{stadtler2014supply} explicitly considers dependent demands.
-Additionally, different components, for example, informative factors
-that impact request. It is also considered a good fit for the Mean
-Absolute Percentage Errors (MAPE) calculation.
-
-ARIMA model helps analyse the time series's probabilistic properties and
-develop sin-gle or multi-equation models \citep{moritz2015comparison}.
-Analysing a time series data is that there is a straightforward yet
-extensive arrangement of useful models that can speak to numerous
-conceivable examples of information found in time series. For a
-stationary time-series, we can envision the information creating process
-as a weighted blend of earlier perceptions in addition to an irregular
-request term. Besides, identifying the ARIMA model parameters (i.e.~p, q
-and d) is often challenging. Thereby, we use R software's statistical
-package; it is widely used in the relevant literature
-\citep{bokde2016psf, dhamo2010using}. Moreover, `R' generated the best
-combination of p, d and q values because it estimates the value of the
-parameter based on AIC (Akaike Information Criterion) and BIC (Bayesian
-Information Criterion).
-
-\#Performance evaluation \{\#sec-performance\}
-
-\hypertarget{forecast-accuracy-measures}{%
-\section{2.4. Forecast accuracy
-measures}\label{forecast-accuracy-measures}}
-
-To evaluate the forecast accuracy and bias, several measures including
-Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), Mean Absolute
-Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE),and Winkler
-are used. We only present the results for RMSSE due to the space limit
-of the journal and also because it is the recommended error metric in M5
-competition \citep{makridakis2022m5}.
-
-Point forecast accuracy is measured via the Mean Squared Scaled Error
-(MSSE) and the Mean Absolute Scaled Error (MASE). The Mean Absolute
-Scaled Error (MASE)
-\citep[hyndman2018forecasting,\citet{article}hyndman2006another,]{book}
-is calculated as ::: \{.cell
-hash=`main\_cache/pdf/unnamed-chunk-28\_a96c898b6cc2c723121cc3be6c515a81'\}
-
-:::
-
-\hypertarget{sec-results}{%
-\section{Results and discussion}\label{sec-results}}
-
-from the total of 37 key pharmaceuticals selected and tested with the
-different forecasting models arima showed that a better forecast
-accuracy for most of the pharmaceuticals followed by mean and
-regresstion models with RMSSE value of 0.850, 0.942 and 0.945
-respectively (Table 1).
-
-\begin{table}[!h]
-\centering
-\begin{tabular}[t]{lrrrrrr}
-\toprule
-.model & RMSE & MAE & MASE & RMSSE & winkler & CRPS\\
-\midrule
-arima & 109214.1 & 91630.74 & 0.925905 & 0.8503067 & 683471.4 & 64771.83\\
-exponential_smoothing & 117985.6 & 98588.74 & 1.048426 & 0.9579627 & 695575.3 & 68638.44\\
-mean & 121312.5 & 107616.64 & 1.087621 & 0.9422224 & 582790.3 & 73334.54\\
-naive & 135300.5 & 116398.62 & 1.268535 & 1.0841642 & 570090.5 & 74691.10\\
-regression & 131416.4 & 114329.68 & 1.065268 & 0.9455063 & 899543.3 & 83315.32\\
-\addlinespace
-snaive & 143861.2 & 122377.80 & 1.165142 & 1.0463095 & 732046.7 & 85266.80\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Then we tested the models by adding different predictors which affects
-the sales of the pharmaceuticals like physical count, budget release and
-budget closure by the government, COVID 19, campaign in to the models as
-predictors which will affects the sales of the key pharmaceuticals
-products and the result showed that still arima have a better forecast
-accuracy with RMSSE of 0.850 (Table 2).
-
-\begin{table}[!h]
-\caption{forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022 after
-considering predictors}\tabularnewline
-
-\centering
-\begin{tabular}[t]{lrrrrrr}
-\toprule
-.model & RMSE & MAE & MASE & RMSSE & winkler & CRPS\\
-\midrule
-arima_predictor & 114064.7 & 95753.88 & 0.9192677 & 0.851753 & 807586.4 & 68631.70\\
-regression_predictor & 132117.6 & 113194.12 & 1.0474410 & 0.941347 & 1000434.4 & 83390.61\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{figure}
-
-{\centering \includegraphics{main_files/figure-pdf/result2-1.pdf}
-
-}
-
-\caption{forecast accuarcy of selected pharmaceuticals in EPSS 2018-2022
-after considering predictors}
-
-\end{figure}
-
-Individual item with different forecasting models We tested 10 models
-namely arima, arima with predictor, combination, exponential smoothing,
-mean, naive, regression, regression with predictor, snaive and stlf.
-Individually on the selected key pharmaceuticals and the result
-indicated that 11 (32.43\%) of the selected pharmaceuticals have better
-forecast accuracy for arima predictor models, followed by naive 6
-(16.21\%), mean, regresstion, arima, and STLF methods each of them with
-3 (8.10\%) (Fig 2)
-
-\includegraphics{main_files/figure-pdf/unnamed-chunk-43-1.pdf}
-
-Those pharmaceuticals who had a better forecast accuracy with Arima
-predictors are Amoxacillin 500mg capsule, Anti RhD, Artehmether +
-Lumefantrine, Atenolol, Atrovastatin 200mg, Ferrosu sulphate with follic
-acid, frusemide injection, Insuline Iosphane bipahsic, Lamivudine
-+Efaverienze + Tenofovir, Lidocaine + HCL and Propyl thiro uracil.The
-six pharmaceuticasl which showed a better forecast accuracy with naive
-models were Ciprofloxacillin 500mg tablet, Magnesoum sulphate injection,
-Omeprazole capsule, Omeprazole injection, Rapid diagnostic test,
-Sulphamethaxone + Trimetoprine tablet.
-
-Discussion:
-
-The objective of this study was to evaluate the forecasting accuracy of
-different models for key selected pharmaceutical products from the
-Ethiopian Pharmaceutical Supply Services (EPSS). The analysis
-encompassed a total of 37 key pharmaceutical products, and the results
-highlight the performance of various forecasting models in predicting
-sales for these products. In the initial assessment of forecasting
-models, the ARIMA model emerged as the most accurate for the majority of
-the pharmaceutical products, exhibiting a Root Mean Squared Scaled Error
-(RMSSE) value of 0.850. This finding suggests that ARIMA has a strong
-predictive capability and outperforms other models, including the mean
-and regression models. The superiority of the ARIMA models over the
-Moving Average might be explained by most of the data set have seasonal
-patern and MA is one of the simplest prediction techniques for making
-projections about time-series without a noticeable seasonal pattern
-\citep{chopra2001supply}.The superiority of the ARIMA model underscores
-its suitability for pharmaceutical sales forecasting within the context
-of EPSS. Similar findings were reported in forecasting ethanol demand in
-India where ARIMA model outperforms other models like linear and
-non-linear regression models \citep{dey2023forecasting}.
-
-To enhance the predictive power of the models, additional predictors
-were incorporated into the analysis. Predictors such as physical count,
-budget release and closure by the government, the impact of COVID-19,
-and promotional campaigns were considered as potential drivers of
-pharmaceutical sales. Despite the inclusion of these predictors, the
-ARIMA model maintained its superior forecasting accuracy with an RMSSE
-of 0.850. This resilience further validates the robustness of the ARIMA
-model and its ability to capture complex interactions between variables
-influencing pharmaceutical sales.
-
-To delve deeper into the individual performance of the forecasting
-models, a comprehensive evaluation was conducted on each of the 37
-pharmaceutical products. Among the ten models tested, ARIMA with
-predictors exhibited the highest forecast accuracy for 32.43\% of the
-products. This was followed by the naive model, which performed well for
-16.21\% of the products. Notably, a few pharmaceuticals demonstrated a
-strong fit with specific models. For instance, Amoxicillin 500mg
-capsule, Anti RhD, Artehmether + Lumefantrine, and others exhibited
-superior forecast accuracy when utilizing the ARIMA model with
-predictors.
-
-Similarly, other pharmaceuticals displayed better forecasting outcomes
-with alternative models. Notably, Ciprofloxacin 500mg tablet, Magnesium
-sulfate injection, Omeprazole capsule, and others showed improved
-accuracy with the naive model. These findings underscore the importance
-of tailoring the choice of forecasting model based on the
-characteristics and dynamics of individual pharmaceutical products.This
-finding confirmed that There is not a single best technique to solve
-time-series forecasting problems \citep{zhang2007quarterly}. In order to
-deal with time-series forecasting, each problem might be solved with a
-different approach \citep{ensafi2022time}.
-
-In conclusion, this study highlights the pivotal role of accurate
-forecasting in the pharmaceutical supply chain management. The ARIMA
-model, particularly when integrated with relevant predictors, emerges as
-a powerful tool for predicting pharmaceutical sales within the EPSS.
-However, the selection of the most appropriate model should consider the
-specific attributes of each pharmaceutical product. Further research
-could explore the optimization of model parameters and the incorporation
-of additional contextual variables to enhance forecasting accuracy and
-supply chain efficiency.
-
-Limitation
-
-One notable limitation of this study revolves around the availability
-and quality of the data used for analysis. While we employed a
-comprehensive dataset from the EPSS, it is essential to acknowledge that
-the accuracy of forecasting models is heavily contingent on the quality,
-granularity, and completeness of historical sales data. Variations in
-data recording practices, potential errors, missing values, or
-inconsistencies within the dataset could impact the precision of the
-forecasts generated by the employed models. Moreover, the inherent
-complexity of pharmaceutical demand, influenced by a myriad of external
-factors such as socioeconomic changes,stock outs, prolonged procurement
-lead time, rationing of avaiable products, pushing of pharmaceuticals to
-different health facility without demand, healthcare policies, and
-unforeseen events, may introduce an additional layer of uncertainty. It
-is important to recognize that despite our best efforts to address
-data-related challenges, the robustness and reliability of our
-forecasting outcomes may still be subject to the limitations inherent in
-the original data.
-
-\hypertarget{sec-conclusion}{%
-\section{Conclusion}\label{sec-conclusion}}
-
-In this study, we embarked on a comprehensive analysis of pharmaceutical
-demand forecasting within the Ethiopian Pharmaceutical Supply Services
-(EPSS) context. By leveraging five years of sales data for 37 key
-pharmaceutical products, we aimed to enhance the accuracy of demand
-predictions and consequently contribute to more effective supply chain
-management. Our findings underscore the significance of accurate demand
-forecasting in optimizing inventory levels, resource allocation, and
-overall supply chain operations.
-
-The application of diverse forecasting models yielded valuable insights
-into their respective performance. Notably, the Autoregressive
-Integrated Moving Average (ARIMA) model emerged as a robust choice,
-consistently exhibiting superior forecast accuracy across a majority of
-the selected pharmaceutical products. Additionally, the incorporation of
-predictor variables, such as physical count, budget release and closure
-by the government, COVID-19 impact, and campaign events, further
-underscored the efficacy of the ARIMA model in accommodating external
-influences and delivering reliable forecasts.
-
-The individualized assessment of forecasting models revealed specific
-strengths across pharmaceutical categories. ARIMA models with predictors
-demonstrated exceptional performance for pharmaceuticals including
-Amoxacillin 500mg capsule, Anti RhD, Artehmether + Lumefantrine,
-Atenolol, and others, affirming their utility in capturing complex
-demand patterns. Moreover, the study shed light on pharmaceuticals best
-suited for alternative models like naive, mean, and regression methods,
-contributing to a comprehensive understanding of forecasting
-applicability.
-
-While these findings contribute significantly to the discourse on
-pharmaceutical demand forecasting, it's essential to acknowledge
-limitations. The accuracy of our forecasts is intrinsically linked to
-the quality and completeness of historical sales data, and the intricate
-interplay of external variables adds complexity to the forecasting
-landscape.
-
-In conclusion, this study serves as a pivotal step towards enhancing
-pharmaceutical supply chain management through refined demand
-forecasting. The demonstrated success of ARIMA models, particularly when
-augmented with predictor variables, holds promise for fostering
-resilient and adaptable supply chains. As the pharmaceutical landscape
-continues to evolve, our insights encourage further exploration of
-advanced forecasting techniques and continuous refinement of strategies
-to bolster supply chain efficiency and ensure consistent availability of
-essential pharmaceuticals.
-
-In light of the study's findings, several actionable recommendations
-emerge. First and foremost, the integration of the Autoregressive
-Integrated Moving Average (ARIMA) model with predictor variables stands
-out as a viable strategy for enhancing pharmaceutical demand forecasting
-accuracy within Ethiopian Pharmaceutical Supply Services (EPSS). This
-entails harnessing the predictive power of external factors to refine
-forecasting precision. To further advance forecasting practices, the
-exploration of additional predictor variables such as public health
-campaigns, disease outbreaks, and economic indicators could offer
-valuable insights into demand dynamics. Moreover, refining the
-granularity of historical sales data, possibly through finer temporal
-intervals and geographic diversity, holds potential for more nuanced and
-accurate predictions. Consideration could also be given to hybrid
-forecasting models that amalgamate ARIMA's short-term accuracy with
-machine learning's predictive prowess for long-term trends. Regular
-model evaluation, sensitivity analysis, and collaborative stakeholder
-engagement are recommended to ensure ongoing accuracy and alignment with
-changing demand patterns.
-
-Capacitating EPSS staff through training programs in forecasting
-methodologies and data interpretation can empower the workforce to make
-informed decisions. Replication of this study in diverse healthcare
-settings and comparative analyses can validate the findings'
-applicability and adaptability. By embracing these recommendations, EPSS
-can fortify its pharmaceutical supply chain with resilient, efficient,
-and responsive operations, ultimately ensuring the consistent
-availability of vital pharmaceuticals. a
-
-\hypertarget{reproducibility}{%
-\section*{Reproducibility}\label{reproducibility}}
+  Large Language Models (LLMs) for time series forecasting, like the one
+  included in this study, offer transformative opportunities for
+  developing countries by making advanced forecasting methods
+  accessible, even in contexts with limited expertise or infrastructure.
+  However, future research should explore their applicability,
+  advantages, limitations, and performance relative to traditional and
+  hybrid forecasting methods.
+\item
+  The use of distribution-based representation of error metrics across
+  different rolling origins or forecast horizons, such as boxplots, to
+  showcase forecast accuracy provides essential insights into
+  variability in accuracy at different point of time in data. Building
+  on this, future work could aim at designing methodologies that
+  identify diagnostic patterns within the distribution of errors. These
+  approaches could suggest model improvements by analyzing error
+  distributions and help practitioners to refine models based on error
+  behavior.
+\item
+  While current study focus on monthly aggregate consumption at the
+  organizational level, there is potential in exploring more granular
+  forecasts. Future research could investigate daily or weekly
+  forecasting and examine the impacts at hierarchical levels, such as
+  specific sites or health facilities. Investigating how fine-grained
+  temporal and hierarchical forecasts influence inventory management and
+  operational efficiency could yield valuable insights for reducing
+  stockouts, minimizing costs and reducing waste.
+\end{itemize}
+
+\section*{Reproducibility}\label{reproducibility}
 \addcontentsline{toc}{section}{Reproducibility}
 
-R code to produce all results in this paper is available at
+We use R and Python for model development and accuracy evaluation. The
+paper is written in Quarto, and all code, data, and files will be shared
+in a GitHub repository upon acceptance for publication.
 

### 2024-12-13 rscript/bahman.R

@@ -60,7 +60,21 @@ fit_univariate <- med_tsb_train_tscv_no_malaria |> model(regression = TSLM(quant
                                                          ets = ETS(quantity)
 )
 
-fcst_univariate <- fit_univariate |> forecast(h=f_horizon, bootstrap = TRUE)
+# fcst_univariate1 <- fit_univariate |> forecast(h=f_horizon, bootstrap = TRUE)
+
+# Generate 1000 future sample paths
+fcst_univariate <- fit_univariate |> generate(h = f_horizon, times = 1000) |>
+  mutate(.sim = if_else(.sim < 0, 0, .sim)) |> 
+  # Compute forecast distributions from future sample paths
+  as_tibble() |>
+  group_by(.id, item, .model, month) |> 
+  summarise(
+    quantity = distributional::dist_sample(list(.sim)), .mean = mean(.sim)
+  ) |>
+  ungroup() |>
+  # Create fable object
+  as_fable(index = month, key = c(.id, item,.model),
+           distribution = "quantity", response = "quantity")
 
 # fcct_and_test <- fcst_univariate |> full_join(med_tsb_test_tscv_stockout, by = c(".id", "item","month"))
 # forecast_nonstock <- fcct_and_test |> select(.id, item, .model, month, quantity= quantity.x,stock_out, .mean) |> filter(stock_out == 0) |> 
@@ -81,7 +95,21 @@ fit_univariate_regg <- med_tsb_train_tscv_no_malaria |> model(regression_reg = T
                                                    arima_reg = ARIMA(quantity ~ stock_replinish  +fical_year_counting )
 )
 
-fcst_predictors <- fit_univariate_regg |> forecast(new_data = med_tsb_test_tscv_no_malaria,bootstrap = TRUE)
+# fcst_predictors1 <- fit_univariate_regg |> 
+#   forecast(new_data = med_tsb_test_tscv_no_malaria,bootstrap = TRUE)
+
+fcst_predictors <- fit_univariate_regg |> generate(new_data = med_tsb_test_tscv_no_malaria, times = 1000) |>
+  mutate(.sim = if_else(.sim < 0, 0, .sim)) |> 
+  # Compute forecast distributions from future sample paths
+  as_tibble() |>
+  group_by(.id, item, .model, month) |> 
+  summarise(
+    quantity = distributional::dist_sample(list(.sim)), .mean = mean(.sim)
+  ) |>
+  ungroup() |>
+  # Create fable object
[…]

### 2025-05-26 diagram/diagram.qmd
revision submitted

```{=latex}
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.85, transform shape, node distance=1.3cm and 2cm]

% Node style
\tikzset{
  process/.style={
    rectangle,
    draw=black,
    thick,
    text width=4.5cm,
    align=center,
    minimum height=1.4cm
  },
  arrow/.style={
    thick,
    ->,
    >=Stealth
  }
}

% Nodes
\node (step1) [process] {1. Expert Identification (with EPSS)};
\node (step2) [process, below=of step1] {2. Onboarding \& Orientation};
\node (step3) [process, below=of step2] {3. Workshop: Collaborative Data Review \& Joint Identification of Anomalies};

\node (step4) [process, right=of step1] {4. Cross Validation Using Bin Cards \& Inventory Records};
\node (step5) [process, below=of step4] {5. Structured Expert Elicitation \& Contextual Clarification};

\node (step6) [process, right=of step4] {6. Expert Elicitation of Predictor Relevance};
\node (step7) [process, below=of step6] {7. Consensus-Based Selection of Predictors};
\node (step8) [process, below=of step7] {8. Variable Encoding \& Time Alignment};

\node (step9) [process, below=of step8, yshift=-0.3cm] {9. Integration into Forecasting Models};

% Arrows
\draw [arrow] (step1) -- (step2);
\draw [arrow] (step2) -- (step3);

\path (step3.east) -- +(0.6,0) coordinate (s3stub);
\draw [arrow] (step3.east) -- (s3stub) |- (step4.west);

\draw [arrow] (step4) -- (step5);

\path (step5.east) -- +(0.6,0) coordinate (s5stub);
\draw [arrow] (step5.east) -- (s5stub) |- (step6.west);

\draw [arrow] (step6) -- (step7);
\draw [arrow] (step7) -- (step8);
\draw [arrow] (step8) -- (step9);

% Group boxes
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step1)(step2)(step3), label={[black]90:Expert Engagement}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step4)(step5), label={[black]90:Validation}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step6)(step7)(step8), label={[black]90:Feature Construction}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step9), label={[black]90:Model Integration}] {};

\end{tikzpicture}
\end{figure}

```

### 2025-05-26 manuscript/02-literature.qmd
revision submitted

@@
 
-<!-- Despite the advancements in forecasting methods, several gaps remain, particularly in addressing uncertainty and incorporating domain knowledge into the modeling process.
-Most existing studies focus on point forecasts, which may not fully capture the range of possible outcomes in pharmaceutical demand.
-This limitation is critical, given the high stakes of inventory management in this industry.
-Moreover, there is a lack of research on the impact of integrating domain knowledge—such as insights from experienced professionals—into forecasting models.
-This integration could potentially improve model accuracy and reliability, especially in complex and rapidly changing environments.
-In summary, the literature on pharmaceutical forecasting has evolved from traditional statistical models to advanced machine learning techniques, each offering unique insights and capabilities.
-However, there is still room for improvement, particularly in addressing uncertainty and incorporating domain knowledge.
--->
-
 @tbl-literature provides a summary of key studies in the literature on forecasting for pharmaceutical products.

### 2025-05-26 manuscript/03-experiment.qmd
revision submitted

@@
 Both measures are on a scale of [0,1].
-the strength of trend and seasonality were calculated using the "STL" (Seasonal and Trend decomposition using Loess) decomposition method, as described by @mstl.
+the strength of trend and seasonality were calculated using the "STL" (Seasonal and Trend decomposition using Loess) decomposition method, as described by @bandara2025mstl
 
@@
 
-### Collaborative expert review to identify factors affecting consumption
+```{r}
+#| label: tbl-feature
+#| echo: false
+#| cache: false
+#| tbl-cap: "Summary statistics of the pharmaceutical product time series"
+#| tbl-pos: "H"
 
-To understand the external factors influencing product consumption, we began by visualizing the time series data for consumption as highlighted in the previous section.
-This initial step helped identify specific instances within the data that showed unusual patterns, such as peaks, unusually high or low observations, and consecutive periods of low consumption.
-These observations indicated potential events or factors affecting consumption levels.
+library(fpp3)
+all_feature <- med_qty_tsb %>%
+  features(quantity, feature_set(pkgs = "feasts")) |> dplyr::select(`ACF lag 1`=acf1,zero_run_mean,nonzero_squared_cv)
 
-To gather domain knowledge about factors affected the consumption of products, we contacted experts from EPSS who work in the distribution unit.
-Next, we collaborated with six experts, each with over 10 years of experience managing pharmaceutical distribution.
-In these sessions, we reviewed the time series of all products together to understand the potential drivers behind these anomalies.
-The experts provided insights based on their extensive experience and knowledge of distribution practices.
-When there was uncertainty or gaps in the data for certain products, additional experts were consulted to gather further context.
-If all experts reached a consensus, the influencing factors were confirmed and included in the analysis.
+summary_stat <- med_qty_tsb %>%
+  features(quantity,list(Mean = mean, Median=median, `Standard deviation`=sd, Miinum=min, Maximum=max)) |> 
+  bind_cols(all_feature) |> mutate(across(where(is.numeric), ~ round(.x, 2)))
+summary_stat |> 
+  knitr::kable(booktabs = T, linesep = "") |>
+  kable_styling(latex_options = c("scale_down"))
+```
 
-As part of this process, we examined the electronic bin cards for 33 pharmaceutical products and conducted detailed data exploration.
-A bin card is a physical or electronic record used to track the inventory levels of individual items in a storage location, such as a warehouse or stockroom.
-The term 'bin' refers to the specific storage location for an item.
-Information typically recorded on a bin card includes the Item ID, unit of measure, product description, account details, date, references, beginning balance, manufacturer, batch number, expiry date, quantity received, quantity issued, and balance.
-In summary, a bin card is a fundamental tool in inventory management, especially in environments where physical stock tracking is essential.
-It provides a straightforward and immediate way to monitor item quantities, helping to ensure effective inventory management.
-Over a period of two weeks, we meticulously traced consumption data through these bin cards.
-Moreover, We ensured accuracy by converting dates from the Ethiopian calendar to the Gregorian calendar and excluded transactions related to internal storage or warehouse relocation within EPSS.
-Through this collaborative approach, we were able to identify several predictors affecting distribution and consumption that are summarised as followings:
+@tbl-feature also provides the summary statistics computed for each pharmaceutical product time series include traditional distributional measures—mean, median, standard deviation, minimum, and maximum—alongside time series–specific and structural characteristics.
+These include the first-order autocorrelation (ACF at lag 1), the mean length of consecutive zero runs (zero_run_mean), and the squared coefficient of variation computed only on non-zero values (nonzero_squared_cv).
+Together, these metrics capture not only central tendency and dispersion, but also temporal dependence, sparsity patterns, and relative variability in periods with non-zero consumption.
 
-- Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
+## Collaborative expert review to identify factors affecting consumption
+
+\textcolor{blue}{To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation.
+In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.}
+
+Figure \ref{fig:expert-pipeline} \textcolor{blue}{presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.}
+
+```{=latex}
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}[scale=0.85, transform shape, node distance=1.3cm and 2cm]
+
+% Node style
+\tikzset{
+  process/.style={
+    rectangle,
+    draw=black,
+    thick,
+    text width=4.5cm,
+    align=center,
+    minimum height=1.4cm
+  },
+  arrow/.style={
+    thick,
+    ->,
+    >=Stealth
+  }
+}
+
+% Nodes
+\node (step1) [process] {1. Expert Identification (with EPSS)};
+\node (step2) [process, below=of step1] {2. Onboarding \& Orientation};
+\node (step3) [process, below=of step2] {3. Workshop: Collaborative Data Review \& Joint Identification of Anomalies};
+
+\node (step4) [process, right=of step1] {4. Cross Validation Using Bin Cards \& Inventory Records};
+\node (step5) [process, below=of step4] {5. Structured Expert Elicitation \& Contextual Clarification};
+
+\node (step6) [process, right=of step4] {6. Expert Elicitation of Predictor Relevance};
+\node (step7) [process, below=of step6] {7. Consensus-Based Selection of Predictors};
+\node (step8) [process, below=of step7] {8. Variable Encoding \& Time Alignment};
+
+\node (step9) [process, below=of step8, yshift=-0.3cm] {9. Integration into Forecasting Models};
+
+% Arrows
+\draw [arrow] (step1) -- (step2);
+\draw [arrow] (step2) -- (step3);
+
+\path (step3.east) -- +(0.6,0) coordinate (s3stub);
+\draw [arrow] (step3.east) -- (s3stub) |- (step4.west);
+
+\draw [arrow] (step4) -- (step5);
+
+\path (step5.east) -- +(0.6,0) coordinate (s5stub);
+\draw [arrow] (step5.east) -- (s5stub) |- (step6.west);
+
+\draw [arrow] (step6) -- (step7);
+\draw [arrow] (step7) -- (step8);
+\draw [arrow] (step8) -- (step9);
+
+% Group boxes
+\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step1)(step2)(step3), label={[black]90:Expert Engagement}] {};
+\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step4)(step5), label={[black]90:Validation}] {};
+\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step6)(step7)(step8), label={[black]90:Feature Construction}] {};
+\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step9), label={[black]90:Model Integration}] {};
+
+\end{tikzpicture}
+\caption{Diagram illustrating the sequential steps involved in collecting, structuring, and utilizing domain knowledge for modeling.}
+\label{fig:expert-pipeline}
+\end{figure}
+
+```
+
+\textcolor{blue}{Through a facilitated half-day workshop, these experts reviewed five years of monthly consumption data for 33 pharmaceutical products.
+Anomalous patterns—such as consumption spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles).
+These insights were cross-validated against bin cards and warehouse records, and internal transfers were excluded to isolate externally relevant events.
+Dates were standardized to the Gregorian calendar.}
+
+\textcolor{blue}{Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor.
+The final predictors included binary indicators for stock replenishment events, categorical variables for fiscal year inventory counts, and dummy variables for malaria seasonality.
+A further description of these variables are summarized as followings:}
+
+- \textcolor{blue}{Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
 Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased.
-This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.
+This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.}
 
-<!-- - Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand.
-This can happen when demand exceeds the available supply, or when there is a delay or failure in the replenishment process.
-Resulted in no transactions and a decline in consumption.
--->
-
-<!-- - COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain.
-The pandemic disrupted global manufacturing and supply, leading to shortages of essential medicines and raw materials, Ethiopia, like many other countries, relies heavily on imported pharmaceutical products.
-With the pandemic causing lockdowns and restrictions in manufacturing hubs like India and China, the supply of essential medicines and raw materials to Ethiopia was significantly disrupted.
-The impact of COVID-19 on consumption was significant.
-After the announcement of COVID-19 in Ethiopia, the central EPSS pushed a six-month supply of products to hubs to prevent stock outs.
-This push was from March to August 2020.
-During this period, if products were not stocked out, consumption continued to hubs and health facilities.
--->
-
-- Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
+- \textcolor{blue}{Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
 The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
@@
 We observed increased consumption before inventory counting periods, as hubs and facilities stocked up.
-July and August were identified as physical counting periods each year.
+July and August were identified as physical counting periods each year.}
 
-<!-- - Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters.
-Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises.
-Conflicts often lead to the destruction of infrastructure, including roads, bridges, and warehouses, making it difficult to transport pharmaceutical products to affected areas.
-This disruption can cause significant delays or prevent the delivery of essential medicines altogether.
-Internal displacements and conflicts in Ethiopia over the past five years also affected consumption and distribution.
-We traced periods of instability and included them as predictors.
-The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.
--->
-
-- Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
+- \textcolor{blue}{Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
 In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease.
@@
 Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks.
-We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.
+We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.}
 
-The key predictors identified are deterministic values, since both past and future values of these predictors are known in advance.
-We then integrated them into the consumption data, creating the complete dataset required for model building and running the experiment.
+\textcolor{blue}{These predictors, encoded with known historical and future values, were added to the modeling dataset.
+This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.}
 
@@
 
+\textcolor{blue}{To improve the robustness of the LSTM models and address overfitting issues, we introduced dropout regularization layers after the LSTM units and employed early stopping based on validation loss during model training.
+Each LSTM model was trained independently for each product series.
+Hyperparameter tuning was performed in a preliminary phase using a subset of products.
+Various configurations of LSTM units (30–100), dense units (50–200), dropout rates (0.1–0.5), and batch sizes (16–64) were evaluated based on validation set performance.
+The final model structure — 50 LSTM units, 100 dense units, a 0.2 dropout rate, and a batch size of 32 — was selected as a trade-off between forecast accuracy and model stability across different demand patterns.}
+
 **TimeGPT**: TimeGPT is the first pre-trained foundational model designed specifically for time series forecasting, developed by Nixtla [@garza2023timegpt].
@@
 
-In all models, probabilistic forecasts were generated using bootstrapping to create 1,000 possible future scenarios for each period within the forecast horizon [@hyndman2021forecasting].
+## Generating probabilistic forecasts
+
+\textcolor{blue}{In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical consumption.
+Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction} [@wang2023].
+
+\textcolor{blue}{In this study, we employed a bootstrapping approach to construct predictive intervals.
+Bootstrapping was chosen primarily for its flexibility and model-agnostic nature, allowing it to be applied uniformly across the diverse range of forecasting methods implemented without requiring strong distributional assumptions.
+Moreover, pharmaceutical consumption data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.}
+
+\textcolor{blue}{Specifically, we assume that future forecast errors will be similar to past forecast errors.
+The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed consumption, and $\hat{y}_t$ denotes the corresponding forecast estimate.
+To simulate future consumption paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates.
+This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.}
+
+\textcolor{blue}{Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts} [@hyndman2021forecasting].
+
+\textcolor{blue}{The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.}
 
@@
 
-To assess the performance of our forecasting methods, we split the data into a series of training and test sets and apply time series cross-validation with a forecast horizon of 6 months.
-Each training set is expanded in monthly increments, allowing model development and hyperparameter tuning to be performed on the training data, with errors evaluated on the corresponding test sets.
-We assess forecasting performance using both point and probabilistic error measures.
+\textcolor{blue}{To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation} [@hyndman2021forecasting].
+\textcolor{blue}{Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods.
+In our setup, the initial training set consisted of all available historical data from December 2017 up to June 2021.
+The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans consumption over a one-year horizon.
+At each iteration, models were trained on an expanding training window and evaluated over a fixed 6-month forecast horizon, aligned with typical EPSS planning cycles.
+After each forecast generation, the training set was expanded by one additional month, and the process was repeated, allowing for rolling assessment across the final 12 months of data.
+This structure ensured that forecasts reflected realistic operational scenarios, where forecasts are continuously updated as new data becomes available.
+This cross-validation design allowed us to evaluate each model's ability to perform across multiple different forecast origins and a variety of demand conditions, providing a more reliable and generalizable understanding of model performance.
+Furthermore, all model development steps, including hyperparameter tuning for the LSTM models, were conducted exclusively using the training data available at each iteration.
+No test set information was used during model selection or tuning.}
 
@@
 It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution.
+For probabilistic evaluation, 1,000 future paths were simulated per series, enabling robust estimation of forecast uncertainty.
+
 $$
@@
 A specific prediction interval could be evaluated using a Winkler score, if required.
-
-<!-- If the  $100(1−\alpha)\%$ prediction interval at time  t is given by $[\ell_{\alpha,t}, u_{\alpha,t}]$, then the Winkler score is defined as the length of the interval plus a penalty if the observation is outside the interval -->
-
-<!-- $$ -->
-<!-- W_{\alpha,t} = \begin{cases} -->
-<!--   (u_{\alpha,t} - \ell_{\alpha,t}) + \frac{2}{\alpha} (\ell_{\alpha,t} - y_t) & \text{if } y_t < \ell_{\alpha,t} \\ -->
-<!--   (u_{\alpha,t} - \ell_{\alpha,t})   & \text{if }  \ell_{\alpha,t} \le y_t \le u_{\alpha,t} \\ -->
-<!--   (u_{\alpha,t} - \ell_{\alpha,t}) + \frac{2}{\alpha} (y_t - u_{\alpha,t}) & \text{if } y_t > u_{\alpha,t}. -->
-<!--   \end{cases} -->
-<!-- $$ -->
-
-<!-- In this study, we report skill scores for winkler aand CRPS, with skill scores, we compute a forecast accuracy measure relative to some benchmark method.
-For example, if we use the naïve method as a benchmark, and also compute forecasts using the ETS model, we can compute the CRPS and Winkler skill score of the ETS model relative to the naïve method.
-If the skill scores are negative, it indicates that the proposed method performs worse than the naïve method; positive scores suggest it performs better.
-In this sense, skill scores are not entirely independent of data scale, similar to the concepts behind MASE and RMSSE for point forecasts.
--->
-
-<!-- $$ -->
-<!-- \frac{\text{CRPS}_{\text{Naïve}} - \text{CRPS}_{\text{ETS}}}{\text{CRPS}_{\text{Naïve}}}. -->
-<!-- $$ -->
-
-<!-- $$ -->
-<!-- \frac{\text{Winkler}_{\text{Naïve}} - \text{Winkler}_{\text{ETS}}}{\text{Winkler}_{\text{Naïve}}}. -->
-<!-- $$ -->

### 2025-05-26 manuscript/04-results.qmd
revision submitted

@@
 The y-axis displays models sorted by their MASE and RMSSE values, with the model exhibiting the lowest error positioned at the bottom.
-This model is the Ariam model, which incorporates exogenous variables.
+This model is the LSTM model, followed by ARIMA, which incorporate exogenous variables.
 Additionally, @fig-point indicates that predictors obtained through interactions with domain experts enhance point forecast accuracy across most models.
@@
 We observed that incorporating domain knowledge improved forecast accuracy for most models, enhancing not only point forecasts but also probabilistic forecasts.
-Notably, ARIMA achieved the highest probabilistic forecast accuracy, while LSTM models had the lowest.
+Notably, LSTM and ARIMA yielded the most accurate probabilistic forecasts when identified predictors were incorporated.
 This aligns with the findings related to point forecast accuracy, reinforcing the earlier explanations for these results.
 
-We observe that LSTM models exhibit a widespread error distribution compared to other models.
-This is attributed to their performance variability, which can significantly depend on the characteristics of demand patterns across different products.
-For example, the product "Amlodipine - 5mg - Tablet" demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero demand.
-Such patterns align well with the strengths of LSTM models, which excel at capturing long-term dependencies and managing complex temporal fluctuations.
-In contrast, the demand for "Anti-Rho (D)" is erratic and sporadic, characterized by frequent, random fluctuations without any discernible long-term trends or seasonal patterns.
-This chaotic nature, combined with inconsistent and often low volumes, poses significant challenges for LSTM models.
-The lack of clear structure or stable patterns limits the models' ability to generalize effectively, resulting in poorer forecast performance.
-Although LSTM models can handle irregular data to some extent, they require sufficient and well-distributed training samples to achieve optimal performance.
-In our study, the majority of products exhibit erratic and sporadic demand patterns similar to "Anti-Rho (D)", resulting in this widespread error distribution.
-
-<!-- In addition to reporting the forecast accuracy presented across different rolling origins, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-horizon.
-this is the accuracy reorted as the average across rolling origins and all products fro each given horizon.
-For both the point forecast and distributional accuracy  we can see that models incoporating information collected by domain experts improves compared to counterparts that that uses only the historical comsuptions.
--->
-
-<!-- Overall, our results indicate that forecast produces using models with infromation colelcted via the domain expert engagement provides reliable forecasts and improves upon the models relying solely on historcial consuption.
--->
+\textcolor{blue}{While LSTM models achieved the best overall forecast accuracy across products, their performance exhibited notable variability depending on the characteristics of individual demand patterns.
+For example, the product $Amlodipine - 5mg - Tablet$ demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero consumption.
+Such patterns align well with the strengths of univariate LSTM models, which are adept at capturing long-term dependencies and managing complex temporal fluctuations.
+In contrast, the demand for $Anti-Rho (D)$ is erratic and sparse, with frequent random fluctuations and little structural consistency.
+This lack of clear temporal patterns can make it challenging for LSTM models to learn generalizable signals, particularly given the limited data length available for training.
+Although LSTM can manage irregular data to some extent, it performs best when patterns are consistent or cyclic.
+These variations across product types contributed to the observed distribution of forecast errors across the product portfolio.
+Moreover, although we expect that multivariate LSTM models benefit from expert-informed predictors, our results show that the univariate LSTM consistently achieved better forecast performance.
+This outcome may be attributed to the nature of the predictors—binary, static, or weakly aligned with short-term temporal dynamics—which can disrupt rather than enhance learning when added to a neural network sensitive to input configuration.
+Moreover, with relatively short historical series and limited training samples, the inclusion of additional variables may have led to overfitting or reduced generalization.
+These findings suggest that while LSTM models can effectively learn temporal patterns from consumption data alone, incorporating structured external knowledge requires careful feature engineering and alignment to be beneficial.}
 
@@
 
-Our analysis also highlights the value of using foundational time series forecasting models like TimeGpt, particularly in developing countries and the global health and humanitarian sectors.
-In these contexts, the lack of advanced analytical skills and systematic data collection can hinder effective forecasting and decision-making.
+## Computational efficiency and resource considerations
+
+\textcolor{blue}{In addition to forecast accuracy, it is also important to consider the computational efficiency of each model, particularly in settings where computational resources and technical expertise are limited.}
+
+@tbl-runtime \textcolor{blue}{shows the total runtime required to train and generate forecasts across all 33 pharmaceutical product time series for each method for all origins.
+All models were implemented using R, except TimeGPT, which was run via Google Colab using a T4 GPU backend.
+All other models were executed on a local machine with 7 CPU cores} (11th Gen Intel(R) Core(TM) i5-1135G7 \@ 2.40 GHz and 8 GB RAM)
+
+```{r}
+#| label: tbl-runtime
+#| echo: false
+#| tbl-cap: "Computation time required for training and generating forecasts for each model across all products."
+#| tbl-pos: "H"
+runtime <- read.csv('../result/runtime_tbl.csv') |> 
+  janitor::clean_names() |> 
+  mutate(category = cut(run_time_in_seconds,
+                        breaks = c(-Inf, 100, 1000, Inf), # Adjust the breaks
+                        labels = c("low", "medium", "high"))) |> 
+  rename(Model = model,
+         'Runtime (Seconds)' = run_time_in_seconds,
+         'Runtime type' = runtime_type)
+
+runtime |> 
+  knitr::kable(booktabs = T, linesep = "") |>
+  kable_styling(latex_options = c("striped"), font_size = 11)
+```
+
+\textcolor{blue}{As shown in Figure 4 and 5, models that incorporated expert-informed regressors generally achieved better forecast accuracy compared to their univariate versions.
+This trend was most evident in classical models such as ARIMA and regression, where the inclusion of regressors led to a noticeable shift toward lower and more concentrated error distributions.
+However, this improvement came with an increase in computational cost.
+In all cases, the addition of regressors increased runtime—by 72\% in the regression model (from 27.48 to 47.38 seconds) and by 43\% in ARIMA (from 96.27 to 138.08 seconds).
+TimeGPT also showed a slight increase in runtime (from 2.57 to 3.72 seconds), although the total processing time remained extremely low overall.
+Moreover, LSTM exhibited the largest increase in runtime when regressors were added, rising from 6,892 to 7,234 seconds.
+This sharp increase reflects the sensitivity of neural networks to input configuration, especially in contexts with limited data, irregular demand, and noisy signals.}
+
+\textcolor{blue}{These results underscore the importance of balancing model performance with implementation cost.
+While deep learning models such as LSTM offer strong potential when well-tuned, they demand significantly more computational resources and may be less robust when integrating static or weakly aligned contextual features.
+By contrast, TimeGPT, a foundational model pretrained on large-scale time series data, provides a compelling alternative.
+It required less than 4 seconds to forecast all products, offered competitive accuracy, and required no tuning or retraining—making it well-suited for practical use in low-resource environments.}
 
@@
 Probabilistic forecasts offer a comprehensive approach by assigning likelihoods to a range of possible outcomes, recognizing that different consumption levels may occur with varying probabilities.
-The primary purpose of probabilistic forecasting, as illustrated in @fig-forecast-density-hstep, is to quantify and communicate uncertainty.
+\textcolor{blue}{The goal is to maximize the sharpness of these predictive distributions—i.e., how concentrated the forecasts are—while maintaining calibration, meaning that the predicted probabilities align well with actual outcomes} [@gneiting2014probabilistic].
+\textcolor{blue}{This approach allows decision-makers to fully leverage available information, incorporating uncertainty in a structured and measurable way.} The primary purpose, as illustrated in @fig-forecast-density-hstep, is to quantify and communicate uncertainty.
 This figure displays the forecast distribution of consumption over a 6-month horizon using a density plot.
@@
 The plot also includes the point forecast alongside 80% and 90% prediction intervals to illustrate potential variability.
+
+Probabilistic forecasts enhance planning and decision-making by offering a comprehensive view of potential future outcomes and their likelihood, rather than relying on a single point estimate.
+A probabilistic forecast takes the form of a predictive probability distribution over future quantities or events of interest.
 
@@
 ```
+
+\textcolor{blue}{To further illustrate the practical relevance, we consider an illustrative case where the mean forecasted consumption for the next month is 1,000 units, with a 90\% prediction interval ranging from 850 to 1,150 units.} @tbl-point_vs_probabilistic \textcolor{blue}{summarizes the decision outcomes under each approach.}
+
+\textcolor{blue}{Under a traditional approach, inventory decisions would rely solely on the point forecast.
+Based on the mean prediction of 1,000 units, the store manager would order precisely that quantity, assuming it would meet the expected demand.
+However, this approach does not incorporate any adjustment for uncertainty, potentially leading to stockouts if actual demand exceeds 1,000 units, or excess inventory if consumption is lower.}
+
+\textcolor{blue}{In contrast, utilizing the full probabilistic forecast allows the decision-maker to take variability into account.
+If a higher service level is required, for example 95\%, the inventory policy may be adjusted by stocking closer to the upper bound of the 90\% prediction interval (e.g., 1,150 units).
+If inventory holding costs are a major concern and the organization can tolerate a modest risk of shortage, the order quantity could remain closer to the median forecast of 1,000 units.}
+
+```{r}
+#| label: tbl-point_vs_probabilistic
+#| tbl-pos: "H"
+#| tbl-cap: "Comparison of ordering decisions under point and probabilistic forecasts - an illustrative example."
+
+library(kableExtra)
+df <- tibble(
+  `Forecast Type` = c("Point Forecast Only", "Probabilistic Forecast (95% service level)"),
+  `Ordering Quantity` = c("1,000 units", "1,150 units"),
+  `Risk Consideration` = c(
+    "No explicit consideration of uncertainty",
+    "Adjusts inventory to account for demand variability"
+  )
+)
+
+  df |> knitr::kable(booktabs = T, linesep = "") |>
+  kable_styling(latex_options = c("scale_down"), font_size = 11)
+
+```
+
+\textcolor{blue}{This illustrative comparison indicates that point forecasts provide a single estimate without adjusting for forecast uncertainty, while probabilistic forecasts allow decision-makers to explicitly align inventory decisions with risk tolerance and service level requirements.
+This is particularly important for pharmaceutical products, where the consequences of stockouts or overstocking can be significant both operationally and clinically.}
+
+## Managerial Implications
+
+\textcolor{blue}{This study offers actionable lessons for supply chain managers, public health planners, and policymakers navigating the uncertainty of pharmaceutical demand—particularly in systems like Ethiopia, where operational constraints and demand volatility are the norm, not the exception.}
+
+\textcolor{blue}{At present, national forecasting at the Ethiopian Pharmaceutical Supply Service (EPSS) still relies heavily on basic extrapolation tools—usually Excel sheets or donor-developed software such as the Quantification Analysis Tool (QAT).
+These tools are easy to use but limited: they assume stable demand, ignore uncertainty, and often miss the operational signals embedded in local experience.
+In practice, forecasts are sometimes adjusted based on gut feeling or anecdotal program insights—not because planners want to—but because the tools don’t offer a better alternative.}
+
+\textcolor{blue}{The models developed here—freely available, open-source, and designed to integrate directly with routine EPSS data—allow planners to make decisions using full probabilistic forecasts, not just single-point estimates.
+Forecasts are delivered with prediction intervals (e.g., 80\% or 90\%) that help teams decide not only how much to order, but how much risk they’re willing to tolerate.
+For products with known seasonality—like antimalarials or diagnostic kits—this matters.
+The system doesn’t just forecast a number; it gives planners a buffer strategy.}
+
+\textcolor{blue}{Beyond accuracy, the models also reflect how medicines actually move through the system.
+Predictors based on warehouse replenishments, fiscal year inventory routines, and known disease cycles are embedded into the forecasts—not hard-coded, but learned from the data in ways that are transparent and reproducible.
+These inputs are drawn from expert operational knowledge and can be updated or modified as the system evolves.}
+
+\textcolor{blue}{Importantly, the entire modeling pipeline is built in R and Python, and is already shared alongside code and data.
+This makes it immediately usable by EPSS analysts and regional logistics teams, even in settings without access to proprietary tools or high-end infrastructure.
+It's not a black box; it's something the system can own, modify, and grow with.}
+
+\textcolor{blue}{To support real uptake, we recommend five practical steps:}
+
+- \textcolor{blue}{Embed these probabilistic models into EPSS quantification rounds to replace rigid extrapolation methods and bring scenario-based planning into monthly and quarterly cycles.}
+
+- \textcolor{blue}{Use prediction intervals to define risk-informed buffer policies, especially for high-volatility products. Not every item needs the same margin of error.}
+
+- \textcolor{blue}{Automate the collection of domain knowledge  (e.g., replenishment events, stock counts, seasonal triggers) into the logistics data stream to reduce reliance on manual elicitation while preserving contextual intelligence.}
+
+- \textcolor{blue}{Train regional and central teams using the open-source tools and shared codebase, turning forecasting into a hands-on, in-country competency rather than a dependency on external tools or consultants.}
+
+- \textcolor{blue}{Establish routine forecast review sessions, supported by visual dashboards that reflect not just what the model says, but how uncertain that estimate is—and what that means for stock levels, procurement plans, and emergency readiness.}
+
+\textcolor{blue}{Taken together, these steps push forecasting beyond formulas and into real decision-making.
+}

### 2025-05-26 manuscript/05-conclusion.qmd
revision submitted

@@
 We evaluated both point and probabilistic forecasts using a range of models, from simple univariate approaches like ARIMA to complex models such as LSTM.
-Notably, we demonstrated that employing complex models does not necessarily lead to more accurate forecasts in the pharmaceutical supply context.
-Additionally, our research revealed that newly developed foundational time series forecasting models have potential applicability, particularly in environments where advanced analytical skills and knowledge are limited, as often seen in pharmaceutical supply chains in developing countries.
+We demonstrated that while the use of complex models may improve forecast accuracy in the pharmaceutical supply context, this comes at the cost of significantly increased computational time—an important consideration for resource-constrained settings in low- and middle-income countries (LMICs).
+Additionally, our research indicates that newly developed foundational time series forecasting models hold promise, particularly in settings with limited computational infrastructure and constrained access to advanced analytical expertise, as often seen in pharmaceutical supply chains in developing countries.
 
@@
 
-- The use of distribution-based representation of error metrics across different rolling origins or forecast horizons, such as boxplots, to showcase forecast accuracy provides essential insights into variability in accuracy at different point of time in data.
-Building on this, future work could aim at designing methodologies that identify diagnostic patterns within the distribution of errors.
-These approaches could suggest model improvements by analyzing error distributions and help practitioners to refine models based on error behavior.
+- \textcolor{blue}{Disruptions such as the COVID-19 pandemic and armed conflicts may alter consumption patterns in time series data, introducing irregularities that challenge the assumptions of traditional forecasting methods.
+In such contexts, it becomes important not only to model observed consumption but also to account for censored demand—i.e., the unmet need that would have been fulfilled under normal conditions and is essential for effective replenishment planning.
+Future research could focus on developing and evaluating forecasting strategies that are specifically designed to handle disrupted time series and estimate censored demand.
+In particular, methods that enhance the robustness of probabilistic forecasts in the presence of structural breaks or exogenous shocks are especially relevant.
+One promising direction is also the development of causal forecasting models that incorporate policy changes, epidemic dynamics, and socio-economic factors—an approach that may be particularly well suited to the realities of pharmaceutical supply chains in low- and middle-income countries (LMICs).}
 
-- While current study focus on monthly aggregate consumption at the organizational level, there is potential in exploring more granular forecasts.
-Future research could investigate daily or weekly forecasting and examine the impacts at hierarchical levels, such as specific sites or health facilities.
-Investigating how fine-grained temporal and hierarchical forecasts influence inventory management and operational efficiency could yield valuable insights for reducing stockouts, minimizing costs and reducing waste.
+- \textcolor{blue}{Future research should explore the integration of probabilistic forecasting into inventory management policies, with a focus on evaluating the operational impact of forecast uncertainty.
+In addition, there is substantial potential in leveraging more granular data—such as daily or weekly consumption patterns—and examining forecasts at hierarchical levels, including individual sites or health facilities.
+Investigating how fine-grained temporal and spatial forecasts influence decision-making in inventory control could provide valuable insights into reducing stockouts, minimizing holding costs, and decreasing pharmaceutical waste.}

### 2025-05-26 manuscript/main.qmd
revision submitted

@@
 title: "The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting"
-# author:
-#   - name: Arebu Issa Bilal
-#     email: arebu.issa@aau.edu.et
-#     affiliations:
-#         - id: Addis 
-#           name: Addis Ababa University
-#           department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
-#           country: Ethiopia.
-#     attributes:
-#         corresponding: true
-#   - name: Bahman Rostami-Tabar
-#     email: rostami-tabarb@cardiff.ac.uk
-#     affiliations:
-#         - id: carbs
-#           name: Cardiff University
-#           department: Data Lab for Social Good Reserach Group, Cardiff Business School 
-#           country: UK.
-#   - name: Harsha Halgamuwe Hewage
-#     email: HalgamuweHewageHR@cardiff.ac.uk
-#     affiliations:
-#         - ref: carbs
-#   - name: Umit Sezer Bititci 
-#     email: U.S.Bititci@hw.ac.uk
-#     affiliations:
-#         - id: herio 
-#           name: Heriot-Watt University
-#           department: Edinburgh Business School, Heriot-Watt University
-#           country: UK.
-#   - name: Teferi Gedif Fenta
-#     email: tgedif@gmail.com
-#     affiliations:
-#         - ref: Addis
+author:
+  - name: Arebu Issa Bilal
+    email: arebu.issa@aau.edu.et
+    affiliations:
+        - id: Addis
+          name: Addis Ababa University
+          department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
+          country: Ethiopia.
+    attributes:
+        corresponding: true
+  - name: Bahman Rostami-Tabar
+    email: rostami-tabarb@cardiff.ac.uk
+    affiliations:
+        - id: carbs
+          name: Cardiff University
+          department: Data Lab for Social Good Reserach Group, Cardiff Business School
+          country: UK.
+  - name: Harsha Halgamuwe Hewage
+    email: HalgamuweHewageHR@cardiff.ac.uk
+    affiliations:
+        - ref: carbs
+  - name: Umit Sezer Bititci
+    email: U.S.Bititci@hw.ac.uk
+    affiliations:
+        - id: herio
+          name: Heriot-Watt University
+          department: Edinburgh Business School, Heriot-Watt University
+          country: UK.
+  - name: Teferi Gedif Fenta
+    email: tgedif@gmail.com
+    affiliations:
+        - ref: Addis
 
@@
   prefer-html: true
-  cache: true
 format:
@@
       text: |
-        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
+        \usepackage{todonotes,mathtools,bm,amsmath,mathpazo,xcolor}
         \mathtoolsset{showonlyrefs}
         \setlength{\parindent}{0cm}
+        \usepackage{tikz}
+        \usetikzlibrary{positioning, shapes.geometric, arrows.meta, fit}
+        \tikzset{
+          process/.style={
+          rectangle,
+          draw=black,
+          thick,
+          text width=4.5cm,
+          align=center,
+          minimum height=1.4cm
+          },
+        groupbox/.style={
+          draw=black,
+          thick,
+          dashed,
+          inner sep=0.3cm,
+          rounded corners,
+          label={[black]90:#1}
+        },
+        arrow/.style={
+          thick,
+          ->,
+          >=Stealth
+        }
+                }
+
     journal:
@@
 ```{r results, child="04-results.qmd"}
-
 ```
@@
 
-We use R and Python for model development and accuracy evaluation. The paper is written in Quarto, and all code, data, and files will be shared in a GitHub repository upon acceptance for publication.
+To enhance transparency and reproducibility, we not only provide data and the code, but also the entire paper that is written in R & Python.
+All materials to reproduce this paper is available at [Insert Github link once paper accepted](Insert Github link once paper accepted).
+
+The repository contains the raw data, all R and Python scripts used in experiments, the results used in the paper, as well as the quarto files for producing this paper.
 

### 2025-05-26 manuscript/main.tex
revision submitted

@@
 \usepackage{xcolor}
-\usepackage{todonotes,mathtools,bm,amsmath,mathpazo}
+\usepackage{todonotes,mathtools,bm,amsmath,mathpazo,xcolor}
 \mathtoolsset{showonlyrefs}
 \setlength{\parindent}{0cm}
+\usepackage{tikz}
+\usetikzlibrary{positioning, shapes.geometric, arrows.meta, fit}
+\tikzset{
+  process/.style={
+  rectangle,
+  draw=black,
+  thick,
+  text width=4.5cm,
+  align=center,
+  minimum height=1.4cm
+  },
+groupbox/.style={
+  draw=black,
+  thick,
+  dashed,
+  inner sep=0.3cm,
+  rounded corners,
+  label={[black]90:#1}
+},
+arrow/.style={
+  thick,
+  ->,
+  >=Stealth
+}
+        }
 \makeatletter
@@
   pdftitle={The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting},
+  pdfauthor={Arebu Issa Bilal; Bahman Rostami-Tabar; Harsha Halgamuwe Hewage; Umit Sezer Bititci; Teferi Gedif Fenta},
   pdfkeywords={Forecasting, Pharmaceutical supply chain, Domain
@@
 pharmaceutical supply chain forecasting}
+\author[1]{Arebu Issa Bilal%
+\corref{cor1}%
+}
+ \ead{arebu.issa@aau.edu.et} 
+\author[2]{Bahman Rostami-Tabar%
+%
+}
+ \ead{rostami-tabarb@cardiff.ac.uk} 
+\author[2]{Harsha Halgamuwe Hewage%
+%
+}
+ \ead{HalgamuweHewageHR@cardiff.ac.uk} 
+\author[3]{Umit Sezer Bititci%
+%
+}
+ \ead{U.S.Bititci@hw.ac.uk} 
+\author[1]{Teferi Gedif Fenta%
+%
+}
+ \ead{tgedif@gmail.com}
+
+\affiliation[1]{organization={Addis Ababa University, Department of
+Pharmaceutics and Social Pharmacy, School of
+Pharmacy},country={Ethiopia.},countrysep={,},postcodesep={}}
+\affiliation[2]{organization={Cardiff University, Data Lab for Social
+Good Reserach Group, Cardiff Business
+School},country={UK.},countrysep={,},postcodesep={}}
+\affiliation[3]{organization={Heriot-Watt University, Edinburgh Business
+School, Heriot-Watt
+University},country={UK.},countrysep={,},postcodesep={}}
 
@@
 calculated using the ``STL'' (Seasonal and Trend decomposition using
-Loess) decomposition method, as described by \citet{mstl}.
+Loess) decomposition method, as described by \citet{bandara2025mstl}
 
@@
 
-\subsubsection{Collaborative expert review to identify factors affecting
+\begin{table}[H]
+
+\caption{\label{tbl-feature}Summary statistics of the pharmaceutical
+product time series}
+
+\centering{
+
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}{lrrrrrrrr}
+\toprule
+item & Mean & Median & Standard deviation & Miinum & Maximum & ACF lag 1 & zero\_run\_mean & nonzero\_squared\_cv\\
+\midrule
+Adrenaline (Epinephrine) - 0.1\% in 1ml Ampoule - Injection & 83077.42 & 69800.00 & 55257.59 & 0 & 204800 & 0.68 & 1.0 & 0.42\\
+Amlodipine - 5mg - Tablet & 330051.67 & 84350.00 & 473283.80 & 0 & 1495700 & 0.82 & 2.0 & 1.75\\
+Amoxicillin - 500mg - Capsule & 25880.97 & 25291.00 & 15751.99 & 1137 & 75305 & 0.30 & 0.0 & 0.37\\
+Anti-Rho (D) & 1911.10 & 1983.00 & 1475.71 & 30 & 8477 & 0.11 & 0.0 & 0.60\\
+Artemether + Lumefanthrine & 7364.28 & 953.00 & 17714.45 & 0 & 88469 & 0.64 & 6.5 & 4.33\\
+Atenolol - 50mg - Tablet & 3695.75 & 3312.50 & 2587.84 & 4 & 12661 & 0.27 & 0.0 & 0.49\\
+Atrovastatin - 20mg - Tablet & 3618.97 & 1302.50 & 5506.25 & 0 & 27973 & 0.44 & 1.0 & 2.26\\
+Ceftriaxone & 542999.03 & 553782.50 & 346067.19 & 0 & 1285708 & 0.47 & 2.0 & 0.36\\
+Dextrose & 312489.42 & 296130.00 & 255750.02 & 0 & 1004850 & 0.53 & 5.0 & 0.39\\
+Frusemide - 10mg/ml in 2ml Ampoule - Injection & 176436.17 & 175360.00 & 137620.74 & 17880 & 940930 & 0.34 & 0.0 & 0.61\\
+Gentamicin & 201360.50 & 182040.00 & 181410.03 & 0 & 654000 & 0.64 & 2.0 & 0.75\\
+Hydralazine - 20mg/ml in 1ml Ampoule - Injection & 2952.17 & 2456.00 & 2741.23 & 6 & 15713 & 0.57 & 0.0 & 0.86\\
+Insulin Isophane Human(Suspension) & 66717.83 & 73816.00 & 46364.04 & 0 & 155631 & 0.60 & 11.0 & 0.21\\
+Insulin Soluble Human & 8973.33 & 8890.50 & 7140.69 & 0 & 28443 & 0.53 & 11.0 & 0.33\\
+Insuline Isophane Biphasic & 12795.87 & 11402.50 & 8475.27 & 860 & 38586 & 0.45 & 0.0 & 0.44\\
+Lamivudine  + Efavirenz +Tenofovir & 145178.95 & 167890.50 & 133576.98 & 0 & 550780 & 0.61 & 11.0 & 0.51\\
+Lamivudine + Zidovudine & 26784.80 & 22579.00 & 22565.25 & 0 & 89910 & 0.55 & 11.0 & 0.39\\
+Lidocaine HCL & 80100.70 & 62702.00 & 105600.48 & 1000 & 791050 & 0.12 & 0.0 & 1.74\\
+Magnesium Sulphate & 3573.83 & 1832.50 & 4465.78 & 0 & 22649 & 0.49 & 11.0 & 1.09\\
+Medroxyprogesterone & 501663.00 & 560692.00 & 328276.14 & 0 & 984774 & 0.65 & 11.0 & 0.16\\
+Metformin - 500mg - Tablet & 26505.25 & 20294.75 & 21695.60 & 654 & 86705 & 0.42 & 0.0 & 0.67\\
+Omeperazole - 20 mg - Capsule (Enclosing Enteric Coated Granules) & 46368.55 & 40012.00 & 38379.32 & 0 & 176219 & 0.62 & 6.0 & 0.52\\
+Omeperazole - 4mg/ml in 10ml - Injection & 11438.88 & 3345.50 & 17643.52 & 0 & 81355 & 0.55 & 5.0 & 1.82\\
+Oral Rehydration Salt & 507220.42 & 558426.00 & 394581.24 & 0 & 1384257 & 0.70 & 6.0 & 0.28\\
+Oxytocin & 180155.23 & 151480.00 & 164309.63 & 0 & 846060 & 0.32 & 11.0 & 0.49\\
+Pentavalent & 422071.73 & 474297.50 & 238940.77 & 0 & 944819 & 0.78 & 11.0 & 0.08\\
+Propylthiouracil - 100mg - Tablet & 7044.46 & 6426.50 & 6112.89 & 0 & 28473 & 0.44 & 2.0 & 0.69\\
+RHZ (Rifampicin + Isoniazid + Pyrazinamide) & 2260.05 & 1759.00 & 2233.00 & 0 & 7095 & 0.70 & 8.0 & 0.45\\
+Rapid Diagnostic Test & 377843.65 & 360187.50 & 264551.71 & 0 & 1004325 & 0.63 & 11.0 & 0.21\\
+Ringer's Injection & 88896.27 & 95227.00 & 54255.34 & 390 & 251283 & 0.53 & 0.0 & 0.37\\
+Sodium Chloride (Normal Saline) & 300706.82 & 289171.50 & 136772.04 & 60759 & 702526 & 0.36 & 0.0 & 0.21\\
+Sulphamethoxazole + Trimethoprim - (200mg + 40mg)/5ml - Suspension & 84216.90 & 65917.50 & 72963.22 & 20 & 259512 & 0.59 & 0.0 & 0.75\\
+Tetracycline - 1\% - Eye Ointment & 199370.65 & 172721.50 & 139041.86 & 7556 & 585987 & 0.02 & 0.0 & 0.49\\
+\bottomrule
+\end{tabular}}
+
+}
+
+\end{table}%
+
+Table~\ref{tbl-feature} also provides the summary statistics computed
+for each pharmaceutical product time series include traditional
+distributional measures---mean, median, standard deviation, minimum, and
+maximum---alongside time series--specific and structural
+characteristics. These include the first-order autocorrelation (ACF at
+lag 1), the mean length of consecutive zero runs (zero\_run\_mean), and
+the squared coefficient of variation computed only on non-zero values
+(nonzero\_squared\_cv). Together, these metrics capture not only central
+tendency and dispersion, but also temporal dependence, sparsity
+patterns, and relative variability in periods with non-zero consumption.
+
+\subsection{Collaborative expert review to identify factors affecting
 consumption}\label{collaborative-expert-review-to-identify-factors-affecting-consumption}
 
-To understand the external factors influencing product consumption, we
-began by visualizing the time series data for consumption as highlighted
-in the previous section. This initial step helped identify specific
-instances within the data that showed unusual patterns, such as peaks,
-unusually high or low observations, and consecutive periods of low
-consumption. These observations indicated potential events or factors
-affecting consumption levels.
-
-To gather domain knowledge about factors affected the consumption of
-products, we contacted experts from EPSS who work in the distribution
-unit. Next, we collaborated with six experts, each with over 10 years of
-experience managing pharmaceutical distribution. In these sessions, we
-reviewed the time series of all products together to understand the
-potential drivers behind these anomalies. The experts provided insights
-based on their extensive experience and knowledge of distribution
-practices. When there was uncertainty or gaps in the data for certain
-products, additional experts were consulted to gather further context.
-If all experts reached a consensus, the influencing factors were
-confirmed and included in the analysis.
-
-As part of this process, we examined the electronic bin cards for 33
-pharmaceutical products and conducted detailed data exploration. A bin
-card is a physical or electronic record used to track the inventory
-levels of individual items in a storage location, such as a warehouse or
-stockroom. The term `bin' refers to the specific storage location for an
-item. Information typically recorded on a bin card includes the Item ID,
-unit of measure, product description, account details, date, references,
-beginning balance, manufacturer, batch number, expiry date, quantity
-received, quantity issued, and balance. In summary, a bin card is a
-fundamental tool in inventory management, especially in environments
-where physical stock tracking is essential. It provides a
-straightforward and immediate way to monitor item quantities, helping to
-ensure effective inventory management. Over a period of two weeks, we
-meticulously traced consumption data through these bin cards. Moreover,
-We ensured accuracy by converting dates from the Ethiopian calendar to
-the Gregorian calendar and excluded transactions related to internal
-storage or warehouse relocation within EPSS. Through this collaborative
-approach, we were able to identify several predictors affecting
-distribution and consumption that are summarised as followings:
+\textcolor{blue}{To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation.
+In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.}
+
+Figure \ref{fig:expert-pipeline}
+\textcolor{blue}{presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.}
+
+\begin{figure}[htbp]
+\centering
+\begin{tikzpicture}[scale=0.85, transform shape, node distance=1.3cm and 2cm]
+
+% Node style
+\tikzset{
+  process/.style={
+    rectangle,
+    draw=black,
+    thick,
+    text width=4.5cm,
+    align=center,
+    minimum height=1.4cm
+  },
+  arrow/.style={
+    thick,
+    ->,
+    >=Stealth
+  }
+}
+
+% Nodes
+\node (step1) [process] {1. Expert Identification (with EPSS)};
+\node (step2) [process, below=of step1] {2. Onboarding \& Orientation};
+\node (step3) [process, below=of step2] {3. Workshop: Collaborative Data Review \& Joint Identification of Anomalies};
+
+\node (step4) [process, right=of step1] {4. Cross Validation Using Bin Cards \& Inventory Records};
+\node (step5) [process, below=of step4] {5. Structured Expert Elicitation \& Contextual Clarification};
+
+\node (step6) [process, right=of step4] {6. Expert Elicitation of Predictor Relevance};
+\node (step7) [process, below=of step6] {7. Consensus-Based Selection of Predictors};
+\node (step8) [process, below=of step7] {8. Variable Encoding \& Time Alignment};
+
+\node (step9) [process, below=of step8, yshift=-0.3cm] {9. Integration into Forecasting Models};
+
+% Arrows
+\draw [arrow] (step1) -- (step2);
+\draw [arrow] (step2) -- (step3);
+
+\path (step3.east) -- +(0.6,0) coordinate (s3stub);
+\draw [arrow] (step3.east) -- (s3stub) |- (step4.west);
+
+\draw [arrow] (step4) -- (step5);
+
+\path (step5.east) -- +(0.6,0) coordinate (s5stub);
+\draw [arrow] (step5.east) -- (s5stub) |- (step6.west);
+
+\draw [arrow] (step6) -- (step7);
+\draw [arrow] (step7) -- (step8);
+\draw [arrow] (step8) -- (step9);
+
+% Group boxes
+\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step1)(step2)(step3), label={[black]90:Expert Engagement}] {};
+\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step4)(step5), label={[black]90:Validation}] {};
+\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step6)(step7)(step8), label={[black]90:Feature Construction}] {};
+\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step9), label={[black]90:Model Integration}] {};
+
+\end{tikzpicture}
+\caption{Diagram illustrating the sequential steps involved in collecting, structuring, and utilizing domain knowledge for modeling.}
+\label{fig:expert-pipeline}
+\end{figure}
+
+\textcolor{blue}{Through a facilitated half-day workshop, these experts reviewed five years of monthly consumption data for 33 pharmaceutical products.
+Anomalous patterns—such as consumption spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles).
+These insights were cross-validated against bin cards and warehouse records, and internal transfers were excluded to isolate externally relevant events.
+Dates were standardized to the Gregorian calendar.}
+
+\textcolor{blue}{Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor.
+The final predictors included binary indicators for stock replenishment events, categorical variables for fiscal year inventory counts, and dummy variables for malaria seasonality.
+A further description of these variables are summarized as followings:}
 
 \begin{itemize}
-\tightlist
 \item
-  Stock replenishment: refers to the process of restocking or refilling
-  inventory to ensure that there are sufficient quantities of products
-  or materials available to meet demand. Whenever there was stock
-  replenishment at the central EPSS, consumption and distribution to
-  hubs and health facilities increased. This increase was attributed to
-  the need to restock depleted inventories and the push from central
-  EPSS to manage space constraints.
+  \textcolor{blue}{Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
+Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased.
+This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.}
+\item
+  \textcolor{blue}{Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
+The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
+Physical inventory counting periods also influenced consumption.
+Stores closed during these periods, halting transactions.
+We observed increased consumption before inventory counting periods, as hubs and facilities stocked up.
+July and August were identified as physical counting periods each year.}
+\item
+  \textcolor{blue}{Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
+In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease.
+Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced.
+During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments.
+Malaria seasonality was another significant predictor.
+Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks.
+We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.}
 \end{itemize}
 
-\begin{itemize}
-\tightlist
-\item
-  Physical fiscal year inventory counting: refers to the process of
-  manually counting and verifying the actual quantities of
-  pharmaceutical products available in stock at a specific location. The
-  process is critical for maintaining the accuracy of inventory records,
-  ensuring that medicines are available when needed, and preventing
-  stockouts or overstocking. Physical inventory counting periods also
-  influenced consumption. Stores closed during these periods, halting
-  transactions. We observed increased consumption before inventory
-  counting periods, as hubs and facilities stocked up. July and August
-  were identified as physical counting periods each year.
-\end{itemize}
-
-\begin{itemize}
-\tightlist
-\item
-  Malaria seasonality: Refers to the predictable patterns and
-  fluctuations in malaria incidence throughout the year, typically
-  influenced by climate and environmental conditions. In many regions,
-  malaria transmission peaks during and shortly after the rainy season,
-  when conditions such as stagnant water pools create ideal breeding
-  sites for the Anopheles mosquitoes that transmit the disease.
-  Conversely, malaria cases often decline during the dry season when
-  mosquito breeding sites are reduced. During peak malaria seasons,
-  there is a significant surge in the demand for antimalarial drugs and
-  other related treatments. Malaria seasonality was another significant
-  predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine
-  and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We
-  identified epidemic periods affecting consumption: September to
-  December 2017, March to May 2018, September to December 2018, March to
-  May 2019, September to December 2019, March to May 2020, September to
-  December 2020, March to May 2021, September to December 2021, and
-  March to May 2022.
-\end{itemize}
-
-The key predictors identified are deterministic values, since both past
-and future values of these predictors are known in advance. We then
-integrated them into the consumption data, creating the complete dataset
-required for model building and running the experiment.
+\textcolor{blue}{These predictors, encoded with known historical and future values, were added to the modeling dataset.
+This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.}
 
@@
 
+\textcolor{blue}{To improve the robustness of the LSTM models and address overfitting issues, we introduced dropout regularization layers after the LSTM units and employed early stopping based on validation loss during model training.
+Each LSTM model was trained independently for each product series.
+Hyperparameter tuning was performed in a preliminary phase using a subset of products.
+Various configurations of LSTM units (30–100), dense units (50–200), dropout rates (0.1–0.5), and batch sizes (16–64) were evaluated based on validation set performance.
+The final model structure — 50 LSTM units, 100 dense units, a 0.2 dropout rate, and a batch size of 32 — was selected as a trade-off between forecast accuracy and model stability across different demand patterns.}
+
 \textbf{TimeGPT}: TimeGPT is the first pre-trained foundational model
@@
 
-In all models, probabilistic forecasts were generated using
-bootstrapping to create 1,000 possible future scenarios for each period
-within the forecast horizon \citep{hyndman2021forecasting}.
+\subsection{Generating probabilistic
+forecasts}\label{generating-probabilistic-forecasts}
+
+\textcolor{blue}{In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical consumption.
+Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction}
+\citep{wang2023}.
+
+\textcolor{blue}{In this study, we employed a bootstrapping approach to construct predictive intervals.
+Bootstrapping was chosen primarily for its flexibility and model-agnostic nature, allowing it to be applied uniformly across the diverse range of forecasting methods implemented without requiring strong distributional assumptions.
+Moreover, pharmaceutical consumption data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.}
+
+\textcolor{blue}{Specifically, we assume that future forecast errors will be similar to past forecast errors.
+The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed consumption, and $\hat{y}_t$ denotes the corresponding forecast estimate.
+To simulate future consumption paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates.
+This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.}
+
+\textcolor{blue}{Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts}
+\citep{hyndman2021forecasting}.
+
+\textcolor{blue}{The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.}
 
@@
 
-To assess the performance of our forecasting methods, we split the data
-into a series of training and test sets and apply time series
-cross-validation with a forecast horizon of 6 months. Each training set
-is expanded in monthly increments, allowing model development and
-hyperparameter tuning to be performed on the training data, with errors
-evaluated on the corresponding test sets. We assess forecasting
-performance using both point and probabilistic error measures.
+\textcolor{blue}{To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation}
+\citep{hyndman2021forecasting}.
+\textcolor{blue}{Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods.
+In our setup, the initial training set consisted of all available historical data from December 2017 up to June 2021.
+The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans consumption over a one-year horizon.
+At each iteration, models were trained on an expanding training window and evaluated over a fixed 6-month forecast horizon, aligned with typical EPSS planning cycles.
+After each forecast generation, the training set was expanded by one additional month, and the process was repeated, allowing for rolling assessment across the final 12 months of data.
+This structure ensured that forecasts reflected realistic operational scenarios, where forecasts are continuously updated as new data becomes available.
+This cross-validation design allowed us to evaluate each model's ability to perform across multiple different forecast origins and a variety of demand conditions, providing a more reliable and generalizable understanding of model performance.
+Furthermore, all model development steps, including hyperparameter tuning for the LSTM models, were conducted exclusively using the training data available at each iteration.
+No test set information was used during model selection or tuning.}
 
@@
 rewards sharpness and penalizes miscalibration, so it measures overall
-performance of the forecast distribution.
+performance of the forecast distribution. For probabilistic evaluation,
+1,000 future paths were simulated per series, enabling robust estimation
+of forecast uncertainty.
+
 \begin{equation}\phantomsection\label{eq-CRPS}{
@@
 exhibiting the lowest error positioned at the bottom. This model is the
-Ariam model, which incorporates exogenous variables. Additionally,
-Figure~\ref{fig-point} indicates that predictors obtained through
-interactions with domain experts enhance point forecast accuracy across
-most models. This underscores the critical importance of systematically
-collecting such expert-informed data alongside transactional consumption
-data.
+LSTM model, followed by ARIMA, which incorporate exogenous variables.
+Additionally, Figure~\ref{fig-point} indicates that predictors obtained
+through interactions with domain experts enhance point forecast accuracy
+across most models. This underscores the critical importance of
+systematically collecting such expert-informed data alongside
+transactional consumption data.
 
@@
 forecast accuracy for most models, enhancing not only point forecasts
-but also probabilistic forecasts. Notably, ARIMA achieved the highest
-probabilistic forecast accuracy, while LSTM models had the lowest. This
-aligns with the findings related to point forecast accuracy, reinforcing
-the earlier explanations for these results.
-
-We observe that LSTM models exhibit a widespread error distribution
-compared to other models. This is attributed to their performance
-variability, which can significantly depend on the characteristics of
-demand patterns across different products. For example, the product
-``Amlodipine - 5mg - Tablet'' demonstrates periods of extreme
-variability, with spikes in demand followed by periods of very low or
-zero demand. Such patterns align well with the strengths of LSTM models,
-which excel at capturing long-term dependencies and managing complex
-temporal fluctuations. In contrast, the demand for ``Anti-Rho (D)'' is
-erratic and sporadic, characterized by frequent, random fluctuations
-without any discernible long-term trends or seasonal patterns. This
-chaotic nature, combined with inconsistent and often low volumes, poses
-significant challenges for LSTM models. The lack of clear structure or
-stable patterns limits the models' ability to generalize effectively,
-resulting in poorer forecast performance. Although LSTM models can
-handle irregular data to some extent, they require sufficient and
-well-distributed training samples to achieve optimal performance. In our
-study, the majority of products exhibit erratic and sporadic demand
-patterns similar to ``Anti-Rho (D)'', resulting in this widespread error
-distribution.
+but also probabilistic forecasts. Notably, LSTM and ARIMA yielded the
+most accurate probabilistic forecasts when identified predictors were
+incorporated. This aligns with the findings related to point forecast
+accuracy, reinforcing the earlier explanations for these results.
+
+\textcolor{blue}{While LSTM models achieved the best overall forecast accuracy across products, their performance exhibited notable variability depending on the characteristics of individual demand patterns.
+For example, the product $Amlodipine - 5mg - Tablet$ demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero consumption.
+Such patterns align well with the strengths of univariate LSTM models, which are adept at capturing long-term dependencies and managing complex temporal fluctuations.
+In contrast, the demand for $Anti-Rho (D)$ is erratic and sparse, with frequent random fluctuations and little structural consistency.
+This lack of clear temporal patterns can make it challenging for LSTM models to learn generalizable signals, particularly given the limited data length available for training.
+Although LSTM can manage irregular data to some extent, it performs best when patterns are consistent or cyclic.
+These variations across product types contributed to the observed distribution of forecast errors across the product portfolio.
+Moreover, although we expect that multivariate LSTM models benefit from expert-informed predictors, our results show that the univariate LSTM consistently achieved better forecast performance.
+This outcome may be attributed to the nature of the predictors—binary, static, or weakly aligned with short-term temporal dynamics—which can disrupt rather than enhance learning when added to a neural network sensitive to input configuration.
+Moreover, with relatively short historical series and limited training samples, the inclusion of additional variables may have led to overfitting or reduced generalization.
+These findings suggest that while LSTM models can effectively learn temporal patterns from consumption data alone, incorporating structured external knowledge requires careful feature engineering and alignment to be beneficial.}
 
@@
 
-Our analysis also highlights the value of using foundational time series
-forecasting models like TimeGpt, particularly in developing countries
-and the global health and humanitarian sectors. In these contexts, the
-lack of advanced analytical skills and systematic data collection can
-hinder effective forecasting and decision-making.
+\subsection{Computational efficiency and resource
+considerations}\label{computational-efficiency-and-resource-considerations}
+
+\textcolor{blue}{In addition to forecast accuracy, it is also important to consider the computational efficiency of each model, particularly in settings where computational resources and technical expertise are limited.}
+
+Table~\ref{tbl-runtime}
+\textcolor{blue}{shows the total runtime required to train and generate forecasts across all 33 pharmaceutical product time series for each method for all origins.
+All models were implemented using R, except TimeGPT, which was run via Google Colab using a T4 GPU backend.
+All other models were executed on a local machine with 7 CPU cores}
+(11th Gen Intel(R) Core(TM) i5-1135G7 @ 2.40 GHz and 8 GB RAM)
+
+\begin{table}[H]
+
+\caption{\label{tbl-runtime}Computation time required for training and
+generating forecasts for each model across all products.}
+
+\centering{
+
+\centering\begingroup\fontsize{11}{13}\selectfont
+
+\begin{tabular}{lrll}
+\toprule
+Model & Runtime (Seconds) & Runtime type & category\\
+\midrule
+\cellcolor{gray!6}{LSTM} & \cellcolor{gray!6}{6892.30} & \cellcolor{gray!6}{CPU with 7 cores} & \cellcolor{gray!6}{high}\\
+LSTM with regressors & 7234.00 & CPU with 7 cores & high\\
+\cellcolor{gray!6}{Regression} & \cellcolor{gray!6}{27.48} & \cellcolor{gray!6}{CPU with 7 cores} & \cellcolor{gray!6}{low}\\
+Regression with regressors & 47.38 & CPU with 7 cores & low\\
+\cellcolor{gray!6}{ARIMA} & \cellcolor{gray!6}{96.27} & \cellcolor{gray!6}{CPU with 7 cores} & \cellcolor{gray!6}{low}\\
+ARIMA with regressors & 138.08 & CPU with 7 cores & medium\\
+\cellcolor{gray!6}{TimeGPT} & \cellcolor{gray!6}{2.57} & \cellcolor{gray!6}{Colab T4 GPU} & \cellcolor{gray!6}{low}\\
+TimeGPT with regressors & 3.72 & Colab T4 GPU & low\\
+\cellcolor{gray!6}{ETS} & \cellcolor{gray!6}{75.71} & \cellcolor{gray!6}{CPU with 7 cores} & \cellcolor{gray!6}{low}\\
+\bottomrule
+\end{tabular}
+\endgroup{}
+
+}
+
+\end{table}%
+
+\textcolor{blue}{As shown in Figure 4 and 5, models that incorporated expert-informed regressors generally achieved better forecast accuracy compared to their univariate versions.
+This trend was most evident in classical models such as ARIMA and regression, where the inclusion of regressors led to a noticeable shift toward lower and more concentrated error distributions.
+However, this improvement came with an increase in computational cost.
+In all cases, the addition of regressors increased runtime—by 72\% in the regression model (from 27.48 to 47.38 seconds) and by 43\% in ARIMA (from 96.27 to 138.08 seconds).
+TimeGPT also showed a slight increase in runtime (from 2.57 to 3.72 seconds), although the total processing time remained extremely low overall.
+Moreover, LSTM exhibited the largest increase in runtime when regressors were added, rising from 6,892 to 7,234 seconds.
+This sharp increase reflects the sensitivity of neural networks to input configuration, especially in contexts with limited data, irregular demand, and noisy signals.}
+
+\textcolor{blue}{These results underscore the importance of balancing model performance with implementation cost.
+While deep learning models such as LSTM offer strong potential when well-tuned, they demand significantly more computational resources and may be less robust when integrating static or weakly aligned contextual features.
+By contrast, TimeGPT, a foundational model pretrained on large-scale time series data, provides a compelling alternative.
+It required less than 4 seconds to forecast all products, offered competitive accuracy, and required no tuning or retraining—making it well-suited for practical use in low-resource environments.}
 
@@
 likelihoods to a range of possible outcomes, recognizing that different
-consumption levels may occur with varying probabilities. The primary
-purpose of probabilistic forecasting, as illustrated in
+consumption levels may occur with varying probabilities.
+\textcolor{blue}{The goal is to maximize the sharpness of these predictive distributions—i.e., how concentrated the forecasts are—while maintaining calibration, meaning that the predicted probabilities align well with actual outcomes}
+\citep{gneiting2014probabilistic}.
+\textcolor{blue}{This approach allows decision-makers to fully leverage available information, incorporating uncertainty in a structured and measurable way.}
+The primary purpose, as illustrated in
 Figure~\ref{fig-forecast-density-hstep}, is to quantify and communicate
@@
 intervals to illustrate potential variability.
+
+Probabilistic forecasts enhance planning and decision-making by offering
+a comprehensive view of potential future outcomes and their likelihood,
+rather than relying on a single point estimate. A probabilistic forecast
+takes the form of a predictive probability distribution over future
+quantities or events of interest.
 
@@
 \end{figure}%
+
+\textcolor{blue}{To further illustrate the practical relevance, we consider an illustrative case where the mean forecasted consumption for the next month is 1,000 units, with a 90\% prediction interval ranging from 850 to 1,150 units.}
+Table~\ref{tbl-point_vs_probabilistic}
+\textcolor{blue}{summarizes the decision outcomes under each approach.}
+
+\textcolor{blue}{Under a traditional approach, inventory decisions would rely solely on the point forecast.
+Based on the mean prediction of 1,000 units, the store manager would order precisely that quantity, assuming it would meet the expected demand.
+However, this approach does not incorporate any adjustment for uncertainty, potentially leading to stockouts if actual demand exceeds 1,000 units, or excess inventory if consumption is lower.}
+
+\textcolor{blue}{In contrast, utilizing the full probabilistic forecast allows the decision-maker to take variability into account.
+If a higher service level is required, for example 95\%, the inventory policy may be adjusted by stocking closer to the upper bound of the 90\% prediction interval (e.g., 1,150 units).
+If inventory holding costs are a major concern and the organization can tolerate a modest risk of shortage, the order quantity could remain closer to the median forecast of 1,000 units.}
+
+\begin{table}[H]
+
+\caption{\label{tbl-point_vs_probabilistic}Comparison of ordering
+decisions under point and probabilistic forecasts - an illustrative
+example.}
+
+\centering{
+
+\centering\begingroup\fontsize{11}{13}\selectfont
+
+\resizebox{\linewidth}{!}{
+\begin{tabular}{lll}
+\toprule
+Forecast Type & Ordering Quantity & Risk Consideration\\
+\midrule
+Point Forecast Only & 1,000 units & No explicit consideration of uncertainty\\
+Probabilistic Forecast (95\% service level) & 1,150 units & Adjusts inventory to account for demand variability\\
+\bottomrule
+\end{tabular}}
+\endgroup{}
+
+}
+
+\end{table}%
+
+\textcolor{blue}{This illustrative comparison indicates that point forecasts provide a single estimate without adjusting for forecast uncertainty, while probabilistic forecasts allow decision-makers to explicitly align inventory decisions with risk tolerance and service level requirements.
+This is particularly important for pharmaceutical products, where the consequences of stockouts or overstocking can be significant both operationally and clinically.}
+
+\subsection{Managerial Implications}\label{managerial-implications}
+
+\textcolor{blue}{This study offers actionable lessons for supply chain managers, public health planners, and policymakers navigating the uncertainty of pharmaceutical demand—particularly in systems like Ethiopia, where operational constraints and demand volatility are the norm, not the exception.}
+
+\textcolor{blue}{At present, national forecasting at the Ethiopian Pharmaceutical Supply Service (EPSS) still relies heavily on basic extrapolation tools—usually Excel sheets or donor-developed software such as the Quantification Analysis Tool (QAT).
+These tools are easy to use but limited: they assume stable demand, ignore uncertainty, and often miss the operational signals embedded in local experience.
+In practice, forecasts are sometimes adjusted based on gut feeling or anecdotal program insights—not because planners want to—but because the tools don’t offer a better alternative.}
+
+\textcolor{blue}{The models developed here—freely available, open-source, and designed to integrate directly with routine EPSS data—allow planners to make decisions using full probabilistic forecasts, not just single-point estimates.
+Forecasts are delivered with prediction intervals (e.g., 80\% or 90\%) that help teams decide not only how much to order, but how much risk they’re willing to tolerate.
+For products with known seasonality—like antimalarials or diagnostic kits—this matters.
+The system doesn’t just forecast a number; it gives planners a buffer strategy.}
+
+\textcolor{blue}{Beyond accuracy, the models also reflect how medicines actually move through the system.
+Predictors based on warehouse replenishments, fiscal year inventory routines, and known disease cycles are embedded into the forecasts—not hard-coded, but learned from the data in ways that are transparent and reproducible.
+These inputs are drawn from expert operational knowledge and can be updated or modified as the system evolves.}
+
+\textcolor{blue}{Importantly, the entire modeling pipeline is built in R and Python, and is already shared alongside code and data.
+This makes it immediately usable by EPSS analysts and regional logistics teams, even in settings without access to proprietary tools or high-end infrastructure.
+It's not a black box; it's something the system can own, modify, and grow with.}
+
+\textcolor{blue}{To support real uptake, we recommend five practical steps:}
+
+\begin{itemize}
+\item
+  \textcolor{blue}{Embed these probabilistic models into EPSS quantification rounds to replace rigid extrapolation methods and bring scenario-based planning into monthly and quarterly cycles.}
+\item
+  \textcolor{blue}{Use prediction intervals to define risk-informed buffer policies, especially for high-volatility products. Not every item needs the same margin of error.}
+\item
+  \textcolor{blue}{Automate the collection of domain knowledge  (e.g., replenishment events, stock counts, seasonal triggers) into the logistics data stream to reduce reliance on manual elicitation while preserving contextual intelligence.}
+\item
+  \textcolor{blue}{Train regional and central teams using the open-source tools and shared codebase, turning forecasting into a hands-on, in-country competency rather than a dependency on external tools or consultants.}
+\item
+  \textcolor{blue}{Establish routine forecast review sessions, supported by visual dashboards that reflect not just what the model says, but how uncertain that estimate is—and what that means for stock levels, procurement plans, and emergency readiness.}
+\end{itemize}
+
+\textcolor{blue}{Taken together, these steps push forecasting beyond formulas and into real decision-making.
+}
 
@@
 point and probabilistic forecasts using a range of models, from simple
-univariate approaches like ARIMA to complex models such as LSTM.
-Notably, we demonstrated that employing complex models does not
-necessarily lead to more accurate forecasts in the pharmaceutical supply
-context. Additionally, our research revealed that newly developed
-foundational time series forecasting models have potential
-applicability, particularly in environments where advanced analytical
-skills and knowledge are limited, as often seen in pharmaceutical supply
-chains in developing countries.
+univariate approaches like ARIMA to complex models such as LSTM. We
+demonstrated that while the use of complex models may improve forecast
+accuracy in the pharmaceutical supply context, this comes at the cost of
+significantly increased computational time---an important consideration
+for resource-constrained settings in low- and middle-income countries
+(LMICs). Additionally, our research indicates that newly developed
+foundational time series forecasting models hold promise, particularly
+in settings with limited computational infrastructure and constrained
+access to advanced analytical expertise, as often seen in pharmaceutical
+supply chains in developing countries.
 
@@
 \item
-  The use of distribution-based representation of error metrics across
-  different rolling origins or forecast horizons, such as boxplots, to
-  showcase forecast accuracy provides essential insights into
-  variability in accuracy at different point of time in data. Building
-  on this, future work could aim at designing methodologies that
-  identify diagnostic patterns within the distribution of errors. These
-  approaches could suggest model improvements by analyzing error
-  distributions and help practitioners to refine models based on error
-  behavior.
+  \textcolor{blue}{Disruptions such as the COVID-19 pandemic and armed conflicts may alter consumption patterns in time series data, introducing irregularities that challenge the assumptions of traditional forecasting methods.
+In such contexts, it becomes important not only to model observed consumption but also to account for censored demand—i.e., the unmet need that would have been fulfilled under normal conditions and is essential for effective replenishment planning.
+Future research could focus on developing and evaluating forecasting strategies that are specifically designed to handle disrupted time series and estimate censored demand.
+In particular, methods that enhance the robustness of probabilistic forecasts in the presence of structural breaks or exogenous shocks are especially relevant.
+One promising direction is also the development of causal forecasting models that incorporate policy changes, epidemic dynamics, and socio-economic factors—an approach that may be particularly well suited to the realities of pharmaceutical supply chains in low- and middle-income countries (LMICs).}
 \item
-  While current study focus on monthly aggregate consumption at the
-  organizational level, there is potential in exploring more granular
-  forecasts. Future research could investigate daily or weekly
-  forecasting and examine the impacts at hierarchical levels, such as
-  specific sites or health facilities. Investigating how fine-grained
-  temporal and hierarchical forecasts influence inventory management and
-  operational efficiency could yield valuable insights for reducing
-  stockouts, minimizing costs and reducing waste.
+  \textcolor{blue}{Future research should explore the integration of probabilistic forecasting into inventory management policies, with a focus on evaluating the operational impact of forecast uncertainty.
+In addition, there is substantial potential in leveraging more granular data—such as daily or weekly consumption patterns—and examining forecasts at hierarchical levels, including individual sites or health facilities.
+Investigating how fine-grained temporal and spatial forecasts influence decision-making in inventory control could provide valuable insights into reducing stockouts, minimizing holding costs, and decreasing pharmaceutical waste.}
 \end{itemize}
@@
 
-We use R and Python for model development and accuracy evaluation. The
-paper is written in Quarto, and all code, data, and files will be shared
-in a GitHub repository upon acceptance for publication.
+To enhance transparency and reproducibility, we not only provide data
+and the code, but also the entire paper that is written in R \& Python.
+All materials to reproduce this paper is available at
+\href{Insert\%20Github\%20link\%20once\%20paper\%20accepted}{Insert
+Github link once paper accepted}.
+
+The repository contains the raw data, all R and Python scripts used in
+experiments, the results used in the paper, as well as the quarto files
+for producing this paper.
 

### 2025-05-26 rscript/03_lstm_revised.R

@@ -1,0 +1,325 @@
+# Load libraries
+library(keras)
+library(keras3)
+library(tensorflow)
+library(tidyverse)
+library(tsibble)
+library(doParallel)
+library(foreach)
+library(abind)
+library(janitor)
+
+# Set seed for reproducibility
+set.seed(123)
+
+# Read and clean data
+med_qty <- read.csv('data/predicotrs_final_tssible_July_04_2024.csv') |>
+  clean_names()
+
+# Prepare tsibble
+med_qty_tsb <- med_qty |>
+  mutate(month = yearmonth(month)) |>
+  group_by(item) |>
+  mutate(item_id = sprintf("id_%02d", cur_group_id())) |>
+  ungroup() |>
+  as_tsibble(index = month, key = c(item_id, item)) |>
+  select(-stock_out, -covid_19, -conflict)
+
+# Set up train/test
+f_horizon <- 6
+test_length <- 12
+
+med_tsb_train_tscv <- med_qty_tsb |>
+  filter_index(. ~ as.character(max(med_qty_tsb$month) - f_horizon)) |>
+  stretch_tsibble(.init = (length(unique(med_qty_tsb$month)) - test_length), .step = 1) |>
+  mutate(unique_id = paste0(item_id, '-', .id)) |>
+  rename(horizon = .id)
+
+med_tsb_test_tscv <- med_qty_tsb |>
+  filter_index(as.character(max(med_qty_tsb$month) - (test_length) + 1) ~ .) |>
+  slide_tsibble(.size = 6, .step = 1) |>
+  mutate(unique_id = paste0(item_id, '-', .id)) |>
+  rename(horizon = .id)
+
+# Define helper functions
+create_sequences <- function(data, target_column, time_steps) {
+  X <- array(dim = c(0, time_steps, ncol(data)))
+  Y <- NULL
+  for (i in 1:(nrow(data) - time_steps)) {
+    sequence <- array(as.matrix(data[i:(i + time_steps - 1), ]), dim = c(1, time_steps, ncol(data)))
+    X <- abind::abind(X, sequence, along = 1)
+    Y <- c(Y, data[i + time_steps, target_column])
+  }
+  return(list(X = X, Y = Y))
+}
+
+normalize <- function(x) {
+  (x - min(x)) / (max(x) - min(x))
+}
+
+rescale <- function(x, min, max) {
+  x * (max - min) + min
+}
+
+calculate_residuals <- function(model, x_data, y_data) {
[…]

### 2025-08-04 manuscript/01-introduction.qmd
3RD REVISION

@@
 This investment is essential and offers significant benefits for improving supply chain delivery.
-However, much of the data being collected focuses on transactional data, such as consumption and sale.
+However, much of the data being collected focuses on transactional data, such as consumption, distribution, and sale.
 Almost all platforms collecting data tend to overlook the critical contextual factors and events that influence these transactions—factors that are just as crucial for accurate forecasting.
 This presents a major barrier, as understanding the broader context, including administration procedures, local policies, disruptions, conflicts, and so on is vital for building reliable forecasting models.
-This gap—known as business context and domain knowledge—requires urgent attention in Pharmaceutical supply chain forecasting.
+This gap — known as business context and domain knowledge — requires urgent attention in pharmaceutical supply chain forecasting.
 Without incorporating this knowledge, any forecasting model is unlikely to produce reliable results.
 We must prioritize collecting and integrating this contextual information to improve the accuracy of demand forecasts in pharmaceutical supply chains.
-Combining statistical models with domain insights offers a more refined forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview].
+Integrating statistical models with domain knowledge leads to a more accurate and context-aware forecasting process, as evidenced by research that highlights the value of judgmental adjustments to enhance forecast accuracy [@fildes2007against; @taylor2022artificial; @soyiri2013overview].
 Domain knowledge enables analysts to incorporate relevant events, such as supply chain disruptions, into the forecasting model.
@@
 
-This research uses consumption data from 33 health commodities, covering the period from December 2017 to July 2022, provided by the Ethiopian Pharmaceutical Supply Service (EPSS), to develop models for six-month-ahead forecasting.
-Initially, we implement univariate models that rely solely on consumption data.
-Following this, we conduct exploratory data analysis and engage in interviews with EPSS experts to gather domain knowledge on events influencing consumption patterns.
-Through these interviews, the experts identified three key factors collectively impacting consumption: stock replenishment period, fiscal calendar, and the seasonal malaria.
-We then develop forecasting models that incorporate the data gathered through the interview process and compare their performance with univariate models that rely solely on consumption data.
+This research uses distribution data from 33 health commodities, covering the period from December 2017 to July 2022, provided by the Ethiopian Pharmaceutical Supply Service (EPSS), to develop models for six-month-ahead forecasting.
+Initially, we implement univariate models that rely solely on distribution data.
+Following this, we conduct exploratory data analysis and engage in interviews with EPSS experts to gather domain knowledge on events influencing distribution patterns.
+Through these interviews, the experts identified three key factors collectively impacting distribution: stock replenishment period, fiscal calendar, and the seasonal malaria.
+We then develop forecasting models that incorporate the data gathered through the interview process and compare their performance with univariate models that rely solely on distribution data.
 

### 2025-08-04 manuscript/02-literature.qmd
3RD REVISION

@@
 
-Accurate sales forecasting is crucial in the pharmaceutical industry, where it directly impacts profit maximization, cost minimization, and the ability to respond to market changes.
+Accurate demand forecasting is crucial in the pharmaceutical industry, where it directly impacts profit maximization, cost minimization, and the ability to respond to market changes.
 Forecasting in healthcare not only influences clinical decisions but also plays a pivotal role in supply chain management, ensuring that the right drugs are available when needed.
@@
 
-In this review, a range of forecasting methods has been applied, from traditional statistical models to advanced machine learning techniques, with many studies using a combination of simple and advanced methods to enhance accuracy [@nikolopoulos2016forecasting;@zhu2021demand;@anusha2014demand].
-Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods like the Holt-Winters method, Single Moving Average (SMA), and Single Exponential Smoothing (SES), which are simpler but effective for certain applications.
+A range of methods is used in pharmaceutical supply chain forecasting, from traditional statistical models to advanced machine learning techniques, with many studies combining both simple and advanced approaches to improve accuracy [@nikolopoulos2016forecasting;@zhu2021demand;@anusha2014demand].
+Some studies, such as those by @newberne2006holt, and @restyana2021analysis, applied traditional forecasting methods.
 In contrast, other researchers explored more advanced approaches, including hybrid models and machine learning techniques, to address complex forecasting challenges, as seen in the studies by @siddiqui2022hybrid, @de2021lead, @kim2015demand, @candan2014demand, and @ribeiro2017improving.
 
-Early research in pharmaceutical forecasting primarily focused on traditional statistical methods.
-These methods, including ARIMA, exponential smoothing, and moving averages, were widely used due to their simplicity and effectiveness in relatively stable environments [@zahra2019forecasting].
+Early research in pharmaceutical forecasting primarily focused on classical statistical methods.
+These methods, including ARIMA (AutoRegressive Integrated Moving Average), exponential smoothing, and moving averages, were widely used due to their simplicity and effectiveness in relatively stable environments [@zahra2019forecasting].
 These models typically assess forecast accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean Squared Error (MSE), and Root Mean Squared Error (RMSE).
@@
 Their findings highlighted the method's utility in short-term planning and resource management, validated through metrics like MAPE, MAD, and MSD.
-Similarly, @restyana2021analysis compared Single Moving Average (SMA) and Single Exponential Smoothing (SES) in drug demand forecasting, concluding that SES provided more accurate results, as indicated by lower MAD and MSE values.
+Similarly, @restyana2021analysis compared  Simple Moving Average (SMA), and Single Exponential Smoothing (SES) in drug demand forecasting, concluding that SES provided more accurate results, as indicated by lower MAD and MSE values.
 
-Recognizing the limitations of traditional models, particularly in capturing the complexities of pharmaceutical demand (e.g., seasonality, external influences), researchers began developing hybrid models.
-These models combine linear methods like ARIMA with nonlinear techniques such as neural networks, aiming to improve forecasting accuracy by addressing diverse data patterns.
-@khalil2014intelligent introduced a hybrid model that integrated ARIMA with neural networks.
-Their study demonstrated that this approach significantly outperformed traditional models, especially in scenarios with limited historical data.
-The hybrid model's effectiveness was confirmed by improvements in key performance indicators like MAPE and RMSE.
-
-The pharmaceutical industry's increasing complexity and the availability of large datasets have led to the adoption of advanced machine learning techniques.
+Recognizing the limitations of traditional models—particularly their inability to capture the complexities of pharmaceutical demand, such as seasonality and external influences—researchers have increasingly turned to more advanced modeling approaches [@khalil2014intelligent].
+Furthermore, the growing complexity of the pharmaceutical industry, coupled with the availability of large datasets, has accelerated the adoption of advanced machine learning techniques.
 These methods, such as Support Vector Regression (SVR), Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer greater flexibility and predictive power.
@@
 1.
-Despite significant investment in data collection technologies and logistics management systems, these systems primarily collect transactional data, such as consumption.
-However, no data is collected on events that influence consumption variability—information that would be crucial for building reliable forecasting models.
+Despite significant investment in data collection technologies and logistics management systems, these systems primarily collect transactional data, such as consumption and distribution.
+However, no data is collected on events that influence their variability — information that would be crucial for building reliable forecasting models.
 No studies to date have explored how domain expertise could enhance pharmaceutical demand forecasting.
@@
 Current research on forecasting for pharmaceutical supply chain predominantly focuses on generating point forecasts.
-There is a lack of studies that consider the entire forecast distribution of monthly consumption, which would better capture the uncertainty of future demand and provide a valuable risk management tool for decision-makers.
+There is a lack of studies that consider the entire forecast distribution, which would better capture the uncertainty of future demand and provide a valuable risk management tool for decision-makers.
 

### 2025-08-04 manuscript/03-experiment.qmd
3RD REVISION

@@
 
-For this study, we use a dataset spanning five years (Dec.
-2017- July 2022) of consumption data obtained from EPSS.
+For this study, we use a dataset spanning five years (December 2017 - July 2022) of distribution data obtained from EPSS.
 Rigorous checks were conducted to ensure the consistency and completeness of the collected data.
@@
 
-### Explaoratory Data Analysis with consumption data
+### Exploratory Data Analysis with distribution data
 
-Given the number of products in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understadn data.
+Given the number of products in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understand data.
 @fig-feature shows the strength of trend versus seasonality.
@@
 #| fig.align: center
-#| fig-cap: "The strength of the trend and seasonality in the time series of pharmaceutical product consumption. The scatter plot shows 33 data points, with each point corresponding to a product."
+#| fig-cap: "The strength of the trend and seasonality in the time series of pharmaceutical product distribution. The scatter plot shows 33 data points, with each point corresponding to a product."
 
@@
 A number of products show pronounced trends, and only a few demonstrate clear seasonal patterns.
-Beyond assessing the strength of trends and seasonality, we also visualized all time series to understand data and various patterns, including trends and instances of erratic consumption behavior during certain months.
-For example, some series show low consumption volumes over consecutive months, followed by peak consumption in specific months, making them more volatile and difficult to forecast.
-This underscores the diversity of monthly pharmaceutical time series patterns within the dataset and highlights the importance of understanding the factors driving these consumption behaviors.
-Figure @fig-dataviz2 illustrates the time plots for a few selected products.
+Beyond assessing the strength of trends and seasonality, we also visualized all time series to understand data and various patterns, including trends and instances of erratic distribution behavior during certain months.
+For example, some series show low distribution volumes over consecutive months, followed by peak distribution in specific months, making them more volatile and difficult to forecast.
+This underscores the diversity of monthly pharmaceutical time series patterns within the dataset and highlights the importance of understanding the factors driving these distribution behaviors.
+@fig-dataviz2 illustrates the time plots for a few selected products.
 
@@
 #| fig-height: 6
-#| fig-cap: "Monthly time plot of consumption.
+#| fig-cap: "Monthly time plot of distribution.
 X-axis shows the month, consisting of 60 data points (months) and y-axis shows the comsumption.
-The panels display data from four products to give a glimpse of the consumption patterns."
+The panels display data from four products to give a glimpse of the distribution patterns."
 
@@
   scale_x_yearmonth(date_breaks = "4 month")+
-  labs(x = "Month", y = "Consumption", title = "Oxytocin")+
+  labs(x = "Month", y = "Distribution", title = "Oxytocin")+
   ggthemes::theme_few()+
@@
   scale_x_yearmonth(date_breaks = "4 month")+
-    labs(x = "Month", y = "Consumption",title = "Amoxicillin - 500mg - Capsule")+
+    labs(x = "Month", y = "Distribution",title = "Amoxicillin - 500mg - Capsule")+
   ggthemes::theme_few()+
@@
   scale_x_yearmonth(date_breaks = "4 month")+
-    labs(x = "Month", y = "Consumption",title = "Ceftriaxone")+
+    labs(x = "Month", y = "Distribution",title = "Ceftriaxone")+
   ggthemes::theme_few()+
@@
   scale_x_yearmonth(date_breaks = "4 month")+
-    labs(x = "Month", y = "Consumption", title = "Atenolol - 50mg - Tablet")+
+    labs(x = "Month", y = "Distribution", title = "Atenolol - 50mg - Tablet")+
   ggthemes::theme_few()+
@@
 
-Visualizing the data revealed that various events significantly impacted the consumption of different products, but these were not reflected in the available data.
+Visualizing the data revealed that various events significantly impacted the distribution of different products, but these were not reflected in the available data.
 Expert insights were crucial for understanding the nature and timing of these events, filling gaps in the system, and incorporating qualitative factors that EPSS logistics system miss.
-Therefore, we conducted interviews with domain experts to collect information on external factors that influence consumption.
-These interviews allowed us to account for unusual patterns and customize our forecasting models to more accurately reflect the unique consumption behaviors of each product.
+Therefore, we conducted interviews with domain experts to collect information on external factors that influence distribution.
+These interviews allowed us to account for unusual patterns and customize our forecasting models to more accurately reflect the unique distribution behaviors of each product.
 
@@
 These include the first-order autocorrelation (ACF at lag 1), the mean length of consecutive zero runs (zero_run_mean), and the squared coefficient of variation computed only on non-zero values (nonzero_squared_cv).
-Together, these metrics capture not only central tendency and dispersion, but also temporal dependence, sparsity patterns, and relative variability in periods with non-zero consumption.
+Together, these metrics capture not only central tendency and dispersion, but also temporal dependence, sparsity patterns, and relative variability in periods with non-zero distribution.
 
-## Collaborative expert review to identify factors affecting consumption
+## Collaborative expert review to identify factors affecting distribution
 
-\textcolor{blue}{To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation.
-In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.}
+To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation.
+In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.
 
-Figure \ref{fig:expert-pipeline} \textcolor{blue}{presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.}
+Figure \ref{fig:expert-pipeline} presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.
 
@@
 
-\textcolor{blue}{Through a facilitated half-day workshop, these experts reviewed five years of monthly consumption data for 33 pharmaceutical products.
-Anomalous patterns—such as consumption spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles).
+Through a facilitated half-day workshop, these experts reviewed five years of monthly distribution data for 33 pharmaceutical products.
+Anomalous patterns—such as distribution spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles).
 These insights were cross-validated against bin cards and warehouse records, and internal transfers were excluded to isolate externally relevant events.
-Dates were standardized to the Gregorian calendar.}
+Dates were standardized to the Gregorian calendar.
 
-\textcolor{blue}{Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor.
+Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor.
 The final predictors included binary indicators for stock replenishment events, categorical variables for fiscal year inventory counts, and dummy variables for malaria seasonality.
-A further description of these variables are summarized as followings:}
+A further description of these variables is summarized as followings:
 
-- \textcolor{blue}{Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
-Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased.
-This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.}
+- Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
+Whenever there was stock replenishment at the central EPSS, distribution to hubs and health facilities increased.
+This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.
 
-- \textcolor{blue}{Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
+- Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
 The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
-Physical inventory counting periods also influenced consumption.
+Physical inventory counting periods also influenced distribution.
 Stores closed during these periods, halting transactions.
-We observed increased consumption before inventory counting periods, as hubs and facilities stocked up.
-July and August were identified as physical counting periods each year.}
+We observed increased distribution before inventory counting periods, as hubs and facilities stocked up.
+July and August were identified as physical counting periods each year.
 
-- \textcolor{blue}{Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
+- Malaria seasonality: refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
 In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease.
@@
 Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks.
-We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.}
+We identified two recurrent epidemic seasons each year—March to May and September to December—that affected distribution from 2017 through 2022.
 
-\textcolor{blue}{These predictors, encoded with known historical and future values, were added to the modeling dataset.
-This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.}
+These predictors, encoded with known historical and future values, were added to the modeling dataset.
+This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.
 
@@
 The trend component can be specified as none ("N"), additive ("A"), or damped additive ("Ad"); the seasonality can be none ("N"), additive ("A"), or multiplicative ("M"); and the error term can be additive ("A") or multiplicative ("M").
-To forecast consumption, we use the ETS() function from the fable package in R, which automatically selects the optimal model for each time series based on the corrected Akaike’s Information Criterion (AICc).
+To forecast distribution, we use the ETS() function from the fable package in R, which automatically selects the optimal model for each time series based on the corrected Akaike’s Information Criterion (AICc).
 In our study, an automated algorithm determines the best configuration for trend, seasonality, and error components across each time series, leveraging the ets() function’s use of AIC to identify the optimal model.
@@
 
-**Multiple Linear Regression (MLR)**: We use Multiple linear regression to model the relationship between a consumption and potential variables influencing its variation.
+**Multiple Linear Regression (MLR)**: We use Multiple linear regression to model the relationship between a distribution and potential variables influencing its variation.
 In our first model, we use multiple linear regression with a trend component to capture the underlying progression over time, i.e., _regression_.
@@
 We then extend this model by introducing additional predictors that include variables such as replenishment cycles, fiscal year indicators, and periods with malaria prevalence.
-These additional predictors allow the model to see if capturing external factors can provide a better understanding of the factors influencing the consumption and result at enhanced accuracy.
+These additional predictors allow the model to see if capturing external factors can provide a better understanding of the factors influencing the distribution and result at enhanced accuracy.
 We produce forecasts using TLSM() function from the fable package in R.
@@
 **ARIMA and ARIMA with regressors**: ARIMA (AutoRegressive Integrated Moving Average) is a powerful statistical model designed to forecast time series by capturing temporal dynamics and are widely used in time series forecasting due to their ability to model complex trends and patterns over time without relying on external predictors.
-ARIMAX (AutoRegressive Integrated Moving Average with eXogenous variables) extends ARIMA by incorporating external variables, or exogenous predictors, into the model.
+ARIMAX (Auto Regressive Integrated Moving Average with eXogenous variables) extends ARIMA by incorporating external variables, or exogenous predictors, into the model.
 This modification allows to include relevant information from external factors such as malaria season, and fiscal year period, and stock replenishement period that may explain variations in the series beyond its internal time dynamics, we refer to this in the result as _arima_reg_.
@@
 
-\textcolor{blue}{To improve the robustness of the LSTM models and address overfitting issues, we introduced dropout regularization layers after the LSTM units and employed early stopping based on validation loss during model training.
+To improve the robustness of the LSTM models and address overfitting issues, we introduced dropout regularization layers after the LSTM units and employed early stopping based on validation loss during model training.
 Each LSTM model was trained independently for each product series.
@@
 Various configurations of LSTM units (30–100), dense units (50–200), dropout rates (0.1–0.5), and batch sizes (16–64) were evaluated based on validation set performance.
-The final model structure — 50 LSTM units, 100 dense units, a 0.2 dropout rate, and a batch size of 32 — was selected as a trade-off between forecast accuracy and model stability across different demand patterns.}
+The final model structure — 50 LSTM units, 100 dense units, a 0.2 dropout rate, and a batch size of 32 — was selected as a trade-off between forecast accuracy and model stability across different demand patterns.
 
@@
 
-\textcolor{blue}{In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical consumption.
-Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction} [@wang2023].
+In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical distribution Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction [@wang2023].
 
-\textcolor{blue}{In this study, we employed a bootstrapping approach to construct predictive intervals.
+In this study, we employed a bootstrapping approach to construct predictive intervals.
 Bootstrapping was chosen primarily for its flexibility and model-agnostic nature, allowing it to be applied uniformly across the diverse range of forecasting methods implemented without requiring strong distributional assumptions.
-Moreover, pharmaceutical consumption data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.}
+Moreover, pharmaceutical distribution data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.
 
-\textcolor{blue}{Specifically, we assume that future forecast errors will be similar to past forecast errors.
-The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed consumption, and $\hat{y}_t$ denotes the corresponding forecast estimate.
-To simulate future consumption paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates.
-This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.}
+Specifically, we assume that future forecast errors will be similar to past forecast errors.
+The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed distribution, and $\hat{y}_t$ denotes the corresponding forecast estimate.
+To simulate future distribution paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates.
+This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.
 
-\textcolor{blue}{Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts} [@hyndman2021forecasting].
+Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts [@hyndman2021forecasting].
 
-\textcolor{blue}{The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.}
+The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.
 
@@
 
-\textcolor{blue}{To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation} [@hyndman2021forecasting].
-\textcolor{blue}{Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods.
+To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation [@hyndman2021forecasting].
+Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods.
 In our setup, the initial training set consisted of all available historical data from December 2017 up to June 2021.
-The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans consumption over a one-year horizon.
+The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans distribution over a one-year horizon.
 At each iteration, models were trained on an expanding training window and evaluated over a fixed 6-month forecast horizon, aligned with typical EPSS planning cycles.
@@
 Furthermore, all model development steps, including hyperparameter tuning for the LSTM models, were conducted exclusively using the training data available at each iteration.
-No test set information was used during model selection or tuning.}
+No test set information was used during model selection or tuning.
 
@@
 
-To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@hyndman2021forecasting].
+To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score (CRPS) [@hyndman2021forecasting].
 It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution.

### 2025-08-04 manuscript/04-results.qmd
3RD REVISION

@@
 
-In this section, we compare the forecasting performance of various approaches, examining models that incorporate expert-identified business context predictors versus those that rely solely on historical consumption data.
+In this section, we compare the forecasting performance of various approaches, examining models that incorporate expert-identified business context predictors versus those that rely solely on historical distribution data.
 Point forecast performance is reported using MASE and RMSSE, while probabilistic forecast accuracy is reported using CRPS.
@@
 Additionally, @fig-point indicates that predictors obtained through interactions with domain experts enhance point forecast accuracy across most models.
-This underscores the critical importance of systematically collecting such expert-informed data alongside transactional consumption data.
+This underscores the critical importance of systematically collecting such expert-informed data alongside transactional distribution data.
 
@@
   geom_boxplot()+
-  labs(y = "Models")+
+  labs(y = "Models", x="MASE")+
   ggthemes::theme_few()
@@
   geom_boxplot()+
-  labs(y = "Models")+
+  labs(y = "Models", x="RMSSE")+
   ggthemes::theme_few()
@@
   geom_boxplot()+
-  labs(y = "Models")+
+  labs(y = "Models", x="CRPS")+
   ggthemes::theme_few()
@@
 
-\textcolor{blue}{While LSTM models achieved the best overall forecast accuracy across products, their performance exhibited notable variability depending on the characteristics of individual demand patterns.
-For example, the product $Amlodipine - 5mg - Tablet$ demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero consumption.
+While LSTM models achieved the best overall forecast accuracy across products, their performance exhibited notable variability depending on the characteristics of individual demand patterns.
+For example, the product $Amlodipine - 5mg - Tablet$ demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero distribution.
 Such patterns align well with the strengths of univariate LSTM models, which are adept at capturing long-term dependencies and managing complex temporal fluctuations.
@@
 Moreover, with relatively short historical series and limited training samples, the inclusion of additional variables may have led to overfitting or reduced generalization.
-These findings suggest that while LSTM models can effectively learn temporal patterns from consumption data alone, incorporating structured external knowledge requires careful feature engineering and alignment to be beneficial.}
+These findings suggest that while LSTM models can effectively learn temporal patterns from distribution data alone, incorporating structured external knowledge requires careful feature engineering and alignment to be beneficial.
 
 The findings emphasize the importance of incorporating relevant domain knowledge into forecasting models.
-In pharmaceutical supply logistics administration systems, data often consists solely of transactional records on consumption and distribution.
+In pharmaceutical supply logistics administration systems, data often consists solely of transactional records on distribution and distribution.
 Including exogenous variables such as administrative procedures, seasonal patterns, or conflicts, or any other relevant factor that could be identified by those with domain knowledge proved essential for reducing forecast errors.
 This approach improved both point and probabilistic forecast accuracy, enabling a more confident assessment of uncertainty.
-Therefore, the systematic collection of information about significant events and their impact on consumption is vital.
-Recording details of events such as policy changes, administration procedures, conflicts, and incorporating this information into forecasting models creates a comprehensive understanding of consumption.
+Therefore, the systematic collection of information about significant events and their impact on distribution is vital.
+Recording details of events such as policy changes, administration procedures, conflicts, and incorporating this information into forecasting models creates a comprehensive understanding of distribution.
 This practice enhances modeling reliability and allows institutions in developing countries and humanitarian organizations to better forecast demand, allocate resources effectively, and respond proactively.
@@
 
-\textcolor{blue}{In addition to forecast accuracy, it is also important to consider the computational efficiency of each model, particularly in settings where computational resources and technical expertise are limited.}
+In addition to forecast accuracy, it is also important to consider the computational efficiency of each model, particularly in settings where computational resources and technical expertise are limited.
 
-@tbl-runtime \textcolor{blue}{shows the total runtime required to train and generate forecasts across all 33 pharmaceutical product time series for each method for all origins.
+@tbl-runtime shows the total runtime required to train and generate forecasts across all 33 pharmaceutical product time series for each method for all origins.
 All models were implemented using R, except TimeGPT, which was run via Google Colab using a T4 GPU backend.
-All other models were executed on a local machine with 7 CPU cores} (11th Gen Intel(R) Core(TM) i5-1135G7 \@ 2.40 GHz and 8 GB RAM)
+All other models were executed on a local machine with 7 CPU cores (11th Gen Intel(R) Core(TM) i5-1135G7 \@ 2.40 GHz and 8 GB RAM)
 
@@
 
-\textcolor{blue}{As shown in Figure 4 and 5, models that incorporated expert-informed regressors generally achieved better forecast accuracy compared to their univariate versions.
+As shown in Figure 4 and 5, models that incorporated expert-informed regressors generally achieved better forecast accuracy compared to their univariate versions.
 This trend was most evident in classical models such as ARIMA and regression, where the inclusion of regressors led to a noticeable shift toward lower and more concentrated error distributions.
@@
 Moreover, LSTM exhibited the largest increase in runtime when regressors were added, rising from 6,892 to 7,234 seconds.
-This sharp increase reflects the sensitivity of neural networks to input configuration, especially in contexts with limited data, irregular demand, and noisy signals.}
+This sharp increase reflects the sensitivity of neural networks to input configuration, especially in contexts with limited data, irregular demand, and noisy signals.
 
-\textcolor{blue}{These results underscore the importance of balancing model performance with implementation cost.
+These results underscore the importance of balancing model performance with implementation cost.
 While deep learning models such as LSTM offer strong potential when well-tuned, they demand significantly more computational resources and may be less robust when integrating static or weakly aligned contextual features.
 By contrast, TimeGPT, a foundational model pretrained on large-scale time series data, provides a compelling alternative.
-It required less than 4 seconds to forecast all products, offered competitive accuracy, and required no tuning or retraining—making it well-suited for practical use in low-resource environments.}
+It required less than 4 seconds to forecast all products, offered competitive accuracy, and required no tuning or retraining—making it well-suited for practical use in low-resource environments.
 
-## An illustration of probabilistic forecast for Pharmacuitical product consumption
+## An illustration of probabilistic forecast for Pharmaceutical product distribution
 
-In this section, we present an illustrative example of a probabilistic forecast for future consumption of _Sodium Chloride (Normal Saline)_ product.
+In this section, we present an illustrative example of a probabilistic forecast for future distribution of _Sodium Chloride (Normal Saline)_ product.
 Due to the complexity of including such visualizations for all products, only one example is shown here.
@@
 The future is inherently uncertain, and effective planning requires considering alternative scenarios.
-Probabilistic forecasts offer a comprehensive approach by assigning likelihoods to a range of possible outcomes, recognizing that different consumption levels may occur with varying probabilities.
-\textcolor{blue}{The goal is to maximize the sharpness of these predictive distributions—i.e., how concentrated the forecasts are—while maintaining calibration, meaning that the predicted probabilities align well with actual outcomes} [@gneiting2014probabilistic].
-\textcolor{blue}{This approach allows decision-makers to fully leverage available information, incorporating uncertainty in a structured and measurable way.} The primary purpose, as illustrated in @fig-forecast-density-hstep, is to quantify and communicate uncertainty.
-This figure displays the forecast distribution of consumption over a 6-month horizon using a density plot.
+Probabilistic forecasts offer a comprehensive approach by assigning likelihoods to a range of possible outcomes, recognizing that different distribution levels may occur with varying probabilities.
+The goal is to maximize the sharpness of these predictive distributions—i.e., how concentrated the forecasts are—while maintaining calibration, meaning that the predicted probabilities align well with actual outcomes [@gneiting2014probabilistic].
+This approach allows decision-makers to fully leverage available information, incorporating uncertainty in a structured and measurable way.
+The primary purpose, as illustrated in @fig-forecast-density-hstep, is to quantify and communicate uncertainty.
+This figure displays the forecast distribution of distribution over a 6-month horizon using a density plot.
 For each month within the forecast period, a separate distribution is generated.
@@
 A single-point forecast cannot inherently provide the probabilistic context needed to capture the range of possible outcomes.
-Although prediction intervals can indicate a range of potential values, they do not convey the detailed probabilities of low or high consumption.
+Although prediction intervals can indicate a range of potential values, they do not convey the detailed probabilities of low or high distribution.
 This distinction highlights the value of probabilistic forecasting in supporting informed decision-making by offering a clearer view of future uncertainties.
@@
 
-\textcolor{blue}{To further illustrate the practical relevance, we consider an illustrative case where the mean forecasted consumption for the next month is 1,000 units, with a 90\% prediction interval ranging from 850 to 1,150 units.} @tbl-point_vs_probabilistic \textcolor{blue}{summarizes the decision outcomes under each approach.}
+To further illustrate the practical relevance, we consider an illustrative case where the mean forecasted distribution for the next month is 1,000 units, with a 90\% prediction interval ranging from 850 to 1,150 units.
+@tbl-point_vs_probabilistic summarizes the decision outcomes under each approach.
 
-\textcolor{blue}{Under a traditional approach, inventory decisions would rely solely on the point forecast.
+Under a traditional approach, inventory decisions would rely solely on the point forecast.
 Based on the mean prediction of 1,000 units, the store manager would order precisely that quantity, assuming it would meet the expected demand.
-However, this approach does not incorporate any adjustment for uncertainty, potentially leading to stockouts if actual demand exceeds 1,000 units, or excess inventory if consumption is lower.}
+However, this approach does not incorporate any adjustment for uncertainty, potentially leading to stockouts if actual demand exceeds 1,000 units, or excess inventory if distribution is lower.
 
-\textcolor{blue}{In contrast, utilizing the full probabilistic forecast allows the decision-maker to take variability into account.
+In contrast, utilizing the full probabilistic forecast allows the decision-maker to take variability into account.
 If a higher service level is required, for example 95\%, the inventory policy may be adjusted by stocking closer to the upper bound of the 90\% prediction interval (e.g., 1,150 units).
-If inventory holding costs are a major concern and the organization can tolerate a modest risk of shortage, the order quantity could remain closer to the median forecast of 1,000 units.}
+If inventory holding costs are a major concern and the organization can tolerate a modest risk of shortage, the order quantity could remain closer to the median forecast of 1,000 units.
 
@@
 
-\textcolor{blue}{This illustrative comparison indicates that point forecasts provide a single estimate without adjusting for forecast uncertainty, while probabilistic forecasts allow decision-makers to explicitly align inventory decisions with risk tolerance and service level requirements.
-This is particularly important for pharmaceutical products, where the consequences of stockouts or overstocking can be significant both operationally and clinically.}
+This illustrative comparison indicates that point forecasts provide a single estimate without adjusting for forecast uncertainty, while probabilistic forecasts allow decision-makers to explicitly align inventory decisions with risk tolerance and service level requirements.
+This is particularly important for pharmaceutical products, where the consequences of stockouts or overstocking can be significant both operationally and clinically.
 
@@
 
-\textcolor{blue}{This study offers actionable lessons for supply chain managers, public health planners, and policymakers navigating the uncertainty of pharmaceutical demand—particularly in systems like Ethiopia, where operational constraints and demand volatility are the norm, not the exception.}
+This study offers actionable lessons for supply chain managers, public health planners, and policymakers navigating the uncertainty of pharmaceutical demand—particularly in systems like Ethiopia, where operational constraints and demand volatility are the norm, not the exception.
 
-\textcolor{blue}{At present, national forecasting at the Ethiopian Pharmaceutical Supply Service (EPSS) still relies heavily on basic extrapolation tools—usually Excel sheets or donor-developed software such as the Quantification Analysis Tool (QAT).
+At present, national forecasting at the EPSS still relies heavily on basic extrapolation tools—usually Excel sheets or donor-developed software such as the Quantification Analysis Tool (QAT).
 These tools are easy to use but limited: they assume stable demand, ignore uncertainty, and often miss the operational signals embedded in local experience.
-In practice, forecasts are sometimes adjusted based on gut feeling or anecdotal program insights—not because planners want to—but because the tools don’t offer a better alternative.}
+In practice, forecasts are sometimes adjusted based on gut feeling or anecdotal program insights—not because planners want to—but because the tools don’t offer a better alternative.
 
-\textcolor{blue}{The models developed here—freely available, open-source, and designed to integrate directly with routine EPSS data—allow planners to make decisions using full probabilistic forecasts, not just single-point estimates.
+The models developed here—freely available, open-source, and designed to integrate directly with routine EPSS data—allow planners to make decisions using full probabilistic forecasts, not just single-point estimates.
 Forecasts are delivered with prediction intervals (e.g., 80\% or 90\%) that help teams decide not only how much to order, but how much risk they’re willing to tolerate.
 For products with known seasonality—like antimalarials or diagnostic kits—this matters.
-The system doesn’t just forecast a number; it gives planners a buffer strategy.}
+The system doesn’t just forecast a number; it gives planners a buffer strategy.
 
-\textcolor{blue}{Beyond accuracy, the models also reflect how medicines actually move through the system.
+Beyond accuracy, the models also reflect how medicines actually move through the system.
 Predictors based on warehouse replenishments, fiscal year inventory routines, and known disease cycles are embedded into the forecasts—not hard-coded, but learned from the data in ways that are transparent and reproducible.
-These inputs are drawn from expert operational knowledge and can be updated or modified as the system evolves.}
+These inputs are drawn from expert operational knowledge and can be updated or modified as the system evolves.
 
-\textcolor{blue}{Importantly, the entire modeling pipeline is built in R and Python, and is already shared alongside code and data.
+Importantly, the entire modeling pipeline is built in R and Python, and is already shared alongside code and data.
 This makes it immediately usable by EPSS analysts and regional logistics teams, even in settings without access to proprietary tools or high-end infrastructure.
-It's not a black box; it's something the system can own, modify, and grow with.}
+It's not a black box; it's something the system can own, modify, and grow with.
 
-\textcolor{blue}{To support real uptake, we recommend five practical steps:}
+To support real uptake, we recommend five practical steps:
 
-- \textcolor{blue}{Embed these probabilistic models into EPSS quantification rounds to replace rigid extrapolation methods and bring scenario-based planning into monthly and quarterly cycles.}
+- Embed these probabilistic models into EPSS quantification rounds to replace rigid extrapolation methods and bring scenario-based planning into monthly and quarterly cycles.
 
-- \textcolor{blue}{Use prediction intervals to define risk-informed buffer policies, especially for high-volatility products. Not every item needs the same margin of error.}
+- Use prediction intervals to define risk-informed buffer policies, especially for high-volatility products. Not every item needs the same margin of error.
 
-- \textcolor{blue}{Automate the collection of domain knowledge  (e.g., replenishment events, stock counts, seasonal triggers) into the logistics data stream to reduce reliance on manual elicitation while preserving contextual intelligence.}
+- Automate the collection of domain knowledge  (e.g., replenishment events, stock counts, seasonal triggers) into the logistics data stream to reduce reliance on manual elicitation while preserving contextual intelligence.
 
-- \textcolor{blue}{Train regional and central teams using the open-source tools and shared codebase, turning forecasting into a hands-on, in-country competency rather than a dependency on external tools or consultants.}
+- Train regional and central teams using the open-source tools and shared codebase, turning forecasting into a hands-on, in-country competency rather than a dependency on external tools or consultants.
 
-- \textcolor{blue}{Establish routine forecast review sessions, supported by visual dashboards that reflect not just what the model says, but how uncertain that estimate is—and what that means for stock levels, procurement plans, and emergency readiness.}
+- Establish routine forecast review sessions, supported by visual dashboards that reflect not just what the model says, but how uncertain that estimate is—and what that means for stock levels, procurement plans, and emergency readiness.
 
-\textcolor{blue}{Taken together, these steps push forecasting beyond formulas and into real decision-making.
-}
+Taken together, these steps push forecasting beyond formulas and into real decision-making.
+For immediate short-term adoption in EPSS, automating the integration of domain knowledge into logistics data streams is critical to improve data quality and reduce manual efforts.
+Establishing routine forecast review sessions supported by visual dashboards can also be implemented in the short term to increase transparency and incorporate uncertainty into decision-making.
+Following these steps, embedding probabilistic forecasting methods and using prediction intervals to design risk-informed buffer policies will enhance planning accuracy and inventory resilience.
+Finally, training regional and central teams using open-source tools should be pursued as a long-term goal to build sustainable, in-country forecasting expertise.

### 2025-08-04 manuscript/05-conclusion.qmd
3RD REVISION

@@
-# Conclusion and future reserach {#sec-conclusion}
+# Conclusion and future research {#sec-conclusion}
 
 In this study, we conducted an extensive analysis of pharmaceutical demand forecasting within the EPSS.
-Using five years of consumption data for 33 key pharmaceutical products, along with additional information gathered through collaboration with domain experts at EPSS, our goal is to contribute to enhancing the demand forecasting process and emphasize the importance of integrating domain knowledge into model building.
+Using five years of distribution data for 33 key pharmaceutical products, along with additional information gathered through collaboration with domain experts at EPSS, our goal is to enhance the demand forecasting process and emphasize the importance of integrating domain knowledge into model building.
 This step is critical, as the phase of understanding the data and incorporating valuable contextual information is often overlooked.
-Too frequently, forecasting efforts rely solely on consumption data, leading to building models with little relevance to reality.
+Too frequently, forecasting efforts rely solely on distribution data, leading to building models with little relevance to reality.
 Our approach highlights the necessity of including domain insights to construct more accurate and effective forecasting models.
@@
 To advance forecasting practices further, exploring additional predictor variables—such as public health campaigns, disease outbreaks, and economic indicators—could provide valuable insights into demand dynamics.
-Improving the granularity of historical consumption data, through finer temporal intervals and geographic differentiation, also holds potential for more precise predictions.
+Improving the granularity of historical distribution data, through finer temporal intervals and geographic differentiation, also holds potential for more precise predictions.
 Additionally, hybrid forecasting models that combine judgmental point forecasts with probabilistic machine learning models could enhance predictive performance.
-Empowering EPSS staff through training in foundational time series models, data interpretation, and collection will enable informed decision-making and strengthen workforce capabilities.
+Empowering EPSS staff through training in foundation models for time series analysis and forecasting, data interpretation, and collection will enable informed decision-making and strengthen workforce capabilities.
 Replicating this study across diverse healthcare settings and conducting comparative analyses would help validate the adaptability and applicability of the findings.
-Additionally, it is recommended that pharmaceutical supply services systematically collect and maintain detailed records of events that may influence consumption, alongside consumption data.
+Additionally, it is recommended that pharmaceutical supply services systematically collect and maintain detailed records of events that may influence distribution, alongside past distribution data.
 This practice is essential for refining demand models and enhancing forecast accuracy.
@@
 Similarly, experts indicated that conflicts would disrupt the distribution of pharmaceuticals.
-While conflict zones did indeed hinder transportation, distribution still occurred whenever roads were temporarily opened, even amidst on-going conflict.
+While conflict zones did indeed hinder transportation, distribution still occurred whenever roads were temporarily opened, even amidst ongoing conflict.
 This created inconsistencies in the data, as periods of halted distribution were followed by rapid replenishment once access was restored.
@@
 
-## Future reserach
+## Future research
 
@@
 
-- \textcolor{blue}{Disruptions such as the COVID-19 pandemic and armed conflicts may alter consumption patterns in time series data, introducing irregularities that challenge the assumptions of traditional forecasting methods.
-In such contexts, it becomes important not only to model observed consumption but also to account for censored demand—i.e., the unmet need that would have been fulfilled under normal conditions and is essential for effective replenishment planning.
+- Disruptions such as the COVID-19 pandemic and armed conflicts may alter distribution patterns in time series data, introducing irregularities that challenge the assumptions of traditional forecasting methods.
+In such contexts, it becomes important not only to model observed distribution but also to account for censored demand—i.e., the unmet need that would have been fulfilled under normal conditions and is essential for effective replenishment planning.
 Future research could focus on developing and evaluating forecasting strategies that are specifically designed to handle disrupted time series and estimate censored demand.
 In particular, methods that enhance the robustness of probabilistic forecasts in the presence of structural breaks or exogenous shocks are especially relevant.
-One promising direction is also the development of causal forecasting models that incorporate policy changes, epidemic dynamics, and socio-economic factors—an approach that may be particularly well suited to the realities of pharmaceutical supply chains in low- and middle-income countries (LMICs).}
+One promising direction is also the development of causal forecasting models that incorporate policy changes, epidemic dynamics, and socio-economic factors—an approach that may be particularly well suited to the realities of pharmaceutical supply chains in low- and middle-income countries (LMICs).
 
-- \textcolor{blue}{Future research should explore the integration of probabilistic forecasting into inventory management policies, with a focus on evaluating the operational impact of forecast uncertainty.
-In addition, there is substantial potential in leveraging more granular data—such as daily or weekly consumption patterns—and examining forecasts at hierarchical levels, including individual sites or health facilities.
-Investigating how fine-grained temporal and spatial forecasts influence decision-making in inventory control could provide valuable insights into reducing stockouts, minimizing holding costs, and decreasing pharmaceutical waste.}
+- Future research should explore the integration of probabilistic forecasting into inventory management policies, with a focus on evaluating the operational impact of forecast uncertainty.
+In addition, there is substantial potential in leveraging more granular data—such as daily or weekly distribution patterns—and examining forecasts at hierarchical levels, including individual sites or health facilities.
+Investigating how fine-grained temporal and spatial forecasts influence decision-making in inventory control could provide valuable insights into reducing stockouts, minimizing holding costs, and decreasing pharmaceutical waste.

### 2025-08-04 manuscript/main.qmd
3RD REVISION

@@
 ---
-title: "The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting"
+title: "The missing puzzle piece: contextual insights for enhanced pharmaceutical supply chain forecasting"
 author:
@@
           name: Addis Ababa University
-          department: Department of Pharmaceutics and Social Pharmacy, School of Pharmacy
+          department: Department of Social and Administrative Pharmacy, School of Pharmacy
           country: Ethiopia.
@@
           name: Cardiff University
-          department: Data Lab for Social Good Reserach Group, Cardiff Business School
+          department: Data Lab for Social Good Research Group, Cardiff Business School
           country: UK.
@@
 Effective demand forecasting supports procurement and inventory management, which in turn could help to prevent stockouts, reduce wastage due to overstocking, and enhance overall healthcare delivery.
-In such settings, reliable forecasts, that acknowledge uncertainties, are essential to ensure that limited resources are used efficiently to meet health needs.
+In such settings, reliable forecasts that acknowledge uncertainties, are essential to ensure that limited resources are used efficiently to meet health needs.
   Despite advancements in forecasting, significant gaps remain in effective forecasting in poor resource countries.
-Most existing literature on pharmaceutical supply chain forecasting focuses primarily on point estimation of consumption while neglecting the inherent uncertainty of these forecasts.
-Additionally, these studies often rely solely on historical consumption data without considering the broader context that influences consumption.
+Most existing literature on pharmaceutical supply chain forecasting focuses primarily on point estimation while neglecting the inherent uncertainty of these forecasts.
+Additionally, these studies often rely solely on historical consumption and distribution data, overlooking the broader contextual factors that shape these patterns.
 Furthermore, many works lack rigorous methodological design, transparency, and reproducibility.
-This study addresses these gaps by integrating domain-specific knowledge, gathered through expert interviews and engagement with key members of the Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models.
+This study addresses these gaps by integrating domain-specific knowledge, gathered through expert interviews and engagement with key members of the Ethiopian Pharmaceutical Supply Service, into forecasting models.
 Using a dataset spanning five years (December 2017 to July 2022) from EPSS, we developed forecasting models that integrate expert-identified variables such as stock replenishment schedules, fiscal inventory counts, and disease outbreaks.
-Evaluation metrics including Mean Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE), and Continuous Ranked Probability Score (CRPS) are used to report forecast accuracy.
+Evaluation metrics including Mean Absolute Scaled Error, Root Mean Squared Scaled Error, and Continuous Ranked Probability Score are used to report forecast accuracy.
 The findings underscore the significance of contextual data in developing robust forecasting models that are adaptable to complex, real-world conditions.
@@
 To enhance transparency and reproducibility, we not only provide data and the code, but also the entire paper that is written in R & Python.
-All materials to reproduce this paper is available at [Insert Github link once paper accepted](Insert Github link once paper accepted).
+All materials to reproduce this paper is available at [this Github repository](https://github.com/arebuissa/Forecasting_EPSS_domain_knowledge).
 

### 2025-08-04 manuscript/main.tex
3RD REVISION

@@
 \hypersetup{
-  pdftitle={The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting},
+  pdftitle={The missing puzzle piece: contextual insights for enhanced pharmaceutical supply chain forecasting},
   pdfauthor={Arebu Issa Bilal; Bahman Rostami-Tabar; Harsha Halgamuwe Hewage; Umit Sezer Bititci; Teferi Gedif Fenta},
@@
 \begin{frontmatter}
-\title{The missing puzzle piece: Contextual insights for enhanced
+\title{The missing puzzle piece: contextual insights for enhanced
 pharmaceutical supply chain forecasting}
@@
 \affiliation[1]{organization={Addis Ababa University, Department of
-Pharmaceutics and Social Pharmacy, School of
+Social and Administrative Pharmacy, School of
 Pharmacy},country={Ethiopia.},countrysep={,},postcodesep={}}
 \affiliation[2]{organization={Cardiff University, Data Lab for Social
-Good Reserach Group, Cardiff Business
+Good Research Group, Cardiff Business
 School},country={UK.},countrysep={,},postcodesep={}}
@@
 to prevent stockouts, reduce wastage due to overstocking, and enhance
-overall healthcare delivery. In such settings, reliable forecasts, that
+overall healthcare delivery. In such settings, reliable forecasts that
 acknowledge uncertainties, are essential to ensure that limited
@@
 pharmaceutical supply chain forecasting focuses primarily on point
-estimation of consumption while neglecting the inherent uncertainty of
-these forecasts. Additionally, these studies often rely solely on
-historical consumption data without considering the broader context that
-influences consumption. Furthermore, many works lack rigorous
+estimation while neglecting the inherent uncertainty of these forecasts.
+Additionally, these studies often rely solely on historical consumption
+and distribution data, overlooking the broader contextual factors that
+shape these patterns. Furthermore, many works lack rigorous
 methodological design, transparency, and reproducibility. This study
@@
 through expert interviews and engagement with key members of the
-Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models.
-Using a dataset spanning five years (December 2017 to July 2022) from
-EPSS, we developed forecasting models that integrate expert-identified
-variables such as stock replenishment schedules, fiscal inventory
-counts, and disease outbreaks. Evaluation metrics including Mean
-Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE),
-and Continuous Ranked Probability Score (CRPS) are used to report
-forecast accuracy. The findings underscore the significance of
-contextual data in developing robust forecasting models that are
-adaptable to complex, real-world conditions. Our results also highlight
-the effectiveness of foundational time series models for forecasting.
-These models are particularly appealing for resource-constrained
-countries that may lack advanced analytical expertise. To promote
-usability, generalizability, and reproducibility, we share the complete
-dataset and code in R and Python, and the entire paper is written in
-Quarto via a GitHub repository to facilitate these practices.
+Ethiopian Pharmaceutical Supply Service, into forecasting models. Using
+a dataset spanning five years (December 2017 to July 2022) from EPSS, we
+developed forecasting models that integrate expert-identified variables
+such as stock replenishment schedules, fiscal inventory counts, and
+disease outbreaks. Evaluation metrics including Mean Absolute Scaled
+Error, Root Mean Squared Scaled Error, and Continuous Ranked Probability
+Score are used to report forecast accuracy. The findings underscore the
+significance of contextual data in developing robust forecasting models
+that are adaptable to complex, real-world conditions. Our results also
+highlight the effectiveness of foundational time series models for
+forecasting. These models are particularly appealing for
+resource-constrained countries that may lack advanced analytical
+expertise. To promote usability, generalizability, and reproducibility,
+we share the complete dataset and code in R and Python, and the entire
+paper is written in Quarto via a GitHub repository to facilitate these
+practices.
 \end{abstract}
@@
 delivery. However, much of the data being collected focuses on
-transactional data, such as consumption and sale. Almost all platforms
-collecting data tend to overlook the critical contextual factors and
-events that influence these transactions---factors that are just as
-crucial for accurate forecasting. This presents a major barrier, as
-understanding the broader context, including administration procedures,
-local policies, disruptions, conflicts, and so on is vital for building
-reliable forecasting models. This gap---known as business context and
-domain knowledge---requires urgent attention in Pharmaceutical supply
-chain forecasting. Without incorporating this knowledge, any forecasting
-model is unlikely to produce reliable results. We must prioritize
-collecting and integrating this contextual information to improve the
-accuracy of demand forecasts in pharmaceutical supply chains. Combining
-statistical models with domain insights offers a more refined
-forecasting process, as evidenced by research that highlights the value
-of judgmental adjustments to enhance forecast accuracy
+transactional data, such as consumption, distribution, and sale. Almost
+all platforms collecting data tend to overlook the critical contextual
+factors and events that influence these transactions---factors that are
+just as crucial for accurate forecasting. This presents a major barrier,
+as understanding the broader context, including administration
+procedures, local policies, disruptions, conflicts, and so on is vital
+for building reliable forecasting models. This gap --- known as business
+context and domain knowledge --- requires urgent attention in
+pharmaceutical supply chain forecasting. Without incorporating this
+knowledge, any forecasting model is unlikely to produce reliable
+results. We must prioritize collecting and integrating this contextual
+information to improve the accuracy of demand forecasts in
+pharmaceutical supply chains. Integrating statistical models with domain
+knowledge leads to a more accurate and context-aware forecasting
+process, as evidenced by research that highlights the value of
+judgmental adjustments to enhance forecast accuracy
 \citep{fildes2007against, taylor2022artificial, soyiri2013overview}.
@@
 
-This research uses consumption data from 33 health commodities, covering
-the period from December 2017 to July 2022, provided by the Ethiopian
-Pharmaceutical Supply Service (EPSS), to develop models for
+This research uses distribution data from 33 health commodities,
+covering the period from December 2017 to July 2022, provided by the
+Ethiopian Pharmaceutical Supply Service (EPSS), to develop models for
 six-month-ahead forecasting. Initially, we implement univariate models
-that rely solely on consumption data. Following this, we conduct
+that rely solely on distribution data. Following this, we conduct
 exploratory data analysis and engage in interviews with EPSS experts to
-gather domain knowledge on events influencing consumption patterns.
+gather domain knowledge on events influencing distribution patterns.
 Through these interviews, the experts identified three key factors
-collectively impacting consumption: stock replenishment period, fiscal
+collectively impacting distribution: stock replenishment period, fiscal
 calendar, and the seasonal malaria. We then develop forecasting models
@@
 compare their performance with univariate models that rely solely on
-consumption data.
+distribution data.
 
@@
 
-Accurate sales forecasting is crucial in the pharmaceutical industry,
+Accurate demand forecasting is crucial in the pharmaceutical industry,
 where it directly impacts profit maximization, cost minimization, and
@@
 
-In this review, a range of forecasting methods has been applied, from
-traditional statistical models to advanced machine learning techniques,
-with many studies using a combination of simple and advanced methods to
-enhance accuracy
+A range of methods is used in pharmaceutical supply chain forecasting,
+from traditional statistical models to advanced machine learning
+techniques, with many studies combining both simple and advanced
+approaches to improve accuracy
 \citep{nikolopoulos2016forecasting, zhu2021demand, anusha2014demand}.
 Some studies, such as those by \citet{newberne2006holt}, and
-\citet{restyana2021analysis}, applied traditional forecasting methods
-like the Holt-Winters method, Single Moving Average (SMA), and Single
-Exponential Smoothing (SES), which are simpler but effective for certain
-applications. In contrast, other researchers explored more advanced
-approaches, including hybrid models and machine learning techniques, to
-address complex forecasting challenges, as seen in the studies by
+\citet{restyana2021analysis}, applied traditional forecasting methods.
+In contrast, other researchers explored more advanced approaches,
+including hybrid models and machine learning techniques, to address
+complex forecasting challenges, as seen in the studies by
 \citet{siddiqui2022hybrid}, \citet{de2021lead}, \citet{kim2015demand},
@@
 Early research in pharmaceutical forecasting primarily focused on
-traditional statistical methods. These methods, including ARIMA,
-exponential smoothing, and moving averages, were widely used due to
-their simplicity and effectiveness in relatively stable environments
+classical statistical methods. These methods, including ARIMA
+(AutoRegressive Integrated Moving Average), exponential smoothing, and
+moving averages, were widely used due to their simplicity and
+effectiveness in relatively stable environments
 \citep{zahra2019forecasting}. These models typically assess forecast
@@
 management, validated through metrics like MAPE, MAD, and MSD.
-Similarly, \citet{restyana2021analysis} compared Single Moving Average
-(SMA) and Single Exponential Smoothing (SES) in drug demand forecasting,
-concluding that SES provided more accurate results, as indicated by
-lower MAD and MSE values.
-
-Recognizing the limitations of traditional models, particularly in
-capturing the complexities of pharmaceutical demand (e.g., seasonality,
-external influences), researchers began developing hybrid models. These
-models combine linear methods like ARIMA with nonlinear techniques such
-as neural networks, aiming to improve forecasting accuracy by addressing
-diverse data patterns. \citet{khalil2014intelligent} introduced a hybrid
-model that integrated ARIMA with neural networks. Their study
-demonstrated that this approach significantly outperformed traditional
-models, especially in scenarios with limited historical data. The hybrid
-model's effectiveness was confirmed by improvements in key performance
-indicators like MAPE and RMSE.
-
-The pharmaceutical industry's increasing complexity and the availability
-of large datasets have led to the adoption of advanced machine learning
+Similarly, \citet{restyana2021analysis} compared Simple Moving Average
+(SMA), and Single Exponential Smoothing (SES) in drug demand
+forecasting, concluding that SES provided more accurate results, as
+indicated by lower MAD and MSE values.
+
+Recognizing the limitations of traditional models---particularly their
+inability to capture the complexities of pharmaceutical demand, such as
+seasonality and external influences---researchers have increasingly
+turned to more advanced modeling approaches
+\citep{khalil2014intelligent}. Furthermore, the growing complexity of
+the pharmaceutical industry, coupled with the availability of large
+datasets, has accelerated the adoption of advanced machine learning
 techniques. These methods, such as Support Vector Regression (SVR),
@@
   logistics management systems, these systems primarily collect
-  transactional data, such as consumption. However, no data is collected
-  on events that influence consumption variability---information that
-  would be crucial for building reliable forecasting models. No studies
-  to date have explored how domain expertise could enhance
-  pharmaceutical demand forecasting.
+  transactional data, such as consumption and distribution. However, no
+  data is collected on events that influence their variability ---
+  information that would be crucial for building reliable forecasting
+  models. No studies to date have explored how domain expertise could
+  enhance pharmaceutical demand forecasting.
 \item
@@
   predominantly focuses on generating point forecasts. There is a lack
-  of studies that consider the entire forecast distribution of monthly
-  consumption, which would better capture the uncertainty of future
-  demand and provide a valuable risk management tool for
-  decision-makers.
+  of studies that consider the entire forecast distribution, which would
+  better capture the uncertainty of future demand and provide a valuable
+  risk management tool for decision-makers.
 \item
@@
 Current study & Monthly & 60 months & Consumption of pharmaceutical products & 6 months & Yes & TimeGpt (with and without predictors), LSTM (with and without predictors), Exponential Smoothing, ARIMA, Dynamic Regression (ARIMAX), Multiple Regression (with and without predictors) & RMSSE,CRIPS,MASE & 33 & Yes & Yes\\
-Siddiqui et al. (2022) & Monthly & 58 months & Sales of pharmaceutical products & Un known & No & ARIMA, HOLT WINTER, ETS,  Theta, ARHOW & MEAN, MAD, MSE, RMSE,  MAPE & 31 & No & No\\
+Siddiqui et al. (2022) & Monthly & 58 months & Sales of pharmaceutical products & Unknown & No & ARIMA, HOLT WINTER, ETS,  Theta, ARHOW & MEAN, MAD, MSE, RMSE,  MAPE & 31 & No & No\\
 de Oliveira et al.(2021) & Daily & 365 days & Lead time & NA & No & k-NN, SVM, RF, LR, MLP & MSE & 11 & Yes & No\\
-Rathipriya et al. (2023) & Weekly & 260 weeks & Sales of pharmaceuticals product & Un known & No & SNN, ARIMA P\_NN GR\_NN RBF\_NN ARIMA P\_NN GR\_NN RBF\_NN & RMSE, Normalized RMSE & 57 & Yes & No\\
+Rathipriya et al. (2023) & Weekly & 260 weeks & Sales of pharmaceuticals product & Unknown & No & SNN, ARIMA P\_NN GR\_NN RBF\_NN ARIMA P\_NN GR\_NN RBF\_NN & RMSE, Normalized RMSE & 57 & Yes & No\\
 Kim et al.(2015) & Monthly & 44 month & Medicines sales data & 1 month & No & VARX & Prediction error rate & 4 & No & No\\
@@
 Zahra and Putra (2019) & Monthly & 36 months & Consumption & 12 months & No & ARIMA, ES & RMSE, MAPE, MAD, MSE & 1 & No & No\\
-Nguyen et  al. (2023) & Un known & Un known & Pharmaceutical products & Un known & No & VARX & ME, MAE,MAPE,RMSE & Unknown & No & No\\
-Christos et al.(2017) & Weekly & 129 weeks & Web-scraping  of analgesic medcines & Un known & Yes & VARX  & AIC, BIC, MSE, MAE & Unknown & No & No\\
+Nguyen et  al. (2023) & Un known & Un known & Pharmaceutical products & Unknown & No & VARX & ME, MAE,MAPE,RMSE & Unknown & No & No\\
+Christos et al.(2017) & Weekly & 129 weeks & Web-scraping  of analgesic medcines & Unknown & Yes & VARX  & AIC, BIC, MSE, MAE & Unknown & No & No\\
 Belghith et al.(2024) & Daily & 1095 days & Sales of pharmaceuticals & 1 year & No & Brown’s method, Holt’s method, and Holt-Winters method, moving average & MAD, MAPE & 125 & No & No\\
@@
 
-For this study, we use a dataset spanning five years (Dec.~2017- July
-2022) of consumption data obtained from EPSS. Rigorous checks were
+For this study, we use a dataset spanning five years (December 2017 -
+July 2022) of distribution data obtained from EPSS. Rigorous checks were
 conducted to ensure the consistency and completeness of the collected
@@
 
-\subsubsection{Explaoratory Data Analysis with consumption
-data}\label{explaoratory-data-analysis-with-consumption-data}
+\subsubsection{Exploratory Data Analysis with distribution
+data}\label{exploratory-data-analysis-with-distribution-data}
 
@@
 and computed features of the time series, including the strength of
-trend and seasonality to better understadn data.
+trend and seasonality to better understand data.
 Figure~\ref{fig-feature} shows the strength of trend versus seasonality.
@@
 \caption{\label{fig-feature}The strength of the trend and seasonality in
-the time series of pharmaceutical product consumption. The scatter plot
+the time series of pharmaceutical product distribution. The scatter plot
 shows 33 data points, with each point corresponding to a product.}
@@
 data and various patterns, including trends and instances of erratic
-consumption behavior during certain months. For example, some series
-show low consumption volumes over consecutive months, followed by peak
-consumption in specific months, making them more volatile and difficult
+distribution behavior during certain months. For example, some series
+show low distribution volumes over consecutive months, followed by peak
+distribution in specific months, making them more volatile and difficult
 to forecast. This underscores the diversity of monthly pharmaceutical
 time series patterns within the dataset and highlights the importance of
-understanding the factors driving these consumption behaviors. Figure
+understanding the factors driving these distribution behaviors.
 Figure~\ref{fig-dataviz2} illustrates the time plots for a few selected
@@
 
-\caption{\label{fig-dataviz2}Monthly time plot of consumption. X-axis
+\caption{\label{fig-dataviz2}Monthly time plot of distribution. X-axis
 shows the month, consisting of 60 data points (months) and y-axis shows
 the comsumption. The panels display data from four products to give a
-glimpse of the consumption patterns.}
+glimpse of the distribution patterns.}
 
@@
 Visualizing the data revealed that various events significantly impacted
-the consumption of different products, but these were not reflected in
+the distribution of different products, but these were not reflected in
 the available data. Expert insights were crucial for understanding the
@@
 Therefore, we conducted interviews with domain experts to collect
-information on external factors that influence consumption. These
+information on external factors that influence distribution. These
 interviews allowed us to account for unusual patterns and customize our
-forecasting models to more accurately reflect the unique consumption
+forecasting models to more accurately reflect the unique distribution
 behaviors of each product.
@@
 tendency and dispersion, but also temporal dependence, sparsity
-patterns, and relative variability in periods with non-zero consumption.
+patterns, and relative variability in periods with non-zero
+distribution.
 
 \subsection{Collaborative expert review to identify factors affecting
-consumption}\label{collaborative-expert-review-to-identify-factors-affecting-consumption}
-
-\textcolor{blue}{To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation.
-In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.}
-
-Figure \ref{fig:expert-pipeline}
-\textcolor{blue}{presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.}
+distribution}\label{collaborative-expert-review-to-identify-factors-affecting-distribution}
+
+To incorporate expert knowledge into time series forecasting in a
+structured and replicable manner, we developed a multi-step process
+involving expert engagement, collaborative data review, and operational
+validation. In partnership with the Forecasting and Market Shaping
+Directorate at EPSS, we identified six experienced professionals from
+the Warehouse and Inventory Management Directorate with deep operational
+knowledge of pharmaceutical logistics.
+
+Figure \ref{fig:expert-pipeline} presents a visual overview of the
+end-to-end approach, highlighting each major stage of the process from
+expert engagement to model integration.
 
@@
 
-\textcolor{blue}{Through a facilitated half-day workshop, these experts reviewed five years of monthly consumption data for 33 pharmaceutical products.
-Anomalous patterns—such as consumption spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles).
-These insights were cross-validated against bin cards and warehouse records, and internal transfers were excluded to isolate externally relevant events.
-Dates were standardized to the Gregorian calendar.}
-
-\textcolor{blue}{Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor.
-The final predictors included binary indicators for stock replenishment events, categorical variables for fiscal year inventory counts, and dummy variables for malaria seasonality.
-A further description of these variables are summarized as followings:}
+Through a facilitated half-day workshop, these experts reviewed five
+years of monthly distribution data for 33 pharmaceutical products.
+Anomalous patterns---such as distribution spikes, prolonged lows, and
+erratic fluctuations---were jointly identified and interpreted in light
+of operational events (e.g., emergency distributions, inventory cycles).
+These insights were cross-validated against bin cards and warehouse
+records, and internal transfers were excluded to isolate externally
+relevant events. Dates were standardized to the Gregorian calendar.
+
+Predictor variables were defined through expert consensus, with
+inclusion contingent on agreement among all six core participants
+regarding the operational relevance and causal validity of each factor.
+The final predictors included binary indicators for stock replenishment
+events, categorical variables for fiscal year inventory counts, and
+dummy variables for malaria seasonality. A further description of these
+variables is summarized as followings:
 
@@
 \item
-  \textcolor{blue}{Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand.
-Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased.
-This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.}
+  Stock replenishment: refers to the process of restocking or refilling
+  inventory to ensure that there are sufficient quantities of products
+  or materials available to meet demand. Whenever there was stock
+  replenishment at the central EPSS, distribution to hubs and health
+  facilities increased. This increase was attributed to the need to
+  restock depleted inventories and the push from central EPSS to manage
+  space constraints.
 \item
-  \textcolor{blue}{Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location.
-The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.
-Physical inventory counting periods also influenced consumption.
-Stores closed during these periods, halting transactions.
-We observed increased consumption before inventory counting periods, as hubs and facilities stocked up.
-July and August were identified as physical counting periods each year.}
+  Physical fiscal year inventory counting: refers to the process of
+  manually counting and verifying the actual quantities of
+  pharmaceutical products available in stock at a specific location. The
+  process is critical for maintaining the accuracy of inventory records,
+  ensuring that medicines are available when needed, and preventing
+  stockouts or overstocking. Physical inventory counting periods also
+  influenced distribution. Stores closed during these periods, halting
+  transactions. We observed increased distribution before inventory
+  counting periods, as hubs and facilities stocked up. July and August
+  were identified as physical counting periods each year.
 \item
-  \textcolor{blue}{Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions.
-In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease.
-Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced.
-During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments.
-Malaria seasonality was another significant predictor.
-Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks.
-We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.}
+  Malaria seasonality: refers to the predictable patterns and
+  fluctuations in malaria incidence throughout the year, typically
+  influenced by climate and environmental conditions. In many regions,
+  malaria transmission peaks during and shortly after the rainy season,
+  when conditions such as stagnant water pools create ideal breeding
+  sites for the Anopheles mosquitoes that transmit the disease.
+  Conversely, malaria cases often decline during the dry season when
+  mosquito breeding sites are reduced. During peak malaria seasons,
+  there is a significant surge in the demand for antimalarial drugs and
+  other related treatments. Malaria seasonality was another significant
+  predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine
+  and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We
+  identified two recurrent epidemic seasons each year---March to May and
+  September to December---that affected distribution from 2017 through
+  2022.
 \end{itemize}
 
-\textcolor{blue}{These predictors, encoded with known historical and future values, were added to the modeling dataset.
-This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.}
+These predictors, encoded with known historical and future values, were
+added to the modeling dataset. This approach ensured consistent
+integration of expert-informed contextual variables across both point
+and probabilistic forecasting models.
 
@@
 multiplicative (``M''); and the error term can be additive (``A'') or
-multiplicative (``M''). To forecast consumption, we use the ETS()
+multiplicative (``M''). To forecast distribution, we use the ETS()
 function from the fable package in R, which automatically selects the
@@
 \textbf{Multiple Linear Regression (MLR)}: We use Multiple linear
-regression to model the relationship between a consumption and potential
-variables influencing its variation. In our first model, we use multiple
-linear regression with a trend component to capture the underlying
-progression over time, i.e., \emph{regression}. We also incorporate
-dummy variables for each month to account for seasonal fluctuations,
-without including any additional predictors, i.e.,
+regression to model the relationship between a distribution and
+potential variables influencing its variation. In our first model, we
+use multiple linear regression with a trend component to capture the
+underlying progression over time, i.e., \emph{regression}. We also
+incorporate dummy variables for each month to account for seasonal
+fluctuations, without including any additional predictors, i.e.,
 \emph{regression\_reg}. This approach helps us establish a baseline
@@
 model to see if capturing external factors can provide a better
-understanding of the factors influencing the consumption and result at
+understanding of the factors influencing the distribution and result at
 enhanced accuracy. We produce forecasts using TLSM() function from the
@@
 and patterns over time without relying on external predictors. ARIMAX
-(AutoRegressive Integrated Moving Average with eXogenous variables)
+(Auto Regressive Integrated Moving Average with eXogenous variables)
 extends ARIMA by incorporating external variables, or exogenous
@@
 
-\textcolor{blue}{To improve the robustness of the LSTM models and address overfitting issues, we introduced dropout regularization layers after the LSTM units and employed early stopping based on validation loss during model training.
-Each LSTM model was trained independently for each product series.
-Hyperparameter tuning was performed in a preliminary phase using a subset of products.
-Various configurations of LSTM units (30–100), dense units (50–200), dropout rates (0.1–0.5), and batch sizes (16–64) were evaluated based on validation set performance.
-The final model structure — 50 LSTM units, 100 dense units, a 0.2 dropout rate, and a batch size of 32 — was selected as a trade-off between forecast accuracy and model stability across different demand patterns.}
+To improve the robustness of the LSTM models and address overfitting
+issues, we introduced dropout regularization layers after the LSTM units
+and employed early stopping based on validation loss during model
+training. Each LSTM model was trained independently for each product
+series. Hyperparameter tuning was performed in a preliminary phase using
+a subset of products. Various configurations of LSTM units (30--100),
+dense units (50--200), dropout rates (0.1--0.5), and batch sizes
+(16--64) were evaluated based on validation set performance. The final
+model structure --- 50 LSTM units, 100 dense units, a 0.2 dropout rate,
+and a batch size of 32 --- was selected as a trade-off between forecast
+accuracy and model stability across different demand patterns.
 
@@
 
-\textcolor{blue}{In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical consumption.
-Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction}
-\citep{wang2023}.
-
-\textcolor{blue}{In this study, we employed a bootstrapping approach to construct predictive intervals.
-Bootstrapping was chosen primarily for its flexibility and model-agnostic nature, allowing it to be applied uniformly across the diverse range of forecasting methods implemented without requiring strong distributional assumptions.
-Moreover, pharmaceutical consumption data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.}
-
-\textcolor{blue}{Specifically, we assume that future forecast errors will be similar to past forecast errors.
-The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed consumption, and $\hat{y}_t$ denotes the corresponding forecast estimate.
-To simulate future consumption paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates.
-This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.}
-
-\textcolor{blue}{Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts}
-\citep{hyndman2021forecasting}.
-
-\textcolor{blue}{The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.}
+In addition to point forecasts, we generated probabilistic forecasts to
+capture the uncertainty surrounding future pharmaceutical distribution
+Several approaches are available for generating probabilistic forecasts,
+including analytical prediction intervals, quantile regression, Bayesian
+modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping,
+and conformal prediction \citep{wang2023}.
+
+In this study, we employed a bootstrapping approach to construct
+predictive intervals. Bootstrapping was chosen primarily for its
+flexibility and model-agnostic nature, allowing it to be applied
+uniformly across the diverse range of forecasting methods implemented
+without requiring strong distributional assumptions. Moreover,
+pharmaceutical distribution data often exhibit irregular and volatile
+patterns, making non-parametric approaches like bootstrapping
+particularly suitable.
+
+Specifically, we assume that future forecast errors will be similar to
+past forecast errors. The forecast error at time \(t\) is defined as:
+\(e_t = y_t - \hat{y}_t\), where \(y_t\) represents the observed
+distribution, and \(\hat{y}_t\) denotes the corresponding forecast
+estimate. To simulate future distribution paths, we randomly sample
+errors with replacement from the historical error distribution and add
+them to the point forecast estimates. This process is repeated multiple
+times (1,000 iterations in our study) to generate a distribution of
+possible outcomes for each forecast horizon.
+
+Prediction intervals at the desired confidence level (e.g., 95\%) are
+then constructed by taking appropriate quantiles from the empirical
+distribution of the simulated forecasts \citep{hyndman2021forecasting}.
+
+The bootstrapping method thus provides a robust and flexible framework
+to quantify forecast uncertainty across a heterogeneous set of
+pharmaceutical products without imposing restrictive parametric
+assumptions.
 
@@
 
-\textcolor{blue}{To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation}
-\citep{hyndman2021forecasting}.
-\textcolor{blue}{Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods.
-In our setup, the initial training set consisted of all available historical data from December 2017 up to June 2021.
-The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans consumption over a one-year horizon.
-At each iteration, models were trained on an expanding training window and evaluated over a fixed 6-month forecast horizon, aligned with typical EPSS planning cycles.
-After each forecast generation, the training set was expanded by one additional month, and the process was repeated, allowing for rolling assessment across the final 12 months of data.
-This structure ensured that forecasts reflected realistic operational scenarios, where forecasts are continuously updated as new data becomes available.
-This cross-validation design allowed us to evaluate each model's ability to perform across multiple different forecast origins and a variety of demand conditions, providing a more reliable and generalizable understanding of model performance.
-Furthermore, all model development steps, including hyperparameter tuning for the LSTM models, were conducted exclusively using the training data available at each iteration.
-No test set information was used during model selection or tuning.}
+To evaluate the performance of our forecasting models, we split the data
+into a series of training and test sets and employed a time series
+cross-validation approach, following best practices for forecasting
+evaluation \citep{hyndman2021forecasting}. Rather than using a fixed
+training and test split, we used a rolling-origin cross-validation
+framework, which allows for a more comprehensive assessment across
+different demand patterns and periods. In our setup, the initial
+training set consisted of all available historical data from December
+2017 up to June 2021. The evaluation period covered the subsequent 12
+months, reflecting the operational needs of the EPSS, which plans
+distribution over a one-year horizon. At each iteration, models were
+trained on an expanding training window and evaluated over a fixed
+6-month forecast horizon, aligned with typical EPSS planning cycles.
+After each forecast generation, the training set was expanded by one
+additional month, and the process was repeated, allowing for rolling
+assessment across the final 12 months of data. This structure ensured
+that forecasts reflected realistic operational scenarios, where
+forecasts are continuously updated as new data becomes available. This
+cross-validation design allowed us to evaluate each model's ability to
+perform across multiple different forecast origins and a variety of
+demand conditions, providing a more reliable and generalizable
+understanding of model performance. Furthermore, all model development
+steps, including hyperparameter tuning for the LSTM models, were
+conducted exclusively using the training data available at each
+iteration. No test set information was used during model selection or
+tuning.
 
@@
 To measure the forecast distribution accuracy, we calculate the
-Continuous Rank Probability Score \citep{hyndman2021forecasting}. It
-rewards sharpness and penalizes miscalibration, so it measures overall
-performance of the forecast distribution. For probabilistic evaluation,
-1,000 future paths were simulated per series, enabling robust estimation
-of forecast uncertainty.
+Continuous Rank Probability Score (CRPS) \citep{hyndman2021forecasting}.
+It rewards sharpness and penalizes miscalibration, so it measures
+overall performance of the forecast distribution. For probabilistic
+evaluation, 1,000 future paths were simulated per series, enabling
+robust estimation of forecast uncertainty.
 
@@
 context predictors versus those that rely solely on historical
-consumption data. Point forecast performance is reported using MASE and
+distribution data. Point forecast performance is reported using MASE and
 RMSSE, while probabilistic forecast accuracy is reported using CRPS.
@@
 systematically collecting such expert-informed data alongside
-transactional consumption data.
+transactional distribution data.
 
@@
 
-\textcolor{blue}{While LSTM models achieved the best overall forecast accuracy across products, their performance exhibited notable variability depending on the characteristics of individual demand patterns.
-For example, the product $Amlodipine - 5mg - Tablet$ demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero consumption.
-Such patterns align well with the strengths of univariate LSTM models, which are adept at capturing long-term dependencies and managing complex temporal fluctuations.
-In contrast, the demand for $Anti-Rho (D)$ is erratic and sparse, with frequent random fluctuations and little structural consistency.
-This lack of clear temporal patterns can make it challenging for LSTM models to learn generalizable signals, particularly given the limited data length available for training.
-Although LSTM can manage irregular data to some extent, it performs best when patterns are consistent or cyclic.
-These variations across product types contributed to the observed distribution of forecast errors across the product portfolio.
-Moreover, although we expect that multivariate LSTM models benefit from expert-informed predictors, our results show that the univariate LSTM consistently achieved better forecast performance.
-This outcome may be attributed to the nature of the predictors—binary, static, or weakly aligned with short-term temporal dynamics—which can disrupt rather than enhance learning when added to a neural network sensitive to input configuration.
-Moreover, with relatively short historical series and limited training samples, the inclusion of additional variables may have led to overfitting or reduced generalization.
-These findings suggest that while LSTM models can effectively learn temporal patterns from consumption data alone, incorporating structured external knowledge requires careful feature engineering and alignment to be beneficial.}
+While LSTM models achieved the best overall forecast accuracy across
+products, their performance exhibited notable variability depending on
+the characteristics of individual demand patterns. For example, the
+product \(Amlodipine - 5mg - Tablet\) demonstrates periods of extreme
+variability, with spikes in demand followed by periods of very low or
+zero distribution. Such patterns align well with the strengths of
+univariate LSTM models, which are adept at capturing long-term
+dependencies and managing complex temporal fluctuations. In contrast,
+the demand for \(Anti-Rho (D)\) is erratic and sparse, with frequent
+random fluctuations and little structural consistency. This lack of
+clear temporal patterns can make it challenging for LSTM models to learn
+generalizable signals, particularly given the limited data length
+available for training. Although LSTM can manage irregular data to some
+extent, it performs best when patterns are consistent or cyclic. These
+variations across product types contributed to the observed distribution
+of forecast errors across the product portfolio. Moreover, although we
+expect that multivariate LSTM models benefit from expert-informed
+predictors, our results show that the univariate LSTM consistently
+achieved better forecast performance. This outcome may be attributed to
+the nature of the predictors---binary, static, or weakly aligned with
+short-term temporal dynamics---which can disrupt rather than enhance
+learning when added to a neural network sensitive to input
+configuration. Moreover, with relatively short historical series and
+limited training samples, the inclusion of additional variables may have
+led to overfitting or reduced generalization. These findings suggest
+that while LSTM models can effectively learn temporal patterns from
+distribution data alone, incorporating structured external knowledge
+requires careful feature engineering and alignment to be beneficial.
 
@@
 administration systems, data often consists solely of transactional
-records on consumption and distribution. Including exogenous variables
+records on distribution and distribution. Including exogenous variables
 such as administrative procedures, seasonal patterns, or conflicts, or
@@
 collection of information about significant events and their impact on
-consumption is vital. Recording details of events such as policy
+distribution is vital. Recording details of events such as policy
 changes, administration procedures, conflicts, and incorporating this
 information into forecasting models creates a comprehensive
-understanding of consumption. This practice enhances modeling
+understanding of distribution. This practice enhances modeling
 reliability and allows institutions in developing countries and
@@
 
-\textcolor{blue}{In addition to forecast accuracy, it is also important to consider the computational efficiency of each model, particularly in settings where computational resources and technical expertise are limited.}
-
-Table~\ref{tbl-runtime}
-\textcolor{blue}{shows the total runtime required to train and generate forecasts across all 33 pharmaceutical product time series for each method for all origins.
-All models were implemented using R, except TimeGPT, which was run via Google Colab using a T4 GPU backend.
-All other models were executed on a local machine with 7 CPU cores}
-(11th Gen Intel(R) Core(TM) i5-1135G7 @ 2.40 GHz and 8 GB RAM)
+In addition to forecast accuracy, it is also important to consider the
+computational efficiency of each model, particularly in settings where
+computational resources and technical expertise are limited.
+
+Table~\ref{tbl-runtime} shows the total runtime required to train and
+generate forecasts across all 33 pharmaceutical product time series for
+each method for all origins. All models were implemented using R, except
+TimeGPT, which was run via Google Colab using a T4 GPU backend. All
+other models were executed on a local machine with 7 CPU cores (11th Gen
+Intel(R) Core(TM) i5-1135G7 @ 2.40 GHz and 8 GB RAM)
 
@@
 
-\textcolor{blue}{As shown in Figure 4 and 5, models that incorporated expert-informed regressors generally achieved better forecast accuracy compared to their univariate versions.
-This trend was most evident in classical models such as ARIMA and regression, where the inclusion of regressors led to a noticeable shift toward lower and more concentrated error distributions.
+As shown in Figure 4 and 5, models that incorporated expert-informed
+regressors generally achieved better forecast accuracy compared to their
+univariate versions. This trend was most evident in classical models
+such as ARIMA and regression, where the inclusion of regressors led to a
+noticeable shift toward lower and more concentrated error distributions.
 However, this improvement came with an increase in computational cost.
-In all cases, the addition of regressors increased runtime—by 72\% in the regression model (from 27.48 to 47.38 seconds) and by 43\% in ARIMA (from 96.27 to 138.08 seconds).
-TimeGPT also showed a slight increase in runtime (from 2.57 to 3.72 seconds), although the total processing time remained extremely low overall.
-Moreover, LSTM exhibited the largest increase in runtime when regressors were added, rising from 6,892 to 7,234 seconds.
-This sharp increase reflects the sensitivity of neural networks to input configuration, especially in contexts with limited data, irregular demand, and noisy signals.}
-
-\textcolor{blue}{These results underscore the importance of balancing model performance with implementation cost.
-While deep learning models such as LSTM offer strong potential when well-tuned, they demand significantly more computational resources and may be less robust when integrating static or weakly aligned contextual features.
-By contrast, TimeGPT, a foundational model pretrained on large-scale time series data, provides a compelling alternative.
-It required less than 4 seconds to forecast all products, offered competitive accuracy, and required no tuning or retraining—making it well-suited for practical use in low-resource environments.}
-
-\subsection{An illustration of probabilistic forecast for Pharmacuitical
+In all cases, the addition of regressors increased runtime---by 72\% in
+the regression model (from 27.48 to 47.38 seconds) and by 43\% in ARIMA
+(from 96.27 to 138.08 seconds). TimeGPT also showed a slight increase in
+runtime (from 2.57 to 3.72 seconds), although the total processing time
+remained extremely low overall. Moreover, LSTM exhibited the largest
+increase in runtime when regressors were added, rising from 6,892 to
+7,234 seconds. This sharp increase reflects the sensitivity of neural
+networks to input configuration, especially in contexts with limited
+data, irregular demand, and noisy signals.
+
+These results underscore the importance of balancing model performance
+with implementation cost. While deep learning models such as LSTM offer
+strong potential when well-tuned, they demand significantly more
+computational resources and may be less robust when integrating static
+or weakly aligned contextual features. By contrast, TimeGPT, a
+foundational model pretrained on large-scale time series data, provides
+a compelling alternative. It required less than 4 seconds to forecast
+all products, offered competitive accuracy, and required no tuning or
+retraining---making it well-suited for practical use in low-resource
+environments.
+
+\subsection{An illustration of probabilistic forecast for Pharmaceutical
 product
-consumption}\label{an-illustration-of-probabilistic-forecast-for-pharmacuitical-product-consumption}
+distribution}\label{an-illustration-of-probabilistic-forecast-for-pharmaceutical-product-distribution}
 
 In this section, we present an illustrative example of a probabilistic
-forecast for future consumption of \emph{Sodium Chloride (Normal
+forecast for future distribution of \emph{Sodium Chloride (Normal
 Saline)} product. Due to the complexity of including such visualizations
@@
 likelihoods to a range of possible outcomes, recognizing that different
-consumption levels may occur with varying probabilities.
-\textcolor{blue}{The goal is to maximize the sharpness of these predictive distributions—i.e., how concentrated the forecasts are—while maintaining calibration, meaning that the predicted probabilities align well with actual outcomes}
-\citep{gneiting2014probabilistic}.
-\textcolor{blue}{This approach allows decision-makers to fully leverage available information, incorporating uncertainty in a structured and measurable way.}
-The primary purpose, as illustrated in
+distribution levels may occur with varying probabilities. The goal is to
+maximize the sharpness of these predictive distributions---i.e., how
+concentrated the forecasts are---while maintaining calibration, meaning
+that the predicted probabilities align well with actual outcomes
+\citep{gneiting2014probabilistic}. This approach allows decision-makers
+to fully leverage available information, incorporating uncertainty in a
+structured and measurable way. The primary purpose, as illustrated in
 Figure~\ref{fig-forecast-density-hstep}, is to quantify and communicate
 uncertainty. This figure displays the forecast distribution of
-consumption over a 6-month horizon using a density plot. For each month
+distribution over a 6-month horizon using a density plot. For each month
 within the forecast period, a separate distribution is generated. The
@@
 they do not convey the detailed probabilities of low or high
-consumption. This distinction highlights the value of probabilistic
+distribution. This distinction highlights the value of probabilistic
 forecasting in supporting informed decision-making by offering a clearer
@@
 
-\textcolor{blue}{To further illustrate the practical relevance, we consider an illustrative case where the mean forecasted consumption for the next month is 1,000 units, with a 90\% prediction interval ranging from 850 to 1,150 units.}
-Table~\ref{tbl-point_vs_probabilistic}
-\textcolor{blue}{summarizes the decision outcomes under each approach.}
-
-\textcolor{blue}{Under a traditional approach, inventory decisions would rely solely on the point forecast.
-Based on the mean prediction of 1,000 units, the store manager would order precisely that quantity, assuming it would meet the expected demand.
-However, this approach does not incorporate any adjustment for uncertainty, potentially leading to stockouts if actual demand exceeds 1,000 units, or excess inventory if consumption is lower.}
-
-\textcolor{blue}{In contrast, utilizing the full probabilistic forecast allows the decision-maker to take variability into account.
-If a higher service level is required, for example 95\%, the inventory policy may be adjusted by stocking closer to the upper bound of the 90\% prediction interval (e.g., 1,150 units).
-If inventory holding costs are a major concern and the organization can tolerate a modest risk of shortage, the order quantity could remain closer to the median forecast of 1,000 units.}
+To further illustrate the practical relevance, we consider an
+illustrative case where the mean forecasted distribution for the next
+month is 1,000 units, with a 90\% prediction interval ranging from 850
+to 1,150 units. Table~\ref{tbl-point_vs_probabilistic} summarizes the
+decision outcomes under each approach.
+
+Under a traditional approach, inventory decisions would rely solely on
+the point forecast. Based on the mean prediction of 1,000 units, the
+store manager would order precisely that quantity, assuming it would
+meet the expected demand. However, this approach does not incorporate
+any adjustment for uncertainty, potentially leading to stockouts if
+actual demand exceeds 1,000 units, or excess inventory if distribution
+is lower.
+
+In contrast, utilizing the full probabilistic forecast allows the
+decision-maker to take variability into account. If a higher service
+level is required, for example 95\%, the inventory policy may be
+adjusted by stocking closer to the upper bound of the 90\% prediction
+interval (e.g., 1,150 units). If inventory holding costs are a major
+concern and the organization can tolerate a modest risk of shortage, the
+order quantity could remain closer to the median forecast of 1,000
+units.
 
@@
 
-\textcolor{blue}{This illustrative comparison indicates that point forecasts provide a single estimate without adjusting for forecast uncertainty, while probabilistic forecasts allow decision-makers to explicitly align inventory decisions with risk tolerance and service level requirements.
-This is particularly important for pharmaceutical products, where the consequences of stockouts or overstocking can be significant both operationally and clinically.}
+This illustrative comparison indicates that point forecasts provide a
+single estimate without adjusting for forecast uncertainty, while
+probabilistic forecasts allow decision-makers to explicitly align
+inventory decisions with risk tolerance and service level requirements.
+This is particularly important for pharmaceutical products, where the
+consequences of stockouts or overstocking can be significant both
+operationally and clinically.
 
@@
 
-\textcolor{blue}{This study offers actionable lessons for supply chain managers, public health planners, and policymakers navigating the uncertainty of pharmaceutical demand—particularly in systems like Ethiopia, where operational constraints and demand volatility are the norm, not the exception.}
-
-\textcolor{blue}{At present, national forecasting at the Ethiopian Pharmaceutical Supply Service (EPSS) still relies heavily on basic extrapolation tools—usually Excel sheets or donor-developed software such as the Quantification Analysis Tool (QAT).
-These tools are easy to use but limited: they assume stable demand, ignore uncertainty, and often miss the operational signals embedded in local experience.
-In practice, forecasts are sometimes adjusted based on gut feeling or anecdotal program insights—not because planners want to—but because the tools don’t offer a better alternative.}
-
-\textcolor{blue}{The models developed here—freely available, open-source, and designed to integrate directly with routine EPSS data—allow planners to make decisions using full probabilistic forecasts, not just single-point estimates.
-Forecasts are delivered with prediction intervals (e.g., 80\% or 90\%) that help teams decide not only how much to order, but how much risk they’re willing to tolerate.
-For products with known seasonality—like antimalarials or diagnostic kits—this matters.
-The system doesn’t just forecast a number; it gives planners a buffer strategy.}
-
-\textcolor{blue}{Beyond accuracy, the models also reflect how medicines actually move through the system.
-Predictors based on warehouse replenishments, fiscal year inventory routines, and known disease cycles are embedded into the forecasts—not hard-coded, but learned from the data in ways that are transparent and reproducible.
-These inputs are drawn from expert operational knowledge and can be updated or modified as the system evolves.}
-
-\textcolor{blue}{Importantly, the entire modeling pipeline is built in R and Python, and is already shared alongside code and data.
-This makes it immediately usable by EPSS analysts and regional logistics teams, even in settings without access to proprietary tools or high-end infrastructure.
-It's not a black box; it's something the system can own, modify, and grow with.}
-
-\textcolor{blue}{To support real uptake, we recommend five practical steps:}
+This study offers actionable lessons for supply chain managers, public
+health planners, and policymakers navigating the uncertainty of
+pharmaceutical demand---particularly in systems like Ethiopia, where
+operational constraints and demand volatility are the norm, not the
+exception.
+
+At present, national forecasting at the EPSS still relies heavily on
+basic extrapolation tools---usually Excel sheets or donor-developed
+software such as the Quantification Analysis Tool (QAT). These tools are
+easy to use but limited: they assume stable demand, ignore uncertainty,
+and often miss the operational signals embedded in local experience. In
+practice, forecasts are sometimes adjusted based on gut feeling or
+anecdotal program insights---not because planners want to---but because
+the tools don't offer a better alternative.
+
+The models developed here---freely available, open-source, and designed
+to integrate directly with routine EPSS data---allow planners to make
+decisions using full probabilistic forecasts, not just single-point
+estimates. Forecasts are delivered with prediction intervals (e.g., 80\%
+or 90\%) that help teams decide not only how much to order, but how much
+risk they're willing to tolerate. For products with known
+seasonality---like antimalarials or diagnostic kits---this matters. The
+system doesn't just forecast a number; it gives planners a buffer
+strategy.
+
+Beyond accuracy, the models also reflect how medicines actually move
+through the system. Predictors based on warehouse replenishments, fiscal
+year inventory routines, and known disease cycles are embedded into the
+forecasts---not hard-coded, but learned from the data in ways that are
+transparent and reproducible. These inputs are drawn from expert
+operational knowledge and can be updated or modified as the system
+evolves.
+
+Importantly, the entire modeling pipeline is built in R and Python, and
+is already shared alongside code and data. This makes it immediately
+usable by EPSS analysts and regional logistics teams, even in settings
+without access to proprietary tools or high-end infrastructure. It's not
+a black box; it's something the system can own, modify, and grow with.
+
+To support real uptake, we recommend five practical steps:
 
@@
 \item
-  \textcolor{blue}{Embed these probabilistic models into EPSS quantification rounds to replace rigid extrapolation methods and bring scenario-based planning into monthly and quarterly cycles.}
+  Embed these probabilistic models into EPSS quantification rounds to
+  replace rigid extrapolation methods and bring scenario-based planning
+  into monthly and quarterly cycles.
 \item
-  \textcolor{blue}{Use prediction intervals to define risk-informed buffer policies, especially for high-volatility products. Not every item needs the same margin of error.}
+  Use prediction intervals to define risk-informed buffer policies,
+  especially for high-volatility products. Not every item needs the same
+  margin of error.
 \item
-  \textcolor{blue}{Automate the collection of domain knowledge  (e.g., replenishment events, stock counts, seasonal triggers) into the logistics data stream to reduce reliance on manual elicitation while preserving contextual intelligence.}
+  Automate the collection of domain knowledge (e.g., replenishment
+  events, stock counts, seasonal triggers) into the logistics data
+  stream to reduce reliance on manual elicitation while preserving
+  contextual intelligence.
 \item
-  \textcolor{blue}{Train regional and central teams using the open-source tools and shared codebase, turning forecasting into a hands-on, in-country competency rather than a dependency on external tools or consultants.}
+  Train regional and central teams using the open-source tools and
+  shared codebase, turning forecasting into a hands-on, in-country
+  competency rather than a dependency on external tools or consultants.
 \item
-  \textcolor{blue}{Establish routine forecast review sessions, supported by visual dashboards that reflect not just what the model says, but how uncertain that estimate is—and what that means for stock levels, procurement plans, and emergency readiness.}
+  Establish routine forecast review sessions, supported by visual
+  dashboards that reflect not just what the model says, but how
+  uncertain that estimate is---and what that means for stock levels,
+  procurement plans, and emergency readiness.
 \end{itemize}
 
-\textcolor{blue}{Taken together, these steps push forecasting beyond formulas and into real decision-making.
-}
-
-\section{Conclusion and future reserach}\label{sec-conclusion}
+Taken together, these steps push forecasting beyond formulas and into
+real decision-making. For immediate short-term adoption in EPSS,
+automating the integration of domain knowledge into logistics data
+streams is critical to improve data quality and reduce manual efforts.
+Establishing routine forecast review sessions supported by visual
+dashboards can also be implemented in the short term to increase
+transparency and incorporate uncertainty into decision-making. Following
+these steps, embedding probabilistic forecasting methods and using
+prediction intervals to design risk-informed buffer policies will
+enhance planning accuracy and inventory resilience. Finally, training
+regional and central teams using open-source tools should be pursued as
+a long-term goal to build sustainable, in-country forecasting expertise.
+
+\section{Conclusion and future research}\label{sec-conclusion}
 
 In this study, we conducted an extensive analysis of pharmaceutical
-demand forecasting within the EPSS. Using five years of consumption data
-for 33 key pharmaceutical products, along with additional information
-gathered through collaboration with domain experts at EPSS, our goal is
-to contribute to enhancing the demand forecasting process and emphasize
-the importance of integrating domain knowledge into model building. This
+demand forecasting within the EPSS. Using five years of distribution
+data for 33 key pharmaceutical products, along with additional
+information gathered through collaboration with domain experts at EPSS,
+our goal is to enhance the demand forecasting process and emphasize the
+importance of integrating domain knowledge into model building. This
 step is critical, as the phase of understanding the data and
 incorporating valuable contextual information is often overlooked. Too
-frequently, forecasting efforts rely solely on consumption data, leading
-to building models with little relevance to reality. Our approach
-highlights the necessity of including domain insights to construct more
-accurate and effective forecasting models.
+frequently, forecasting efforts rely solely on distribution data,
+leading to building models with little relevance to reality. Our
+approach highlights the necessity of including domain insights to
+construct more accurate and effective forecasting models.
 
@@
 economic indicators---could provide valuable insights into demand
-dynamics. Improving the granularity of historical consumption data,
+dynamics. Improving the granularity of historical distribution data,
 through finer temporal intervals and geographic differentiation, also
@@
 probabilistic machine learning models could enhance predictive
-performance. Empowering EPSS staff through training in foundational time
-series models, data interpretation, and collection will enable informed
-decision-making and strengthen workforce capabilities. Replicating this
-study across diverse healthcare settings and conducting comparative
-analyses would help validate the adaptability and applicability of the
-findings. Additionally, it is recommended that pharmaceutical supply
-services systematically collect and maintain detailed records of events
-that may influence consumption, alongside consumption data. This
-practice is essential for refining demand models and enhancing forecast
-accuracy.
+performance. Empowering EPSS staff through training in foundation models
+for time series analysis and forecasting, data interpretation, and
+collection will enable informed decision-making and strengthen workforce
+capabilities. Replicating this study across diverse healthcare settings
+and conducting comparative analyses would help validate the adaptability
+and applicability of the findings. Additionally, it is recommended that
+pharmaceutical supply services systematically collect and maintain
+detailed records of events that may influence distribution, alongside
+past distribution data. This practice is essential for refining demand
+models and enhancing forecast accuracy.
 
@@
 conflict zones did indeed hinder transportation, distribution still
-occurred whenever roads were temporarily opened, even amidst on-going
+occurred whenever roads were temporarily opened, even amidst ongoing
 conflict. This created inconsistencies in the data, as periods of halted
@@
 
-\subsection{Future reserach}\label{future-reserach}
+\subsection{Future research}\label{future-research}
 
@@
 \item
-  \textcolor{blue}{Disruptions such as the COVID-19 pandemic and armed conflicts may alter consumption patterns in time series data, introducing irregularities that challenge the assumptions of traditional forecasting methods.
-In such contexts, it becomes important not only to model observed consumption but also to account for censored demand—i.e., the unmet need that would have been fulfilled under normal conditions and is essential for effective replenishment planning.
-Future research could focus on developing and evaluating forecasting strategies that are specifically designed to handle disrupted time series and estimate censored demand.
-In particular, methods that enhance the robustness of probabilistic forecasts in the presence of structural breaks or exogenous shocks are especially relevant.
-One promising direction is also the development of causal forecasting models that incorporate policy changes, epidemic dynamics, and socio-economic factors—an approach that may be particularly well suited to the realities of pharmaceutical supply chains in low- and middle-income countries (LMICs).}
+  Disruptions such as the COVID-19 pandemic and armed conflicts may
+  alter distribution patterns in time series data, introducing
+  irregularities that challenge the assumptions of traditional
+  forecasting methods. In such contexts, it becomes important not only
+  to model observed distribution but also to account for censored
+  demand---i.e., the unmet need that would have been fulfilled under
+  normal conditions and is essential for effective replenishment
+  planning. Future research could focus on developing and evaluating
+  forecasting strategies that are specifically designed to handle
+  disrupted time series and estimate censored demand. In particular,
+  methods that enhance the robustness of probabilistic forecasts in the
+  presence of structural breaks or exogenous shocks are especially
+  relevant. One promising direction is also the development of causal
+  forecasting models that incorporate policy changes, epidemic dynamics,
+  and socio-economic factors---an approach that may be particularly well
+  suited to the realities of pharmaceutical supply chains in low- and
+  middle-income countries (LMICs).
 \item
-  \textcolor{blue}{Future research should explore the integration of probabilistic forecasting into inventory management policies, with a focus on evaluating the operational impact of forecast uncertainty.
-In addition, there is substantial potential in leveraging more granular data—such as daily or weekly consumption patterns—and examining forecasts at hierarchical levels, including individual sites or health facilities.
-Investigating how fine-grained temporal and spatial forecasts influence decision-making in inventory control could provide valuable insights into reducing stockouts, minimizing holding costs, and decreasing pharmaceutical waste.}
+  Future research should explore the integration of probabilistic
+  forecasting into inventory management policies, with a focus on
+  evaluating the operational impact of forecast uncertainty. In
+  addition, there is substantial potential in leveraging more granular
+  data---such as daily or weekly distribution patterns---and examining
+  forecasts at hierarchical levels, including individual sites or health
+  facilities. Investigating how fine-grained temporal and spatial
+  forecasts influence decision-making in inventory control could provide
+  valuable insights into reducing stockouts, minimizing holding costs,
+  and decreasing pharmaceutical waste.
 \end{itemize}
@@
 All materials to reproduce this paper is available at
-\href{Insert\%20Github\%20link\%20once\%20paper\%20accepted}{Insert
-Github link once paper accepted}.
+\href{https://github.com/arebuissa/Forecasting_EPSS_domain_knowledge}{this
+Github repository}.
 
